---
title: "Occupational Mobility Is Not One Thing: Evidence from the Canadian Labour Market" 
author: "Richard Martin"
date: today
format:
  pdf:
    documentclass: article
    number-sections: true
    geometry: margin=1in
bibliography: references.bib
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{threeparttable}
  - \DeclareMathOperator*{\argmin}{argmin}
editor: 
  markdown: 
    wrap: 72
execute:
  echo: false
  message: false
  warning: false
abstract: |
  Workers predominantly move between proximate occupations, but proximity means different things across the labour market. We use entropic regularized optimal   transport to conduct a horse race among three distance metrics — skill, hierarchy,   and binary — and show that no single geometry governs occupational mobility. Instead, the appropriate metric depends on each occupation's education specificity: skill where skills are highly transferable, hierarchy where silos (occupational broad   groups) and platforms (education and experience) structure mobility, and binary where   human capital is so occupation-specific that all alternative occupations appear equally distant. Each regime implies a distinct policy lever for addressing labour shortages.
---



```{r, include=FALSE}
library(tidyverse)
library(here)
library(knitr)
library(kableExtra)
library(patchwork)
library(conflicted)
conflicts_prefer(dplyr::filter)
cdfs <- read_rds(here("out","dist_cdfs.rds"))
calibration <- read_rds(here("out", "calibration.rds"))
```

## Pre-registration Notice {.unnumbered}

This paper is posted as a pre-registration of study design, prior to Statistics Canada Virtual Data Lab access.  Anything that looks like a result should be interpreted as a prior. Pre-registration is intended to defend against claims of post-hoc rationalization.

## Introduction

Occupational mobility is not one thing. Workers tend to move between proximate occupations, but proximity is not self-evident: occupations differ along multiple dimensions simultaneously --- canonically the tasks undertaken [@gathmann2010general], but also by required skills, hierarchical position, and selective education requirements. Each implies a different geometry of the labour market, and different predictions about who moves where. We test which geometry governs mobility --- and find that the answer depends on where in the labour market you look.

We separate local mobility (transitions that occur without new 
highest educational attainment) from non-local mobility (transitions that occur when a new highest educational attainment is achieved). For local mobility, distance is the explanatory variable — we ask which of three candidate metrics best rationalizes observed occupational transitions: skill (Euclidean in PCA-reduced O\*NET space), hierarchical (tree metric on the NOC taxonomy), and binary (discrete metric). For non-local mobility, the role of distance is reversed: rather than asking which metric explains where workers go, we ask what determines how far they travel, treating transition distance as the outcome and measures of education and occupation specificity as its determinants.

Regarding local mobility, we anticipate that no single metric will dominate globally. A labour market in which doctors, managers, and labourers all follow the same distance geometry would be surprising. We therefore partition the transition data into three sub-regimes based on the *origin* occupation's education specificity, each mapping to a hypothesized winning metric --- skill distance where human capital is general, hierarchical distance where silo structure binds, and binary switching costs where human capital is so occupation-specific that all alternative destinations appear equally distant. @sec-cf develops this partition in detail.

The policy relevance is direct. Labour shortages in different segments of the market require different adjustment mechanisms, and the right mechanism depends on which distance metric governs mobility in that sub-regime. Shortages where human capital is general may be addressed through reskilling, building on transferable skills. Shortages where silo structure binds are best addressed by increasing supply at the bottom platform of the relevant silo, facilitating upward movement to address shortages at higher levels. Shortages where human capital is occupation-specific cannot be addressed through either mechanism: the relevant human capital is acquired only through specific education programs, so interventions must target access to and capacity within those programs directly. Getting the geometry right is therefore not merely a methodological contribution --- it determines which policy lever to pull.

For local mobility, we use entropic regularized optimal transport (EROT) [@peyre2019computational; @cuturi2013sinkhorn] to conduct horse races between distance metrics, evaluating how well each rationalizes the full occupation-to-occupation transition matrix using relative KL divergence and relative Cost-Weighted Total Variation Distance (CWTVD). For non-local mobility, we estimate at the individual level the reduced-form relationship between transition distance and two specificity measures --- the individual's education occupation specificity (EOS), and their *destination* occupation's education specificity (OES). These measures proxy for the intensity of education-driven jumps: a new highest attainment with a high EOS opens doors to particular distant occupations, and a destination occupation with a high OES draws selectively from particular educations. 

The local vs non-local models provide complementary cross-regime falsification evidence: they should rationalize the mobility data in-regime, worsen in the presence of measurement error, and fail completely out of regime.

[FINDINGS PLACEHOLDER]

## Conceptual Framework {#sec-cf}

The central structural choice in this paper is the partition of 
occupational transitions into two qualitatively distinct regimes. A 
local transition occurs when a worker changes occupation without 
acquiring new educational attainment --- the worker's highest 
educational attainment in 2021 is the same as in 2016. A non-local 
transition occurs when a worker's highest educational attainment 
increases between the two census years. This partition reflects a 
fundamental difference in the mechanism governing mobility. For local 
transitions, the 2016 occupation is an informative starting point: 
human capital is relatively stable over the inter-census period, and 
distance from that starting point governs destination choice. For 
non-local transitions, the 2016 occupation is largely uninformative: 
a new highest educational attainment shifts the worker's position in 
occupational space in a way that is not captured by the 2016 
occupation, acting as a wormhole --- enabling discontinuous jumps that 
would otherwise be difficult to reconcile with their 2016 occupation.

Two symmetric specificity measures play central roles in the analysis, 
one defined over occupation (NOC) and one over *education*: the cross product 
of field of study and highest level of attainment. These 
measures are derived from Statistics Canada Table 98-10-0404-01 --- a 
one-quarter sample of the Canadian population, of which at most 
one-quarter overlaps with the linked microdata used for estimation.

To calculate an education's occupation specificity (EOS), we begin by 
calculating the KL divergence of $P(\text{NOC}|\text{education})$ from 
$P(\text{NOC})$. A higher value indicates that an education is 
distinctive in its occupation proportions. Because the empirical 
distribution can only place mass on a limited number of occupations for 
niche educations, KL divergence has a mechanical floor; we therefore 
residualize $\log(\text{KL})$ on $\log(T)$ where $T$ is the total 
number of workers with that education. The residual from this 
regression is defined to be EOS. This measure enters only the 
non-local partition as a continuous regressor proxying for wormhole 
entrance intensity --- a high EOS value increases the probability of 
a jump.

A symmetric procedure is followed to calculate an occupation's 
education specificity (OES). We begin by calculating the KL divergence 
of $P(\text{education}|\text{NOC})$ from $P(\text{education})$. A 
higher value indicates that an occupation is distinctive in its 
education proportions. Because the empirical distribution can only 
place mass on a limited number of educations for niche occupations, KL 
divergence has a mechanical floor; we therefore residualize 
$\log(\text{KL})$ on $\log(T)$ where $T$ is the total number of 
workers with that occupation. The residual from this regression is 
defined to be OES.  In the local partition, OES of the \emph{origin} occupation is used as a classifier defining the three sub-regimes; in the non-local partition, OES of the \emph{destination} occupation serves as a continuous regressor proxying for wormhole exit intensity --- conditional on a high EOS, a high OES increases the probability of a long jump.

OES captures where an occupation sits between the poles of general and specific human capital [@becker1964human]. Occupations drawing broadly from the education distribution require general human capital in Becker's sense --- skills whose returns are portable across employers and occupations --- while occupations drawing narrowly from specific credential pipelines require occupation-specific human capital, valuable within the occupation but not transferable beyond it. Occupational licensing [@kleiner2013] represents an explicit form of OES; licensing lists capture credential selectivity directly, but rely on curated registries that are often incomplete, jurisdiction-specific, and require manual updating. OES measures the same underlying phenomenon from the equilibrium education mix observed in the data, without requiring a curated list.

Occupational mobility tends to decline with age, but age is a proxy for the underlying mechanism rather than the mechanism itself. @mincer1981labor show empirically that age effects on mobility largely disappear once tenure is controlled, and attribute the residual to specific human capital accumulation rather than aging per se: as workers commit to an occupational path, the wedge between their current wage and outside options widens, making mobility increasingly costly. Tenure is their operational measure of that wedge, but the wedge is unobservable in principle — what they estimate is the hazard of separation as a function of job duration, not specific capital directly. OES approaches the same underlying quantity from the supply side: rather than inferring specific capital from the duration of job spells, it reads off the credential selectivity of the occupation from the equilibrium education mix. A newly trained and a 20-year veteran physician occupy the same high-OES occupation, and should be similarly immobile; a newly appointed and a long-tenured senior manager occupy the same low-OES occupation and should remain similarly mobile. Tenure varies within each pair but the mobility prediction does not — because what constrains mobility is the credential selectivity of the occupation, not the worker's time in it. This is a distinction tenure cannot identify as a stable property of the occupation, but OES can. If OES is the better proxy for latent occupation-specific human capital, OES tertiles should produce sharper metric separation than age tertiles. A placebo partition test using age tertiles in place of OES tertiles is reported in Section \ref{sec-res}.

Canada's National Occupational Classification (NOC) organizes 
occupations into a five-digit hierarchy. The first digit identifies one of ten broad occupational categories (e.g., Natural and Applied 
Sciences, Health, Trades). The remaining digits subdivide each category into progressively finer groupings down to the unit group level. 
Crosscutting this hierarchy is the Training, Education, Experience and 
Responsibilities (TEER) system, which assigns each occupation a level 
from 0 to 5 reflecting the degree of training and experience required: 
TEER 0 denotes senior management, TEER 1 requires university education, and TEER 5 requires short-term work demonstration only. The broad 
category defines the silo boundary in the analysis; the TEER levels 
define the platforms within each silo.

The three sub-regimes are named after their hypothesized winning metric 
rather than after the OES dimension that defines them --- a convention 
adopted for clarity, not a presumption of results. The tertile 
boundaries are not theoretically derived; if the three-regime structure 
is genuine, the horse race results should be robust to moderate 
misclassification at the boundaries.

The three sub-regimes are defined primarily by OES tertile, but two groups are reassigned regardless of their tertile: TEER 0 (senior management) and TEER 5 (labour) occupations are both assigned to the Skill sub-regime. The reassignment is best understood through the silo metaphor that motivates the Hierarchy sub-regime itself.  Think of each NOC broad group as a silo, with TEER levels defining elevation within it. Workers at ground level --- TEER 5 --- can walk freely to the base of any silo: entry points are at ground level, so no wall blocks lateral movement. Workers who have climbed some elevation within a silo face a different geometry: a lateral exit requires descent, a cross-ground traverse, and re-entry into another silo at its base. Workers who climb high enough eventually clear the walls entirely --- TEER 0 --- and move laterally again, through the open space of management skill above the silos.

The Hierarchy sub-regime therefore requires a specific condition: occupation-specific human capital sufficient to generate silo entry costs, but not so high as to foreclose within-silo movement. TEER 0 and TEER 5 bracket this middle ground from opposite ends. For TEER 5, there is no specific human capital to generate entry costs; ground-level movement is governed by skill distance, which remains meaningful even at low skill levels --- O*NET measures are standardized to mean zero and unit variance prior to PCA, so low-skill occupations occupy a distinct region of the skill space with real internal variation rather than collapsing to a degenerate cluster. For TEER 0, the walls have been cleared; managerial human capital is portable across silos by nature, and the relevant geometry is in management skill space rather than within any single broad group.

The Binary sub-regime represents the limiting case of the silo metaphor: each occupation is its own silo that lacks any internal structure.  The occupation's education specificity is so high that the richer geometry of skill or hierarchical distance adds no predictive power over a simple stay-or-leave cost. A high OES therefore predicts both a long initial jump --- observable in the non-local sample --- and subsequent immobility --- observable in the local sample. At the worker level only one of the above can be observed, but the two predictions are tested in separate samples.

The local/non-local partition has a natural falsification structure. 
Each model is tested both where it should work and where it should not, 
and the regime-specific predictions provide stronger evidence for the 
causal interpretation of the underlying mechanisms than either 
falsification alone would. A single confounder would have to operate 
in opposite directions across regimes to account for both patterns; 
multiple confounders would each have to align fortuitously with the 
predicted cross-regime structure. Neither is impossible, but both are 
less parsimonious than the theoretical mechanism itself. The design 
does not eliminate the possibility of confounding, but it raises the 
bar considerably: the model's validity rests not on the absence of 
confounders but on the implausibility of a story that generates the 
predicted pattern of opposing failures across regimes.

## Cost Metrics

The skill distance matrix $C_{\text{skill}}$ is constructed from 161 
O\*NET measures spanning skills, abilities, knowledge, and work 
activities, mapped to NOC 2021 occupations via a purpose-built 
crosswalk @onet_to_noc. An alternative approach would be to 
use OaSIS occupation-level skill profiles, but we construct our own 
crosswalk for a specific reason: falsifiability. The horse race 
requires that measurement error in the skill distance matrix be 
attributable to a known source --- namely, imprecision in the O\*NET 
to NOC mapping. OaSIS obscures the structure of that mapping, making 
it impossible to characterize where and how measurement error enters 
$C_{\text{skill}}$; our crosswalk makes this structure transparent 
and therefore makes the identification strategy falsifiable. Each NOC 
is represented as a weighted average of its constituent O\*NET 
occupations, with weights determined by the crosswalk mapping.  For more details on the mapping see Appendix A.

O\*NET provides two scales for each measure: Importance (how important the attribute is to the occupation) and Level (the degree to which the attribute is required). Both scales are standardized to a common 0--10 range and combined via geometric mean, yielding a single score that reflects both the relevance and the intensity of each attribute. The geometric mean is preferred over the arithmetic mean as it penalizes cases where one scale is near zero — an attribute that is important but required at negligible level, or required at high level but deemed unimportant, receives a lower composite score than one scoring moderately on both dimensions. The two scales are highly correlated in practice, so this choice is unlikely to be consequential for the resulting distance matrix, but we note it for transparency.

In high-dimensional spaces, Euclidean distances become increasingly uniform across pairs as dimensions accumulate, reducing their discriminating power. Principal components analysis is applied to the (standardized) 161-dimensional skill profiles, and the number of components retained is determined by a search over $k$: starting from $k=1$ we increment $k$ until the Spearman rank correlation between pairwise distances in the k-dimensional PCA space and pairwise distances in the full 161-dimensional space exceeds 0.99. This criterion --- rank preservation of pairwise distances rather than variance explained --- is chosen because the object of interest is the distance matrix itself, not the underlying skill dimensions. Twelve principal components satisfy this criterion. The skill distance between any two occupations is then the Euclidean distance between their coordinates in the retained 12-dimensional PCA space.  An interactive [skill distance heatmap](https://rpmartin.github.io/not-one-thing-companion/#sec-skill-heatmap) of the resulting matrix is available in the online companion.

The binary switching cost matrix $C_{\text{binary}}$ assigns a cost of zero to remaining in the same occupation and one to any transition. Structurally, the labour market is comprised of as many distinct silos as there are occupations, each with no internal structure of any kind. $C_{\text{binary}}$ serves two roles in the analysis. First, it is the theoretical straw man in the global horse race: if occupational mobility costs are purely a function of whether a switch occurs rather than how far, then $C_{\text{binary}}$ should rationalize observed mobility patterns better than $C_{\text{skill}}$ or $C_{\text{hierarchy}}$. Second, and more subtly,  $C_{\text{binary}}$ is best suited to settings where mobility is limited --- where most workers stay rather than move. In such settings, the information contained in a richer cost structure can cause model fit to worsen. This second role motivates the Binary sub-regime hypothesis: in occupations where human capital is so specific that all exits appear equally distant, binary is not merely the straw man but the correct description of the mobility geometry.

The hierarchical distance matrix $C_{\text{hierarchy}}$ is derived directly from the NOC 2021 taxonomy, treating the occupational classification structure as a tree metric. Distance between any two occupations is defined as follows: zero if all five digits match (same unit group); one if the first four digits match (same minor group, different unit group); two if the first three digits match (same sub-major group, different minor group); three plus the absolute difference in TEER level if only the first digit matches (same broad occupational category, different sub-major group); and nine for any cross-category transition. The absolute TEER difference ranges from zero to five, so within-category distances range from three to eight, preserving a strict separation from the cross-category sentinel value of nine.  An interactive [hierarchical distance heatmap](https://rpmartin.github.io/not-one-thing-companion/#sec-hier-heatmap) of the resulting matrix is available in the online companion.

NOC broad groups function as silos — bounded mobility clusters with limited entry points and internal progression along skill levels [@doeringer1971internal] — and TEER levels function as platforms within each silo, capturing the vertical skill stratification that governs upward movement. By construction, any between-platform move (minimum distance 4) is further than any within-platform move (maximum distance 3), and any cross-silo move (distance 9) is further than any within-silo move (maximum distance 8). The NOC broad group is a coarser unit than the occupational clusters described by @althauser1989, but the appropriate level of aggregation for the horse race: a finer partition would increase the preponderance of cross-category sentinel values in the hierarchical distance distribution, collapsing the metric's geometry toward binary and undermining its role as a distinct candidate in the horse race.

## Cost Metric Normalization

The three cost matrices are defined on different scales by construction: skill distances are continuous and spread over a wide range, hierarchical distances are discrete with a sentinel value of nine reserved for cross-category transitions, and binary distances take only the values zero and one. Direct comparison in the horse race requires that the matrices be anchored to a common scale; without normalization, the regularization parameter implies very different levels of concentration for each matrix, making the horse race results uninterpretable.

We normalize skill and hierarchical distances to a common scale by anchoring both at a substantively meaningful unit of hierarchical distance. The anchor is hierarchical distance 3 — the maximum distance without changing platforms. Distance 3 falls at the fourth percentile of the hierarchical distribution, reflecting the preponderance of cross-silo pairs. The corresponding fourth percentile of the skill distance distribution is 7.145. Dividing hierarchical distances by 3 and skill distances by 7.145 anchors both matrices at the same point in their respective distributions. @fig-cdfs shows the empirical CDFs of all three matrices before and after normalization. The raw distributions differ substantially in scale and shape; after normalization, skill and hierarchical CDFs overlap closely over the range of distances relevant to observed mobility, diverging only in the right tail. The sensitivity of horse race results to different anchors is examined in @sec-AC.


```{r}
#| warning: false
#| echo: false
#| label: fig-cdfs
#| fig-cap: "Cost Matrix Normalization: Raw vs. Normalized CDFs"
cdfs
```

## Mobility Data {#sec-data}

The analysis uses the Census 2016--2021 longitudinal file, constructed 
by linking long-form respondents across the two census cycles via the 
Derived Record Depository (DRD), Statistics Canada's national 
longitudinal database of personal identifiers. Linkage is 
probabilistic; details of the matching keys and overall linkage rate 
will be verified upon data access.

[Linkage Details Here]

In the following analysis we make use of a \emph{measurement complete} 
subset of the linked data where we observe occupation, field of study, 
highest educational attainment, class of worker and age in both 2016 and 2021.

[Missingness Details Here]

Canada's NOC 2021 hierarchy nominally contains 516 unit groups, but 
the five Senior Manager NOCs (00011--00015) are aggregated into NOC 
00018 in the CIP-NOC table used to derive EOS and OES, leaving 512 
occupations. Eight unit groups have no admissible O*NET mapping and 
are excluded from all estimation samples, leaving 504 occupations in 
the analysis.

There are two core subsamples used in both the local and non-local 
mobility analyses. Their role reverses depending on whether they serve 
as the primary identification sample or the falsification sample.

The \emph{local clean sample} comprises individuals with no increase 
in highest educational attainment between the two census years, no 
self-employment in either year, no TEER--attainment mismatch in 2016, 
and an O*NET-to-NOC mapping quality score above the bottom decile for 
their occupations in both 2016 and 2021. This is the sample where 
distance-based models should best rationalize mobility, and where the 
specificity regression should find nothing given that highest 
educational attainment was acquired prior to 2016.

The \emph{non-local clean sample} comprises individuals who acquire a 
new highest educational attainment between 2016 and 2021, subject to 
the same measurement restrictions as the local clean sample: no 
self-employment in either year, no TEER--attainment 
mismatch\footnote{For the non-local sample the TEER--attainment 
mismatch filter is applied to the 2021 occupation, where a credential 
acquired late in the inter-census window may not yet be reflected in 
the reported occupation.}, and an O*NET-to-NOC mapping quality score 
above the bottom decile for occupations in both 2016 and 2021. This 
is the sample where the specificity regression should best explain 
distance, and where the local model should struggle to rationalize 
long distance jumps.

In addition, for both models we investigate the effect of measurement error in occupation, when the data is otherwise appropriate for the model being fit. We assume that occupation measurement error is likely when either individuals are self-employed, or exhibit a TEER--attainment mismatch. For the local model the TEER mismatch filter is applied to the 2016 occupation; for the non-local model it is applied to the 2021 occupation.\footnote{For 
details regarding TEER--attainment mismatch criteria, see Appendix B.} 
This occupation mismeasurement introduces noise into the measurement 
of distance regardless of metric, producing symmetric fit degradation.

For non-local mobility we can perform one further diagnostic: the effect of poor skill mapping from O\*NET to NOC, when the data is otherwise appropriate for the model being fit. Poor skill mapping is identified by the bottom decile of the O\*NET-to-NOC mapping quality composite. Poor mapping introduces noise specifically into $C_{\text{skill}}$ while leaving 
$C_{\text{hierarchy}}$ unaffected --- a metric-specific prediction 
that serves as a falsification of the mapping quality composite 
itself. This diagnostic is not available for local mobility: the 
mapping quality split partitions the 504 NOC codes into 454 cleanly 
mapped and 50 poorly mapped, so the local clean subset (transitions 
among the 454 clean NOCs) and the mapping quality subset (transitions 
involving at least one poorly mapped NOC) have disjoint cells by 
construction. Since $\Delta KL$ and $\Delta CWTVD$ require a common 
set of possible origin-destination pairs to be comparable, no valid 
comparison is possible between the local clean and mapping quality 
subsets. For the two remaining cross-sample comparisons, fit metrics are 
evaluated on the intersection of observed origin-destination pairs 
across the two samples being compared.

## Local Model {#sec-local}

Mobility costs are latent: workers' switching costs are not observable 
in principle, so any distance matrix is necessarily a specification of 
an unobservable quantity. The three candidate matrices --- skill, 
hierarchical, and binary --- represent distinct hypotheses about the 
structure of those latent costs.

The entropy-regularized optimal transport problem minimizes the following Lagrangian, where the first term is the transport cost, the second is the entropic regularization penalizing deterministic transport plans, and the third and fourth enforce the marginal constraints $\sum_j P_{ij} = a_i$ and $\sum_i P_{ij} = b_j$ via dual variables $f_i$ and
$g_j$.

$$\mathcal{L} = \sum_{ij} P_{ij} C_{ij} + \varepsilon \sum_{ij} P_{ij}(\log P_{ij} - 1) + \sum_i f_i(a_i - \sum_j P_{ij}) + \sum_j g_j(b_j - \sum_i P_{ij})$$

where $a_i$ and $b_j$ are the exact occupation stocks in 2016 and 2021
respectively. 

The first-order conditions yield:
\begin{equation}
\log P_{ij} = \frac{f_i}{\varepsilon} + \frac{g_j}{\varepsilon} - 
\frac{C_{ij}}{\varepsilon}
\end{equation}
Exponentiating, the optimal transport plan takes the form 
$P^\star_{ij} = u_i \, e^{-C_{ij}/\varepsilon} \, v_j$, where 
$u_i = e^{f_i/\varepsilon}$ and $v_j = e^{g_j/\varepsilon}$ are 
scaling vectors. Strict convexity of the regularized objective 
guarantees a unique solution. The solution is therefore a row- and 
column-rescaling of the exponential kernel $K_{ij} = e^{-C_{ij}/\varepsilon}$, where the Sinkhorn algorithm recovers it by alternating row and column 
normalization until both marginal constraints are satisfied to within a specified tolerance. All computations use a log-domain Sinkhorn solver 
for numerical stability, implemented in the \texttt{safesink} R package [@safesink].

In log space, the optimal transport plan decomposes into additive origin effects ($\log u_i$), additive destination effects ($\log v_j$), and a dyadic cost term ($-C_{ij}/\varepsilon$). The scaling vectors adjust to ensure that the implied joint distribution matches the observed origin and destination marginals, absorbing any components of the data that operate at the level of origins or destinations alone given the specified cost structure. The cost matrix captures the residual structure that depends on origin--destination pairing. The horse race asks which specification of that dyadic structure best rationalizes observed transitions.

The scaling vectors $u_i$ and $v_j$ correspond to Lagrange multipliers associated with the marginal constraints and can be interpreted as dual shadow values in the entropy-regularized transport problem. We focus on mobility as the primary object of the model: wages are not used in estimation. However in @sec-wd, we report that the induced OT dual potentials exhibit a stable relationship with 2019 occupational wages across a wide range of $\varepsilon$, suggesting that the inferred latent structure has economic content beyond fitting mobility flows.

The natural alternative for evaluating occupational mobility costs is 
the gravity model [@anderson_vanwincoop_2003; @silva2006log], which 
estimates origin and destination fixed effects jointly with the cost 
parameter(s). This joint estimation is precisely the problem in our 
context: fixed effects absorb variation in mobility rates that may 
reflect true cost structure, allowing misspecified cost matrices to 
achieve good fit by compensating through the fixed effects. Using a gravity model would be akin to randomly dosing the horses with various performance enhancing drugs — all competitors run faster, but the winner may be the fastest horse or simply the one whose treatment happened to help most, and the two cannot be disentangled.  EROT avoids this issue by holding the marginal structure fixed at observed values, separating the question of how much mobility occurs --- determined by the margins --- from the question of where mobility flows --- shaped by the cost matrix.

The regularization parameter $\varepsilon$ has a natural interpretation
in the random utility framework. The optimal transport plan $P^\star_{ij} = u_i \, e^{-C_{ij}/\varepsilon} \, v_j$ is the static equilibrium of a discrete choice model in which workers draw idiosyncratic utility shocks scaled by $\varepsilon$ over destination occupations. Low values of $\varepsilon$ concentrate the transport plan on cost-minimizing assignments, approaching a deterministic reallocation in the limit; high values diffuse mobility across destinations, approaching the independence baseline $P_{ij} = a_i b_j$ as $\varepsilon \to \infty$. The RUM analogy governs the interior structure of the transport plan --- how probability mass is allocated across origin-destination pairs --- but does not extend to the margins. In a standard logit model the choice probabilities are outcomes of utility maximization with no constraint on their aggregates; in EROT the origin and destination occupation distributions are hard constraints that the transport plan must exactly reproduce. This distinction is methodologically consequential: because the margins cannot adjust to absorb cost misspecification, a poorly chosen cost matrix cannot hide behind fixed effects or scale recalibration --- its misfit appears directly in the objective.  

A critical identification constraint follows directly from the form of 
the optimal transport plan: only the ratio $C_{ij}/\varepsilon$ enters 
$P^\star$. The scale of the cost matrix and the value of $\varepsilon$ 
are therefore not separately identified, and results reported at a 
single value of $\varepsilon$ are uninterpretable without an implicit 
normalization assumption. We evaluate across a log-spaced grid of 
$\varepsilon$ values ranging from 0.5 to 2, spanning from relatively 
concentrated to relatively diffuse transport plans, and look for 
dominance that is robust across scales.

Interpreting the sweep requires care at low values of $\varepsilon$. 
As $\varepsilon \to 0$ the transport plan concentrates aggressively on 
cost-minimizing assignments. At low $\varepsilon$ both skill and hierarchy over-concentrate mass on the transitions their geometry predicts as 
cheapest --- leaving insufficient mass for transitions that actually 
occur but that their cost structure scores as expensive.  Binary avoids 
this failure because its cost structure does not discriminate among 
non-zero transitions: all switches cost the same, so mass is not 
over-concentrated on any subset of destinations. In the limit, binary 
wins regardless of the true data generating process --- this is a 
property of the mathematics, not the data. The [online companion](https://rpmartin.github.io/not-one-thing-companion/#sec-simulation) 
demonstrates this directly using simulation, showing that binary 
dominates at low $\varepsilon$ even when the DGP is known to be skill 
or hierarchy.

The boundary between the artifact region and the informative region 
cannot be determined in advance: it depends on the cost matrix scale, 
the concentration of the observed transition matrix, and the 
normalization. The decision rule is therefore stated in terms of the 
observable sweep pattern rather than a pre-specified threshold. Binary 
is declared the winner only if it dominates both non-binary metrics 
across the full $\varepsilon$ grid --- a full sweep is the minimum 
credible evidence for binary, because it is the only pattern that 
guarantees a win somewhere in the informative region, wherever that 
region turns out to be. A low-$\varepsilon$-only binary lead is 
uninformative by construction and is not counted as evidence for 
binary. A non-binary metric is declared the winner if it dominates the 
other non-binary competitor across the full grid, even if it loses to 
binary at low $\varepsilon$ --- the non-binary competition is not 
distorted by the artifact, so that comparison is informative at all 
values of $\varepsilon$.  Finally, if neither non-binary metric dominates 
the other across the full grid, the result for that sub-regime is declared a tie.

The horse race is evaluated using two complementary fit metrics, both
normalized relative to the independence baseline $P^{\text{ind}}_{ij} =
a_i b_j$, so that 0 indicates no improvement over independence, 1
indicates perfect fit, and negative values indicate a transport plan that fits observed mobility worse than independence.

The first metric is relative KL divergence improvement:

$$\Delta KL = \frac{KL(P^{\text{obs}} \parallel P^{\text{ind}}) - KL(P^{\text{obs}} \parallel \hat{P})}{KL(P^{\text{obs}} \parallel P^{\text{ind}})}$$

where $KL(P \parallel Q) = \sum_{ij} P_{ij} \log(P_{ij} / Q_{ij})$.
This measures the share of the divergence between observed transitions
and independence that is explained by the fitted transport plan. KL
divergence is sensitive to probability mass wherever it is misallocated, but is geometry-blind: misallocating mass to a nearby occupation is penalized identically to misallocating it to a distant one.

The second metric is relative Cost-Weighted Total Variation Distance (CWTVD) improvement:

$$\Delta CWTVD = \frac{CWTVD(P^{\text{obs}}, P^{\text{ind}}) - CWTVD(P^{\text{obs}}, \hat{P})}{CWTVD(P^{\text{obs}}, P^{\text{ind}})}$$

where $CWTVD(P^{\text{obs}}, \hat{P}) = \sum_{ij}
\frac{|P^{\text{obs}}_{ij} - \hat{P}_{ij}|}{2} \cdot C_{ij}$. 

CWTVD is a cost-weighted extension of total variation distance: the 
classical TVD between discrete distributions, 
$\frac{1}{2}\sum_{ij}|P_{ij} - Q_{ij}|$, is extended by weighting 
each absolute residual by the corresponding entry of a cost matrix, 
penalizing probability mass misallocated to geometrically distant 
origin-destination pairs more heavily than mass misallocated to nearby 
ones. Unlike KL divergence, CWTVD respects the metric structure of the 
outcome space; unlike the Earth Mover's Distance [@rubner2000earth], it does not solve an additional optimal transport problem over the residual mass, making it computationally trivial given an existing transport plan.

The scoring cost matrix $C$ used in CWTVD is the arithmetic mean of the
three candidate normalized cost matrices: $C_{\text{skill}}$,
$C_{\text{hierarchy}}$, and $C_{\text{binary}}$. Ideally we would use the latent cost matrix to score the models, but if we had this, what is the point of the horse race in the first place?  Using a neutral common metric ensures that no candidate model is advantaged by the choice of scoring matrix. The two metrics are complementary by design: KL divergence captures overall probability mass misallocation regardless of geometry, while CWTVD captures whether misallocation is concentrated in nearby or distant pairs. A transport plan that scores well on both metrics fits observed mobility both accurately and geometrically.

## Non-Local Model {#sec-non_local}

The non-local model estimates the conditional-on-positive relationship between $\log(d_{ij})$ --- the log distance between 2016 and 2021 occupations --- and the specificity measures EOS and OES. The outcome is computed separately using skill and hierarchical distance; the wormhole mechanism should operate independently of which metric measures the jump, providing a robustness check on the distance measure choice.

The theoretical mechanism is asymmetric. A high value of EOS is the 
primary condition for a wormhole jump --- the entrance is logically 
prior to the exit. Conditional on a sufficiently high EOS, OES 
determines the distance of the jump. This suggests a joint 
specification that includes both the EOS main effect and an EOS 
$\times$ OES interaction. Whether the joint model can separately 
identify entrance and exit intensity depends on the worker-level 
correlation between EOS and OES: when the two measures are highly 
correlated, the interaction term EOS $\times$ OES approximates a 
quadratic in a single specificity index, and the coefficients on EOS 
and EOS $\times$ OES can no longer be cleanly interpreted as entrance 
and exit intensity separately. The univariate models remain 
interpretable regardless of the correlation --- each estimates the 
reduced-form association between a single specificity measure and 
distance traveled, without attempting to decompose entrance from exit 
intensity. We therefore estimate three specifications --- two 
univariate models and the joint model --- and report the Spearman 
correlation between EOS and OES as a diagnostic. A low correlation 
supports a structural reading of the joint model's coefficients; a 
high correlation limits interpretation to the univariate models, with 
the joint model reported for completeness. The correlation is itself 
informative about the structure of education-occupation matching in 
the Canadian labour market independently of its implications for 
identification.

\begin{equation}
\log(d_{ij}) = \alpha + \beta_1 \, \text{EOS} + \delta_i + \varepsilon_{ij}
\label{eq-eos}
\end{equation}

\begin{equation}
\log(d_{ij}) = \alpha + \beta_2 \, \text{OES} + \delta_i + \varepsilon_{ij}
\label{eq-oes}
\end{equation}

\begin{equation}
\log(d_{ij}) = \alpha + \beta_3 \, \text{EOS} + \beta_4 \, \text{EOS} \times \text{OES} + \delta_i + \varepsilon_{ij}
\label{eq-joint}
\end{equation}

Prior to estimation, EOS and OES are standardized to mean zero and unit variance using the full linked sample. This ensures that coefficients are on a common scale both within and across the four subsamples, making the parameter estimates interpretable as a comparison of effect sizes, and the degradation pattern across samples interpretable as a comparison of fit rather than a reflection of differences in variable scale.

The primary hypothesis is that all $\beta$s are greater than zero. A secondary hypothesis posits that $\beta_1 > \beta_2$: the logical priority of wormhole entrance over exit predicts that education specificity has a stronger reduced-form association with distance traveled than occupation specificity. Age might seem a natural control given the well-documented empirical regularity that occupational mobility is decreasing in age [@topel1992job]. It is excluded from the main specification, however, because its correlation with jump distance operates primarily through origin occupation quality: younger workers with a new highest attainment tend to hold low-quality 2016 occupations, generating large jumps that genuinely reflect larger human capital increments, rather than a confound. Controlling for age would partially obscure the mechanism of interest. The results confirm that including it attenuates the EOS and OES coefficients and produces a negative age coefficient, consistent with older non-local movers starting closer to their destination.

The model is evaluated across four samples as described in Section \ref{sec-data}, with results reported for both skill and hierarchical distance outcomes. The within-$R^2$ --- the share of variance in $\log(d_{ij})$ explained by the specificity measures after partialling out origin occupation fixed effects --- is not expected to be large in absolute terms: transition distance is inherently noisy and EOS and OES are aggregate measures that capture only the systematic component of specificity-driven mobility. The relevant comparison is the ordering of within-$R^2$ across samples, not its absolute level.

The positive distance restriction removes many apprenticeship completers, since NOC makes no distinction between apprentices and journeypeople of the same occupation --- those already working in the target trade in 2016 show zero measured distance upon completion and are excluded by construction. Rather than threatening identification, this restriction improves measurement validity: the excluded zeros reflect NOC coarseness rather than true immobility, and their inclusion would introduce noise into the outcome variable. Workers who complete an entire apprenticeship program within the five-year window from a different 2016 occupation remain in the sample, and for these workers the wormhole mechanism is cleanly observable. The limitation is one of external validity rather than internal validity: estimates do not generalise to within-trade apprenticeship completers, whose human capital increment is real but invisible to NOC.


## Results \[PLACEHOLDER --- awaiting data\]  {#sec-res}

### Local

\[KL and CWTVD improvement figures by cost metric, $\varepsilon$ and
subsample\]

\[KL and CWTVD improvement by 3 OES sub regimes, cost metric, $\varepsilon$ and
subsample\]

\[KL and CWTVD improvement by 3 age sub regimes, cost metric, $\varepsilon$ and
subsample\]

### Non-local

\[tie-fighter plot for the $\beta$\]

\[ table for the $R^2$\]

## Limitations

### Local Mobility

Two sample restriction mechanisms introduce a potential selection 
concern for local mobility. Non-linkage and non-observability of 
occupation are both disproportionately concentrated among 
low-attachment workers --- individuals with casual, seasonal, or 
intermittent employment whose identifying information is also less 
stable across census cycles --- so the two filters largely co-select 
the same low-skill population rather than introducing independent 
exclusions. This shared exclusion affects the representativeness of 
population-level transition estimates (external validity) but does not 
threaten the internal validity of the horse race. For the horse race
results to be biased, the excluded workers' transitions would need to be 
systematically more or less rationalizable by a particular cost 
matrix than those of their observed counterparts in the same region 
of occupational space. No such mechanism is apparent: exclusion is 
driven by attachment instability and identifier churn, neither of 
which has a clear relationship to how well transitions conform to 
any particular cost structure conditional on occupation. The residual 
unlinked population among otherwise-observable workers --- recent 
immigrants, young workers, and high-movers --- similarly lacks any 
obvious occupational-space concentration that would differentially 
favour or penalize a particular metric.

A notable omitted factor for local mobility  is \emph{path-specific} wage differentials. The additively separable component of wage gains --- recoverable from origin and destination wage levels alone --- is absorbed by the 
margins insofar as high-wage destinations attract larger inflows and 
low-wage origins larger outflows. The residual dyadic component, 
reflecting transition-specific wage premia and discounts beyond what 
origin and destination wage levels predict, is an omitted factor. 
Short-distance transitions are within a similar occupational 
neighbourhood, reducing employer uncertainty about candidate fit and 
productivity; long-distance transitions cross larger skill or 
hierarchical gaps, and employers may discount wage offers to 
compensate for that uncertainty, or workers may accept lower wages 
to gain entry to a distant occupation.

Thus, path-specific wage differentials are better understood as an 
alternative mechanism, rather than a confounder. If employers use the same metric that governs mobility to assess candidate fit uncertainty --- discounting wages for transitions that are long under the winning metric --- that discount reinforces worker selection toward shorter transitions under 
the same metric. Both channels predict identical patterns of 
transitions and are equally consistent with the winning metric being 
the correct description of the mobility geometry. The horse race 
identifies which distance geometry best rationalizes observed 
transitions in a reduced-form sense, but cannot separately attribute 
that rationalization to worker or employer preferences regarding 
distance.

### Non-Local Mobility

The non-local partition is restricted to individuals who acquire a new highest educational attainment between 2016 and 2021 and report an occupation in both periods. A 2016 occupation is observed if and only if the individual is working in May 2016 --- the census reference week. @tbl-working-while-studying reports concurrent employment during the 2015/2016 school year averaging around 50%, with a notable gap between college and university students consistent with more intensive programs suppressing concurrent employment. The May reference week provides a partial offset --- falling after the typical post-secondary academic year ends, it captures individuals who have recently finished exams and may have already begun summer employment --- but this offset is weaker for intensive professional programs that are more likely to run year-round. The resulting truncation is not random: missingness is concentrated at the top of the EOS distribution, precisely where the wormhole mechanism is strongest and jumps are longest. Thus estimates are attenuated toward zero and should be interpreted as lower bounds on the true relationship, given that truncation is concentrated precisely where the signal is strongest. Notably, this missingness pattern is not purely a limitation: disproportionate absence of 2016 occupations among high-EOS educations is itself consistent with the wormhole mechanism, since program intensity and specificity are precisely what suppresses concurrent employment. In the online companion we plot EOS against the share of workers missing a 2016 occupation by education type, quantifying the magnitude of this truncation.

```{r}
#| label: tbl-working-while-studying
#| tbl-cap: "Share of students employed during program, by age and institution type"

tbbl <- cansim::get_cansim("3710010401")|>
  janitor::clean_names()|>
  mutate(age = str_extract(age, "\\d+"))|>
  filter(ref_date=="2015/2016",
         type_of_institution_attended %in% c("College","University"),
         age %in% 18:24)|>
  select(age, type_of_institution_attended, value)|>
  pivot_wider(id_cols = age, names_from = type_of_institution_attended, values_from = value)

tbbl_mean <- tbbl|>
      summarise(age = "Unweighted Mean", across(c(College, University), mean))

bind_rows(tbbl, tbbl_mean)|>
  kbl(
    digits = 1,
    col.names = c("Age", "College (%)", "University (%)"),
    booktabs = TRUE,
    linesep = ""
  ) |>
  kable_styling(latex_options = c("HOLD_position")) |>
 footnote(
  general = "Statistics Canada Table 37-10-0104-01, 2015/2016.",
  general_title = "Source:",
  threeparttable = TRUE,
  escape = FALSE
)|>
#  row_spec(nrow(tbbl), hline_after = TRUE) |>
  row_spec(nrow(tbbl)+1, bold = TRUE)  # bold mean row
```


The specificity regression faces an analogous identification 
challenge to the horse race. EOS and OES are observed and enter the 
regression, but cannot distinguish between two mechanisms that 
predict identical patterns. A specific education opens doors to 
particular distant occupations that a general education does not --- 
the human capital match is narrow, so only a subset of distant 
destinations become reachable following attainment (enablement). At 
the same time, workers with specific educations may face stronger 
incentives to seek out distant occupations where those skills are 
scarce and rewarded, making long-distance moves financially attractive 
relative to staying within a crowded local neighbourhood (motivation). 
Because both channels operate through the same EOS and OES measures, 
the reduced-form estimates conflate the two and cannot be separately 
attributed to either. The regression should therefore be interpreted 
as identifying the reduced-form predictive relationship between 
specificity and distance traveled, without attribution to a specific 
mechanism --- though the estimate remains informative: if either 
channel is operative, the regression will detect it.

## Conclusion

- What we found \[PLACEHOLDER --- awaiting data\]

The central policy takeaway from this analysis is that labour market adjustment mechanisms are not one-size-fits-all: the appropriate intervention depends on which mobility regime contains the occupation experiencing a shortage. The three mobility sub-regimes suggest three distinct policy levers.

In the Skill sub-regime, reskilling and short-cycle training are the natural adjustment mechanisms. Workers in adjacent occupations already possess much of the relevant task content, and targeted reskilling can accelerate the reallocation of labour toward occupations in shortage without requiring formal retraining. The policy implication is that investment in modular, portable training that builds on existing skill profiles will generate the largest mobility response in this segment.

In the Hierarchy sub-regime, the relevant bottleneck is not skill transferability but the hierarchy's silo and platform structure. An occupational shortage within one silo cannot be addressed by upskilling workers from a different silo, even if the occupations share a near-identical skill profile. Instead, shortages must be addressed within-silo, through either lateral movement on a platform or vertical movement between platforms. The best way to address a shortage at a higher platform is by increasing the supply of workers to lower platforms, freeing incumbents to pursue upward movement through the silo — the occupational analogue of the internal labour market adjustment mechanism described by @doeringer1971internal. Bridging programs, mentorship structures, and targeted recruitment into within-silo entry-level positions are the appropriate instruments.

In the Binary sub-regime, either a worker has the required education, or they do not. Neither reskilling nor within-silo movement can substitute for education. Adjustment in this sub-regime depends entirely on expanding formal training pipelines: increasing capacity in the relevant educational programs, reducing barriers to program entry, and shortening time-to-completion where possible. These are also the occupations that immigration policy should target, both in terms of admission and rapid credential recognition. 

If the hypothesized sub-regime structure is confirmed, the geometry of occupational mobility determines which policy lever to pull --- and the wrong lever is not just ineffective but potentially counterproductive. 

```{=latex}
\appendix
```

## O\*NET to NOC 2021 Skill Mapping {#sec-AA}

OaSIS provides occupation-level skill profiles for Canadian
classifications and would be an attractive data source for constructing
$C_{\text{skill}}$. However, OaSIS does not expose how those profiles
are constructed: the mapping between O\*NET and NOC occupations is not
observable, the weights linking underlying occupations are not published, and there is no information on how multiple O\*NET occupations are combined to form a NOC profile. This appendix describes an alternative mapping constructed explicitly from published crosswalks, allowing full transparency of inputs and weights, diagnostic evaluation of mapping quality, and reproducibility of all outputs.

The mapping chains four classifications: O\*NET 2019 $\rightarrow$ SOC
2018 $\rightarrow$ NOC 2016 $\rightarrow$ NOC 2021. Each stage introduces potential one-to-many relationships. A naive approach would assign equal weight to all admissible paths, but this leads to over-weighting of occupations that appear frequently in crosswalk chains and inclusion of unrelated occupations. Instead, path weights are constructed to reflect the degree of splitting at each stage: if a source occupation splits into $n_s$ destination occupations at stage $s$, each path through that stage receives weight $w_s = 1/n_s$, and the overall path weight is $w_{\text{path}} = w_1 \times w_2 \times w_3$. Path weights are then aggregated by origin O\*NET occupation and destination NOC 2021 code and normalized to sum to one within each NOC to yield base weights. The consequences of naive versus path weighting can be severe: for NOC 72600 (Air pilots, flight engineers and flying instructors), naive weights assign five-eighths of the skill profile to agricultural workers, while path weights assign 78\% to pilots.

A power transformation and renormalization is then applied to further
reduce the influence of mapping artifacts: $\tilde{w}_k \propto w_k^2$.
This concentrates weight on the dominant O\*NET contributors and
diminishes the influence of peripheral occupations that enter through
crosswalk chains with many splits. For NOC 72600, downweighting
increases the pilot share of the skill profile to 96\%. Each O\*NET
occupation is then represented as a point in a reduced skill space using the first 10 principal components of the O\*NET skill content variables, and for each NOC 2021 code the weighted centroid of its contributing O\*NET occupations is computed using the downweighted mapping, yielding a 10-dimensional skill profile for each NOC occupation.

Two diagnostics assess the quality of each NOC's mapping in skill space. Herfindahl concentration measures how concentrated the mapping weights are across contributing O\*NET occupations: $H = \sum_k \tilde{w}_k^2$. For example, a Herfindahl of $\frac{2}{3}$ indicates that the NOC skill profile is effectively dominated by a single O\*NET occupation with a weight of $\approx .8$.  In contrast, when $H=\frac{1}{3}$ the largest contributor's share is between $\frac13$ and $\sqrt{\frac13}=.58$. Scaled Similarity measures the internal coherence of the contributing occupations in skill space: the weighted average distance of each contributing O\*NET occupation from the NOC centroid is computed, then normalized to a $[0.05, 1]$ interval by dividing by the maximum value across all NOCs, multiplying by 0.95, and subtracting from 1, so that $SS = 1 - 0.95 \times \bar{d} / \max_j \bar{d}_j$. The lower bound of 0.05 is chosen to match the approximate lower bound of the Herfindahl index, making the two measures' floors commensurable. The mapping quality composite is the product $H \times SS$, which is high only when the mapping is both concentrated and internally coherent. The three interpretation cases are: high $H$ with any $SS$ (no problem --- almost all weight on a single O\*NET occupation); low $H$ with high $SS$ (no problem --- diffuse but coherent); and low $H$ with low $SS$ (problematic --- a diffuse average of disparate occupations whose skill profile is unreliable). NOC 41200 (University professors and lecturers) is a canonical example of the third case.

Occupations in the bottom decile of $H \times SS$ are excluded
from the local clean and non-local clean samples. Because the exclusion criterion is based on mapping quality rather than occupation characteristics, it
introduces noise into $C_{\text{skill}}$ selectively while leaving
$C_{\text{hierarchy}}$ and $C_{\text{binary}}$ unaffected --- a property exploited in the mapping quality falsification design described in @sec-data. The full mapping pipeline is implemented in
[@onet_to_noc] and is constructed entirely from O\*NET data and published crosswalks.

## 2016 TEER mismatch {#sec-AB}

Admissible TEERs are the Target TEER $\pm1$ plus TEER 0 (management), where the assumption is that management roles are independent of educational attainment. 

\begin{table}[H]
\centering
\begin{tabular}{lcr}
\toprule
Highest Attainment & Target TEER & Admissible TEERs \\
\midrule
No certificate, diploma or degree & 5 & 4, 5, 0 \\
Secondary (high) school diploma & 4 & 3, 4, 5, 0 \\
Apprenticeship or trades certificate or diploma & 3 & 2, 3, 4, 0 \\
College / CEGEP / other non-university & 2 & 1, 2, 3, 0 \\
University certificate or diploma below bachelor & 2 & 1, 2, 3, 0 \\
Bachelor's degree & 1 & 1, 2, 0 \\
Above bachelor (Master's, PhD, MD, etc.) & 1 & 1, 2, 0 \\
\bottomrule
\end{tabular}
\caption{Admissible TEERs by highest attainment level}
\label{tab:teer_mismatch}
\end{table}

## Normalization Robustness  {#sec-AC}

In the main body of the paper, we anchor all distance metrics at hierarchical distance 3 — the maximum within-silo horizontal distance, representing movement within a broad occupational group without crossing TEER levels. In this appendix we examine robustness to anchoring at hierarchical distance 2 (same sub-major group, different minor group) and hierarchical distance 4 (same broad group, one TEER step), representing one step down into shorter horizontal movement and one step up into vertical movement respectively. The unconditional percentile locates each anchor in the full pairwise distribution. 

```{r}
#| echo: false
calibration |>
  mutate(
    p_uncond = scales::percent(p_uncond, accuracy = 0.1),
    h_anchor = as.integer(h_anchor),
    s_anchor = round(s_anchor, 2)
  ) |>
  rename(
    `Hierarchical Anchor` = h_anchor,
    `Unconditional Percentile` = p_uncond,
    `Skill Anchor` = s_anchor
  ) |>
  kable(
    format = "latex",
    caption = "Normalization anchors under alternative calibration quantiles",
    align = c("c", "c", "c"),
    booktabs = TRUE
  ) |>
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE
  )
```

## Wages vs Dual Potentials  {#sec-wd}

The EROT dual potentials $\log u_i$ and $\log v_j$ are Lagrange multipliers on the origin and 
destination marginal constraints respectively. They absorb everything about an occupation that 
affects its aggregate attractiveness as a source or destination \textit{given} the cost structure 
--- that is, the component of occupational desirability that operates at the level of origins or 
destinations alone rather than through origin--destination pairing. In a competitive labour market, 
wages are the price that clears occupation-level supply and demand, and are therefore a natural 
candidate for an occupation-level correlate of the dual potentials.

The theoretical relationship is not unambiguous in sign across the full occupational distribution. 
A high $\log u_i$ indicates that an occupation generates more outflows than the cost structure 
alone predicts. This could reflect low wages driving workers out, but equally a well-connected 
occupation in skill or hierarchical space --- one with many cheap transitions available --- 
generating large outflows for structural rather than compensatory reasons. A high $\log v_j$ 
indicates that an occupation attracts more inflows than the cost structure predicts, which could 
reflect high wages as a pull factor, but also low entry costs generating large inflows at the base 
of a silo regardless of wage level. In both cases the potential conflates the wage signal with 
structural features of the occupation's position in the mobility graph, and the relationship with 
wages is an empirical question rather than a theoretical prediction.

We estimate the following regressions at the occupation level:

\begin{equation}
    \log(w_j) = \alpha + \beta_u \log u_j + \epsilon_j
\end{equation}

\begin{equation}
    \log(w_j) = \alpha + \beta_v \log v_j + \epsilon_j
\end{equation}

\noindent where $w_j$ is the 2019 occupational median wage. 2019 wages are used as the wage 
measure for two reasons: they represent the midpoint of the 2016--2021 transition window, and 
they avoid the transitory disruption to occupational wage structures introduced by the COVID-19 
pandemic in 2020--2021. Both regressions are estimated by WLS with weights proportional to 2016 
occupation size, separately for each cost matrix and across the full $\varepsilon$ grid. The 
$\varepsilon$ stability of $\beta_u$, $\beta_v$, and their $R^2$ values is the primary object 
of interest.

What is theoretically cleaner is precisely this stability claim. The dual potentials are 
determined by the marginal constraints, which are held fixed at observed occupation stocks 
regardless of the regularization level. As $\varepsilon$ varies, the interior structure of the 
transport plan changes --- mass is redistributed across origin--destination pairs --- but the 
potentials must always rationalize the same aggregate occupation flows. Stable $\beta$ estimates 
across $\varepsilon$ therefore indicate that the relationship reflects the structural content of 
the occupation-level flow constraints rather than a feature of the transport plan's geometry at 
a particular scale. Instability would conversely suggest the correlation is a regularization 
artifact.

## References
