---
title: "Occupational Mobility Is Not One Thing: Evidence from the Canadian Labour Market" 
author: "Richard Martin"
date: today
format:
  pdf:
    documentclass: article
    number-sections: true
    geometry: margin=1in
bibliography: references.bib
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \DeclareMathOperator*{\argmin}{argmin}
editor: 
  markdown: 
    wrap: 72
---

```{r, include=FALSE}
library(tidyverse)
library(here)
library(knitr)
library(kableExtra)
library(patchwork)
library(conflicted)
cdfs <- read_rds("out/dist_cdfs.rds")
calibration <- read_rds(here("out", "calibration.rds"))
```

## Pre-registration Notice

This paper is posted as a pre-registration of study design, prior to Statistics Canada Virtual Data Lab access.  Anything that looks like a result should be interpreted as a prior. Pre-registration is intended to defend against claims of post-hoc rationalization.

## Abstract

Workers predominantly move between proximate occupations, but proximity means different things in different parts of the labour market. Using entropic regularized optimal transport to conduct a horse race among three distance metrics — skill, hierarchy, and binary — we show that no single geometry governs occupational mobility. Instead, the appropriate metric depends on each occupation's education specificity: skill where skills are broadly transferable (low specificity), hierarchy where silos and platforms structure mobility (medium specificity), and binary where human capital is so occupation-specific that all alternative occupations appear equally distant (high specificity). Each regime implies a distinct policy lever for addressing labour shortages.

## Introduction

Occupational mobility is not one thing. Workers tend to move between proximate occupations, but proximity is not self-evident: occupations differ along multiple dimensions simultaneously — the skills they require, their position within a hierarchy, and the degree to which they selectively admit workers with particular educational backgrounds. Each implies a different geometry of the labour market, and different predictions about who moves where. We test which geometry governs mobility — and find that the answer depends on where in the labour market you look.

We separate local mobility — transitions that occur without new highest educational attainment — from non-local mobility — transitions that occur when a new highest educational attainment is achieved. For local mobility, distance is the explanatory variable: we ask which of three candidate metrics (skill, hierarchy, and binary) best rationalizes observed occupational transitions. For non-local mobility, the role of distance is reversed: rather than asking which metric explains where workers go, we ask what determines how far they travel, treating transition distance as the outcome and measures of education and occupation specificity as its determinants.

We anticipate that no single metric will dominate globally. A labour market 
in which doctors, managers, and labourers all follow the same distance geometry would be surprising. To that end, we construct a measure of an occupation's education specificity (OES) --- using KL divergence to capture how different an occupation's education distribution is from that of the labour market as a whole --- and split 504\footnote{8 occupations can not be mapped to O*NET skill data.} occupations defined by Canada's National Occupational Classification (NOC) into three tertile sub-regimes, mapping to our three hypothesized cost metrics.

OES is in effect an empirical proxy for where an occupation sits on the general-to-specific human capital spectrum [@becker1964human]: occupations with low OES draw broadly from the education space, revealing general human capital requirements, while occupations with high OES draw narrowly, revealing a specific human capital requirement that is acquired through a particular education.

The tertile boundaries are not theoretically derived: the general-to-specific human capital spectrum is continuous and the true transition points between regimes are unknown. If the three-regime structure is a genuine feature of the labour market, the horse race results should be robust to moderate misclassification at the boundaries --- occupations near the tertile thresholds may be incorrectly assigned, but the bulk of the distribution will be correctly classified and the predicted metric dominance should survive.

The three sub-regimes are named after their hypothesized winning distance 
metric --- Skill, Hierarchy, and Binary --- rather than after the education 
specificity dimension that defines them. This naming convention is adopted 
for clarity: the regime name directly signals the predicted geometry of 
mobility, making the horse race hypotheses easier to track throughout the 
paper. The naming does not presuppose the result; the hypotheses are 
pre-committed predictions that the data may or may not confirm.

The Skill sub-regime (low specificity) is hypothesized to be governed by 
skill distance: skills are generic and mobility is not constrained by 
hierarchical structure. The Hierarchy sub-regime (middle specificity) is 
hypothesized to be governed by hierarchical distance: skills are specific 
enough that hierarchical structure binds, with mobility predominantly 
occurring within labour market silos defined by the NOC broad group. The 
Binary sub-regime (high specificity) is hypothesized to be governed by 
binary switching costs: human capital is so occupation-specific that the 
richer distance structure of the other metrics is counterproductive, and 
mobility is better described as either staying or, rarely, leaving 
at equal cost regardless of destination.

The policy relevance is direct. Labour shortages in different segments of the market require different adjustment mechanisms, and the right mechanism depends on which distance metric governs mobility in that sub-regime. Shortages in the Skill sub-regime may be addressed through reskilling, building on transferable skills. Shortages in the Hierarchy sub-regime are best addressed by increasing 
supply at the bottom platform of the relevant silo, facilitating  upward movement through the silo to address shortages at higher levels. Shortages in the Binary sub-regime cannot be addressed through either of the above mechanisms: the relevant human capital is acquired only through specific education programs. Thus interventions must target access to and capacity within those programs directly.  Getting the geometry right is therefore not merely a methodological contribution — it determines which policy lever to pull.

For local mobility, we use entropic regularized optimal transport (EROT) to conduct the horse race. We evaluate EROT fit using two complementary measures: relative KL divergence, which captures whether the predicted transition matrix assigns appropriate probability mass to observed transitions; and relative Cost-Weighted Total Variation Distance (CWTVD), which complements this by weighting misallocated mass by a neutral cost matrix so that errors involving geometrically distant pairs are penalized more heavily than errors involving nearby ones. A model that scores well on both has correctly allocated mass and allocated it to the right part of occupational space.

[FINDINGS PLACEHOLDER]

## Conceptual Framework

The central structural choice in this paper is the partition of occupational transitions into two qualitatively distinct regimes. A local transition occurs when a worker changes occupation without acquiring new educational attainment --- the worker's highest educational attainment in 2021 is the same as in 2016. A non-local transition occurs when a worker's highest educational attainment increases between the two census years.  This partition reflects a fundamental difference in the mechanism governing mobility. For local transitions, the 2016 occupation is an informative starting point: human capital is relatively stable over the inter-census period, and distance from that starting point governs destination choice. For non-local transitions, the 2016 occupation is largely uninformative: a new highest educational attainment shifts the worker's position in occupational space in a way that is not captured by the 2016 occupation, acting as a wormhole --- enabling discontinuous jumps that would otherwise be difficult to reconcile with their 2016 occupation.

Two symmetric specificity measures play central roles in the analysis, one defined over occupations and one over \textit{education}: the cross product of field of study (CIP) and highest level of attainment. These measures are derived from Statistics Canada Table 98-10-0404-01 --- a one-quarter sample of the Canadian population, of which at most one-quarter overlaps with the linked microdata used for estimation.

To calculate an education's occupation specificity (EOS), we begin by calculating the KL divergence of $P(\text{NOC}|\text{education})$ from $P(\text{NOC})$. A higher value indicates that an education is distinctive in its occupation proportions. Because the empirical distribution can only place mass on a limited number of occupations for niche educations, KL divergence has a mechanical floor; we therefore residualize $\log(\text{KL})$ on $\log(T)$ where $T$ is the total number of workers with that education. The residual from this regression is defined to be EOS.  This measure enters only the non-local partition as a continuous regressor proxying for wormhole entrance intensity 
--- a high EOS value increases the probability of \emph{any} jump.

A symmetric procedure is followed to calculate an occupation's education specificity (OES). We begin by calculating the KL divergence of $P(\text{education}|\text{NOC})$ from $P(\text{education})$.  A higher value 
indicates that an occupation is distinctive in its education proportions. Because the empirical distribution can only place mass on a limited number of educations for niche occupations, KL divergence has a mechanical floor; we therefore residualize $\log(\text{KL})$ on $\log(T)$ where $T$ is the total number of workers with that occupation. The residual from this regression is defined to be OES. In the local partition OES is used as a classifier defining the three sub-regimes; in the non-local partition it serves as a continuous regressor proxying for wormhole exit intensity --- conditional on a high EOS, 
a high OES increases the probability of a \emph{long} jump.  Occupational licensing [@kleiner2013labor] represents an explicit form of OES; licensing lists capture credential selectivity directly, but rely on curated registries that are often incomplete, jurisdiction-specific, and require manual updating. OES captures the same underlying phenomenon from the equilibrium education mix observed in the data, without requiring a curated list.

Occupations are assigned to sub-regimes based on OES tertile, subject to a TEER override rule. The Skill sub-regime includes occupations with bottom tertile education specificity, expanded to include all TEER 0 (senior management) and TEER 5 (labour) occupations regardless of specificity tertile. This override reflects two distinct mechanisms: TEER 0 occupations exhibit portable managerial human capital that is not confined to a single occupational silo — a school principal or library manager can move laterally across organizational contexts in a way that a physician or dentist cannot; TEER 5 occupations have sufficiently low barriers to entry that education specificity does not meaningfully constrain mobility. The Hierarchy sub-regime includes the remaining middle tertile occupations, where education specificity is sufficient to generate silo structure but not so high as to foreclose within-silo movement across platforms. The Binary sub-regime includes the remaining top tertile occupations.

Workers in Binary sub-regime occupations find themselves marooned: their human capital is too specific to support further local mobility. Workers in the Skill sub-regime, by contrast, can move horizontally across occupational space, building on transferable skills. The wormhole and marooned metaphors thus describe the two ends of the same mechanism: acquiring a highly specific education places a worker in the Binary sub-regime, predicting both an initial long-distance jump and subsequent immobility.

The local/non-local partition has a natural falsification structure. If the 
partition is meaningful, distance metrics should improve on the independence 
baseline within the local regime but not when applied out of regime to 
non-local transitions --- where new educational attainment, not distance, is 
doing the explanatory work. Conversely, the specificity regression should 
predict distance within the non-local regime but fail to do so out of regime, 
where human capital is relatively stable and movement is local. Each model is 
tested both where it should work and where it should not, and the predicted 
double dissociation [@shadish2002experimental] provides stronger evidence 
for the causal interpretation of the underlying mechanisms than either 
falsification alone would. A single confounder would have to operate in 
opposite directions across regimes to account for both dissociations; multiple 
confounders would each have to align fortuitously with the predicted pattern. 
Neither is impossible, but both are less parsimonious than the theoretical 
mechanism itself. The design does not eliminate the possibility of confounding,
but it raises the bar considerably: the model's validity rests not on the 
absence of confounders but on the implausibility of a story that generates the 
predicted pattern of double dissociation across regimes.

The natural alternative to EROT for evaluating occupational mobility costs is the gravity model [@anderson_vanwincoop_2003; @silva2006log], which estimates origin and destination fixed effects jointly with the cost parameter(s). This joint estimation is precisely the problem in our context: fixed effects absorb variation in mobility rates that may reflect true cost structure, allowing misspecified cost matrices to achieve good fit by compensating through the fixed effects. Using a gravity model would be akin to randomly hobbling the horses before the race. EROT avoids this issue by holding the marginal structure fixed at observed values, separating the question of how much mobility occurs — determined by the margins — from the question of where mobility flows — shaped by the cost matrix. EROT is described in detail in Section \ref{sec-local}.

## Cost Metrics

The skill distance matrix $C_{\text{skill}}$ is constructed from 161 O\*NET
measures spanning skills, abilities, knowledge, and work activities,
mapped to NOC 2021 occupations via a purpose-built crosswalk [@onet_to_noc]. 
Each NOC is represented as a weighted average of its constituent O\*NET
occupations, with weights determined by the crosswalk mapping.

O\*NET provides two scales for each measure: Importance (how important
the attribute is to the occupation) and Level (the degree to which the
attribute is required). Both scales are standardized to a common 0--10
range and combined via geometric mean, yielding a single score that
reflects both the relevance and the intensity of each attribute. The
geometric mean is preferred over the arithmetic mean as it penalizes
cases where one scale is near zero — an attribute that is important but
required at negligible level, or required at high level but deemed
unimportant, receives a lower composite score than one scoring
moderately on both dimensions. The two scales are highly correlated in
practice, so this choice is unlikely to be consequential for the
resulting distance matrix, but we note it for transparency.

Raw O\*NET measures are standardized prior to dimensionality reduction.
Principal components analysis is applied to the 161-dimensional skill
profiles, and the number of components retained is determined
empirically: we identify the minimum k such that the Spearman rank
correlation between pairwise distances in the k-dimensional PCA space
and pairwise distances in the full 161-dimensional space exceeds 0.99.
This criterion — rank preservation of pairwise distances rather than
variance explained — is chosen because the object of interest is the
distance matrix itself, not the underlying skill dimensions. Twelve
principal components satisfy this criterion.

The skill distance between any two occupations is then the Euclidean
distance between their coordinates in the retained 12-dimensional PCA
space. An interactive heatmap of the resulting skill distance matrix 
can be found 
[here](https://rpmartin.github.io/not-one-thing-companion/#sec-skill-heatmap).

A note on the crosswalk is warranted. An alternative approach would be
to use OaSIS occupation-level skill profiles. We construct our own
crosswalk for a specific reason: falsifiability. The horse race requires
that measurement error in the skill distance matrix be attributable to a
known source — namely, imprecision in the O\*NET to NOC mapping. OaSIS
obscures the structure of that mapping, making it impossible to
characterize where and how measurement error enters $C_{\text{skill}}$.
Our crosswalk makes this structure transparent and therefore makes the
identification strategy falsifiable. 

The hierarchical distance matrix $C_{\text{hier}}$ is derived directly
from the NOC 2021 taxonomy, treating the occupational classification
structure as a tree metric. Distance between any two occupations is
defined as follows: zero if all five digits match (same unit group); one
if the first four digits match (same minor group, different unit group);
two if the first three digits match (same sub-major group, different
minor group); three plus the absolute difference in TEER level if only
the first digit matches (same broad occupational category, different
sub-major group); and nine for any cross-category transition. The
absolute TEER difference ranges from zero to five, so within-category
distances range from three to eight, preserving a strict separation from
the cross-category sentinel value of nine.

Intuitively, the hierarchical view of the labour market is that it is comprised of ten largely separate silos, each containing up to six platforms corresponding to TEER levels. Each platform has internal geography --- occupations within the same silo \emph{and} platform are at distances of one, two, or three depending on how closely related they are within the NOC hierarchy. Moving between platforms always costs more than any within-platform move: inter-platform distances start at four and increase with the number of platforms traversed, up to eight. Moving between silos always costs more than any within-silo move: cross-silo transitions are assigned a sentinel value of nine, the minimum integer exceeding the maximum within-silo distance of eight. An interactive heatmap of the resulting hierarchical distance matrix can be found 
[here](https://rpmartin.github.io/not-one-thing-companion/#sec-hier-heatmap).

The binary switching cost matrix $C_{\text{binary}}$ assigns a cost of zero 
to remaining in the same occupation and one to any transition. It is a 
limiting case, even more extreme than the hierarchical distance --- 
topologically, the labour market is comprised of 504 distinct silos, each 
with no internal structure of any kind. $C_{\text{binary}}$ serves two roles in the analysis. First, it is the theoretical straw man in the global horse race: if occupational mobility costs are purely a function of whether a switch occurs rather than how far, then $C_{\text{binary}}$ should rationalize observed mobility patterns better than $C_{\text{skill}}$ or $C_{\text{hier}}$. Second, and more subtly,  $C_{\text{binary}}$ is best suited to settings where mobility is limited --- where most workers stay rather than move. In such settings, the information contained in a richer cost structure can cause model fit to worsen. This second role motivates the Binary sub-regime hypothesis: in occupations where human capital is so specific that all exits appear equally distant, binary is not merely the straw man but the correct description of the mobility geometry.

## Cost Metric Normalization

The three cost matrices are defined on different scales by construction:
skill distances are continuous and spread over a wide range, hierarchical
distances are discrete with a sentinel value of nine reserved for
cross-category transitions, and binary distances take only the values
zero and one. Direct comparison in the horse race requires that the matrices be anchored to a common scale; without normalization, a given value of $\varepsilon$ implies very different levels of concentration for each matrix, making the horse race results uninterpretable across the $\varepsilon$ grid.

We normalize skill and hierarchical distances to a common scale using a
quantile-matching approach, with binary serving as the reference. The
normalization proceeds as follows. Cross-category sentinel values of nine
are excluded from the hierarchical distribution, as they represent a
qualitatively distinct mobility regime rather than a point on the
continuous within-category distance scale. The conditional median of the
remaining hierarchical distances is four. We then locate four in the
unconditional hierarchical distribution — including the sentinel values —
which falls at the eighth percentile, reflecting the dominance of
cross-category pairs in the full pairwise distribution. Finally, we
locate the eighth percentile of the skill distance distribution, which is
8.59. Hierarchical distances are divided by four and skill distances by
8.59, anchoring both matrices at the same quantile of their respective
unconditional distributions.

```{r}
#| warning: false
#| echo: false
#| label: fig-cdfs
#| fig-cap: "Cost Matrix Normalization: Raw vs. Normalized CDFs"
cdfs
```

@fig-cdfs shows the empirical CDFs of all three matrices before
and after normalization. The raw distributions differ substantially in
scale and shape; after normalization, skill and hierarchical CDFs overlap
closely over the range of distances relevant to observed mobility,
diverging only in the right tail. The sensitivity of horse race results
to the normalization anchor is examined in @sec-AC

\pagebreak

## Mobility Data {#sec-data}

The analysis uses the Census 2016--2021 longitudinal file, constructed by linking long-form respondents across the two census cycles via the Derived Record Depository (DRD), Statistics Canada's national longitudinal database of personal identifiers. Linkage is probabilistic; details of the matching keys and overall linkage rate will be verified upon data access and reported in the final paper.

The estimation sample is partitioned into subsets that vary in two dimensions:
whether the local/non-local regime assignment is correct, and whether occupation measurement is clean. This design generates a natural ordering of expected model fit that serves as an internal validation of the partition's theoretical content.

The local clean sample is the primary identification sample for the horse race, and the falsification sample for the specificity regression. It comprises individuals with an observed occupation in both 2016 and 2021, no increase in highest educational attainment between the two census years, no self-employment in 2016, no TEER--attainment mismatch in 2016, and an \textsc{o*net}-to-\textsc{noc} mapping quality score above the bottom decile. This is the sample where the horse race should work best and where the specificity regression should find nothing.

The non-local clean sample comprises individuals who acquire a new highest educational attainment between 2016 and 2021 and report an observed occupation in both periods, subject to the same measurement restrictions: no self-employment in 2016, no TEER--attainment mismatch, and an \textsc{o*net}-to-\textsc{noc} mapping quality score above the bottom decile. It is the primary estimation sample for the specificity regression, and the falsification sample for the horse race. This is the sample where the specificity regression should work best and where the horse race should find nothing. 

The mismeasurement diagnostic samples introduce measurement problems one at a time, holding regime assignment correct. The occupational mismeasurement sample retains individuals whose 2016 occupation is likely mismeasured, identified by two criteria. Self-employed workers are included in the mismeasurement partition on the grounds that self-employment is associated with role multiplicity — owners simultaneously performing managerial, technical, and administrative functions — making the reported NOC an unreliable anchor for measuring occupational distance. The second is TEER--attainment mismatch, where the reported occupation is inconsistent with highest educational attainment. Attainment mismatch could easily occur for either recent attainments or for those awaiting credential recognition.\footnote{For details regarding TEER--attainment mismatch criteria, see Appendix B.} Both criteria introduce noise into the origin occupation, and their effects on model fit are expected to be symmetric across cost matrices.

The mapping quality sample retains individuals whose 2016 occupation falls in the bottom decile of the \textsc{o*net}-to-\textsc{noc} mapping quality composite; poor mapping introduces noise specifically into $C_{\text{skill}}$ while leaving $C_{\text{hier}}$ unaffected, providing a difference-in-differences diagnostic on metric-specific fit degradation. Each of these samples is expected to show worse fit than the corresponding clean sample, with the mapping quality sample showing selective degradation of $C_{\text{skill}}$ relative to $C_{\text{hier}}$.

The mismeasurement residual samples comprise all observations not assigned to any of the named samples above, split by regime. The local mismeasured residual contains individuals in the local regime with multiple simultaneous measurement problems --- self-employment, TEER--attainment mismatch, and/or poor \textsc{o*net}-to-\textsc{noc} mapping in combination. The non-local mismeasured residual contains individuals in the non-local regime with multiple simultaneous measurement problems. Within each regime the source of fit degradation is measurement noise rather than regime misassignment, preserving the interpretability of the regime axis while acknowledging that the specific contribution of each noise source cannot be separately identified. Expected fit is worse than the corresponding single-problem diagnostic samples but, crucially, better than the out-of-regime clean samples --- where the only problem is regime misassignment. If that ordering holds, it constitutes direct evidence that regime misassignment is more damaging to model fit than the accumulation of measurement problems, strengthening the theoretical case for the local/non-local partition.

Two sample restriction mechanisms introduce a potential selection concern for the horse race. Non-linkage and non-observability of occupation are both
disproportionately concentrated among low-attachment workers --- individuals with casual, seasonal, or intermittent employment whose identifying information is also less stable across census cycles --- so the two filters largely co-select the same low-skill population rather than introducing independent exclusions. This shared exclusion affects the representativeness of population-level transition estimates but does not threaten the internal validity of the horse race. For the horse race results to be biased, the excluded workers' transitions would need to be systematically more or less rationalizable by a particular cost matrix than those of their observed
counterparts in the same region of occupational space. No such mechanism is apparent: exclusion is driven by attachment instability and identifier churn, neither of which has a clear relationship to how well transitions conform to any particular cost structure conditional on occupation. The residual unlinked population among otherwise-observable workers --- recent immigrants, young workers, and high-movers--- similarly lacks any obvious occupational-space concentration that would differentially favour or penalize a particular metric.

The non-local partition is restricted to individuals who acquire a new highest educational attainment between 2016 and 2021 and report an occupation in both periods. The selection concern here is distinct from the local partition: the excluded population is not disproportionately low-skill, but rather disproportionately those whose human capital increased the most. The most demanding programs — medicine, law, dentistry— require sufficiently intensive study that concurrent employment is largely precluded, meaning students in these programs are less likely to report a 2016 occupation and thus are excluded from the sample.

This selection mechanism has a specific implication for the non-local
partition's central empirical question: the relationship between education and occupation specificities, and distance traveled. The most
demanding programs tend to also be the most specific, and the human
capital increment they confer is large — generating the longest
new attainment driven occupational jumps. Excluding these observations
therefore truncates the upper tail of the distance distribution
precisely where education specificity is highest. Estimates of the
relationship between education specificity and distance traveled should
accordingly be interpreted as lower bounds on the true relationship in
the full population of new educational attainment.

## Local Model {#sec-local}

The entropy-regularized optimal transport problem minimizes the following
Lagrangian, where the first term is the transport cost, the second is
the entropic regularization penalizing deterministic transport plans,
and the third and fourth enforce the marginal constraints $\sum_j
P_{ij} = a_i$ and $\sum_i P_{ij} = b_j$ via dual variables $f_i$ and
$g_j$.

$$\mathcal{L} = \sum_{ij} P_{ij} C_{ij} + \varepsilon \sum_{ij} P_{ij}(\log P_{ij} - 1) + \sum_i f_i(a_i - \sum_j P_{ij}) + \sum_j g_j(b_j - \sum_i P_{ij})$$

where $a_i$ and $b_j$ are the exact occupation stocks in 2016 and 2021
respectively. The framework is implicitly first-order Markov: the
margins absorb any factor that operates additively at the level of the
origin or destination occupation, but not factors that depend on the
specific origin-destination pair. The cost matrix is itself dyadic ---
it encodes the hypothesis that the specific pairing matters --- and the
horse race is asking which specification of dyadic structure best
captures the non-Markovian residual after the margins have absorbed the
first-order effects.

The first-order conditions yield:

$$\log P_{ij} = \frac{f_i}{\varepsilon} + \frac{g_j}{\varepsilon} - \frac{C_{ij}}{\varepsilon}$$

Exponentiating, the optimal transport plan takes the form $P^\star_{ij}
= u_i \, e^{-C_{ij}/\varepsilon} \, v_j$, where $u_i =
e^{f_i/\varepsilon}$ and $v_j = e^{g_j/\varepsilon}$ are scaling
vectors. Strict convexity of the regularized objective guarantees a
unique solution. The solution is therefore a row- and column-rescaling
of the Gibbs kernel $K_{ij} = e^{-C_{ij}/\varepsilon}$ where the Sinkhorn algorithm recovers it by alternating row and column normalization until both marginal constraints are satisfied to within a specified tolerance. All computations use a log-domain Sinkhorn solver
for numerical stability, implemented in the \texttt{safesink} R package
[@safesink].

The regularization parameter $\varepsilon$ has a natural interpretation
in the random utility framework. The optimal transport plan $P^\star_{ij}
= u_i \, e^{-C_{ij}/\varepsilon} \, v_j$ is the static equilibrium of a
discrete choice model in which workers draw idiosyncratic utility shocks
scaled by $\varepsilon$ over destination occupations. Low values of
$\varepsilon$ concentrate the transport plan on cost-minimizing
assignments, approaching a deterministic reallocation in the limit; high
values diffuse mobility across destinations, approaching the independence
baseline $P_{ij} = a_i b_j$ as $\varepsilon \to \infty$. The RUM
analogy governs the interior structure of the transport plan --- how
probability mass is allocated across origin-destination pairs --- but
does not extend to the margins. In a standard logit model the choice
probabilities are outcomes of utility maximization with no constraint on
their aggregates; in EROT the origin and destination occupation
distributions are hard constraints that the transport plan must exactly
reproduce. This distinction is methodologically consequential: because
the margins cannot adjust to absorb cost misspecification, a poorly
chosen cost matrix cannot hide behind fixed effects or scale
recalibration --- its misfit appears directly in the objective.

A critical identification constraint follows directly from the form of
the optimal transport plan: only the ratio $C_{ij}/\varepsilon$ enters
$P^\star$. The scale of the cost matrix and the value of $\varepsilon$
are therefore not separately identified. This means that results
reported at a single value of $\varepsilon$ are uninterpretable without
an implicit normalization assumption --- and that assumption can be
gamed, since any cost matrix can be made to fit well by tuning
$\varepsilon$ to its scale. A horse race conducted at a single value is not a horse race; it is a calibration exercise.

Evaluating across a grid of $\varepsilon$ values sidesteps this by looking 
for dominance that is robust across scales — a metric that wins at multiple 
values cannot be attributed to a fortuitous scale alignment. 
We evaluate across a log-spaced grid of $\varepsilon$ values ranging from 
0.5 to 2, spanning from relatively concentrated to relatively diffuse transport plans while remaining in the region where the fit metrics retain discriminating power. The criterion for a robust winner requires some care. At low values of
$\varepsilon$, cost-sensitive matrices concentrate probability mass too
aggressively relative to observed mobility, incurring an overconfidence
penalty that mechanically favours the flat binary matrix regardless of
which metric truly best rationalizes mobility. This low-$\varepsilon$
binary lead is therefore uninformative about genuine descriptive
accuracy and should not be interpreted as evidence that binary is the
superior metric. The operative criterion is dominance over the full
epsilon grid after discounting this artefact: a cost matrix is declared
the winner if it dominates all competitors across the sweep, or
dominates all non-binary competitors across the sweep while losing to
binary only at low values of $\varepsilon$ where the overconfidence penalty is
operative. Conversely, when binary genuinely best rationalizes mobility
--- as in sub-regimes where occupational transitions are not structured
by continuous distance --- it will dominate not just at low values of $\varepsilon$, but across the entire grid and by a margin that dwarfs the overconfidence effect. The $\varepsilon$ grid thus serves a dual purpose: identifying the winning metric and distinguishing genuine binary dominance from penalty artefact.

The horse race is evaluated using two complementary fit metrics, both
normalized relative to the independence baseline $P^{\text{ind}}_{ij} =
a_i b_j$, so that 0 indicates no improvement over independence, 1
indicates perfect fit, and negative values indicate a transport plan that
fits observed mobility worse than independence.

The first metric is relative KL divergence improvement:

$$\Delta KL = \frac{KL(P^{\text{obs}} \parallel P^{\text{ind}}) - KL(P^{\text{obs}} \parallel \hat{P})}{KL(P^{\text{obs}} \parallel P^{\text{ind}})}$$

where $KL(P \parallel Q) = \sum_{ij} P_{ij} \log(P_{ij} / Q_{ij})$.
This measures the share of the divergence between observed transitions
and independence that is explained by the fitted transport plan. KL
divergence is sensitive to probability mass wherever it is misallocated,
but is geometry-blind: misallocating mass to a nearby occupation is
penalized identically to misallocating it to a distant one.

The second metric is relative Cost-Weighted Total Variation Distance
(CWTVD) improvement [@villani2009optimal; @rubner2000earth]:

$$\Delta CWTVD = \frac{CWTVD(P^{\text{obs}}, P^{\text{ind}}) - CWTVD(P^{\text{obs}}, \hat{P})}{CWTVD(P^{\text{obs}}, P^{\text{ind}})}$$

where $CWTVD(P^{\text{obs}}, \hat{P}) = \sum_{ij}
\frac{|P^{\text{obs}}_{ij} - \hat{P}_{ij}|}{2} \cdot C_{ij}$. CWTVD
extends classical total variation distance by weighting absolute
residuals elementwise by a cost matrix, penalizing probability mass
misallocated to geometrically distant origin-destination pairs more
heavily than mass misallocated to nearby ones. Unlike KL divergence,
CWTVD respects the metric structure of the outcome space; unlike the
Wasserstein distance, it does not solve an additional optimal transport
problem over the residual mass, making it computationally trivial given
an existing transport plan.

The scoring cost matrix $C$ used in CWTVD is the arithmetic mean of the
three candidate normalized cost matrices: $C_{\text{skill}}$,
$C_{\text{hier}}$, and $C_{\text{binary}}$. Using a neutral common metric
ensures that no candidate model is advantaged by the choice of scoring
matrix. The two metrics are complementary by design: KL divergence
captures overall probability mass misallocation regardless of geometry,
while CWTVD captures whether misallocation is concentrated in nearby or
distant pairs. A transport plan that scores well on both metrics fits
observed mobility both accurately and geometrically.

## Non-Local Model {#sec-non_local}

The non-local model estimates the conditional-on-positive relationship between 
$\log(d_{ij})$ --- the log distance between 2016 and 2021 occupations --- and 
the specificity measures EOS and OES. The outcome is computed separately using 
skill and hierarchical distance; the wormhole mechanism should operate 
independently of which metric measures the jump, providing a robustness check 
on the distance measure choice.

The theoretical mechanism is asymmetric. A high value of $\text{EOS}$ is the 
primary condition for a wormhole jump --- the entrance is logically prior to 
the exit. Conditional on a sufficiently high $\text{EOS}$, $\text{OES}$ 
determines the distance of the jump. A full specification including the EOS 
main effect and an EOS $\times$ OES interaction is not pre-registered: at the 
worker level, EOS and OES are anticipated to be highly correlated, creating an 
identification problem for any joint specification that attempts to separately 
estimate entrance and exit effects.

The two estimating equations are:

\begin{align}
\log(d_{ij}) &= \alpha + \beta_1 \, \text{EOS} + \delta_i + \varepsilon_{ij} \tag{1} \\[6pt]
\log(d_{ij}) &= \alpha + \beta_2 \, \text{OES} + \delta_i + \varepsilon_{ij} \tag{2}
\end{align}

where $\delta_i$ are origin occupation fixed effects, included to absorb the backdoor path from prior educational trajectory to the specificity measures. 
The univariate specifications should be interpreted as reduced-form estimates of the total association between each specificity measure and distance traveled, rather than structural identification of the entrance and exit effects separately. The pre-registered hypothesis is $\beta_1 > \beta_2 > 0$ in the non-local clean sample. The Spearman correlation between $\text{EOS}$ and $\text{OES}$ is reported as a descriptive diagnostic; high correlation attenuates but does not eliminate the discriminating power of the comparison, and is itself informative about the structure of education-occupation matching in the Canadian labour market.

The model is evaluated across six samples, ordered by expected fit. The non-local clean sample is the primary identification sample, where the wormhole
mechanism is operative and measurement is clean. The local clean sample is the 
falsification sample, where human capital is relatively stable and specificity has no purchase on distance traveled; all coefficients are predicted to be approximately zero. The four remaining samples introduce measurement problems one at a time or in combination, and are expected to show progressive degradation in fit relative to the non-local clean sample. The $R^2$ of each regression is tracked across all six samples to document this degradation pattern, which constitutes an internal validation of the local/non-local partition. Results are reported for both skill and hierarchical distance outcomes.

The positive distance restriction removes many apprenticeship completers, since NOC makes no distinction between apprentices and journeypeople of the same occupation --- those already working in the target trade in 2016 show zero measured distance upon completion and are excluded by construction. Rather than threatening identification, this restriction improves measurement validity: the excluded zeros reflect NOC coarseness rather than true immobility, and their inclusion would introduce noise into the outcome variable. Workers who complete an entire apprenticeship program within the five-year window from a different 2016 occupation remain in the sample, and for these workers the wormhole mechanism is cleanly observable. The limitation is one of external validity rather than internal validity: estimates do not generalise to within-trade apprenticeship completers, whose human capital increment is real but invisible to NOC.


## Results  \[PLACEHOLDER --- awaiting data\]  {#sec-res}

\[KL and CWTVD improvement figures by cost metric, $\varepsilon$ and
subsample\]

\[KL and CWTVD improvement by sub regime, cost metric, $\varepsilon$ and
subsample\]

\[p-values and R squared from regressing log(distance>0) on OES and EOS separately plus origin fixed effects.\]

## Limitations

For local mobility, a notable omitted factor is *path-specific* wage
differentials. The additively separable component of wage gains ---
recoverable from origin and destination wage levels alone --- is absorbed
by the margins insofar as high-wage destinations attract larger inflows,
and low-wage origins larger outflows. The residual dyadic component,
reflecting transition-specific wage premia and discounts beyond what
origin and destination wage levels predict, is an omitted factor.
Short-distance transitions are within a similar occupational
neighbourhood, reducing employer uncertainty about candidate fit and
productivity; long-distance transitions cross larger skill or
hierarchical gaps, and employers may discount wage offers to compensate
for that uncertainty, or workers may accept lower wages to gain entry to
a distant occupation.

The threat to the horse race would require that these residual dyadic
wage effects are differentially correlated with the competing cost
matrices --- that they track one metric more closely than the others.
There is no obvious mechanism by which employer uncertainty would operate
through skill distance systematically differently than through
hierarchical distance; the dyadic wage residual story is essentially
metric-agnostic.

Path-specific wage differentials therefore attenuate the observed
distance signal but are unlikely to distort the horse race result, which
depends only on relative fit across metrics rather than absolute fit
levels. Regardless, incorporating path-specific wage differentials into the cost matrix or the estimation is not feasible with the Canadian census data: 
occupation is recorded at the census date, but wages are derived from 
prior-year tax records. For occupational switchers the timing of the 
transition relative to the tax year determines whether the observed wage 
reflects the origin occupation, the destination, or a mix of both --- 
a source of measurement error that cannot be resolved without longitudinal 
wage data at higher frequency than the census provides.

For non-local mobility, the specificity regression faces an analogous but 
distinct identification challenge. The regression estimates the reduced-form 
relationship between EOS, OES, and distance traveled following an increase 
in highest educational attainment. Both the enablement and motivation channels 
predict a positive relationship between specificity and distance: a specific 
education opens doors to particular distant occupations that a general education does not (enablement), and workers with specific educations tend to command wage premia at those destinations, making long-distance moves financially attractive (motivation). Because both channels operate through the same EOS and OES measures, the reduced-form estimates conflate the two and cannot be separately attributed to enablement or motivation. Separating the enablement channel from the motivation channel would require either an instrument that shifts wage premia without shifting human capital reallocation, or a structural model of education choice and occupational mobility. Neither is feasible with the available data. The regression should therefore be interpreted as identifying the reduced-form predictive relationship between specificity and distance traveled, without attribution to a specific mechanism.

## Conclusion

- What we found \[PLACEHOLDER --- awaiting data\]

The central policy takeaway from this analysis is that labour market
adjustment mechanisms are not one-size-fits-all: the appropriate
intervention depends on which mobility regime contains the occupation
experiencing a shortage. The three mobility sub-regimes suggest three distinct policy levers.

In the Skill sub-regime, reskilling and short-cycle training are the natural adjustment mechanisms. Workers in adjacent occupations already possess much of the relevant task content, and targeted reskilling can accelerate the
reallocation of labour toward occupations in shortage without requiring
formal retraining. The policy implication is that investment
in modular, portable training that builds on existing skill profiles
will generate the largest mobility response in this segment.

In the Hierarchy sub-regime, the relevant bottleneck is not skill 
transferability but the hierarchy's silo and platform structure. An 
occupational shortage within one silo cannot be addressed by upskilling 
workers from a different silo, even if the occupations share a near-identical 
skill profile. Instead, shortages must be addressed within-silo, through 
either lateral movement on a platform or vertical movement between 
platforms. The best way to address a shortage at a higher platform is by increasing the supply of workers to lower platforms, freeing incumbents to pursue upward movement through the silo [@doeringer1971internal]. Bridging programs, mentorship structures, and targeted recruitment into within-silo entry-level positions are the appropriate instruments.

In the Binary sub-regime, either a worker has the required education, or they do not. Neither reskilling nor within-silo movement can substitute for education. Adjustment in this sub-regime depends entirely on expanding formal training pipelines: increasing capacity in the relevant educational programs, reducing barriers to program entry, and shortening time-to-completion where possible. These are also the occupations that immigration policy should target, both in terms of admission and rapid credential recognition. 

If the hypothesized sub-regime structure is confirmed, the geometry of 
occupational mobility determines which policy lever to pull --- and the 
wrong lever is not just ineffective but potentially counterproductive.


## Appendix A: O\*NET to NOC 2021 Skill Mapping

OaSIS provides occupation-level skill profiles for Canadian
classifications and would be an attractive data source for constructing
$C_{\text{skill}}$. However, OaSIS does not expose how those profiles
are constructed: the mapping between O\*NET and NOC occupations is not
observable, the weights linking underlying occupations are not published,
and there is no information on how multiple O\*NET occupations are
combined to form a NOC profile. This appendix describes an alternative
mapping constructed explicitly from published crosswalks, allowing full
transparency of inputs and weights, diagnostic evaluation of mapping
quality, and reproducibility of all outputs.

The mapping chains four classifications: O\*NET 2019 $\rightarrow$ SOC
2018 $\rightarrow$ NOC 2016 $\rightarrow$ NOC 2021. Each stage introduces
potential one-to-many relationships. A naive approach would assign equal
weight to all admissible paths, but this leads to over-weighting of
occupations that appear frequently in crosswalk chains and inclusion of
unrelated occupations. Instead, path weights are constructed to reflect
the degree of splitting at each stage: if a source occupation splits into
$n_s$ destination occupations at stage $s$, each path through that stage
receives weight $w_s = 1/n_s$, and the overall path weight is
$w_{\text{path}} = w_1 \times w_2 \times w_3$. Path weights are then
aggregated by origin O\*NET occupation and destination NOC 2021 code and
normalized to sum to one within each NOC to yield base weights. The
consequences of naive versus path weighting can be severe: for NOC 72600
(Air pilots, flight engineers and flying instructors), naive weights
assign five-eighths of the skill profile to agricultural workers, while
path weights assign 78\% to pilots.

A power transformation and renormalization is then applied to further
reduce the influence of mapping artifacts: $\tilde{w}_k \propto w_k^2$.
This concentrates weight on the dominant O\*NET contributors and
diminishes the influence of peripheral occupations that enter through
crosswalk chains with many splits. For NOC 72600, downweighting
increases the pilot share of the skill profile to 96\%. Each O\*NET
occupation is then represented as a point in a reduced skill space using
the first 10 principal components of the O\*NET skill content variables,
and for each NOC 2021 code the weighted centroid of its contributing
O\*NET occupations is computed using the downweighted mapping, yielding a
10-dimensional skill profile for each NOC occupation.

Two diagnostics assess the quality of each NOC's mapping in skill space.
Herfindahl concentration measures how concentrated the mapping weights
are across contributing O\*NET occupations: $H = \sum_k \tilde{w}_k^2$.
For example, a Herfindahl in excess of $\frac{2}{3}$ indicates that the
NOC skill profile is effectively dominated by a single O\*NET occupation
with a minimum weight of 0.79; low values indicate a diffuse mapping.
Scaled Similarity measures the internal coherence of the contributing
occupations in skill space: the weighted average distance of each
contributing O\*NET occupation from the NOC centroid is computed, then
normalized to a $[0.05, 1]$ interval by dividing by the maximum value
across all NOCs, multiplying by 0.95, and subtracting from 1, so that
$SS = 1 - 0.95 \times \bar{d} / \max_j \bar{d}_j$. The lower bound of
0.05 is chosen to match the approximate lower bound of the Herfindahl
index, making the two measures' floors commensurable. The mapping quality
composite is the product $H \times SS$, which is high only when the
mapping is both concentrated and internally coherent. The three
interpretation cases are: high $H$ with any $SS$ (no problem --- almost
all weight on a single O\*NET occupation); low $H$ with high $SS$ (no
problem --- diffuse but coherent); and low $H$ with low $SS$
(problematic --- a diffuse average of disparate occupations whose skill
profile is unreliable). NOC 41200 (University professors and lecturers)
is a canonical example of the third case.

NOC 2021 occupations in the bottom decile of $H \times SS$ are excluded
from the local clean and non-local clean samples and isolated in the
mapping quality diagnostic subsample. Because the exclusion criterion is
based on mapping quality rather than occupation characteristics, it
introduces noise into $C_{\text{skill}}$ selectively while leaving
$C_{\text{hier}}$ and $C_{\text{binary}}$ unaffected --- a property
exploited in the mapping quality falsification design described in @sec-data. The full mapping pipeline is implemented in
[@onet_to_noc] and is
constructed entirely from O\*NET data and published crosswalks.

## Appendix B: 2016 TEER mismatch

Admissible TEERs are the Target TEER $\pm1$ plus TEER 0 (management), where the
assumption is that management roles are independent of educational attainment. 

\begin{table}[H]
\centering
\begin{tabular}{lcr}
\toprule
Highest Attainment & Target TEER & Admissible TEERs \\
\midrule
No certificate, diploma or degree & 5 & 4, 5, 0 \\
Secondary (high) school diploma & 4 & 3, 4, 5, 0 \\
Apprenticeship or trades certificate or diploma & 3 & 2, 3, 4, 0 \\
College / CEGEP / other non-university & 2 & 1, 2, 3, 0 \\
University certificate or diploma below bachelor & 2 & 1, 2, 3, 0 \\
Bachelor's degree & 1 & 1, 2, 0 \\
Above bachelor (Master's, PhD, MD, etc.) & 1 & 1, 2, 0 \\
\bottomrule
\end{tabular}
\caption{Admissible TEERs by highest attainment level}
\label{tab:teer_mismatch}
\end{table}

## Appendix C: Normalization Robustness  {#sec-AC}

Recall that in the main body of the paper, we anchored all distance metrics by 
the conditional median hierarchical distance.  In this appendix we explore whether the results are robust to using instead either the 25th or 75th conditional quantiles. Conditional quantiles are computed over within-category hierarchical distances, excluding the cross-category sentinel value of nine. The unconditional percentile locates the same value in the full pairwise distribution including sentinel values. Skill anchor is the corresponding quantile of the skill distance distribution.

```{r}
#| echo: false
calibration |>
  mutate(
    q_cond = scales::percent(q_cond),
    p_uncond = scales::percent(p_uncond, accuracy = 0.1),
    h_anchor = as.integer(h_anchor),
    s_anchor = round(s_anchor, 2)
  ) |>
  rename(
    `Conditional Quantile` = q_cond,
    `Hierarchical Anchor` = h_anchor,
    `Unconditional Percentile` = p_uncond,
    `Skill Anchor` = s_anchor
  ) |>
  kable(
    format = "latex",
    caption = "Normalization anchors under alternative calibration quantiles",
    align = c("c", "c", "c", "c"),
    booktabs = TRUE
  ) |>
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE
  )
```


## References
