---
title: "Occupational Mobility Is Not One Thing: Evidence from the Canadian Labour Market" 
author: "Richard Martin"
date: today
format:
  pdf:
    documentclass: article
    number-sections: true
    geometry: margin=1in
bibliography: references.bib
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \DeclareMathOperator*{\argmin}{argmin}
editor: 
  markdown: 
    wrap: 72
abstract: |
  Workers predominantly move between proximate occupations, but proximity means
  different things across the labour market. We use entropic regularized optimal
  transport to conduct a horse race among three distance metrics — skill, hierarchy,
  and binary — and show that no single geometry governs occupational mobility.
  Instead, the appropriate metric depends on each occupation's education specificity:
  skill where skills are highly transferable, hierarchy where silos (occupational broad
  groups) and platforms (education and experience) structure mobility, and binary where
  human capital is so occupation-specific that all alternative occupations appear equally
  distant. Each regime implies a distinct policy lever for addressing labour shortages.
---



```{r, include=FALSE}
library(tidyverse)
library(here)
library(knitr)
library(kableExtra)
library(patchwork)
library(conflicted)
cdfs <- read_rds(here("out","dist_cdfs.rds"))
calibration <- read_rds(here("out", "calibration.rds"))
```

## Pre-registration Notice {.unnumbered}

This paper is posted as a pre-registration of study design, prior to Statistics Canada Virtual Data Lab access.  Anything that looks like a result should be interpreted as a prior. Pre-registration is intended to defend against claims of post-hoc rationalization.

## Introduction

Occupational mobility is not one thing. Workers tend to move between proximate occupations, but proximity is not self-evident: occupations differ along multiple dimensions simultaneously --- canonically the tasks undertaken [@gathmann2010general], but also by required skills, hierarchical position, and selective education requirements. Each implies a different geometry of the labour market, and different predictions about who moves where. We test which geometry governs mobility --- and find that the answer depends on where in the labour market you look.

We separate local mobility (transitions that occur without new 
highest educational attainment) from non-local mobility (transitions that occur when a new highest educational attainment is achieved). For local mobility, distance is the explanatory variable — we ask which of three candidate metrics best rationalizes observed occupational transitions: skill (Euclidean in PCA-reduced O*NET space), hierarchical (tree metric on the NOC taxonomy), and binary (discrete metric). For non-local mobility, the role of distance is reversed: rather than asking which metric explains where workers go, we ask what determines how far they travel, treating transition distance as the outcome and measures of education and occupation specificity as its determinants.

We anticipate that no single metric will dominate globally. A labour 
market in which doctors, managers, and labourers all follow the same 
distance geometry would be surprising. We therefore partition 
the transition data into three sub-regimes based on the origin occupation's 
education specificity, each mapping to a hypothesized winning metric --- 
skill distance where human capital is general, hierarchical distance where 
silo structure binds, and binary switching costs where human capital is so 
occupation-specific that all alternative destinations appear equally 
distant. The Conceptual Framework develops this partition in detail.

The policy relevance is direct. Labour shortages in different segments 
of the market require different adjustment mechanisms, and the right 
mechanism depends on which distance metric governs mobility in that 
sub-regime. Shortages where human capital is general may be addressed 
through reskilling, building on transferable skills. Shortages where 
silo structure binds are best addressed by increasing supply at the 
bottom platform of the relevant silo, facilitating upward movement to 
address shortages at higher levels. Shortages where human capital is 
occupation-specific cannot be addressed through either mechanism: the 
relevant human capital is acquired only through specific education 
programs, so interventions must target access to and capacity within 
those programs directly. Getting the geometry right is therefore not 
merely a methodological contribution --- it determines which policy 
lever to pull.

For local mobility, we use entropic regularized optimal transport (EROT) [@peyre2019computational; @cuturi2013sinkhorn] to conduct the horse race, evaluating fit using relative KL divergence and relative Cost-Weighted Total Variation Distance (CWTVD). For non-local mobility, we estimate the relationship between transition distance and two specificity measures --- education's occupation specificity (EOS) and occupation's education specificity (OES) --- which proxy for the intensity of education-driven jumps: EOS opens doors to particular distant occupations, and OES draws selectively from particular educations.

The two models provide a cross-regime falsification design: distance metrics should improve on the independence baseline within the local regime but not when applied out of regime to non-local transitions, where new educational attainment rather than distance is doing the explanatory work; conversely, measures of specificity should predict distance within the non-local regime but fail to do so in the local regime, where human capital is relatively stable.

[FINDINGS PLACEHOLDER]

## Conceptual Framework

The central structural choice in this paper is the partition of 
occupational transitions into two qualitatively distinct regimes. A 
local transition occurs when a worker changes occupation without 
acquiring new educational attainment --- the worker's highest 
educational attainment in 2021 is the same as in 2016. A non-local 
transition occurs when a worker's highest educational attainment 
increases between the two census years. This partition reflects a 
fundamental difference in the mechanism governing mobility. For local 
transitions, the 2016 occupation is an informative starting point: 
human capital is relatively stable over the inter-census period, and 
distance from that starting point governs destination choice. For 
non-local transitions, the 2016 occupation is largely uninformative: 
a new highest educational attainment shifts the worker's position in 
occupational space in a way that is not captured by the 2016 
occupation, acting as a wormhole --- enabling discontinuous jumps that 
would otherwise be difficult to reconcile with their 2016 occupation.

Two symmetric specificity measures play central roles in the analysis, 
one defined over occupations and one over \textit{education}: the 
cross product of field of study (CIP) and highest level of attainment. 
These measures are derived from Statistics Canada Table 98-10-0404-01 
--- a one-quarter sample of the Canadian population, of which at most 
one-quarter overlaps with the linked microdata used for estimation.

To calculate an education's occupation specificity (EOS), we begin by 
calculating the KL divergence of $P(\text{NOC}|\text{education})$ from 
$P(\text{NOC})$. A higher value indicates that an education is 
distinctive in its occupation proportions. Because the empirical 
distribution can only place mass on a limited number of occupations for 
niche educations, KL divergence has a mechanical floor; we therefore 
residualize $\log(\text{KL})$ on $\log(T)$ where $T$ is the total 
number of workers with that education. The residual from this 
regression is defined to be EOS. This measure enters only the 
non-local partition as a continuous regressor proxying for wormhole 
entrance intensity --- a high EOS value increases the probability of 
\emph{any} jump.

A symmetric procedure is followed to calculate an occupation's 
education specificity (OES). We begin by calculating the KL divergence 
of $P(\text{education}|\text{NOC})$ from $P(\text{education})$. A 
higher value indicates that an occupation is distinctive in its 
education proportions. Because the empirical distribution can only 
place mass on a limited number of educations for niche occupations, KL 
divergence has a mechanical floor; we therefore residualize 
$\log(\text{KL})$ on $\log(T)$ where $T$ is the total number of 
workers with that occupation. The residual from this regression is 
defined to be OES. In the local partition OES is used as a classifier 
defining the three sub-regimes; in the non-local partition it serves 
as a continuous regressor proxying for wormhole exit intensity --- 
conditional on a high EOS, a high OES increases the probability of a 
\emph{long} jump.

OES captures where an occupation sits between the poles of general and specific human capital [@becker1964human]: occupations drawing broadly from the education distribution resemble general training, whose returns are portable across employers, while occupations drawing narrowly from specific credential pipelines resemble the occupation-specific form of specific training Becker identifies — skills valuable within an occupation or industry but not transferable beyond it. Occupational licensing [@kleiner2013] represents an explicit form of OES; licensing lists capture credential selectivity directly, but rely on curated registries that are often incomplete, jurisdiction-specific, and require manual updating. OES captures the same underlying phenomenon from the equilibrium education mix observed in the data, without requiring a curated list.

The three sub-regimes are named after their hypothesized winning metric 
rather than after the OES dimension that defines them --- a convention 
adopted for clarity, not a presumption of results. The tertile 
boundaries are not theoretically derived; if the three-regime structure 
is genuine, the horse race results should be robust to moderate 
misclassification at the boundaries. The Skill sub-regime (low OES) is 
hypothesized to be governed by skill distance: skills are generic and 
mobility is not constrained by hierarchical structure. The Hierarchy 
sub-regime (middle OES) is hypothesized to be governed by hierarchical 
distance: skills are specific enough that hierarchical structure binds, 
with mobility predominantly occurring within labour market silos 
defined by NOC broad groups. The Binary sub-regime (high OES) is 
hypothesized to be governed by binary switching costs: human capital is 
so occupation-specific that the richer distance structure of the other 
metrics is counterproductive, and mobility is better described as 
either staying or, rarely, leaving at equal cost regardless of 
destination.

Occupations are assigned to sub-regimes based on OES tertile, subject 
to a TEER override rule. The Skill sub-regime includes occupations with 
bottom tertile education specificity, expanded to include all TEER 0 
(senior management) and TEER 5 (labour) occupations regardless of 
specificity tertile. This override reflects two distinct mechanisms: 
TEER 0 occupations exhibit portable managerial human capital that is 
not confined to a single occupational silo --- a school principal or 
library manager can move laterally across organizational contexts in a 
way that a physician or dentist cannot; TEER 5 occupations have 
sufficiently low barriers to entry that education specificity does not 
meaningfully constrain mobility. The Hierarchy sub-regime includes the 
remaining middle tertile occupations, where education specificity is 
sufficient to generate silo structure but not so high as to foreclose 
within-silo movement across platforms. The Binary sub-regime comprises 
the remaining top-tertile occupations. Workers in these occupations are 
in an absorbing state: their human capital is too specific to support 
further local mobility. The wormhole and absorbing-state outcomes thus 
describe the two ends of the same mechanism --- acquiring a highly 
specific education predicts both an initial long-distance jump and 
subsequent immobility.

The local/non-local partition has a natural falsification structure. Each model is tested both where it should work and where it should not, and the opposing sign predictions across regimes provide stronger evidence for the causal interpretation of the underlying mechanisms than either falsification alone would. A single confounder would have to operate in opposite directions across regimes to account for both patterns; multiple confounders would each have to align fortuitously with the predicted cross-regime structure. Neither is impossible, but both are less parsimonious than the theoretical mechanism itself. The design does not eliminate the possibility of confounding, but it raises the bar considerably: the model's validity rests not on the absence of confounders but on the implausibility of a story that generates the predicted pattern of opposing failures across regimes.

## Cost Metrics

The skill distance matrix $C_{\text{skill}}$ is constructed from 161 O\*NET
measures spanning skills, abilities, knowledge, and work activities,
mapped to NOC 2021 occupations via a purpose-built crosswalk [@onet_to_noc]. 
Each NOC is represented as a weighted average of its constituent O\*NET
occupations, with weights determined by the crosswalk mapping.

O\*NET provides two scales for each measure: Importance (how important
the attribute is to the occupation) and Level (the degree to which the
attribute is required). Both scales are standardized to a common 0--10
range and combined via geometric mean, yielding a single score that
reflects both the relevance and the intensity of each attribute. The
geometric mean is preferred over the arithmetic mean as it penalizes
cases where one scale is near zero — an attribute that is important but
required at negligible level, or required at high level but deemed
unimportant, receives a lower composite score than one scoring
moderately on both dimensions. The two scales are highly correlated in
practice, so this choice is unlikely to be consequential for the
resulting distance matrix, but we note it for transparency.

Raw O\*NET measures are standardized prior to dimensionality reduction.
Principal components analysis is applied to the 161-dimensional skill
profiles, and the number of components retained is determined
empirically: we identify the minimum k such that the Spearman rank
correlation between pairwise distances in the k-dimensional PCA space
and pairwise distances in the full 161-dimensional space exceeds 0.99.
This criterion — rank preservation of pairwise distances rather than
variance explained — is chosen because the object of interest is the
distance matrix itself, not the underlying skill dimensions. Twelve
principal components satisfy this criterion.

The skill distance between any two occupations is then the Euclidean
distance between their coordinates in the retained 12-dimensional PCA
space. An interactive heatmap of the resulting skill distance matrix 
can be found 
[here](https://rpmartin.github.io/not-one-thing-companion/#sec-skill-heatmap).

A note on the crosswalk is warranted. An alternative approach would be
to use OaSIS occupation-level skill profiles. We construct our own
crosswalk for a specific reason: falsifiability. The horse race requires
that measurement error in the skill distance matrix be attributable to a
known source — namely, imprecision in the O\*NET to NOC mapping. OaSIS
obscures the structure of that mapping, making it impossible to
characterize where and how measurement error enters $C_{\text{skill}}$.
Our crosswalk makes this structure transparent and therefore makes the
identification strategy falsifiable. 

The hierarchical distance matrix $C_{\text{hierarchy}}$ is derived directly from the NOC 2021 taxonomy, treating the occupational classification structure as a tree metric. NOC broad groups function as silos — bounded mobility clusters with limited entry points and internal progression along skill levels [@doeringer1971internal] — and TEER levels function as platforms within each silo, capturing the vertical skill stratification that governs upward movement. Cross-silo transitions require bypassing the entry port logic of an entirely different occupational internal labour market, motivating the sentinel value assigned to such pairs.  The NOC broad group is a coarser unit than the occupational clusters described by @althauser1989, but captures the same boundary structure at a level of aggregation tractable with census data.

Distance between any two occupations is defined as follows: zero if all five digits match (same unit group); one if the first four digits match (same minor group, different unit group); two if the first three digits match (same sub-major group, different minor group); three plus the absolute difference in TEER level if only the first digit matches (same broad occupational category, different sub-major group); and nine for any cross-category transition. The absolute TEER difference ranges from zero to five, so within-category distances range from three to eight, preserving a strict separation from the cross-category sentinel value of nine. By construction, any between-platform move costs more than any within-platform move, and any cross-silo move costs more than any within-silo move. An interactive heatmap of the resulting hierarchical distance matrix can be found [here](https://rpmartin.github.io/not-one-thing-companion/#sec-hier-heatmap). 

The binary switching cost matrix $C_{\text{binary}}$ assigns a cost of zero 
to remaining in the same occupation and one to any transition. It is a 
limiting case, even more extreme than the hierarchical distance --- 
topologically, the labour market is comprised of 504 distinct silos, each 
with no internal structure of any kind. $C_{\text{binary}}$ serves two roles in the analysis. First, it is the theoretical straw man in the global horse race: if occupational mobility costs are purely a function of whether a switch occurs rather than how far, then $C_{\text{binary}}$ should rationalize observed mobility patterns better than $C_{\text{skill}}$ or $C_{\text{hierarchy}}$. Second, and more subtly,  $C_{\text{binary}}$ is best suited to settings where mobility is limited --- where most workers stay rather than move. In such settings, the information contained in a richer cost structure can cause model fit to worsen. This second role motivates the Binary sub-regime hypothesis: in occupations where human capital is so specific that all exits appear equally distant, binary is not merely the straw man but the correct description of the mobility geometry.

## Cost Metric Normalization

The three cost matrices are defined on different scales by construction:
skill distances are continuous and spread over a wide range, hierarchical
distances are discrete with a sentinel value of nine reserved for
cross-category transitions, and binary distances take only the values
zero and one. Direct comparison in the horse race requires that the matrices be anchored to a common scale; without normalization, a given value of $\varepsilon$ implies very different levels of concentration for each matrix, making the horse race results uninterpretable across the $\varepsilon$ grid.

We normalize skill and hierarchical distances to a common scale by anchoring both at a substantively meaningful unit of hierarchical distance. The anchor is hierarchical distance 3 — the maximum within-silo horizontal distance, representing movement within a broad occupational group without crossing TEER levels. Distance 3 falls at the fourth percentile of the unconditional hierarchical distribution, reflecting the dominance of cross-silo pairs in the full pairwise distribution. The corresponding fourth percentile of the skill distance distribution is 7.145. Hierarchical distances are divided by 3 and skill distances by 7.145, anchoring both matrices at the same point in their respective unconditional distributions.


```{r}
#| warning: false
#| echo: false
#| label: fig-cdfs
#| fig-cap: "Cost Matrix Normalization: Raw vs. Normalized CDFs"
cdfs
```

@fig-cdfs shows the empirical CDFs of all three matrices before
and after normalization. The raw distributions differ substantially in
scale and shape; after normalization, skill and hierarchical CDFs overlap
closely over the range of distances relevant to observed mobility,
diverging only in the right tail. The sensitivity of horse race results
to the normalization anchor is examined in @sec-AC


## Mobility Data {#sec-data}

The analysis uses the Census 2016--2021 longitudinal file, constructed by linking long-form respondents across the two census cycles via the Derived Record Depository (DRD), Statistics Canada's national longitudinal database of personal identifiers. Linkage is probabilistic; details of the matching keys and overall linkage rate will be verified upon data access and reported in the final paper.

The estimation sample is partitioned into subsets that vary in two dimensions:
whether the local/non-local regime assignment is correct, and whether occupation measurement is clean. This design generates a natural ordering of expected model fit that serves as an internal validation of the partition's theoretical content.

The local clean sample is the primary identification sample for the horse race, and the falsification sample for the specificity regression. It comprises individuals with an observed occupation in both 2016 and 2021, no increase in highest educational attainment between the two census years, no self-employment in either year, no TEER--attainment mismatch in 2016, and an O*NET-to-NOC mapping quality score above the bottom decile. This is the sample where the horse race should work best and where the specificity regression should find nothing.

The non-local clean sample comprises individuals who acquire a new highest educational attainment between 2016 and 2021 and report an observed occupation in both periods, subject to the same measurement restrictions: no self-employment in either year, no TEER--attainment mismatch, and an  O*NET-to-NOC  mapping quality score above the bottom decile. It is the primary estimation sample for the specificity regression, and the falsification sample for the horse race. This is the sample where the specificity regression should work best and where the horse race should find nothing. 

The mismeasurement diagnostic samples introduce measurement problems one at a time, holding regime assignment correct. The occupational mismeasurement sample is identified by two criteria: First, workers that are self-employed on the grounds that self-employment is associated with role multiplicity — owners simultaneously performing managerial, technical, and administrative functions — making the reported NOC(s) unreliable for measuring occupational distance. Second, TEER--attainment mismatch where the 2016 occupation is inconsistent with their highest educational attainment. Attainment mismatch could easily occur for either recent attainments or for those awaiting credential recognition.\footnote{For details regarding TEER--attainment mismatch criteria, see Appendix B.} Both criteria make distance measurement noisy: we expect the quality of fit to degrade regardless of metric.

The mapping quality sample retains individuals whose 2016 occupation falls in the bottom decile of the  O*NET-to-NOC  mapping quality composite; poor mapping introduces noise specifically into $C_{\text{skill}}$ while leaving $C_{\text{hierarchy}}$ unaffected, providing a difference-in-differences diagnostic on metric-specific fit degradation.

The mismeasurement residual samples comprise all observations not assigned to any of the named samples above, split by regime. The local mismeasured residual contains individuals in the local regime with multiple simultaneous measurement problems --- self-employment, TEER--attainment mismatch, and/or poor  O*NET-to-NOC  mapping in combination. The non-local mismeasured residual contains individuals in the non-local regime with multiple simultaneous measurement problems. Within each regime the source of fit degradation is measurement noise rather than regime misassignment, preserving the interpretability of the regime axis while acknowledging that the specific contribution of each noise source cannot be separately identified. Expected fit is worse than the corresponding single-problem diagnostic samples but, crucially, better than the out-of-regime clean samples --- where the only problem is regime misassignment. If that ordering holds, it constitutes direct evidence that regime misassignment is more damaging to model fit than the accumulation of measurement problems, strengthening the theoretical case for the local/non-local partition.

Two sample restriction mechanisms introduce a potential selection concern for the horse race. Non-linkage and non-observability of occupation are both
disproportionately concentrated among low-attachment workers --- individuals with casual, seasonal, or intermittent employment whose identifying information is also less stable across census cycles --- so the two filters largely co-select the same low-skill population rather than introducing independent exclusions. This shared exclusion affects the representativeness of population-level transition estimates but does not threaten the internal validity of the horse race. For the horse race results to be biased, the excluded workers' transitions would need to be systematically more or less rationalizable by a particular cost matrix than those of their observed
counterparts in the same region of occupational space. No such mechanism is apparent: exclusion is driven by attachment instability and identifier churn, neither of which has a clear relationship to how well transitions conform to any particular cost structure conditional on occupation. The residual unlinked population among otherwise-observable workers --- recent immigrants, young workers, and high-movers--- similarly lacks any obvious occupational-space concentration that would differentially favour or penalize a particular metric.

The non-local partition is restricted to individuals who acquire a new highest educational attainment between 2016 and 2021 and report an occupation in both periods. The selection concern here is distinct from the local partition: the excluded population is not disproportionately low-skill, but rather disproportionately those whose human capital increased the most. The most demanding programs — medicine, law, dentistry — require sufficiently intensive study that concurrent employment is largely precluded, meaning students in these programs are less likely to report a 2016 occupation and thus are excluded from the sample.

This selection mechanism has a specific implication for the non-local partition's central empirical question: the relationship between education and occupation specificities and distance traveled. The most demanding programs tend also to be the most specific, and the human capital increment they confer is large — generating the longest new-attainment-driven occupational jumps. Excluding these observations truncates the upper tail of the distance distribution precisely where education specificity is highest. Estimates of the relationship between education specificity and distance traveled should accordingly be interpreted as lower bounds on the true relationship in the full population of new educational attainment. Once the linked data is available, a diagnostic plot of EOS against the proportion of workers missing a 2016 occupation — computed by education type — will quantify the magnitude of this truncation.

## Local Model {#sec-local}

Mobility costs are latent: workers' switching costs are not observable 
in principle, so any distance matrix is necessarily a specification of 
an unobservable quantity. The three candidate matrices --- skill, 
hierarchical, and binary --- represent distinct hypotheses about the 
structure of those latent costs. The horse race asks which specification 
best rationalizes observed transitions, not which one correctly measures 
an observable.

The natural alternative for evaluating occupational mobility costs is 
the gravity model [@anderson_vanwincoop_2003; @silva2006log], which 
estimates origin and destination fixed effects jointly with the cost 
parameter(s). This joint estimation is precisely the problem in our 
context: fixed effects absorb variation in mobility rates that may 
reflect true cost structure, allowing misspecified cost matrices to 
achieve good fit by compensating through the fixed effects. Using a 
gravity model would be akin to randomly hobbling the horses before the 
race. EROT avoids this issue by holding the marginal structure fixed at 
observed values, separating the question of how much mobility occurs --- 
determined by the margins --- from the question of where mobility flows 
--- shaped by the cost matrix.

The entropy-regularized optimal transport problem minimizes the following
Lagrangian, where the first term is the transport cost, the second is
the entropic regularization penalizing deterministic transport plans,
and the third and fourth enforce the marginal constraints $\sum_j
P_{ij} = a_i$ and $\sum_i P_{ij} = b_j$ via dual variables $f_i$ and
$g_j$.

$$\mathcal{L} = \sum_{ij} P_{ij} C_{ij} + \varepsilon \sum_{ij} P_{ij}(\log P_{ij} - 1) + \sum_i f_i(a_i - \sum_j P_{ij}) + \sum_j g_j(b_j - \sum_i P_{ij})$$

where $a_i$ and $b_j$ are the exact occupation stocks in 2016 and 2021
respectively. The framework is implicitly first-order Markov: the
margins absorb any factor that operates additively at the level of the
origin or destination occupation, but not factors that depend on the
specific origin-destination pair. The cost matrix is itself dyadic ---
it encodes the hypothesis that the specific pairing matters --- and the
horse race is asking which specification of dyadic structure best
captures the non-Markovian residual after the margins have absorbed the
first-order effects.

The first-order conditions yield:

$$\log P_{ij} = \frac{f_i}{\varepsilon} + \frac{g_j}{\varepsilon} - \frac{C_{ij}}{\varepsilon}$$

Exponentiating, the optimal transport plan takes the form $P^\star_{ij}
= u_i \, e^{-C_{ij}/\varepsilon} \, v_j$, where $u_i =
e^{f_i/\varepsilon}$ and $v_j = e^{g_j/\varepsilon}$ are scaling
vectors. Strict convexity of the regularized objective guarantees a
unique solution. The solution is therefore a row- and column-rescaling
of the Gibbs kernel $K_{ij} = e^{-C_{ij}/\varepsilon}$ where the Sinkhorn algorithm recovers it by alternating row and column normalization until both marginal constraints are satisfied to within a specified tolerance. All computations use a log-domain Sinkhorn solver
for numerical stability, implemented in the \texttt{safesink} R package
[@safesink].

The regularization parameter $\varepsilon$ has a natural interpretation
in the random utility framework. The optimal transport plan $P^\star_{ij}
= u_i \, e^{-C_{ij}/\varepsilon} \, v_j$ is the static equilibrium of a
discrete choice model in which workers draw idiosyncratic utility shocks
scaled by $\varepsilon$ over destination occupations. Low values of
$\varepsilon$ concentrate the transport plan on cost-minimizing
assignments, approaching a deterministic reallocation in the limit; high
values diffuse mobility across destinations, approaching the independence
baseline $P_{ij} = a_i b_j$ as $\varepsilon \to \infty$. The RUM
analogy governs the interior structure of the transport plan --- how
probability mass is allocated across origin-destination pairs --- but
does not extend to the margins. In a standard logit model the choice
probabilities are outcomes of utility maximization with no constraint on
their aggregates; in EROT the origin and destination occupation
distributions are hard constraints that the transport plan must exactly
reproduce. This distinction is methodologically consequential: because
the margins cannot adjust to absorb cost misspecification, a poorly
chosen cost matrix cannot hide behind fixed effects or scale
recalibration --- its misfit appears directly in the objective.

A critical identification constraint follows directly from the form of
the optimal transport plan: only the ratio $C_{ij}/\varepsilon$ enters
$P^\star$. The scale of the cost matrix and the value of $\varepsilon$
are therefore not separately identified. This means that results
reported at a single value of $\varepsilon$ are uninterpretable without
an implicit normalization assumption --- and that assumption can be
gamed, since any cost matrix can be made to fit well by tuning
$\varepsilon$ to its scale. A horse race conducted at a single value is not a horse race; it is a calibration exercise.

Evaluating across a grid of $\varepsilon$ values sidesteps this by looking 
for dominance that is robust across scales — a metric that wins at multiple 
values cannot be attributed to a fortuitous scale alignment. 
We evaluate across a log-spaced grid of $\varepsilon$ values ranging from 
0.5 to 2, spanning from relatively concentrated to relatively diffuse transport plans while remaining in the region where the fit metrics retain discriminating power. The criterion for a robust winner requires some care. At low values of
$\varepsilon$, cost-sensitive matrices concentrate probability mass too
aggressively relative to observed mobility, incurring an overconfidence
penalty that mechanically favours the flat binary matrix regardless of
which metric truly best rationalizes mobility. This low-$\varepsilon$
binary lead is therefore uninformative about genuine descriptive
accuracy and should not be interpreted as evidence that binary is the
superior metric. The operative criterion is dominance over the full
epsilon grid after discounting this artefact: a cost matrix is declared
the winner if it dominates all competitors across the sweep, or
dominates all non-binary competitors across the sweep while losing to
binary only at low values of $\varepsilon$ where the overconfidence penalty is
operative. Conversely, when binary genuinely best rationalizes mobility
--- as in sub-regimes where occupational transitions are not structured
by continuous distance --- it will dominate not just at low values of $\varepsilon$, but across the entire grid and by a margin that dwarfs the overconfidence effect. The $\varepsilon$ grid thus serves a dual purpose: identifying the winning metric and distinguishing genuine binary dominance from penalty artefact. For a demonstration of this effect with simulation data (where the DGP is known), 
can be found [here](https://rpmartin.github.io/not-one-thing-companion/#sec-simulation).

The horse race is evaluated using two complementary fit metrics, both
normalized relative to the independence baseline $P^{\text{ind}}_{ij} =
a_i b_j$, so that 0 indicates no improvement over independence, 1
indicates perfect fit, and negative values indicate a transport plan that
fits observed mobility worse than independence.

The first metric is relative KL divergence improvement:

$$\Delta KL = \frac{KL(P^{\text{obs}} \parallel P^{\text{ind}}) - KL(P^{\text{obs}} \parallel \hat{P})}{KL(P^{\text{obs}} \parallel P^{\text{ind}})}$$

where $KL(P \parallel Q) = \sum_{ij} P_{ij} \log(P_{ij} / Q_{ij})$.
This measures the share of the divergence between observed transitions
and independence that is explained by the fitted transport plan. KL
divergence is sensitive to probability mass wherever it is misallocated,
but is geometry-blind: misallocating mass to a nearby occupation is
penalized identically to misallocating it to a distant one.

The second metric is relative Cost-Weighted Total Variation Distance (CWTVD) improvement:

$$\Delta CWTVD = \frac{CWTVD(P^{\text{obs}}, P^{\text{ind}}) - CWTVD(P^{\text{obs}}, \hat{P})}{CWTVD(P^{\text{obs}}, P^{\text{ind}})}$$

where $CWTVD(P^{\text{obs}}, \hat{P}) = \sum_{ij}
\frac{|P^{\text{obs}}_{ij} - \hat{P}_{ij}|}{2} \cdot C_{ij}$. 

CWTVD is a cost-weighted extension of total variation distance: the 
classical TVD between discrete distributions, 
$\frac{1}{2}\sum_{ij}|P_{ij} - Q_{ij}|$, is extended by weighting 
each absolute residual by the corresponding entry of a cost matrix, 
penalizing probability mass misallocated to geometrically distant 
origin-destination pairs more heavily than mass misallocated to nearby 
ones. Unlike KL divergence, CWTVD respects the metric structure of the 
outcome space; unlike the Earth Mover's Distance 
[@rubner2000earth], it does not solve an additional optimal 
transport problem over the residual mass, making it computationally 
trivial given an existing transport plan.


The scoring cost matrix $C$ used in CWTVD is the arithmetic mean of the
three candidate normalized cost matrices: $C_{\text{skill}}$,
$C_{\text{hierarchy}}$, and $C_{\text{binary}}$. Using a neutral common metric
ensures that no candidate model is advantaged by the choice of scoring
matrix. The two metrics are complementary by design: KL divergence
captures overall probability mass misallocation regardless of geometry,
while CWTVD captures whether misallocation is concentrated in nearby or
distant pairs. A transport plan that scores well on both metrics fits
observed mobility both accurately and geometrically.

## Non-Local Model {#sec-non_local}

The non-local model estimates the conditional-on-positive relationship between 
$\log(d_{ij})$ --- the log distance between 2016 and 2021 occupations --- and 
the specificity measures EOS and OES. The outcome is computed separately using 
skill and hierarchical distance; the wormhole mechanism should operate 
independently of which metric measures the jump, providing a robustness check 
on the distance measure choice.

The theoretical mechanism is asymmetric. A high value of EOS is the primary condition for a wormhole jump — the entrance is logically prior to the exit. Conditional on a sufficiently high EOS, OES determines the distance of the jump. This suggests a joint specification that includes both the EOS main effect and an EOS × OES interaction. Whether the joint model can separately identify entrance and exit intensity depends on the worker-level correlation between EOS and OES: when the two measures are highly correlated, the interaction term EOS × OES approximates a quadratic in a single specificity index, and the coefficients on EOS and EOS × OES can no longer be cleanly interpreted as entrance and exit intensity separately. We therefore estimate three specifications — two univariate models and the joint model — and report the Spearman correlation between EOS and OES as a diagnostic informing how much structural weight to place on the joint model's coefficients. The three estimating equations are:

$$
\log(d_{ij}) = \alpha + \beta_1 \, \text{EOS} + \delta_i + \varepsilon_{ij}
$$ {#eq-eos}

$$
\log(d_{ij}) = \alpha + \beta_2 \, \text{OES} + \delta_i + \varepsilon_{ij}
$$ {#eq-oes}

$$
\log(d_{ij}) = \alpha + \beta_3 \, \text{EOS} + \beta_4 \, \text{EOS} \times \text{OES} + \delta_i + \varepsilon_{ij}
$$ {#eq-joint}


Prior to estimation, EOS and OES are standardized to mean zero and unit variance using the full linked sample. This ensures that coefficients are on a common scale both within and across six subsamples, making the parameter estimates interpretable as a comparison of effect sizes, and the degradation pattern across samples interpretable as a comparison of fit rather than a reflection of differences in variable scale.

The primary hypothesis is that all $\beta$s are greater than zero. A secondary hypothesis posits that $\beta_1 > \beta_2$: the logical priority of wormhole entrance over exit predicts that education specificity has a stronger reduced-form association with distance traveled than occupation specificity. The Spearman correlation between $\text{EOS}$ and $\text{OES}$ is itself informative about the structure of education-occupation matching in the Canadian labour market, and is reported as a descriptive diagnostic alongside the regression results.

The model is evaluated across six samples, ordered by expected fit (within-$R^2$ --- measuring the share of variance in $\log(d_{ij})$ explained by the specificity measures after partialling out origin occupation fixed effects). The within-$R^2$ is not expected to be large in absolute terms — transition distance is inherently noisy and EOS and OES are aggregate measures that capture only the systematic component of specificity-driven mobility. The relevant comparison is the ordering of within-$R^2$ across samples, not its absolute level.  The non-local clean sample is the primary identification sample, where the wormhole mechanism is operative and measurement is clean. The local clean sample is the falsification sample, where human capital is relatively stable and specificity has no purchase on distance traveled; all $\beta$s are predicted to be approximately zero.  The four remaining samples introduce measurement problems one at a time or in combination, and are expected to show progressive degradation in fit relative to the non-local clean sample. The within-$R^2$ of each regression is tracked across all six samples to document this degradation pattern, which constitutes an internal validation of the local/non-local partition. Results are reported for both skill and hierarchical distance outcomes.

The positive distance restriction removes many apprenticeship completers, since NOC makes no distinction between apprentices and journeypeople of the same occupation --- those already working in the target trade in 2016 show zero measured distance upon completion and are excluded by construction. Rather than threatening identification, this restriction improves measurement validity: the excluded zeros reflect NOC coarseness rather than true immobility, and their inclusion would introduce noise into the outcome variable. Workers who complete an entire apprenticeship program within the five-year window from a different 2016 occupation remain in the sample, and for these workers the wormhole mechanism is cleanly observable. The limitation is one of external validity rather than internal validity: estimates do not generalise to within-trade apprenticeship completers, whose human capital increment is real but invisible to NOC.


## Results  \[PLACEHOLDER --- awaiting data\]  {#sec-res}

### Local

\[KL and CWTVD improvement figures by cost metric, $\varepsilon$ and
subsample\]

\[KL and CWTVD improvement by 3 sub regimes, cost metric, $\varepsilon$ and
subsample\]

### Non-local

\[tie-fighter plot for the $\beta$\]

\[ table for the $R^2$\]

## Limitations

For local mobility, a notable omitted factor is *path-specific* wage
differentials. The additively separable component of wage gains ---
recoverable from origin and destination wage levels alone --- is absorbed
by the margins insofar as high-wage destinations attract larger inflows,
and low-wage origins larger outflows. The residual dyadic component,
reflecting transition-specific wage premia and discounts beyond what
origin and destination wage levels predict, is an omitted factor.
Short-distance transitions are within a similar occupational
neighbourhood, reducing employer uncertainty about candidate fit and
productivity; long-distance transitions cross larger skill or
hierarchical gaps, and employers may discount wage offers to compensate
for that uncertainty, or workers may accept lower wages to gain entry to
a distant occupation.

The threat to the horse race would require that these residual dyadic
wage effects are differentially correlated with the competing cost
matrices --- that they track one metric more closely than the others.
There is no obvious mechanism by which employer uncertainty would operate
through skill distance systematically differently than through
hierarchical distance; the dyadic wage residual story is essentially
metric-agnostic.

Path-specific wage differentials therefore attenuate the observed
distance signal but are unlikely to distort the horse race result, which
depends only on relative fit across metrics rather than absolute fit
levels. Regardless, incorporating path-specific wage differentials into the cost matrix or the estimation is not feasible with the Canadian census data: 
occupation is recorded at the census date, but wages are derived from 
prior-year tax records. For occupational switchers the timing of the 
transition relative to the tax year determines whether the observed wage 
reflects the origin occupation, the destination, or a mix of both --- 
a source of measurement error that cannot be resolved without longitudinal 
wage data at higher frequency than the census provides.

For non-local mobility, the specificity regression faces an analogous but 
distinct identification challenge. The regression estimates the reduced-form 
relationship between EOS, OES, and distance traveled following an increase 
in highest educational attainment. Both the enablement and motivation channels 
predict a positive relationship between specificity and distance: a specific 
education opens doors to particular distant occupations that a general education does not (enablement), and workers with specific educations tend to command wage premia at those destinations, making long-distance moves financially attractive (motivation). Because both channels operate through the same EOS and OES measures, the reduced-form estimates conflate the two and cannot be separately attributed to enablement or motivation. Separating the enablement channel from the motivation channel would require either an instrument that shifts wage premia without shifting human capital reallocation, or a structural model of education choice and occupational mobility. Neither is feasible with the available data. The regression should therefore be interpreted as identifying the reduced-form predictive relationship between specificity and distance traveled, without attribution to a specific mechanism.

## Conclusion

- What we found \[PLACEHOLDER --- awaiting data\]

The central policy takeaway from this analysis is that labour market adjustment mechanisms are not one-size-fits-all: the appropriate intervention depends on which mobility regime contains the occupation experiencing a shortage. The three mobility sub-regimes suggest three distinct policy levers.

In the Skill sub-regime, reskilling and short-cycle training are the natural adjustment mechanisms. Workers in adjacent occupations already possess much of the relevant task content, and targeted reskilling can accelerate the
reallocation of labour toward occupations in shortage without requiring formal retraining. The policy implication is that investment in modular, portable training that builds on existing skill profiles will generate the largest mobility response in this segment.

In the Hierarchy sub-regime, the relevant bottleneck is not skill transferability but the hierarchy's silo and platform structure. An occupational shortage within one silo cannot be addressed by upskilling workers from a different silo, even if the occupations share a near-identical skill profile. Instead, shortages must be addressed within-silo, through either lateral movement on a platform or vertical movement between platforms. The best way to address a shortage at a higher platform is by increasing the supply of workers to lower platforms, freeing incumbents to pursue upward movement through the silo — the occupational analogue of the internal labour market adjustment mechanism described by @doeringer1971internal. Bridging programs, mentorship structures, and targeted recruitment into within-silo entry-level positions are the appropriate instruments.

In the Binary sub-regime, either a worker has the required education, or they do not. Neither reskilling nor within-silo movement can substitute for education. Adjustment in this sub-regime depends entirely on expanding formal training pipelines: increasing capacity in the relevant educational programs, reducing barriers to program entry, and shortening time-to-completion where possible. These are also the occupations that immigration policy should target, both in terms of admission and rapid credential recognition. 

If the hypothesized sub-regime structure is confirmed, the geometry of occupational mobility determines which policy lever to pull --- and the wrong lever is not just ineffective but potentially counterproductive.

## Appendix A: O\*NET to NOC 2021 Skill Mapping

OaSIS provides occupation-level skill profiles for Canadian
classifications and would be an attractive data source for constructing
$C_{\text{skill}}$. However, OaSIS does not expose how those profiles
are constructed: the mapping between O\*NET and NOC occupations is not
observable, the weights linking underlying occupations are not published,
and there is no information on how multiple O\*NET occupations are
combined to form a NOC profile. This appendix describes an alternative
mapping constructed explicitly from published crosswalks, allowing full
transparency of inputs and weights, diagnostic evaluation of mapping
quality, and reproducibility of all outputs.

The mapping chains four classifications: O\*NET 2019 $\rightarrow$ SOC
2018 $\rightarrow$ NOC 2016 $\rightarrow$ NOC 2021. Each stage introduces
potential one-to-many relationships. A naive approach would assign equal
weight to all admissible paths, but this leads to over-weighting of
occupations that appear frequently in crosswalk chains and inclusion of
unrelated occupations. Instead, path weights are constructed to reflect
the degree of splitting at each stage: if a source occupation splits into
$n_s$ destination occupations at stage $s$, each path through that stage
receives weight $w_s = 1/n_s$, and the overall path weight is
$w_{\text{path}} = w_1 \times w_2 \times w_3$. Path weights are then
aggregated by origin O\*NET occupation and destination NOC 2021 code and
normalized to sum to one within each NOC to yield base weights. The
consequences of naive versus path weighting can be severe: for NOC 72600
(Air pilots, flight engineers and flying instructors), naive weights
assign five-eighths of the skill profile to agricultural workers, while
path weights assign 78\% to pilots.

A power transformation and renormalization is then applied to further
reduce the influence of mapping artifacts: $\tilde{w}_k \propto w_k^2$.
This concentrates weight on the dominant O\*NET contributors and
diminishes the influence of peripheral occupations that enter through
crosswalk chains with many splits. For NOC 72600, downweighting
increases the pilot share of the skill profile to 96\%. Each O\*NET
occupation is then represented as a point in a reduced skill space using
the first 10 principal components of the O\*NET skill content variables,
and for each NOC 2021 code the weighted centroid of its contributing
O\*NET occupations is computed using the downweighted mapping, yielding a
10-dimensional skill profile for each NOC occupation.

Two diagnostics assess the quality of each NOC's mapping in skill space.
Herfindahl concentration measures how concentrated the mapping weights
are across contributing O\*NET occupations: $H = \sum_k \tilde{w}_k^2$.
For example, a Herfindahl in excess of $\frac{2}{3}$ indicates that the
NOC skill profile is effectively dominated by a single O\*NET occupation
with a minimum weight of 0.79; low values indicate a diffuse mapping.
Scaled Similarity measures the internal coherence of the contributing
occupations in skill space: the weighted average distance of each
contributing O\*NET occupation from the NOC centroid is computed, then
normalized to a $[0.05, 1]$ interval by dividing by the maximum value
across all NOCs, multiplying by 0.95, and subtracting from 1, so that
$SS = 1 - 0.95 \times \bar{d} / \max_j \bar{d}_j$. The lower bound of
0.05 is chosen to match the approximate lower bound of the Herfindahl
index, making the two measures' floors commensurable. The mapping quality
composite is the product $H \times SS$, which is high only when the
mapping is both concentrated and internally coherent. The three
interpretation cases are: high $H$ with any $SS$ (no problem --- almost
all weight on a single O\*NET occupation); low $H$ with high $SS$ (no
problem --- diffuse but coherent); and low $H$ with low $SS$
(problematic --- a diffuse average of disparate occupations whose skill
profile is unreliable). NOC 41200 (University professors and lecturers)
is a canonical example of the third case.

NOC 2021 occupations in the bottom decile of $H \times SS$ are excluded
from the local clean and non-local clean samples and isolated in the
mapping quality diagnostic subsample. Because the exclusion criterion is
based on mapping quality rather than occupation characteristics, it
introduces noise into $C_{\text{skill}}$ selectively while leaving
$C_{\text{hierarchy}}$ and $C_{\text{binary}}$ unaffected --- a property
exploited in the mapping quality falsification design described in @sec-data. The full mapping pipeline is implemented in
[@onet_to_noc] and is
constructed entirely from O\*NET data and published crosswalks.

## Appendix B: 2016 TEER mismatch

Admissible TEERs are the Target TEER $\pm1$ plus TEER 0 (management), where the
assumption is that management roles are independent of educational attainment. 

\begin{table}[H]
\centering
\begin{tabular}{lcr}
\toprule
Highest Attainment & Target TEER & Admissible TEERs \\
\midrule
No certificate, diploma or degree & 5 & 4, 5, 0 \\
Secondary (high) school diploma & 4 & 3, 4, 5, 0 \\
Apprenticeship or trades certificate or diploma & 3 & 2, 3, 4, 0 \\
College / CEGEP / other non-university & 2 & 1, 2, 3, 0 \\
University certificate or diploma below bachelor & 2 & 1, 2, 3, 0 \\
Bachelor's degree & 1 & 1, 2, 0 \\
Above bachelor (Master's, PhD, MD, etc.) & 1 & 1, 2, 0 \\
\bottomrule
\end{tabular}
\caption{Admissible TEERs by highest attainment level}
\label{tab:teer_mismatch}
\end{table}

## Appendix C: Normalization Robustness  {#sec-AC}

In the main body of the paper, we anchor all distance metrics at hierarchical distance 3 — the maximum within-silo horizontal distance, representing movement within a broad occupational group without crossing TEER levels. In this appendix we examine robustness to anchoring at hierarchical distance 2 (same sub-major group, different minor group) and hierarchical distance 4 (same broad group, one TEER step), representing one step down into shorter horizontal movement and one step up into vertical movement respectively. The unconditional percentile locates each anchor in the full pairwise distribution. 

```{r}
#| echo: false
calibration |>
  mutate(
    p_uncond = scales::percent(p_uncond, accuracy = 0.1),
    h_anchor = as.integer(h_anchor),
    s_anchor = round(s_anchor, 2)
  ) |>
  rename(
    `Hierarchical Anchor` = h_anchor,
    `Unconditional Percentile` = p_uncond,
    `Skill Anchor` = s_anchor
  ) |>
  kable(
    format = "latex",
    caption = "Normalization anchors under alternative calibration quantiles",
    align = c("c", "c", "c"),
    booktabs = TRUE
  ) |>
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE
  )
```


## References
