---
title: "Occupational Mobility Is Not One Thing: Evidence from the Canadian Labour Market" 
author: "Richard Martin"
date: today
format:
  pdf:
    documentclass: article
    number-sections: true
    geometry: margin=1in
bibliography: references.bib
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{threeparttable}
  - \DeclareMathOperator*{\argmin}{argmin}
editor: 
  markdown: 
    wrap: 72
execute:
  echo: false
  message: false
  warning: false
abstract: |
  Education-occupation pipelines sort workers into occupations by specificity, and how specific an occupation a worker lands in governs which geometry rules their subsequent moves. Using entropic regularized optimal transport, we run a horse race among three distance metrics and find that no single geometry governs local mobility: the metric depends on an occupation's education specificity, skill where human capital is general, hierarchy where intermediate, binary where so occupation-specific that neither predicts destinations. The specificity that governs local mobility is thus endogenous, set by the pipeline that previously placed the worker. Each regime implies a distinct policy lever for labour shortages.
--- 

```{r, include=FALSE}
#packages
library(tidyverse)
library(here)
library(knitr)
library(kableExtra)
library(patchwork)
library(gt)
library(janitor)
library(irr)
library(kableExtra)
library(conflicted)
conflicts_prefer(dplyr::filter)
#functions
format_counts <- function(tbbl) {
  tbbl %>%
    ungroup() %>%
    mutate(part = paste(n, `TEER`)) %>%
    pull(part) %>%
    (\(x) {
      paste0(
        "(",
        paste(x[-length(x)], collapse = ", "),
        ", and ",
        x[length(x)],
        ")"
      )
    })()
}
#constants
lvls <- c("Skill", "Hierarchy", "Binary")   # low -> high education specificity
shade <- "#EAF0F4"
#data
cdfs <- read_rds(here("out","dist_cdfs.rds"))
calibration <- read_rds(here("out", "calibration.rds"))
noc_edu_spec  <- read_rds(here("out", "noc_edu_spec.rds"))
teer_counts <- tabyl(noc_edu_spec$`Based on TEER`)|>
  rename(TEER="noc_edu_spec$`Based on TEER`")
kappa <- round(kappa2(noc_edu_spec[,c("Based on TEER", 
                                "Based on Occupation Education Specificity")])$value,2)
confusion <- noc_edu_spec |>
  mutate(across(c(`Based on TEER`,
                  `Based on Occupation Education Specificity`),
                \(x) factor(x, levels = lvls))) |>
  with(table(`Based on TEER`, `Based on Occupation Education Specificity`)) 
prop_same <- sum(diag(confusion))/sum(confusion)
m <- confusion|>
  as.data.frame.matrix() |>
  tibble::rownames_to_column("TEER")
disagree <- noc_edu_spec|>filter(`Based on TEER`=="Binary", `Based on Occupation Education Specificity`=="Skill")|>pull(noc)
zeros_components <- read_rds(here("out","zeros_components.rds"))
zeros_tabyl <- janitor::tabyl(zeros_components$csize)
colnames(zeros_tabyl)[1] <- "group_size"
zeros_tabyl <- zeros_tabyl|>
  mutate(product=group_size*n,
         total=sum(product)
         )
big_zero_dist_nocs <- tibble(noc_plus_title=names(zeros_components$membership)[zeros_components$membership ==  which.max(zeros_components$csize)])
off_diag_elements <-  nrow(noc_edu_spec)^2-nrow(noc_edu_spec)
```

## Pre-registration Notice {.unnumbered}

This paper is posted as a pre-registration of study design, prior to Statistics Canada Virtual Data Lab access.  Anything that resembles a result should be interpreted as a prior. Pre-registration is intended to defend against claims of post-hoc rationalization. The firewall is between the design and the interior transition flows, which reside in the Virtual Data Lab: the estimand, the regime partition rule, the adjudication thresholds, and the committed predictions are all fixed here, before those flows are observed. The cost matrices and lift are built instead from public data (O\*NET, published crosswalks, and the census education$\times$NOC table) and are reproducible from it; their construction is therefore open to inspection rather than pre-registered in the same sense, and choices internal to it are noted as such where they arise.

## Introduction

Occupational mobility is not one thing. Workers tend to move between proximate occupations, but proximity is not self-evident: occupations differ along multiple dimensions simultaneously, canonically the tasks undertaken [@gathmann2010general], but also by required skills, hierarchical position, and selective education requirements. Each implies a different geometry of the labour market, and different predictions about who moves where. We test which geometry governs mobility, and find that the answer depends on where in the labour market one looks.

We separate local mobility, the shorter transitions that occur without a substantial gain in human capital, from non-local mobility, the potentially much longer transitions that follow a large gain in human capital. The two are not parallel phenomena: non-local mobility is the mechanism behind local mobility's heterogeneity. We hypothesize that the specificity of an individual's education-occupation pipeline determines which metric governs their subsequent movement. 

For local mobility, we ask which of three candidate metrics best rationalizes observed occupational transitions: skill (Euclidean in PCA-reduced O*NET space), hierarchical (tree metric on occupation taxonomy), and binary (degenerate metric). For non-local mobility, we ask whether the specificity of the acquired education is matched by the specificity of the destination occupation, and whether those matches are genuine pipelines rather than a coincidental pairing.

Canada's National Occupational Classification (NOC) organizes occupations into a five-digit hierarchy. The first digit identifies one of ten broad occupational categories (e.g., Natural and Applied Sciences, Health, Trades). Crosscutting this hierarchy, the second digit is the Training, Education, Experience and Responsibilities (TEER): a level from 0 to 5 reflecting vertical stratification. TEER 0 indicates management, whereas TEER 1–5 reflect decreasing levels of required education (TEER 1 university education at the high end, and TEER 5 no formal education at the low end). The remaining three digits subdivide each category into progressively finer groupings down to the unit group level. For hierarchical distance, occupation broad categories delineate separate labour market silos, and TEER levels define platforms within each silo.

Regarding local mobility, we anticipate that no single metric will dominate globally. A labour market in which cashiers, cooks and chemists all follow the same distance geometry would be surprising. We therefore partition the transition data into three sub-regimes based on the origin occupation’s education specificity (OES), corresponding to three hypothesised distance structures: skill distance where human capital is general (e.g. cashiers), hierarchical distance where silo structure binds (e.g. cooks), and a degenerate binary regime where human capital is so occupation-specific that distance no longer meaningfully orders destinations (e.g. chemists).

The policy relevance is direct. Labour shortages in different segments of the market require different adjustment mechanisms, and the right mechanism depends on which distance metric governs. Shortages where human capital is general may be addressed through reskilling, augmenting existing transferable skills. Shortages where silo structure binds are addressed within the silo, through lateral moves across a platform and, where the step up requires no credential, moves to an adjacent one. Shortages where human capital is occupation-specific are different in kind: the relevant capital is acquired overwhelmingly through dedicated education programs, so it cannot be grown at scale through experience or lateral hiring, and the shortage cannot be filled by moving workers directly from other occupations. Interventions must target access to and capacity within the programs directly. Getting the geometry right is therefore not merely a methodological contribution: it determines which policy lever to pull.

For local mobility we use entropic regularized optimal transport (EROT) [@cuturi2013sinkhorn; @peyre2019computational] to conduct horse races among candidate distance metrics, evaluating how well each rationalizes the full occupation-to-occupation transition matrix. A gravity model [@anderson_vanwincoop_2003], estimated by PPML [@silva2006log], shares EROT's functional form but estimates its origin and destination effects having observed the full transition matrix. Because these fixed effects are fit using the same interior transitions on which model performance is evaluated, they can absorb discrepancies arising from a misspecified geometry, compressing differences in fit across competing distance metrics.

In contrast, EROT is solved using only a candidate cost metric and the margins of the transition matrix, without access to the interior flows. Its scaling vectors are pinned to the margins rather than fit to the cells, so discrepancies due to a misspecified geometry must appear in the predicted interior. No metric is shown the transition matrix when determining these scaling factors: the entire matrix, diagonal and off-diagonal alike, is predicted from the cost structure and the margins. Regarding the diagonal, it carries no discriminating signal because all metrics assign zero self-distance and therefore treat staying identically. Metrics differ only in their allocation of off-diagonal mass, which is where the comparison is determined. Each geometry's score is then a verdict on the cost structure alone, based on the allocation it did not use in estimation, which is what makes the comparison clean; we develop this in @sec-local.

For non-local mobility we work at the individual level with two specificity measures drawn from the public education$\times$NOC table: the same OES used to partition the sub-regimes, and its mirror on the education side, Education's Occupation Specificity (EOS). A wormhole is a path a credential opens through occupation space, a specific education leading to a specific occupation, and the claim is that the two specificities are the same structure seen at the two ends of a worker's realized history, the education acquired and the occupation later held. The public table cannot establish that they are connected: it is a standing stock, and a specific education and a specific occupation may top their respective rankings for reasons that never meet in any worker's history. The linked file supplies the connection. For each worker who acquires a new credential, we pair the specificity of the acquired education with the specificity of the occupation they hold after, and ask whether the two co-vary. That they do is the demonstration. That the credential's specificity attaches to where the worker arrived and not to where they began is the identification: a time-invariant disposition toward specificity would attach to both ends alike, so the gap between them is what marks the pattern as credential-driven rather than a fixed trait of the worker. A disposition that itself drifts forward over a career shares the credential's direction and is not differenced out by timing alone; it is separated instead by the local-mover falsification, since a credential-independent drift would surface where no credential was acquired. We make no causal claim, only that the two specificities travel together through realized transitions in a direction the credential's timing fixes; we develop this in @sec-non_local.


[FINDINGS PLACEHOLDER]

## Regime Partition {#sec-rp}

The central structural choice in this paper is the partition of occupational transitions into two regimes, by the size of the human-capital increment that becomes *operative* for a worker between census years. What matters is not when a credential was earned but when its specific human capital becomes operative, reaching an occupation that uses it, so a credential earned pre-2016 but put to use only within the 2016-2021 window is a large increment on the same footing as one earned in the window. The operative increment is latent; its proxy is the change in the education record between censuses.  We define education as the cross product of the level of highest attainment and field of study (CIP). A recorded rise in level, or a same-level field change across a two-digit CIP boundary, represent a significant increment and route non-local; all others route local. Near-field changes within a two-digit CIP at the same attainment level are assumed to be negligible increments whether real or, more plausibly, misreported.^[For example, there are 22 ways (4 digit CIPS) to report a Bachelor of Commerce degree: [Statistics Canada](https://www23.statcan.gc.ca/imdb/p3VD.pl?Function=getVD&TVD=1420413&CVD=1420414&CPV=52.&CST=15052022&CLV=1&MLV=3) ]

The two regimes differ in the mechanism governing mobility, not merely in the size of the increment. For local transitions the 2016 occupation anchors destination choice; for non-local transitions the education's human capital increment repositions the worker in a way the 2016 occupation does not capture, the wormhole of the introduction.

Three quantities recur throughout the paper which are derived from a single public source, the census education$\times$NOC table (98-10-0404-01) described in @sec-data-public. Their single origin raises a circularity question wherever they are combined: this is taken up in @sec-data-independence. We define them once here. Write $P(n\mid e)$ for the probability that a worker with education $e$ is found in occupation $n$, $P(e\mid n)$ for its conjugate, and $P(n)$, $P(e)$ for the margins. *Lift* is $m(e,n) = P(n\mid e)/P(n)$, how distinctively education $e$ feeds occupation $n$; we read it in natural logs, so $\log m = 0$ is the baseline at which an education feeds an occupation no more than the occupation's prevalence predicts. Its construction, with the Dirichlet smoothing and lift-floor the reclassification below relies on, is given in @sec-AB. *Occupation's Education Specificity* (OES) is the size-residualized KL divergence of $P(e\mid n)$ from the education margin $P(e)$, residualized on the log occupation total to remove the mechanical floor by which a niche occupation's finite education distribution inflates the raw divergence; it measures how selective an occupation's educational intake is, and is signed and centred near zero. *Education's Occupation Specificity* (EOS) is its mirror, the same divergence with the conditioning reversed, of $P(n\mid e)$ from the occupation margin, residualized on the log education total; it measures how narrowly a field of study sends its holders across occupations. Lift scores a single education--occupation cell; OES and EOS are the two marginal concentrations of the same table, read from the occupation side and the education side. All three are cross-sectional: they describe how education and occupation co-occur in a single standing stock, so the language of an education "feeding" an occupation is a statement of association, not of movement. None of them, alone, is evidence about any worker's history: the table shows that an education and an occupation co-occur, but not in what order, since a standing stock records no sequence. The linked file adds the order the table lacks, observing a worker acquire the education and then be found in the occupation. What it supplies is thus the sequence of the two endpoints, acquired-then-held, not the path between them; intervening occupations across the five-year window are unobserved. 

We study non-negative operative increments only, which puts two kinds of case outside our object of interest. A recorded *decrease* in attainment level, whatever lies behind it, is excluded from every specification by a rule fixed in advance on the recorded level alone. The exclusion is a precaution: a decrease might accompany an unpredictable career change unrelated to any credential, which would distort the local horse race if routed local and contaminate the association between education specificity and occupation specificity if routed non-local, and the level rule removes it before either can happen. *Substitutive retraining*, an over-credentialed worker who repositions on a newly earned lower credential, is likewise not the additive mobility we study, and needs no separate handling. Recorded as a decrease, through the confusion of highest attainment with the most recent or most salient credential, it is caught by the rule above. Recorded correctly, its highest attainment is unchanged, so it routes local, where it behaves like any other unrecorded activation: an unpredictable, geometry-free move that flatters binary, which prices every off-diagonal move alike, while the graded metrics that bet on where movers go score it as a miss. That is a contaminant the local model already carries and reasons about, not a new one this case introduces. Neither fate lets it into the non-local sample, so it never contaminates the EOS--OES association.

### Proxy error and reclassification

Within scope, the latent increment and its proxy diverge in two directions, and lift detects some of each.

A worker whose education record is unchanged may nonetheless have activated a standing credential within the window: the proxy routes them local, but the mechanism is non-local. The signature is a credential dormant at the origin and feeding the destination strongly, so we reinstate to non-local when
$$\log m(e, n_{2016}) < 0 \quad\text{and}\quad \log m(e, n_{2021}) \ge \tau .$$
The first condition is a gate at baseline; the second is a threshold on pipeline strength, $\tau$, the pre-committed parameter, fixed at $\tau = 2$ on log lift as its primary value and swept over $\{2, 3, 4\}$ in @sec-AD.

Conversely, a worker whose record shows an education change may not have activated it: the proxy routes them non-local, but the mechanism is local. The signature is a destination the education actively fails to feed: we reinstate to local when $\log m(e, n_{2021}) < 0$. Educations that feed a destination at roughly the base rate sit near zero and are left non-local, a deliberately conservative threshold: it reinstates only unambiguous cases and errs toward retaining workers as non-local. The Dirichlet smoothing of @sec-AB reinforces this, shrinking $\log m$ toward zero from both sides and so pulling marginal cells away from reinstatement.

Both rules read lift only where the destination cell clears the lift-floor of @sec-AB, a precondition on reading the quantity at all rather than a criterion of either rule. Both errors are the same failure of the proxy to track the latent increment, and both are handled the same way: reclassify on lift and re-estimate, reporting the result beside the uncorrected one. Only rescue from local to non-local commits to a parameter, $\tau$; its sensitivity to $\tau$ is set out in @sec-AD, the lift-floor being held fixed at its justified value (@sec-AB) rather than varied.^[The reverse rule turns on $\log m = 0$, the neutral point of the ratio itself, at which an education feeds an occupation no more than the occupation's prevalence would predict; it is the point the smoothing of @sec-AB shrinks toward by construction, so it is a property of the measure rather than a value we chose. $\tau$, by contrast, selects a strength above baseline, which is why it is the axis the sweep varies.]

Lift can not rescue the classification errors that leave no signature. A worker with unchanged education stays local when the credential activated but carried the worker into an occupation it does not distinctively feed. Low destination lift is precisely the condition that the credential opened no distinctive path, so these workers face the destination options any local worker faces, and we expect their moves to resemble local moves in distribution: the contaminant is then a local move in all but name, and the fit the local geometry takes for it is not credit it has failed to earn.^[Should that expectation fail, the residual is two-sided rather than conservative, because binary's off-diagonal is flat where the informative geometries are graded. Where such a worker lands far from their origin, the move is one the cost metrics assign low probability, degrading the fit of skill and hierarchy while leaving binary unchanged, so that tail favours binary; where the move is short, skill and hierarchy score a cheap cell while binary scores an ordinary one, so that tail favours them. The two tails offset, and the net turns on the distance profile of off-pipeline activators. The proxy-versus-adjusted contrast of @sec-local-predictions is designed to detect that bias.] The opposite case, a worker with an education change, stays non-local when the credential did not activate but nevertheless appears operative at the destination: $\log m(e, n_{2021}) \ge 0$. This is a genuine local move sitting in the non-local sample, a worker whose newly acquired credential did no repositioning, so it dilutes the EOS--OES association the non-local sample is meant to carry, weakening the very pattern the demonstration predicts.

### Design and Identifying Assumptions

The local/non-local partition has a natural falsification structure. Each model is tested both where it should work and where it should not, and the regime-specific predictions provide stronger evidence for the causal interpretation of the underlying mechanisms than either falsification alone would. A confounder that could account for both patterns would have to operate in opposite directions across regimes, or, if several, each would have to align with the predicted cross-regime structure. Neither is impossible, and one class of confounder, selection into retraining on unobserved mobility propensity, is aligned with the regime boundary by construction. We therefore treat it not as one item in a controls inventory but as the single assumption the design cannot discharge on its own logic, and we trace its direction through each result it touches below, since that direction, not its elimination, is what the identification finally rests on.

Every other confounder the partition faces must act in opposite directions across the two regimes to survive the cross-regime pattern, and that is the demand the design is built to make; any other confound story is less parsimonious than the theoretical mechanism itself. The design does not eliminate the possibility of confounding, but it raises the bar considerably: the model's validity rests not on the absence of confounders but on the implausibility of a story that generates the predicted pattern of opposing failures across regimes.

This is a different basis for causal interpretation from conditioning on observables, and it implies a different role for controls. Where identification rests on selection on observables, every additional control closes another backdoor and a sparse specification is a liability. Here the evidential weight rests on a cross-regime pattern, and a confound that would threaten it must behave in opposite directions across the two samples rather than merely correlate with the outcome within one. Controls do not address that threat: adding covariates to a within-sample estimate cannot make a symmetric confound asymmetric. Our specifications are therefore spare by derivation rather than by default, and the design leans on contrasts that difference confounds out, the cross-regime falsification and the within-worker gap, rather than on an inventory of everything the data happen to record.

One confound the design cannot condition away is selection into retraining on an unobserved propensity. The canonical case is latent mobility propensity: a worker inclined to move is also inclined to acquire the credential that routes them non-local, so acquisition and mobility are confounded, and because the propensity is never observed, no covariate removes it. The two sides of the design face this differently. On the local side, it is what the cross-regime falsification of @sec-local-falsification must reckon with, and we trace its direction through each result it touches. On the non-local side, the EOS--OES gap of @sec-non_local is largely insulated from it. A propensity that is a time-invariant trait of the worker loads on origin and destination alike and differences out of the gap. What survives differencing is only a propensity that attaches to specific destinations in particular, whether present all at once or arriving as a forward drift, and that is the mechanism the design describes, not a confound distinct from it. The single assumption the design does not discharge is therefore narrower than a general selection-into-retraining concern: it is the possibility of a disposition that reaches asymmetrically forward, which is indistinguishable from the wormhole by construction.

The horse race is affected differently, and not uniformly across the three metrics. It does not estimate whether credentials drive distance; it asks which geometry best predicts the flows, skill against hierarchy against binary. The selection has two margins that push the same way. Workers who dislike both retraining and moving --- a single stay-put disposition --- select into the local sample, and their few reluctant moves are short ones that obey proximity in occupation space. Workers at the opposite pole, who retrain precisely to reach a specific distant occupation, select out of it; that margin is not a behavioural assumption but the mechanism itself, since the wormhole is a credential opening a path to a specific occupation, typically far from the origin. The local sample therefore retains the short, proximity-obeying transitions and loses the long, diffuse ones.

For the skill-versus-hierarchy contrast this is harmless. A stay-put worker, conditional on leaving, realizes moves near their origin, and near-origin cells tend to be short under both graded geometries rather than short under one and long under the other, so the selection adds proximity mass the two metrics share. The geometries agree only partially, so some near cells are short under one and not the other; but the selection keys on magnitude, not shape, and has no reason to favour skill's divergent cells over hierarchy's, so the added mass stays symmetric in expectation.

This symmetry is a scope condition we state rather than test. We cannot check it directly, because the selection works through an unobserved trait: no move is marked as one the selection added, so its contribution cannot be separated from the observed flows. The disposition does cast one observed shadow, but it is the wrong one to help here. Because the stay-put trait keys on a distaste for both moving and retraining, its visible expression is the acquire-or-not decision, and it is *non*-acquisition, routing a worker local, that marks the high-stay-put end; acquiring marks the low end. So the trait is not wholly hidden. But that shadow is a single bit read off a continuous disposition: it records whether a worker retrained, not how strong their reservation cost is, and the selection that concerns us operates among the non-acquirers, on where each sits along the continuous dial. The binary flag cannot resolve that, so knowing who acquired does not tell us which of the retained local moves are the short ones the selection added. The observability of acquisition therefore does not make the within-local proximity selection observable, and the scope condition stands. What licenses the symmetry instead is what the trait selects on. A stay-put disposition keys on how far a move is, and skill and hierarchy are two measures of that same distance. The trait keys on distance itself, not on which of the two rulers reports it, so however much or little the two rulers agree, it has no way to prefer moves that are short under skill to moves that are short under hierarchy. It adds short moves to both alike.

The graded-versus-binary contrast is different, because there the trait does have a side to take. Skill and hierarchy both read distance and reward short moves; binary reads no distance and treats every move alike. So a disposition that keys on shortness helps skill and hierarchy together and does nothing for binary, and the selection is directional, which is why binary is set aside as a downward-contaminated rival in the graded sub-regimes below. Between skill and hierarchy there is no such side to take, and absent something to tilt it, a lopsided result would reflect the underlying mobility structure, not selection into the sample.

For the graded-versus-binary contrast it is not harmless, and the bias is directional. Binary prices every off-diagonal move alike; its comparative advantage is precisely where movement is unpredictable, where no distance gradient organizes the flows and a flat cost fits as well as any. Short proximity-obeying moves are the opposite case, the moves a graded metric fits and binary cannot, and both selection margins enrich the local sample for exactly them. Skill and hierarchy therefore look better relative to binary in the local sample than they would in a preference-neutral sample, one not selected on the stay-put disposition. This weakens a graded win over binary in the pooled local sample, which the selection would partly manufacture, but it strengthens the sub-regime prediction it runs against: a binary win in the high-OES cell (@sec-subregime-predictions) occurs despite a sample stacked toward the graded metrics, so an observed one is evidence against the grain of the selection rather than with it.

## OES and the Sub-regimes {#sec-cf}

OES, defined above, measures how selective an occupation's educational intake is. Being a size-residualized regression residual it is signed and centred near zero, a high value marking a divergence above what the occupation's size predicts. It classifies the three local sub-regimes on *origin* OES, a proxy for how credential-specific the worker's 2016 occupation is.

OES proxies for where an occupation sits between the poles of general and 
specific human capital [@becker1964human]. Occupations drawing broadly from 
the education distribution require general human capital in Becker's sense, 
skills whose returns are portable across employers and occupations, while 
occupations drawing narrowly from specific credential pipelines require 
occupation-specific human capital, valuable within the occupation but not 
transferable beyond it. Occupational licensing [@kleiner2013] is one 
institutional source of high specificity, and licensing lists offer a direct proxy for credential selectivity, but they rely on curated registries that 
are often incomplete, jurisdiction-specific, and require manual updating. OES proxies for the same underlying selectivity from the equilibrium education mix observed in the data, without requiring a curated list.

The three sub-regimes are named for their hypothesised winning metric
rather than for the dimension that defines them, a labelling
convention adopted for readability, not a presumption of results. The
primary partition is by OES rank. The group sizes are imported from the
TEER groupings described below `r format_counts(teer_counts)`, fixing the 
boundaries before any transition data is examined. This borrowing is purely 
a sizing convention: TEER supplies
the three counts, not which occupations populate them. Assignment
within those counts is determined entirely by OES rank, with the 
`r teer_counts$n[teer_counts$TEER=="Binary"]` occupations with the
highest OES forming the Binary sub-regime, the `r teer_counts$n[teer_counts$TEER=="Skill"]`
with the lowest forming the Skill sub-regime, and the remaining `r teer_counts$n[teer_counts$TEER=="Hierarchy"]`
forming the Hierarchy sub-regime. Importing the sizes from TEER rather
than cutting the OES distribution at natural breakpoints is a
pre-commitment device: it prevents the boundaries from being chosen,
consciously or otherwise, to fit the transition data. @fig-regimes
confirms that OES varies continuously across the occupational
distribution with no visible clustering, so no natural breakpoints
exist to exploit in any case.

The partition follows directly from the human capital specificity argument. Occupations with high OES draw narrowly from specific credential pipelines and map to the Binary sub-regime, where the prediction is that no occupational geometry orders destinations. For example, absent a new highest attainment, the occupational movement an Optometrist (NOC 31111) is hard to predict: most remain Optometrists, and among the few who do move, neither skill nor hierarchical distance predicts where they go. In contrast, occupations with low OES draw broadly from the education distribution; their human capital is general in Becker's sense, and skill similarity governs mobility. For example, absent a new highest attainment, a Retail and wholesale trade managers (NOC 60020) subsequent mobility is rationalized by skill distance: the skill profile of most managers (even across NOC silos) is quite similar, yielding many short-distance options.

The remaining occupations sit between these poles, where the silo and platform structure of the NOC hierarchy governs transitions. A Nurse Aide (NOC 33102) is an example: the closest destinations are the same-platform occupations within its minor group, such as Dental Assistant (NOC 33100), Medical Laboratory Assistant (NOC 33101), and Pharmacy Assistant (NOC 33103), all lateral moves among TEER 3 assisting occupations in health. Hierarchical distance also predicts, at higher cost, upward movement to a different platform within the health silo, such as licensed practical nurse (NOC 32101). Where the platform above is credential-gated, as licensed practical nurse is, the jump necessarily involves a specific educational increment, and that increment can go unrecorded: the new credential may be omitted from a self-reported education record, or it may be real but sub-maximal, correctly leaving the highest attainment field unchanged even though it enabled the jump. In either case a credential-gated move the human-capital story would call non-local appears local in the data, a source of regime contamination we return to in @sec-non_local-predictions. Where the platform above is instead reachable only through accumulated experience, no such increment exists and no contamination arises; the missing attainment change is correct.

The TEER partition that supplies the group sizes has an independent theoretical motivation paralleling the OES logic. TEER 1 occupations, such as Physicians (NOC 31102), Dentists (NOC 31110), and Lawyers (NOC 41101), carry the most occupation-specific human capital, which makes it difficult to predict alternative occupations.  At the other end of the spectrum, we hypothesize that TEER 0, 4, and 5 occupations carry so little occupation-specific human capital that movement can freely traverse the broad occupational groups. TEER 2 and 3 fall in between, with human capital specific enough to impede cross-silo transitions yet not so specific as to produce occupational lock-in. The two orderings rest on different logics, yet both motivate the same three-way conceptual split. The TEER partition serves as a robustness check on the primary OES-based partition. The partitions agree on `r scales::percent(prop_same)` of occupations, with Cohen's $\kappa$=`r kappa`. The agreement is modest, which is what the design expects and all it requires: TEER and OES are built on different logics, so the check is whether the horse-race results hold under the TEER assignment, not whether the two assignments coincide.

We treat the OES partition as the primary specification and the TEER partition as a confirmatory check. TEER is a coarse, discrete proxy for specificity, built for an unrelated administrative purpose; OES is a continuous measure derived directly from the equilibrium education mix observed for each occupation. The two partitions agree where TEER's administrative categories happen to track specificity, and diverge where they do not. @tbl-partition-crosstab quantifies that divergence.  In particular, there are `r length(disagree)` occupations that, based on being TEER 1 would be in the binary sub-regime, but based on OES they belong in the skill sub-regime: `r disagree|>paste(collapse = ", ")`. We expect these to see substantially more movement than, say, Optometrists, with mobility governed by skill similarity rather than credential lock-in.

```{r}
#| label: tbl-partition-crosstab
#| tbl-cap: "Sub-regime confusion matrix."

kbl(m, booktabs = TRUE, align = "lccc", row.names = FALSE,
    col.names = c("TEER", "Skill", "Hierarchy", "Binary"), linesep = "") |>
  add_header_above(c(" " = 1, "Occupation education specificity" = 3),
                   bold = TRUE) |>
  kable_styling(latex_options = "hold_position",
                full_width = FALSE, position = "center") |>
  column_spec(2, bold       = c(TRUE,    FALSE,   FALSE)) |>
  column_spec(3, bold       = c(FALSE,   TRUE,    FALSE)) |>
  column_spec(4, bold       = c(FALSE,   FALSE,   TRUE)) |>
  footnote(
    general = paste(
      "Counts are occupations (n = 498).",
      "Both partitions share the same group sizes by construction"
    ),
    general_title = "", footnote_as_chunk = TRUE,
    escape = FALSE, threeparttable = TRUE
  )
```

```{r}
#| warning: false
#| echo: false
#| label: fig-regimes
#| fig-cap: "Mobility Sub-Regimes"

noc_edu_spec |>
  pivot_longer(cols = contains("Based on"), names_to = "criteria", values_to = "Sub-Regime") |>
  ggplot(aes(T, specificity, text = noc, fill = `Sub-Regime`)) +
  scale_fill_viridis_d() +
  geom_point(shape = 21, size = 1.5, stroke = .5) +
  scale_x_log10(labels = scales::comma) +
  facet_wrap(~criteria) +
  theme(text = element_text(size = 9)) +
  labs(x = "Occupation Size",
       y = "Occupation's Education Specificity")
```


Occupational mobility tends to decline with age, but age is a proxy for the underlying mechanism rather than the mechanism itself. @mincer1981labor show empirically that age effects on mobility largely disappear once tenure is controlled, and attribute the disappearance to specific human capital accumulation rather than aging per se: as workers commit to an occupational path, the wedge between their current wage and outside options widens, making mobility increasingly costly. Tenure is their operational measure of that wedge, but the wedge is unobservable in principle; what they estimate is the hazard of separation as a function of job duration, not specific capital directly. OES approaches the same underlying quantity from the supply side: rather than inferring specific capital from the duration of job spells, it reads off the credential selectivity of the occupation from the equilibrium education mix. A newly trained and a 20-year veteran physician occupy the same high-OES occupation, and should be similarly immobile; a newly appointed and a long-tenured senior manager occupy the same low-OES occupation and should remain similarly mobile. Tenure varies within each pair but the mobility prediction does not, because what constrains mobility is the credential selectivity of the occupation, not the worker's time in it. This is a distinction tenure cannot identify as a stable property of the occupation, but OES can.

The rival partition the theory calls for is one on tenure, the variable Mincer and Jovanovic actually control. Tenure is unavailable in the census linkage, and age is the only observable that stands in for it, a proxy for a proxy: the worker's clock for tenure, tenure for specific capital. Whether an age partition would add anything to the OES partition depends on how strongly occupation-level OES and worker age correlate in the realized data, which we cannot know before access. If the two are highly correlated, an age partition cuts the sample much as the OES partition does and yields little beyond it; our results in that case cannot be cleanly separated from an age effect, and we report that as a scope condition. If they are weakly correlated, the two cuts cross-cut, the OES result is identified against age, and an age partition becomes genuinely informative rather than a restatement of the OES cut.

We therefore do not pre-commit to the age partition unconditionally, since its informativeness depends on a correlation we cannot see before access. We pre-commit instead to reporting the occupation-level OES-age correlation directly at access, and to running the age partition as a reported robustness check in the event that correlation is low enough for the two cuts to diverge. A partition that adds nothing where the cuts coincide is not evidence for OES; a partition that OES survives where the cuts diverge is. Should age resolve the horse race as sharply as OES where the two cross-cut, we could not separate the credential-selectivity mechanism from accumulated specific capital, and we would report the OES result as confounded with an age effect rather than as independent evidence for the mechanism.



## Cost Metrics {#sec-cost-metrics}

The skill distance matrix $C_{\text{skill}}$ is constructed from 161 O\*NET measures spanning skills, abilities, knowledge, and work activities, mapped to NOC 2021 occupations via a purpose-built crosswalk [@onet_to_noc]. An alternative would be to use OaSIS occupation-level skill profiles, but we construct our own crosswalk for two reasons. First, it is built entirely from public O\*NET data and published crosswalks, so the mapping is fully reproducible and open to inspection, where a curated external profile is not. Second, constructing the mapping ourselves yields the intermediate weights from which mapping coherence can be assessed, which is what lets us identify and exclude the small number of occupations O\*NET cannot coherently represent. Each NOC is represented as a weighted average of its constituent O*NET occupations, with weights determined by the crosswalk mapping. For more details on the mapping see @sec-AA.

O\*NET provides two scales for each measure: Importance (how important the attribute is to the occupation) and Level (the degree to which the attribute is required). We standardize both scales to a common 0--10 range and combined via geometric mean, yielding a single score that reflects both the relevance and the intensity of each attribute. The geometric mean is preferred over the arithmetic mean as it penalizes cases where one scale is near zero, an attribute that is important but required at negligible level, or required at high level but deemed unimportant, receives a lower composite score than one scoring moderately on both dimensions. The two scales are highly correlated ($\rho=.96$), so this choice is unlikely to be consequential for the resulting distance matrix but we note it for transparency.

The 161 O\*NET dimensions are highly correlated, so the full item space is far higher-dimensional than the variation it captures. We reduce it by PCA on the standardised profiles, retaining the number of components needed to reproduce the pairwise distance structure rather than a target share of variance: starting from $k = 1$, we increment until the Spearman rank correlation between pairwise distances in the $k$-dimensional space and in the full 161-dimensional space exceeds 0.99. Rank preservation is the right criterion because the object of interest is the distance matrix itself, not the skill dimensions. Twelve components satisfy it. That the ranking stabilises by twelve is the signal that the remaining components carry item-level noise rather than meaningful variation: retaining them would reproduce the same ranking while adding noise the full space accumulates across 149 low-variance directions.  The Skill distance between two occupations is the Euclidean distance between their coordinates in this 12-dimensional space. An interactive heatmap of the resulting distance matrix is available in the [online companion.](https://rpmartin.github.io/not-one-thing-companion/#sec-skill-heatmap)

A small number of off-diagonal cells of $C_{\text{skill}}$ are exactly zero, because a few NOC unit groups map to the same O\*NET occupation: NOC resolves distinctions that O\*NET does not, so occupations O\*NET treats as one are assigned identical skill profiles and zero mutual skill distance. This affects `r zeros_components$num_zero_off_diagonal` of the roughly `r round(off_diag_elements, -2)` ordered off-diagonal pairs (`r scales::percent(zeros_components$num_zero_off_diagonal/off_diag_elements, accuracy=.01)`, falling into `r sum(zeros_tabyl$n)` groups of mutually skill-identical occupations. The zeros are not measurement error: where O\*NET does not separate two occupations, no O\*NET-based skill measure can, and zero distance is the correct statement of that limit. We keep them, and characterize the groups in @sec-AA.

The binary switching cost matrix $C_{\text{binary}}$ assigns a cost of zero to remaining in the same occupation and one to any transition. Under entropically regularized optimal transport, this implies that all off-diagonal transitions share the same switching cost, so conditional on leaving an occupation, relative transition probabilities are governed entirely by the destination scaling factors, which are pinned by the destination margins. In other words, the origin occupation provides no additional information for ordering alternative destinations once a move occurs.

$C_{\text{binary}}$ serves two roles in the analysis. First, it is the theoretical straw man in the global horse race: the hypothesis that mobility costs depend only on whether a switch occurs rather than on where it leads. If this hypothesis holds, $C_{\text{binary}}$ should rationalize observed mobility better than either $C_{\text{skill}}$ or $C_{\text{hierarchy}}$. Second, $C_{\text{binary}}$ is a plausible description of labour markets in which occupational distance carries little information about the destinations of movers. In such settings, conditioning on a move, destinations are allocated according to the destination scalings, which track overall prevalence rather than proximity to the origin occupation, so the additional geometry imposed by the structured metrics does not improve predictive performance.

The horse race therefore turns on destination structure rather than mobility rates, since all three metrics accommodate the observed propensity to remain in the same occupation. This second role motivates the Binary sub-regime hypothesis: where human capital is sufficiently specific that occupational distance does not systematically order exits, binary is not merely a degenerate benchmark but a substantive model of mobility. Because it imposes no origin-specific structure on off-diagonal transitions, its only free mechanism is the aggregate stay--move split, governed by the single parameter $\varepsilon$, which sets the overall degree of mobility. @sec-local uses this characterization when comparing where the three metrics peak.

The hierarchical distance matrix $C_{\text{hierarchy}}$ is derived directly from the NOC 2021 taxonomy, treating the occupational classification structure as a tree metric. Distance between any two occupations is defined as follows: zero if all five digits match (same unit group); one if the first four digits match (same minor group, different unit group); two if the first three digits match (same sub-major group, different minor group); three plus the absolute difference in TEER level if only the first digit matches (same broad occupational category, different sub-major group); and nine for any cross-category transition. The absolute TEER difference ranges from zero to five, so within-category distances range from three to eight, preserving a strict separation from the cross-category sentinel value of nine.  An interactive hierarchical distance heatmap of the resulting matrix is available in the [online companion.](https://rpmartin.github.io/not-one-thing-companion/#sec-hier-heatmap)

Occupation broad groups function as silos (bounded mobility clusters with limited entry points and internal progression along skill levels [@doeringer1971internal]), and TEER levels function as platforms within each silo, capturing the vertical skill stratification that governs upward movement. By construction, any between-platform move (minimum distance 4) is further than any within-platform move (maximum distance 3), and any cross-silo move (distance 9) is further than any within-silo move (maximum distance 8). The occupation broad group is a coarser unit than the occupational clusters described by @althauser1989, but the appropriate level of aggregation for the horse race: a finer partition would increase the preponderance of cross-category sentinel values in the hierarchical distance distribution, collapsing the metric's geometry toward binary and undermining its role as a distinct candidate in the horse race.

## Cost Metric Normalization

The three cost matrices are defined on different scales by construction: skill 
distances are continuous and spread over a wide range, hierarchical distances 
are discrete with a sentinel value of nine reserved for cross-category 
transitions, and binary distances take only the values zero and one. Because 
the cost matrix scale and $\varepsilon$ are not separately identified, the 
choice of anchor shifts where a metric's peak falls on the $\varepsilon$ axis 
but not the peak fit it attains; the winning metric is therefore invariant to 
normalization. Normalization matters instead for the legibility of the sweep 
plots: on a common $\varepsilon$ grid, matrices on different scales imply 
different effective $\varepsilon$, so the curves are not visually comparable 
unless first placed on a common scale. For the same reason we anchor at a 
substantively meaningful unit, so that position on the $\varepsilon$ axis 
remains interpretable.

We normalize skill and hierarchical distances to a common scale by anchoring 
both at a meaningful unit of hierarchical distance: distance 3, the maximum 
distance without changing platforms. Distance 3 falls at the fourth percentile 
of the hierarchical distribution, reflecting the preponderance of cross-silo 
pairs. The corresponding fourth percentile of the skill distance distribution 
is 7.145. Dividing hierarchical distances by 3 and skill distances by 7.145 
anchors both matrices at the same point in their respective distributions. 
Binary distances are not subject to this anchoring; their two-point support is 
shown for reference. @fig-cdfs gives the empirical CDFs before and after 
normalization. Normalization pins the skill and hierarchical distributions to 
the same point at the anchor, but their shapes remain distinct: hierarchical 
distance is discrete, with the small minority of within-platform pairs spread 
below the anchor and the bulk of pairs concentrated at the cross-category 
sentinel, while skill distance is continuous and dispersed across a wider 
range. This residual difference in shape is the separability the horse race 
exploits; normalization equalizes scale, not geometry.

```{r}
#| warning: false
#| echo: false
#| label: fig-cdfs
#| fig-cap: "Empirical CDFs of the three cost matrices before and after normalization; normalization places skill and hierarchical on a common scale for the shared $\\varepsilon$ sweeps without equalizing their shape."
cdfs
```

## Data {#sec-data}

### Public inputs {#sec-data-public}

The three constructs of @sec-rp, OES, EOS, and lift, are all computed from the publicly available 2021 Census table 98-10-0404-01, a random one-quarter sample of Canadian households that cross-classifies field of study, highest educational attainment, and occupation. With *education* defined as the field-of-study by attainment composite (@sec-rp), this is the education$\times$NOC table the constructs draw on.

Canada's NOC 2021 hierarchy nominally contains 516 unit groups, but
the five Senior Manager NOCs (00011--00015) are aggregated into NOC
00018 in table 98-10-0404-01, leaving 512
occupations. Eight unit groups have no admissible O*NET mapping and
are excluded, leaving 504 occupations.

We further exclude six occupations whose O\*NET skill mapping is incoherent.
Mapping quality is assessed with an ordinal index (the product of Herfindahl
concentration and scaled similarity, @sec-AA).  We calibrate it against a benchmark
occupation, 72600 (air pilots, flight engineers and flying instructors), which was
verified as well-mapped by direct inspection when constructing the crosswalk:
its raw crosswalk is dominated by unrelated agricultural occupations, but after
path-weighting and downweighting its profile is 96% pilots. This "verified-good"
occupation ranks seventh from the bottom, so only six occupations rank below a
case we know to be adequately mapped, which indicates high mapping quality
overall. We retain 72600 and all better-ranked occupations, and exclude the six
ranked below it: university professors and lecturers (41200), college and other
vocational instructors (41210), specialists in clinical and laboratory medicine
(31100), specialists in surgery (31101), police investigators and other
investigative occupations (41310), and labourers in textile processing and
cutting (95105). Five of the six aggregate heterogeneous specialties or roles
under a single unit group, so no single O*NET profile coherently represents
them. The retention of 72600 and the exclusion of these six is an asymmetry of
evidence: the benchmark is cleared by inspection, while the six rank below it
and none has been cleared. The boundary is drawn conservatively, since some of
the six may in fact be adequately mapped, but retaining marginal mappings would
differentially disadvantage the skill metric.

These 498 occupations are the single universe for the analysis: the OES ranking
that fixes the three sub-regimes `r format_counts(teer_counts)`, the partition
in @tbl-partition-crosstab and @fig-regimes, and the local and non-local
estimation samples are all defined over them.

### Virtual Data Lab {#sec-data-vdl}

The analysis uses the 2016 and 2021 census master files, linked via the Census 2016-2021 Longitudinal File (CEN_2016_2021). The linkage file is derived from the
Derived Record Depository (DRD), Statistics Canada's national 
longitudinal database of personal identifiers. Linkage is 
probabilistic; details of the matching keys and overall linkage rate 
will be verified upon data access.

The entire analysis is conducted on the NOC 2021 occupation classification, the single occupation space in which the cost matrices, the education$\times$NOC constructs, and the transition matrix are all defined. This requires the 2016 occupation to be expressed on NOC 2021, and we expect the longitudinal file to deliver it so, since a file built for cross-cycle occupational comparison would otherwise place the two cycles in incompatible classifications; the 2016 census occupation was originally coded to NOC 2016, and Statistics Canada retains the coded census microdata and publishes a NOC 2016-to-NOC 2021 concordance. This expectation is verified at access. Should the 2016 occupation instead be delivered on NOC 2016, it is concorded to NOC 2021 before any analysis, using the Statistics Canada concordance and excluding, by the conservative criterion already applied to occupations O\*NET cannot coherently map (@sec-data-public), any 2016 unit group that does not concord cleanly to a single NOC 2021 occupation. The concordance is a fixed, published object applied blind to the interior flows, so it introduces no researcher degree of freedom over the transition data. This same restatement mechanism is what a later NOC 2026 wave would rely on to be brought onto the NOC 2021 universe, should the design be extended to a third cycle.

Requiring an occupation in both cycles fixes the estimand as a scope condition worth stating plainly: throughout the paper, occupational mobility means mobility within the population enumerated on the long form and occupationally observed in both 2016 and 2021, a resident, twice-linked, twice-employed population. This excludes anyone who emigrated or died between the cycles, anyone who immigrated after 2016, anyone who entered or left the labour force across the window, and the low-attachment workers whose identifiers do not stably link (@sec-limitations). The restriction bears on the two halves of the paper unequally. For the local horse race, which asks only which geometry rationalizes the transitions that are observed, it is an external-validity caveat rather than a threat to the comparison. For the non-local wormhole it is sharper, since credential-driven repositioning is exactly what recent immigrants do on arrival, and the design cannot see it: a naturalized physician re-entering practice, or an internationally trained engineer requalifying, has no 2016 Canadian occupation to anchor the increment. The wormhole result therefore speaks to domestic credential acquisition, and its silence on immigrant repositioning is a limit on the generality of the mechanism, not evidence against it \textemdash{} a limit worth flagging precisely because the Binary sub-regime's policy reading turns to immigration and credential recognition, which this sample structurally underweights.

[Linkage Details Here]

In the following analysis we make use of a \emph{measurement complete}
subset of the linked data where we observe occupation, field of study,
highest educational attainment and class of worker in both 2016 and 2021.
Field of study is defined only for postsecondary credentials: for a worker
whose highest attainment is high school or less, the census records no
field, and this not-applicable value is a complete observation rather than
a missing one. We retain these workers, coding them to a single
``no postsecondary field'' education cell; they are the least
specific educations and anchor the low-specificity end of the distribution,
so dropping them as missing would silently remove exactly the observations
the specificity contrast is measured against. Completeness is therefore a
requirement that attainment, occupation, and class of worker be observed,
and that field of study be observed \emph{where it is defined}, not that
every worker carry a postsecondary field.

[Missingness Details Here]

The only remaining exclusion is individuals that report being self employed in either year: self-employed workers frequently perform multiple roles simultaneously, making a single occupation an unreliable summary of their occupational position.

The analysis runs on two samples, local and non-local, defined by the regime partition of @sec-rp: the local sample holds transitions with no substantial operative human-capital increment, the non-local sample holds those with a large one. Each sample is classified two ways. The proxy classification routes workers on the recorded education change alone. The rescued classification is the same workers with the lift-based reclassification of @sec-rp applied, moving the two kinds of misrouted worker across the local/non-local boundary in both directions: a proxy-local worker whose lift shows a dormant credential activating within the window is reinstated to non-local, and a proxy-non-local worker whose destination lift is low is reinstated to local. The rescued classification is not a separate sample but a corrected version of the same two, differing only by the workers reclassification moves; the proxy classification is primary and the rescued classification is reported beside it as the robustness check of @sec-rp. Two samples estimated under two classifications yield the four sets of estimates referenced throughout.

The global local horse race between the three cost metrics is run on all four sample-classifications: the local sample under both classifications, and the non-local sample under both classifications. In regime, each metric is scored on all three fit measures, peak $\Delta KL$ and peak $\Delta CWTVD$ under each of the two informative judges, the adjudication of @sec-local-predictions. Out of regime the runs establish a dissociation rather than falsify a pooled winner (@sec-local-falsification): the graded metrics are predicted to fall below independence on $\Delta CWTVD$ once a new attainment has reset the worker's position, while binary, making no distance bet, holds flat across the boundary, its flatness the control that marks the graded collapse as geometric rather than the non-local sample being harder for every metric alike. Thus we report the results of 12 model fits (not including the $\varepsilon$ grid). 

The sub-regime analysis runs two three-by-three grids, one within the local sample and one on the non-local sample, each under both classifications. We do not assume the reclassification and the OES partition are orthogonal: leaked non-local transitions favour the binary metric wherever they land (@sec-local-falsification), and the proxy-versus-adjusted contrast, run on each sub-regime, measures where in the OES distribution that leakage sorts rather than presuming it.

The first grid is within-local. Each of the three cost metrics is scored on each of the three OES sub-regime subsets: the diagonal carries the prediction that each metric wins its own subset (skill in the Skill subset, hierarchy in Hierarchy, binary in Binary), and the off-diagonal carries the prediction that each metric fails away from it. The winner tracking the diagonal as the subset changes is the local dissociation; a metric that won across the whole grid would show no regime dependence.

The second grid is on the non-local sample, stripped by origin. A worker's 2016 occupation is observed even when a new attainment has reset where their human capital is deployed, so the non-local transitions can be split by origin sub-regime, and each sub-regime's out-of-regime test is run on its own origin stripe: the metric a subset hypothesises as its governor is scored on that subset's origin-stripped non-local transitions. The committed predictions sit on this grid's diagonal, and they are of two kinds. For the graded governors the prediction is falsification: skill on the Skill-origin non-locals and hierarchy on the Hierarchy-origin non-locals are predicted to fall below independence, the credential reset having voided the origin their geometry priced. For binary the prediction is persistence, not falsification: binary made no distance bet in-regime, so there is none to break out of regime, and we predict instead that its fit on the Binary-origin non-locals resembles its fit on the Binary-origin locals, the sub-regime's defiance of distance surviving the reset. The off-diagonal cells of this grid carry no committed prediction; we score them so that a subset whose realised winner is not its hypothesised governor can still be read out of regime, as reported exploration rather than committed test.

The two grids run on different axes, the OES subset within local and the increment across the local/non-local boundary, and both bear on regime dependence; the hypothesised governor is fixed in advance, so each out-of-regime test is well defined whether or not that governor wins its in-regime cell, though the dissociation it belongs to is interpretable only where the in-regime leg holds. Thus we report the result of 36 model fits (not including the $\varepsilon$ grid).

The non-local sample serves the two analyses differently. The out-of-regime horse race above strips it by origin sub-regime; the EOS--OES association of @sec-non_local does not, being a worker-level estimate pooled across all credential-acquirers rather than a per-regime one. That association is estimated on the pooled non-local sample under both classifications, where the acquired education's EOS is predicted to track the destination occupation's OES more than the origin's, the forward gap that marks the pattern as credential-driven. It is also estimated on the local sample restricted to movers under both classifications, as the cross-sample falsification of @sec-non_local-predictions, where the gap is predicted to collapse because a stale credential links to both origin and destination occupations alike. The primary estimate is the within-worker gap from the stacked origin--destination specification, reported with the inference its cluster counts license (@sec-non_local-inference), alongside the two-panel display of the destination and origin associations. The demonstration, the two falsifications, and the classification contrast are the committed results of the non-local section.

### Independence of the two sources {#sec-data-independence}

The 2021 census table used to create the lift, EOS and OES measures is largely independent of the linked file used for estimation. The linked 2016--2021 file is constructed by linking long-form respondents across the two census cycles, and since each cycle independently samples one quarter of households, it covers approximately one sixteenth of all Canadian households; the overlap between the table and the linked file is therefore at most one quarter, the share of 2021 long-form respondents who also appeared in the 2016 long-form. Because the constructs are derived largely from data independent of the estimation sample, the risk that either mechanically predicts transition outcomes through shared sampling rather than through the credential-selectivity mechanism is reduced.

That argument concerns shared sampling. A separate question is construct dependence: OES, EOS, and the lift used for reclassification are all built from the one education$\times$NOC table, so it is worth stating, for each place they are combined, why the shared origin does not circularly drive the result.

The EOS--OES association of @sec-non_local is the one place two marginals of the table are set against each other. The association is between the acquired education's EOS and the destination occupation's OES, two different marginals, one conditioned on education and one on occupation, neither a function of the other. The two sides are distinct marginals, so a positive association between them is a fact about the joint structure of the table realized in transitions, not an artifact of construction. The gap is immune for the opposite reason. Where the association relies on EOS and OES being *different* marginals, the gap relies on origin and destination OES being the *same* one: both are scored identically from the same occupation margin, so any circularity in how OES is built loads on both ends alike and differences out. This is the same cancellation that makes the gap immune to a latent specificity disposition, and it holds regardless of how OES is constructed.


The partition and the reclassification are two table-derived operations, and the concern is that they interact. The operations are crossed with each other in a two-by-two table: the partition is either the proxy or lift-rescued, and the analysis is either the pooled global race or the OES sub-regimes. A shared origin can only manufacture a result where two table-derived operations act on the same workers at once, so any cell using at most one is clean by inspection: the proxy partition uses no occupation margin, and the lift rescue in the pooled global race meets no OES partition. Only the interaction term, the lift-rescued classification within the OES sub-regimes, puts two of them on the same workers, and it alone needs certifying.

The one cell where two table-derived operations interact is the lift-rescued classification run within the OES sub-regimes: there the reclassification and the sub-regime partitions are based on the same table. This cell is  certified by the proxy-versus-rescued contrast. The proxy classification, clean by the argument above, is primary, and the contrast is run within each sub-regime (@sec-subregime-predictions), so a consistent sub-regime verdict across proxy and lift rescued samples is one the composition did not manufacture. This is what answers the composition worry: it shows the conclusion does not depend on the rescue. A separate worry, about the rescue itself rather than its effect on the verdict, is that the reinstated set might be an artifact of where the activation threshold is set; that is handled independently by the @sec-AD sweep over $\tau$, which shows the rescue is stable across the threshold rather than tuned to it.

## Local Model {#sec-local}

Mobility costs are latent: workers' switching costs are not observable 
in principle, so any distance matrix is necessarily a specification of 
an unobservable quantity. The three candidate matrices, skill, 
hierarchical, and binary, represent distinct hypotheses about the 
structure of those latent costs.

### The transport model

The entropy-regularized optimal transport problem minimizes the following Lagrangian, where the first term is the transport cost, the second is the entropic regularization penalizing deterministic transport plans, and the third and fourth enforce the marginal constraints $\sum_j P_{ij} = a_i$ and $\sum_i P_{ij} = b_j$ via dual variables $f_i$ and
$g_j$.

$$\mathcal{L} = \sum_{ij} P_{ij} C_{ij} + \varepsilon \sum_{ij} P_{ij}(\log P_{ij} - 1) + \sum_i f_i(a_i - \sum_j P_{ij}) + \sum_j g_j(b_j - \sum_i P_{ij})$$

where $a_i$ and $b_j$ are the empirical marginal distributions in 2016 and 2021 respectively. 

The first-order conditions yield:
\begin{equation}
\log P_{ij} = \frac{f_i}{\varepsilon} + \frac{g_j}{\varepsilon} - 
\frac{C_{ij}}{\varepsilon}
\end{equation}
Exponentiating, the optimal transport plan takes the form 
$P^\star_{ij} = u_i \, e^{-C_{ij}/\varepsilon} \, v_j$, where 
$u_i = e^{f_i/\varepsilon}$ and $v_j = e^{g_j/\varepsilon}$ are 
scaling vectors. Strict convexity of the regularized objective 
guarantees a unique solution. The solution is therefore a row- and 
column-rescaling of the exponential kernel $K_{ij} = e^{-C_{ij}/\varepsilon}$, where the Sinkhorn algorithm recovers it by alternating row and column 
normalization until both marginal constraints are satisfied to within a specified tolerance. All computations use a log-domain Sinkhorn solver 
for numerical stability, implemented in the \texttt{safesink} R package [@safesink].

In log space, the optimal transport plan decomposes into additive origin effects
($\log u_i$), additive destination effects ($\log v_j$), and a dyadic cost term
($-C_{ij}/\varepsilon$). The scaling vectors correspond to Lagrange multipliers on
the marginal constraints and adjust to ensure the implied joint distribution matches
the observed origin and destination occupation stocks, absorbing any components of
the data that operate at the level of origins or destinations alone given the
specified cost structure. The cost matrix captures the residual structure that
depends on origin--destination pairing. The horse race asks which specification of
that dyadic structure best rationalizes observed transitions.

The regularization parameter $\varepsilon$ has a natural interpretation
in the random utility framework. The optimal transport plan $P^\star_{ij} = u_i \, e^{-C_{ij}/\varepsilon} \, v_j$ is the static equilibrium of a discrete choice model in which workers draw idiosyncratic utility shocks scaled by $\varepsilon$ over destination occupations. Low values of $\varepsilon$ concentrate the transport plan on cost-minimizing assignments, approaching a deterministic reallocation in the limit; high values diffuse mobility across destinations, approaching the independence baseline $P_{ij} = a_i b_j$ as $\varepsilon \to \infty$. The RUM analogy governs the interior structure of the transport plan, how probability mass is allocated across origin-destination pairs, but does not extend to the margins. In a standard logit model the choice probabilities are outcomes of utility maximization with no constraint on their aggregates; in EROT the origin and destination occupation distributions are hard constraints that the transport plan must exactly reproduce. This distinction is methodologically consequential: because the margins cannot adjust to absorb cost misspecification, a poorly chosen cost matrix cannot hide behind fixed effects or scale recalibration, its misfit appears directly in the objective. 

A critical identification constraint follows directly from the form of 
the optimal transport plan: only the ratio $C_{ij}/\varepsilon$ enters 
$P^\star$. The scale of the cost matrix and the value of $\varepsilon$ 
are therefore not separately identified, and results reported at a 
single value of $\varepsilon$ are uninterpretable without an implicit 
normalization assumption. We evaluate across a wide log-spaced grid of 
$\varepsilon$ values from 0.1 to 1000, deliberately broader than the range 
expected to be informative, to ensure the informative region is fully 
contained within the evaluated range and its boundaries are visually apparent.

<!-- The same non-identification forecloses a natural alternative to the horse race, namely nesting the candidate geometries in an encompassing model, estimating a blend $C = \alpha C_{\text{skill}} + (1-\alpha)C_{\text{hierarchy}}$ and reading $\alpha$. Because only the ratio $C/\varepsilon$ enters the plan, $\alpha$ is not a parameter the transport problem estimates: EROT solves for a plan given a single cost matrix, so $\alpha$ would sit outside the problem, indexing a family of matrices each of which is fitted afresh. What the encompassing model reduces to is therefore a grid search over $(\alpha, \varepsilon)$ scored by the same criterion the horse race already uses, and we decline it for two reasons.  The criterion set out in the next section answers the question the grid would answer, and with fewer moving parts, locating the geometry as intermediate where it is intermediate without a fitted weight. And a maximizing $\alpha$ would not license the interpretation the exercise invites, since reading it as the true mixture presupposes the truth lies on the segment between these two geometries, which is a functional-form commitment the design does not make and the data cannot check. -->

### Fit criteria and adjudication

The horse race is evaluated using two complementary fit metrics, both
normalized relative to the independence baseline $P^{\text{ind}}_{ij} =
a_i b_j$, so that 0 indicates no improvement over independence, 1
indicates perfect fit, and negative values indicate a transport plan that fits observed mobility worse than independence. 

The first metric is relative KL divergence improvement:

$$\Delta KL = \frac{KL(P^{\text{obs}} \parallel P^{\text{ind}}) - KL(P^{\text{obs}} \parallel \hat{P})}{KL(P^{\text{obs}} \parallel P^{\text{ind}})}$$

where $KL(P \parallel Q) = \sum_{ij} P_{ij} \log(P_{ij} / Q_{ij})$.
This measures the share of the divergence between observed transitions
and independence that is explained by the fitted transport plan. KL
divergence is sensitive to probability mass wherever it is misallocated, but is geometry-blind: misallocating mass to a nearby occupation is penalized identically to misallocating it to a distant one.

The second metric is relative Cost-Weighted Total Variation Distance (CWTVD) improvement:

$$\Delta CWTVD = \frac{CWTVD(P^{\text{obs}}, P^{\text{ind}}) - CWTVD(P^{\text{obs}}, \hat{P})}{CWTVD(P^{\text{obs}}, P^{\text{ind}})}$$

where $CWTVD(P^{\text{obs}}, \hat{P}) = \sum_{ij}
\frac{|P^{\text{obs}}_{ij} - \hat{P}_{ij}|}{2} \cdot C_{ij}$. 

CWTVD is a cost-weighted extension of total variation distance: the 
classical TVD between discrete distributions, 
$\frac{1}{2}\sum_{ij}|P_{ij} - Q_{ij}|$, is extended by weighting 
each absolute residual by the corresponding entry of a cost matrix, 
penalizing probability mass misallocated to geometrically distant 
origin-destination pairs more heavily than mass misallocated to nearby 
ones. For adjudication we score $\Delta CWTVD$ under each of the two informative cost matrices in turn, $C_{\text{skill}}$ and $C_{\text{hierarchy}}$, treating each as sole judge. These two are the extremes of the family of blends the score could use: because $C_{\text{binary}}$ is flat off-diagonal and contributes no off-diagonal weighting, that family is the convex hull of $C_{\text{skill}}$ and $C_{\text{hierarchy}}$, so the two informative geometries are its endpoints and every geometry-aware judge is a blend of them. At a fixed $\varepsilon$ the score is linear in the blend weight, so at fixed $\varepsilon$ a metric leading at both endpoints leads at every blend between them; we adjudicate on the peak over $\varepsilon$ rather than at a fixed one, and do not claim the fixed-$\varepsilon$ interpolation survives the peak. What the two endpoints give us is therefore not a certificate over the whole interior but the two most separated informative judges, scored and reported jointly: where they agree the geometry-aware verdict is unambiguous, and where they disagree that disagreement is itself an outcome (the endpoint split below), which a single interior judge would have concealed by averaging the two into one score. The two informative matrices are genuinely separable: the cell-level correlation between $C_{\text{skill}}$ and $C_{\text{hierarchy}}$ is about 0.3, so the endpoints are far enough apart to bracket a real interval rather than two near-identical judges.

We adjudicate the horse race on the likelihood criterion and the
geometry-aware criterion, the latter evaluated at both informative
endpoints, and a metric wins only by leading on all of them. The
likelihood criterion is peak relative KL improvement. For each
candidate cost matrix we maximise $\Delta KL$ over the $\varepsilon$
grid, and the reason for using it is that the EROT plan
$P^\star_{ij} = u_i\, e^{-C_{ij}/\varepsilon}\, v_j$ is the equilibrium of
a multinomial-logit discrete-choice model. Comparing candidates by their
maximised likelihood is then identical to comparing their peak relative
KL improvement, so peak $\Delta KL$ is maximum-likelihood model
selection rather than a fit heuristic. The geometry-aware criterion is
peak relative CWTVD improvement over the same grid, scored under each
informative judge, $C_{\text{skill}}$ and $C_{\text{hierarchy}}$, in
turn. KL is blind to the geometry of the outcome space, scoring a unit
of misallocated mass the same whether it lands on a near pair or a far
one. CWTVD weights each misallocation by distance, and so separates the
two correlated informative geometries more sharply than KL does
(@sec-AC).
 
The two justifications are independent. KL carries the likelihood
interpretation but cannot see geometry; CWTVD sees geometry but carries no
likelihood interpretation. Neither contains the other, so we hold them on
equal footing rather than naming one primary, and a win must satisfy the
likelihood criterion and the geometry-aware criterion at both endpoints.

 
The case for CWTVD rests differently from the case for KL. Its resolving-power property, that the peak $\Delta CWTVD$ gap between the correlated geometries tracks the true mixture more sharply than the KL gap does, is a claim about the mapping from a known mixture to the measured gap, and can be established only under known data-generating processes, not read off a single observed labour market. @sec-AC establishes it by simulation. The estimation sample then takes the measurement; we make no prediction that the Canadian data will confirm CWTVD's advantage, since the data answers a different question, which geometry governs Canadian mobility, and the resolving-power property is what licenses trusting CWTVD's answer when it parts from KL's.
 
We compare peaks rather than improvement at any single $\varepsilon$
because of the same non-identification that motivated the grid above:
only the ratio $C_{ij}/\varepsilon$ enters $P^\star$, so a metric's
improvement at a fixed $\varepsilon$ has no interpretation without an
implicit normalization, and the peak is the scale-free summary that
removes the ambiguity. Each candidate traces an
interior maximum over the grid: at low $\varepsilon$ the plan
over-concentrates on the diagonal relative to the observed move rate, and
as $\varepsilon \to \infty$ it diffuses to the independence baseline
$P_{ij} = a_i b_j$, where improvement returns to zero. The grid from $0.1$
to $1000$ is wide enough that all three candidates peak in its interior on
each criterion, with the uninformative tails visible at each end. Binary
peaks in the interior too. As the one-parameter stay/move model described
in @sec-cost-metrics, its only lever is the aggregate stay/move split, so
its peak sits at the $\varepsilon$ where the diffusion of mass off the
diagonal matches the empirical move rate, and it peaks earlier than skill
and hierarchy, which keep improving as falling $\varepsilon$ sharpens an
off-diagonal allocation that pays off.
 
The two peaks rest on different footings. Peak $\Delta KL$ is a profile-likelihood model-selection criterion: because $\Delta KL$ is monotone in the maximised log-likelihood, maximising it over the grid profiles the likelihood over $\varepsilon$, the single unidentified scale parameter, so comparing peaks compares maximised likelihoods. Peak $\Delta CWTVD$ carries no such derivation. CWTVD has no likelihood for the peak to profile, so taking its maximum is the same scale-free summary applied by analogy, not an operation a nuisance-parameter argument singles out. Two things license the peak for CWTVD in place of that derivation, and they are distinct. The first is the resolving-power property of @sec-AC: that the peak gap between the two correlated geometries tracks the true mixture, so the maximised CWTVD gap carries the geometric signal the criterion is meant to read. This is a property of the mapping from mixture to gap and holds independent of sample size. The second is that the peak is an $\mathrm{argmax}$ over a curve realized at finite $N$, and a noisy curve can place its maximum where sampling alone opens a gap between geometries the truth does not separate. Resolving power does not speak to that: it says the gap responds to a real mixture imbalance, not that the $\mathrm{argmax}$ stays quiet under none. What certifies the finite-$N$ behaviour is the calibration of @sec-AE, which draws at the realized count and margins under a balanced truth and reads off how often the peak gap clears the threshold by noise alone. We run it where the concern is acute, the sparse high-OES Binary subset; in the well-populated subsets, which hold thousands of transitions, the argmax is stable for the same reason peak $\Delta KL$ is at those counts, and we do not extend the Binary calibration to them, since the noise depends on each subset's margins and not on its count alone. The justification for peaking CWTVD is thus empirical where the justification for peaking KL is analytic; we adopt the peak for both, holding them
co-equal, while noting the asymmetry in what each peak rests on.
 
The winner of the race is the cost matrix whose peak exceeds the nearest
rival's by at least 10\% on both criteria at once. There are two: a
likelihood criterion, peak $\Delta KL$, and a geometry-aware criterion,
peak $\Delta CWTVD$. The geometry-aware criterion is not a single score but
a one-parameter family, $\Delta CWTVD$ scored under the blend
$C_\lambda = \lambda C_{\text{skill}} + (1-\lambda) C_{\text{hierarchy}}$
for $\lambda \in [0,1]$; binary contributes no off-diagonal weight and so
drops out, leaving this segment as the whole space of geometry-aware
judges. We evaluate the family at its two endpoints, $C_{\text{skill}}$ and
$C_{\text{hierarchy}}$, because they are its most separated members: at a
fixed $\varepsilon$ the score is linear in $\lambda$, so the endpoints are
where a lead is hardest to hold across the family, but we adjudicate on the
peak over $\varepsilon$ and do not claim the fixed-$\varepsilon$
interpolation carries through the peak to certify the interior. The two
endpoints are therefore the two extreme informative judges, reported
jointly, not a proof of soundness at every interior blend. The two CWTVD
scorings are the two extremes of one criterion, not two independent
confirmations to be counted alongside $\Delta KL$: requiring a lead at
both is a joint condition on the one geometry-aware criterion, not a
second and third hurdle. A metric wins only if the likelihood endorses its
fit and the geometry-aware criterion endorses it at both endpoints, so a
plan that puts mass in roughly the right cells but at the wrong distances,
scoring high on $\Delta KL$ and low on either geometric endpoint, does not
win. Because the nearest rival can differ between the two judges, a winner
must lead whichever rival is closest under each.

Three outcomes follow, and
we report whichever obtains:

\begin{enumerate}
  \item \textbf{Decisive win.} One metric leads the nearest rival by 10\%
  at all three scoring points, peak $\Delta KL$ and peak $\Delta CWTVD$
  under each of the two informative judges, which between them instantiate
  the two criteria (the two CWTVD scores being the two endpoints of the one
  geometry-aware criterion). The race resolves in its favour.
  \item \textbf{Endpoint split.} The two geometric endpoints favour
  different metrics: one leads under $C_{\text{skill}}$-scoring, another
  under $C_{\text{hierarchy}}$-scoring. The winner is blend-dependent.
  Rather than a defect this is the design's most informative outcome, since
  a single blended judge would have crowned whichever metric led at the
  midpoint and concealed that the endpoints disagree. We report both, and
  the disagreement locates the mobility geometry of that regime as
  genuinely intermediate between the two rather than governed by either
  alone.
  \item \textbf{Unresolved.} No metric clears 10\% at all three scoring
  points and the two geometric endpoints do not cleanly favour different
  metrics.
  This covers both the case where the likelihood and geometry-aware
  criteria disagree about a single winner and the case where no candidate
  leads by 10\% on any criterion. We report which criteria favour which
  metric, since even an unresolved race describes how observed mobility
  departs from each candidate, but we name no governor.
\end{enumerate}
 
The 10\% threshold is an effect-size threshold, set in advance and applied
uniformly at all three scoring points. We treat two peak improvements differing
by less than ten percentage points as substantively equivalent regardless
of whether the difference is statistically distinguishable, on the grounds
that a smaller difference does not warrant naming one geometry the governor
of mobility in that regime. The threshold is substantive, not statistical, so
it does not flex with the number of scoring points: a gap below 10\% on a
geometric endpoint is substantively unimpressive whether it stands alone
or beside the others, and correcting the bar for the count of scores
would confuse a substantive effect-size threshold with a statistical
significance threshold. At the
sample sizes the linked census file provides, even the smallest sub-regime
holds thousands of transitions, so sampling noise in each peak gap is well
below this threshold and the binding constraint is the substantive one. The one
regime where this could fail is the high-OES Binary sub-regime, whose low
mobility leaves its off-diagonal cells sparse.

Once the transition data are available we run the following check. (see recipe) Take the high-OES Binary sub-regime's realized origin and destination margins, $a$ and $b$, and draw 1000 transition samples of size $N = \mathbf{1}^\top a$ from the entropic plan under a cost matrix that is a balanced mixture of the normalized skill and hierarchy geometries, $C = \tfrac{1}{2}(C_{\text{skill}}/7.145) + \tfrac{1}{2}(C_{\text{hierarchy}}/3)$, the two placed on a common scale by the fourth-percentile anchoring of @sec-cf. Each sample is scored with both geometric criteria exactly as the horse race does. This gives the distribution of the peak gap the criteria produce when the truth is a balanced mixture of the two geometries. Such a mixture is an intermediate cost structure that neither endpoint approximates, so neither should win decisively, and any gap on a given draw is manufactured by sampling noise at this regime's count. If fewer than 5% of draws produce a peak gap above 10%, a gap above 10% in the real data is informative and we read the Binary race normally. If balanced mixtures routinely clear 10%, the regime is noise-limited and we report it as such rather than reading its verdict.
 
An unresolved race is an informative no-decision rather than a coin to be
flipped. A disagreement between the likelihood-grounded and the
geometry-aware criteria, or between the two geometric endpoints, localizes
how a metric misfits, and breaking it by fiat would discard exactly the
signal the criteria were chosen to expose.

### Classifications and the lift adjustment

The horse race is run under two classifications of the same samples, the proxy classification of @sec-rp and the lift-adjusted (rescued) classification, and the lift adjustment behaves the same way wherever it is applied, so we state its logic once here as shared machinery before the global and sub-regime predictions draw on it.

The lift adjustment improves the fit of both skill and hierarchy on the local data and does not improve it on the non-local data. On the local side the adjustment removes leaked non-local transitions, the long credential-driven moves that the cost metrics assign low probability and whose presence degrades their fit, so their peak rises. On the non-local side the adjustment adds rescued wormholes and removes non-activating moves, if anything making the sample more movement-heavy, so skill and hierarchy fit it no better and may fit it worse. The prediction is therefore that the local-versus-non-local gap for these metrics widens under adjustment, strengthening the dissociation.

The lift adjustment does not help the binary metric on either sample. The transitions the adjustment moves between samples are off-diagonal moves, and binary does not model where movers go, so shuffling them changes neither its local nor its non-local fit. This is the control that makes the widening interpretable: the adjustment moves the structured metrics and leaves binary fixed, the same dissociation the global predictions below establish, seen a second time through the reclassification.

### The global local race {#sec-local-predictions}

The global local race pools all origin occupations and asks which single geometry best rationalizes the pooled local transitions, and, out of regime, whether that geometry fails on the non-local sample. It is the setup that motivates the partition rather than a result we defend: we do not believe a single geometry governs, and the pooled no-dominance we expect is what licenses cutting the sample into sub-regimes. This subsection is self-contained; the sub-regime grids of @sec-subregime-predictions build on the same machinery but run on a second axis this one does not have.

We commit in advance to two predictions, each following from the mechanism rather than fitted to the data, and separately state the motivating expectation the partition rests on.

First, the skill and hierarchy metrics fit the local sample, which carries no substantial increment proxy, better than the non-local sample, which carries a positive increment. The dissociation is a cross-sample comparison, so it is read on a single common measure, the neutral-blend $\Delta CWTVD$ against independence of @sec-local-falsification, rather than on the two-judge adjudication that decides the in-regime race: their peak on that measure sits above independence in the local sample and falls below it in the non-local sample. Local mobility is proximity-governed and these cost metrics predict short moves; non-local mobility is the credential-driven long jump they predict against.

Second, the binary metric does not show the systematic local-versus-non-local shift that skill and hierarchy do. Binary makes no off-diagonal bet, so its fit depends only on the aggregate stay/move split and not on where movers go, and it tunes $\varepsilon$ to each sample's move rate. Its peak fit is low in both samples, since it models none of the off-diagonal structure either sample carries; what matters for the dissociation is that it does not fall from the local to the non-local sample the way a metric that had made a geometric bet does, because binary made no bet to be caught out on. We therefore predict binary's peak fit is comparable across the two samples, not that it equals any particular level, and in any case does not participate in the directional collapse that identifies the structured metrics as regime-contingent.

The motivating expectation is that neither skill nor hierarchy is a clear winner on the pooled local data. Each governs its own sub-regime, so pooled across sub-regimes the two are close and neither clears the 10\% threshold. This no-dominance is not a prediction we defend on equal footing with the dissociation predictions above; it is the absence that motivates the partition, since a single global winner would itself count against the thesis that mobility is not one thing. The selection of @sec-rp bears on it in a specific direction: because it enriches the local sample with the short proximity-obeying moves that the graded metrics fit and binary cannot, it pushes the pooled race toward a spurious graded winner, against the no-dominance expectation rather than with it. A pooled result of no dominance is therefore conservative, having survived a selection that works to manufacture the global winner it denies. What we do if a clear winner appears despite this is set out in the contingency below, not treated as a refutation.

#### Out of regime: the non-local falsification {#sec-local-falsification}

The global race has a single out-of-regime axis, the non-local sample, where a new attainment has reset the worker's position. Because we expect no pooled winner, there is no single governing metric to carry out of regime; what the non-local runs establish is the dissociation itself. The graded metrics, which bet on where movers go, are predicted to fall below independence there, position having been reset; binary, which makes no such bet, is predicted to hold flat across the boundary, and that flatness is the control that marks the graded collapse as a geometric failure rather than non-local transitions being harder for every metric alike. The three metrics are therefore all scored on the non-local sample under both classifications. We state the below-independence method in full here; the sub-regime non-local leg of @sec-subregime-predictions refers back to it, where a per-regime governor does exist to be carried out of regime.

The falsification scores $\Delta CWTVD$ under a single neutral cost matrix
rather than under the two informative judges used to adjudicate the race.
That matrix is $C$, the arithmetic mean of the three normalized cost
matrices $C_{\text{skill}}$, $C_{\text{hierarchy}}$, and
$C_{\text{binary}}$, in which no candidate geometry is privileged. The race
needs two judges because it resolves a close comparative contest between
correlated geometries, and reading the verdict at both extreme informative
judges is what keeps a winner from being an artifact of one scoring choice. The falsification needs the opposite: not a comparison
but a single fixed line that a predicted metric either crosses or does not.
A neutral blend supplies exactly one such line, and scoring the crossing
under two judges instead would reintroduce the grey zone the fixed line is
meant to remove. This is the one role for which the blend, retired from
adjudication, remains the right instrument.

The out-of-regime criterion is stated against independence, on this
neutral geometry-aware measure. We state it for a single metric predicted
to govern, which is the sub-regime case; at the global level, where no one
metric governs, it applies to each graded metric in turn. In regime, the
metric predicted to govern
peaks above independence on $\Delta KL$ and on $\Delta CWTVD$ under the
neutral blend, since its geometry rationalizes the transitions and
improvement over independence is what the horse race measures. Out of
regime, where a new highest attainment resets the worker's position and
renders the 2016 occupation uninformative, that metric is predicted to
fall below independence on $\Delta CWTVD$.
 
The out-of-regime criterion rests on $\Delta CWTVD$ alone, and the reason
is that $\Delta KL$ cannot register the failure being tested. The
falsification asks whether the predicted metric has made the wrong
geometric bet, placing mass at the wrong distances. $\Delta CWTVD$ weights
each misallocation by distance and so penalises exactly this, falling below
independence when a short-move metric meets a movement-heavy matrix.
$\Delta KL$ is blind to geometry and scores any plan that captures the
diagonal stay/move structure, which every metric shares. A wrong-geometry
metric therefore collects partial KL credit for the diagonal even while its
off-diagonal bet is wrong, and its peak $\Delta KL$ can sit slightly above
independence rather than below it. This is visible in the simulations: in
the [online companion](https://rpmartin.github.io/not-one-thing-companion), under a binary data-generating process the skill and
hierarchy metrics peak below independence on $\Delta CWTVD$ but rise to a
small positive peak on $\Delta KL$ at moderate $\varepsilon$. Requiring the
geometry-blind measure to confirm a geometric failure would ask the wrong
instrument to register it. We therefore state the out-of-regime criterion
on $\Delta CWTVD$, the measure built to see what the falsification is about.
 
Setting the line at independence rather than at binary discharges the
conditionality that a binary benchmark would leave open. A wrongly
predicting metric always sits below binary out of regime, but whether its
$\Delta CWTVD$ also falls below independence depends on the non-local
transition matrix being movement-heavy: only when most credential-changers
move, and move far, does a short-move metric place enough mass on vacated
near pairs to fit worse than independence. We predict the non-local matrix
is movement-heavy, because the wormhole mechanism is precisely the claim
that a new credential drives long discontinuous jumps.

Because the crossing depends on that prediction holding, we separate it
from the verdict rather than fold it in. The non-local move rate is
directly observed, so we report it first, as a descriptive prior to any
$\Delta CWTVD$ score: whether the non-local matrix is movement-heavy is
settled by looking, not by the falsification. The below-independence test
is then read conditional on that first stage. If the matrix is
movement-heavy and the predicted metric crosses below independence, the
crossing confirms two things the mechanism asserts together, that the
metric made the wrong geometric bet out of regime and that the bet was
wrong because the jumps are long, and the falsification fires as designed.
If the matrix is movement-heavy and the metric nonetheless stays above
independence, the falsification fires against the design: the local
geometry is rationalizing transitions in a regime where a reset position
should have voided it, the outcome the double dissociation cannot survive.
If instead the first stage shows the matrix is not movement-heavy, the
below-independence test has no purchase to begin with, since a short-move
metric meeting a light matrix need not fit worse than independence; we
report the test as inapplicable rather than as either a pass or a design
failure, and the local model is neither falsified nor confirmed out of
regime by this route. Bundling the light-matrix case with the
governing-out-of-regime case would read a failed precondition as a
substantive failure, which the first stage exists to prevent.
 
Where the test applies, independence is a fixed point, so the crossing
itself is unambiguous. The predicted metric's $\Delta CWTVD$ peak is either
below zero or it is not, with no margin to calibrate and no grey zone to
adjudicate, where the race resolves a close contest between two correlated
geometries by asking whether one peak exceeds the other by 10\%. The two sides of the
falsification are also weaker bars than winning the race. A metric clears
the in-regime side by improving on independence at all, and fails the
out-of-regime side by crossing below it; the race is won only by leading
every rival by 10\% on both measures. Both skill and hierarchy peak well
above independence in either of their sub-regimes, since they are
correlated and each partially rationalizes the other's transitions, yet
only one wins. Improving on independence is necessary for a geometry to be
governing at all; leading by 10\% on both measures is what crowns it.
 
The mechanism that drives the predicted metric's $\Delta CWTVD$ below
independence is the same one that makes binary a useful point of
comparison. All three cost matrices assign zero self-distance,
$C_{ii} = 0$, so all three concentrate mass on the diagonal and capture the
dominant stay/move feature of the transition matrix. That diagonal
contribution is common to every plan, and what distinguishes the metrics is
the off-diagonal allocation. Binary makes no off-diagonal bet, since every
switch is equidistant under it, so it cannot be wrong about where movers
go. Skill and hierarchy do make that bet, and out of regime it is the wrong
one, because they place mass on geometrically proximate pairs that wormhole
movers avoid while the data concentrates on far pairs. A metric predicting
short moves into a movement-heavy matrix withholds mass from exactly where
the data sits, and $\Delta CWTVD$ penalises that misallocation by distance,
driving it below independence. The signature of this failure on
$\Delta CWTVD$ is a curve that rises monotonically toward the independence
baseline from below without crossing it, improving as $\varepsilon$ grows
only because diffusion toward independence is the one thing that helps a
metric whose geometry points the wrong way.
 

#### If a clear winner appears

The thesis rests on the sub-regime dissociation of @sec-subregime-predictions, not on the pooled race, so a clear global winner is not a refutation but a result to be read against the sub-regime grid, and we fix that reading in advance. If a metric wins the pooled race while the grid diagonal still holds, each metric governing its own subset, we reject the single-geometry claim and affirm regime-specific governance: the diagonal holding is the dissociation, so governance is regime-specific and the pooled winner is a compositional artifact of where movers concentrate across subsets, not evidence that one geometry governs. We report it as such, and the aggregate tilt toward the pooled winner is noted rather than read as global governance. If the diagonal collapses and the pooled winner is a graded metric, the thesis of regime-specific governance is falsified, and the pooled win does not establish global governance in its place: with the diagonal collapsed a pooled graded win is equally consistent with a compositional artifact, the selection above being one route to it, so we report it as classification- and selection-dependent rather than as global governance. The strongest form of this collapse is a single graded metric winning both the pooled race and every diagonal cell, the same geometry governing each sub-regime it was predicted to fail in; this is the one outcome that would affirmatively support single-geometry governance rather than merely fail to refute it, and we read it as such, against the thesis, subject to two conditions. The first is the selection caveat: the selection of @sec-rp works to manufacture a graded win, so a sweep is credited only because it survives that selection rather than being produced by it. The second is the global out-of-regime leg: a graded metric genuinely governing must fall below independence on the non-local sample once a new attainment resets the worker's position, while binary holds flat across that boundary, so we credit a graded sweep as global governance only when it is accompanied by this collapse across the local/non-local increment. A sweep that wins every diagonal cell but does not collapse out of regime is winning for some reason that survives the credential reset, which is not the geometry pricing proximity, and we do not read it as global governance. A graded sweep is therefore reported as evidence for global governance by that metric where a pooled graded win alone is not, the difference being that the sweep has won the sub-regime cells the compositional-artifact reading is built to explain, leaving no subset in which a rival geometry governs for composition to aggregate over. If the diagonal collapses and the winner is binary, geometry does not order destinations even within the sub-regimes designed to isolate where it should, and the finding is the negative one, that occupational distance does not organize local mobility in these data; we report it as such. Binary can also lead the pooled race while losing the graded subsets, if the movers concentrate in the segment binary governs, in which case the diagonal holds and the first reading applies. The one outcome the design cannot absorb is binary sweeping the diagonal, and because binary is scored as a genuine model throughout rather than as a straw man, that outcome would be a fair defeat of the geometric hypothesis rather than an artifact to explain away.

### The sub-regime grids {#sec-subregime-predictions}

The sub-regime analysis runs on two out-of-regime axes where the global race had one. The first is within the local sample, across the OES sub-regime boundaries: each metric is scored on each sub-regime subset, and a metric is predicted to fail away from its own. The second is across the local/non-local boundary, on the non-local transitions stripped by origin sub-regime: each sub-regime's hypothesised governor is scored on its own origin stripe, the graded governors predicted to fail there and binary predicted to persist. The two axes are distinct, the OES subset within local and the increment across the local/non-local boundary, and both bear on regime dependence. This subsection is self-contained on the first axis and refers back to @sec-local-predictions for the below-independence method the second axis shares.

#### Axis one: the off-diagonal grid

Within the local sample we score all three metrics on each of the three OES sub-regime subsets, the full three-by-three grid, and we do this under both the proxy and the lift-adjusted classifications, yielding two grids. We commit to the following.

First, on the diagonal each metric wins its own subset: skill in the Skill subset, hierarchy in the Hierarchy subset, binary in the Binary subset. Binary winning its own subset is a substantive result, the claim that where human capital is occupation-specific no distance geometry orders destinations, not a null.

Second, off the diagonal each metric fails away from its own subset. The winner tracking the diagonal as the subset changes is the within-local dissociation; a metric that led across the whole grid would show no regime dependence and would count against the thesis.

Third, the two grids exist to guard against a specific bias, not as a generic robustness check. Undetected leakage of non-local transitions into the local sample favours the binary metric, since a leaked wormhole is a long move that degrades skill and hierarchy while binary, making no off-diagonal bet, is unchanged (@sec-rp). Running both classifications is what makes this bias visible and correctable: the proxy grid carries the leakage, the lift-adjusted grid removes the part of it that leaves a lift signature, and the difference between the grids is the signature of the bias. That signature is committed in advance, skill and hierarchy rise and binary stays flat under adjustment, never the reverse; its location is not. A subset stable across the two grids is uncontaminated; a subset where the structured metrics rise is where leakage sorted. We do not presume leakage concentrates in any particular subset, since no origin-OES mechanism predicts that it should; we measure where it sorts.

Fourth, the adjustment corrects only the detectable leakage. The reclassification fires on a lift signature, so leaks that leave none, a credential that activated but carried the worker into an occupation it does not distinctively feed, remain in both grids. The proxy-versus-adjusted difference is therefore a lower bound on the total leakage, and the adjusted grid is de-contaminated but not clean. The residual signature-less leakage is handled not by the reclassification but by the residual-bias argument of @sec-rp, which shows it biases conservatively.

Fifth, we pre-commit to the reading of the Binary subset, where the favouring-binary bias would most damage a result we care about. If binary's win there narrows under adjustment but still clears the ten percent threshold, the win is robust against detectable leakage and against sparse-cell noise per @sec-AE, with the residual signature-less leakage arguing conservatively per @sec-rp rather than removed. If binary's win does not survive adjustment, the detectable leakage alone was enough to manufacture it, and we report the Binary subset's result as classification-dependent rather than claiming binary governs there. The don't-survive reading is decisive; the survive reading carries the residual caveat.

As a worker-level extension of the same logic, we pre-register a partition on age in place of OES, reported at access. The specificity OES reads at the level of the occupation has a counterpart in the individual career. Early on, when little occupation-specific capital has accumulated, the binding constraint on mobility is capability, and moves run along skill adjacency, near in skill space while dispersed across silos. By mid-career the worker has settled into a silo and moves within it, so the binding constraint is the silo boundary that hierarchical distance prices. Late in the career the reservation wage has risen far enough that few moves clear it and the survivors carry no systematic geometry, the condition the binary model describes. We cut the worker age distribution at 15 to 25, 25 to 40, and 40 and over, fixed in advance, and predict the winning metric shifts across these bands in the order skill, hierarchy, binary. The prediction is the order of the shift, not its sharpness. Age reaches the governing constraint only as a proxy for a proxy, the worker's clock for tenure and tenure for accumulated specific capital, and each band is internally heterogeneous in career stage, so we expect the partition to resolve mobility far less decisively than OES and margins within any band may be thin. A confirmation of the ordering through so degraded an instrument is independent corroboration of the specificity mechanism rather than a second decisive result; the disconfirming outcomes are a different order, no ordering, or a winner that does not move across the bands. The oldest band is populous in workers but sparse in moves, since the move rate falls with age, so we read its binary result only against the sparse-cell power check of @sec-AE, which establishes whether a graded signal of the size seen in the younger bands would have been detectable there had it been present.

One qualification to the nearest-rival rule is forced by the selection analysis of @sec-rp, and it is fixed here rather than chosen at access. In a sub-regime whose hypothesised governor is a graded metric, the selection into the local sample depresses binary's measured fit below what a preference-neutral sample would produce, so a graded metric's lead over binary is inflated by the selection rather than earned against it. We therefore do not count a graded metric's margin over binary toward its decisive-win condition in the Skill and Hierarchy sub-regimes: the graded governor must lead the other graded metric by 10\% at all three scoring points, and binary is set aside there as a downward-contaminated rival. The exclusion runs one way and only where the contamination does. It does not touch the Binary sub-regime, where binary is the hypothesised governor and the same selection runs against it, so its win over the graded metrics is earned against the grain of the bias and is scored against both (@sec-subregime-predictions). It does not touch the pooled global race, where binary's competitiveness is the phenomenon under test and its exclusion would delete the contest whose null result motivates the partition. The standard is mechanical, not discretionary: a rival is set aside only where its off-diagonal commitment leaves it exposed to the proximity mass the selection adds while the candidate's does not, which is the graded-over-binary case alone.

#### Axis two: the non-local grid, stripped by origin

The second grid tests each sub-regime's governor across the local/non-local boundary, on the non-local transitions stripped to that sub-regime's own origin occupations. A worker's 2016 occupation is observed even after a new attainment resets where their human capital is deployed, so the non-local sample splits by origin sub-regime, and a governor is tested only against the origins it was claimed to govern. The governor tested is the one the subset hypothesises in advance, not the one it realises, so the test is well defined even where the in-regime cell does not resolve; what such a case costs is interpretive, since the dissociation the test belongs to needs the in-regime leg to hold, not the running of the test itself.

The committed predictions sit on the diagonal and come in two kinds. For the graded sub-regimes the prediction is falsification, by the below-independence method of @sec-local-predictions: skill on the Skill-origin non-locals and hierarchy on the Hierarchy-origin non-locals are predicted to fall below independence, the credential reset having voided the origin their geometry priced. In regime each graded governor peaks above independence on both measures; out of regime its $\Delta CWTVD$ peak is predicted to fall below it.

For the Binary sub-regime the prediction is persistence, not falsification. Binary makes no distance bet in regime, so there is no geometric bet for the reset to break; but the Binary hypothesis is that these occupations defy distance-based rationalization, and that is a claim we hold rather than know, in regime and out alike. The out-of-regime form of it is invariance: we predict binary's fit on the Binary-origin non-locals resembles its fit on the Binary-origin locals, the sub-regime's defiance of distance surviving the one perturbation that resets everyone else. This is the sub-regime-localised counterpart of binary's role as the flat control in the global race, and it is corroboration by invariance rather than a below-independence crossing, since binary has no geometry to be caught pricing wrongly.

The Binary-origin non-local cell carries a second, independent prediction on the entry side, and it is a prediction about a rate, not about presence. The lock-in mechanism that defines the sub-regime predicts a low re-credentialing rate out of high-OES occupations, because workers already in a high-OES occupation rarely acquire a new highest attainment: occupation-specific human capital is reached through a single credentialing pipeline and confers no incentive to re-credential. Rarely is not never, so the mechanism predicts the cell is sparse, not empty, and a few workers in it are exactly what a low rate produces and confirm the premise rather than breaking it. The falsifying observation is accordingly a rate, not a body count: the cell falsifies lock-in only if workers leave Binary-origin occupations to re-credential at a rate comparable to the rate out of the Skill and Hierarchy origins, so the test is the Binary-origin re-credentialing rate against those graded-origin baselines, with the Binary-origin stock as the denominator, and a single worker or a handful reads as confirmation of near-emptiness. This entry-side rate prediction and the persistence prediction the populated part of the cell carries pull in opposite directions on the same cell, and the reconciliation is that they speak to different quantities: lock-in wants a low rate relative to the graded origins, while persistence wants enough absolute moves to score the horse race, and a small count can be both a low rate against a large Binary-origin stock and enough moves to fit a race. Where the cell is too sparse even for that, the persistence leg is reported as underpowered rather than run, per @sec-AE, while the rate reading stands on its own. The remaining out-of-regime content of the Binary hypothesis lies further on the entry side and is carried by the EOS--OES association of the non-local model rather than by a horse race.

The off-diagonal cells of this grid carry no committed prediction. We score them so that a subset whose realised winner is not its hypothesised governor can still be read out of regime: if skill unexpectedly leads the Hierarchy locals, its behaviour on the Hierarchy-origin non-locals is then available as reported exploration rather than a post-hoc scramble. Scoring the full grid on both axes, within-local and origin-stripped non-local, is what keeps every result, expected or surprising, legible out of regime.

## Non-Local Model {#sec-non_local}

The non-local regime is where the horse race predicts occupational geometry to fail: once a new attainment resets the worker's position, distance in skill or hierarchy space no longer orders where they go, and the geometric metrics collapse toward independence. That failure is a negative result on its own, a regime the geometry does not govern, and it leaves a question the horse race cannot answer, what governs there instead. This section is the paper's positive account of that regime, and the reason to care about the relationship it tests is that account rather than the relationship taken by itself.

The stakes are threefold. First, the paper's thesis is that mobility is not one thing, and a dissociation is far stronger with both arms than with one: the horse race supplies the negative arm, that no single geometry governs non-local moves, and an EOS--OES linkage supplies the positive arm, that credential specificity orders those moves where distance does not. Without it the non-local regime is a hole in the geometry; with it the regime is governed by a second organizing principle orthogonal to the first, and *not one thing* becomes the presence of two structures rather than the absence of one. Second, the linkage is policy-actionable in a way distance is not: occupations cannot be moved closer in skill space, but credentials can be made more or less specific, and if specificity-matching governs the non-local regime then the design of credentials is a lever on where retrained workers land and how tightly they are channelled there, which speaks directly to apprenticeship and credentialing policy. Third, the strength of the relationship is itself the quantity of interest, a weak linkage meaning credentials are loose signposts and a strong one meaning they are rails.

All three stakes depend on the relationship carrying information beyond a tautology. That specific credentials pair with specific occupations is nearly true by construction, since a credential is defined by the occupation it gates; the finding earns its stakes only if it says more than that credentialed workers hold the jobs their credentials are for. The gap and collapse below are what supply the more: they establish that the linkage is forward-directed and triggered by acquisition rather than a standing feature of who holds what, and it is that forward, credential-triggered content, not the bare pairing, that the three stakes rest on. Should the gap come in near zero, the relationship reduces to the tautology, the non-local regime has no positive account, and the honest reading is that it is governed by nothing the design can name, a null that is informative for the thesis but a materially weaker and different result.

The non-local model asks whether the specificity of a worker's newly acquired education and the specificity of the occupation they reach are linked at the individual level. It is built from the two marginal specificities defined at @sec-rp, OES on the occupation side and EOS on the education side, read off the public education$\times$NOC table, together with a pairing that only the linked file supplies.

Each marginal is defensible on its own and neither is derivable from the other. What the public table cannot establish is that they are connected: that the high-EOS educations and the high-OES occupations are the same specificity showing up at the two ends of a worker's history, the education acquired and the occupation later held. The top of the EOS ranking and the top of the OES ranking contain fields and occupations whose names echo one another, but a cross-section is a standing stock, not a record of sequence, and co-occurrence in a snapshot cannot distinguish an education acquired before the occupation was reached from two independently specific margins that happen to share a vocabulary. Establishing the link requires observing the same worker acquire the education and then arrive at the occupation, which is what the linked file provides and the public table cannot.

For each credential-acquirer we attach the EOS of the education acquired within the window and the OES of both their 2016 and 2021 occupations, and ask first whether the acquired education's EOS co-varies with the destination occupation's OES across workers. The prediction is a positive individual-level association: workers who acquire highly specific educations arrive in highly specific occupations. We make no causal claim. The association is a pattern, that the two specificities travel together through realized transitions, and the mechanism we offer for it, occupation-specific human capital opening a path to the occupation that uses it, is an interpretation of the pattern rather than something the estimate identifies.

Because EOS is a property of the education and OES a property of the occupation, every worker sharing an education carries the identical EOS and every worker sharing an occupation the identical OES, so the individual-level association is a pattern over education--occupation pairings weighted by how many workers traverse each. We estimate it as the slope of a standardized regression, which recovers the correlation, and cluster standard errors three ways, on education, on occupation, and on worker, to respect that dependence: education and occupation carry the shared-value dependence just described, while worker enters because the stacked origin--destination specification of @sec-non_local-predictions contributes two rows per worker, and the two share the same person. The estimator we read from this clustering is selected by the effective cluster counts at access, the analytic three-way variance where the counts permit and a permutation test where they do not, as set out in @sec-non_local-inference.

**The gap.** A positive EOS--OES association on its own is weak evidence, because a latent disposition toward specificity, a worker drawn to specific credentials who also sorts into specific occupations for reasons unrelated to any path between them, would produce it. We separate the mechanism from that disposition with a within-worker contrast: we estimate the acquired education's EOS association with the destination occupation's OES and with the origin occupation's OES, and compare them. The mechanism predicts the association is stronger for the destination than for the origin, and the gap between the two is what the newly acquired credential adds.

The gap is robust to the latent-disposition confound for a reason of timing. A time-invariant disposition toward specificity loads on origin and destination alike, since it is a fixed trait of the worker present at both the origin and destination occupations, so it raises both associations together and differences out of the gap. The newly acquired credential cannot load on the origin, because the origin predates it: the credential did not exist, operatively, when the worker occupied the 2016 occupation, so whatever forward reach it has toward the destination it cannot have toward the origin. The gap therefore isolates the credential's forward reach from any fixed specificity the worker carried into the transition.

One form of the disposition is not time-invariant and so is not differenced out by this argument: a disposition that drifts forward over a career, creeping up the specificity ladder through ordinary opportunity rather than through any credential. Such a drift separates origin from destination in the same direction the credential does, because both are functions of elapsed time within the window, so timing alone cannot tell a forward-reaching trait from a forward-reaching credential. What separates them is the cross-sample falsification of @sec-non_local-predictions. A credential-independent drift operates whether or not a credential is acquired, so it should raise the destination association above the origin association among local movers too, where no credential was acquired within the window. The mechanism predicts the forward gap collapses there; a forward drift predicts it persists. The local-mover check therefore discriminates the two: a collapse rules out a drift that operates alike whether or not a credential is acquired, and what remains is a drift concentrated in the credential-acquirers themselves, either switched on by the credential, which is not an alternative to the mechanism but the mechanism itself, or running faster in the workers who retrain, which is the asymmetric-forward disposition set aside among the identifying assumptions as the one the design does not discharge. The only confounder the gap does not remove is thus the same one it never claimed to, a trait that links the acquired credential specifically to the destination and not the origin, which is the mechanism the design describes rather than an alternative to it.

We estimate the gap in a single model, stacking each worker's origin and destination OES into one outcome column with a role indicator and interacting EOS with role; the interaction coefficient is the gap, and its standard error, clustered on education, occupation, and worker, is the inference. The companion display is a two-panel traffic-weighted scatter, destination OES against EOS beside origin OES against EOS, in which the gap is the difference in slope between the panels.

#### Inference {#sec-non_local-inference}

The association is estimated over individuals, but EOS is a property of the education and OES of the occupation, so all workers sharing an education carry the identical EOS and all sharing an occupation the identical OES. Treating workers as independent would badly overstate precision. We keep the worker as the unit of observation, which fixes the point estimate at the traffic-weighted pattern, and place the dependence in the variance estimator; aggregating to education--occupation pairings first would not help, since pairings still share educations and occupations and would themselves require the same clustering. The variance is clustered three ways. Education and occupation carry the shared-value dependence just described; worker enters as a third margin because the stacked origin--destination specification contributes two rows per worker, sharing the same person, so the worker margin is required for correctness even though it is not a fragile one, its clusters being uniformly of size two and hence numerous and balanced. An occupation forms a single cluster spanning every row in which it appears, whether as a worker's origin or another's destination: the shared OES value it induces is the same number regardless of role, so the residual correlation it carries spans both roles, and splitting the origin and destination appearances of one occupation into separate clusters would miss exactly that cross-role dependence. The occupation cluster is therefore the occupation, not the occupation-by-role.

The point estimate is well powered but the inference is not automatically so, and the two turn on different quantities. Every variable the non-local model requires, occupation in both cycles, field of education, and work activity, is long-form only, the short form being a demographic head count that carries none of them, so the estimand lives on the intersection of the two 25\% long-form samples, on the order of a sixteenth of the population. In worker terms this is still a large sample, hundreds of thousands of usable transitions even after the credential-acquisition condition and the observed-occupation requirement, so the point estimates are well powered; scarcity of workers is not the concern. The concern is the cluster count, a different and far smaller quantity: inference runs on the number of distinct educations and occupations, not the number of workers, since everyone sharing an education carries the identical EOS and their residuals are correlated. A margin can hold hundreds of thousands of workers across a few hundred clusters. Cluster-robust asymptotics depend on the effective number of clusters, well approximated by the inverse-Herfindahl count $(\sum_g n_g)^2 / \sum_g n_g^2$ over cluster sizes $n_g$, rather than on the nominal number of educations or occupations; a skewed size distribution that concentrates workers in a few large clusters drives the effective count far below the nominal one, and on a populated but concentrated margin it can land in genuinely few-clusters territory rather than merely below its nominal value.

Which of education and occupation is the more fragile margin is not settled by the construction, and we do not claim to predict it. What the sampling does support is a one-sided claim about occupation. The sample conditions on credential acquisition, and among acquirers are a set of very high-traffic generic fields, the bachelor's degrees in arts, commerce, and general science that dominate credential flows. These fields fan out across the labour market, dispersing their holders over many destination occupations, so the occupation margin receives workers from many educations and no single occupation swallows a dominant share. That dispersion is what supports the occupation margin's effective cluster count: the fanning-out educations spread destination mass widely, which is the condition a healthy occupation margin needs. The education margin is a different matter and we leave it open. It is a mixture, the same high-traffic generic fields alongside a tail of specific credentials that funnel narrowly, and whether its effective count clears the few-clusters range depends on the whole size distribution across educations, which the fan-out argument does not characterize. The presence of a few large generic fields does not by itself settle the education count, since a populous spread of mid-sized educations can carry it even when a handful of fields are large. We therefore expect occupation to be well-populated in effective terms and make no matching prediction for education.

This occupation claim is worth pressing on directly, since it is the one the regime choice can lean on. For occupation to concentrate despite the fan-out, the large generic credentials would have to feed a small common set of high-inflow catch-all destinations, general managers, administrative officers, retail supervisors, piling occupation-margin mass onto a few large clusters. This is not impossible, since fan-out requires destinations to be numerous but not that inflows be even, and generic credentials do feed catch-all occupations disproportionately. But it requires destinations to be many yet lopsided; the alternative, a single credential feeding a single occupation, concentrates occupation only in the specific-credential profile that is by construction the low-traffic tail and cannot dominate the margin. High traffic and tight destination concentration are close to mutually exclusive in this sample, which is why we expect occupation to clear the few-clusters range. Nothing beyond that rides on it: both effective counts are computed at access and the regime is chosen by them, so the fan-out argument supplies only a reason to expect enough effective occupation clusters, and the permutation fallback covers whatever the education count turns out to be.

We do not rest on the prior, since the effective counts are unknown before access, and we commit to the full decision rule now. We report the effective cluster count, by the inverse-Herfindahl measure above, for education, occupation, and worker at access, and the estimator we read is selected by those counts before the gap is seen. Where all three effective counts clear the few-clusters range, primary inference is the analytic three-way cluster-robust variance, which handles all three margins at once and delivers a confidence interval on the gap, the quantity the section is built to read. Where any margin's effective count falls into few-clusters territory, the analytic variance is untrustworthy on that margin, and no one-margin resampling repairs it cleanly: a wild cluster bootstrap resampling the fragile margin corrects that margin only by ceasing to cluster the others, trading a broken leg for a dropped one rather than fixing the estimator, and there is no cleanly specified multiway wild cluster bootstrap that resamples the fragile margin while still clustering the rest. We therefore do not use the bootstrap. Primary inference in the fragile regime is instead a permutation test on the EOS--role interaction. Holding OES, occupation, role, and the within-worker pairing fixed, we permute the acquired-education labels across workers, both of a worker's stacked rows carrying the same reassigned EOS, and rebuild the null distribution of the gap directly. This handles the two dependences in a single operation and for a single reason: permuting education labels as whole-worker blocks preserves the education tie structure exactly, so the test never invokes a large-cluster asymptotic on education and its validity is indifferent to how few effective education clusters there are; and because occupation, OES, and role are never disturbed, the occupation dependence is carried into every permuted dataset identically to the real data rather than dropped, so it too is handled without an asymptotic. The permutation test rests on exchangeability of EOS across workers under the null, which is the same maintained identifying assumption the design carries elsewhere, the asymmetric-forward disposition of the identifying-assumptions section, and not a new or weaker one; it fixes the inference problem the cluster counts pose without touching identification. What it delivers is a valid $p$-value for the null of no forward gap rather than an interval on the gap's magnitude: we accept that trade, reading significance rather than a trustworthy interval in the fragile regime, while the point estimate, which is well powered regardless of cluster counts, still carries the magnitude. One limit on the fallback is worth stating in advance: if the education margin is concentrated, the number of distinct whole-worker relabelings is limited, and the achievable $p$-value granularity is bounded by that concentration, so in a highly concentrated case the permutation floor is coarser than a continuous $p$-value; we report the realized granularity at access rather than presume it fine. We report which regime obtained and which estimator it selected, rather than choosing after seeing the gap.

To keep each cluster non-degenerate and to bound the leverage of thinly-acquired cells, the association is estimated over educations acquired by at least ten workers in the non-local sample and, symmetrically, over occupations reached by at least ten, a cluster-floor of ten set in advance and applied to both clustering margins at access. Ten is a convention rather than a theorem: it clears with margin the range below which a cluster contributes negligible independent information and singleton and doubleton clusters destabilize the cluster-robust variance, and it is round enough to carry no hidden researcher degree of freedom. What makes it defensible here is that it is fixed before access on a data-independent rationale, not that any external standard blesses the number; it is an analytic choice about what enters the estimation, distinct from and prior to any decision about what output may be released. The cluster-floor applies symmetrically because the floor guards against a different failure than concentration and either margin can suffer it: singleton and doubleton clusters destabilize the cluster-robust variance on whichever margin they fall, and neither margin is exempt from a thin tail, education from rarely-acquired credentials and occupation from rarely-reached destinations, so protecting cluster count on only one would leave the other exposed. The same cluster-floor incidentally removes the rarest and most misreport-prone credentials, attenuating the masking discussed in @sec-non_local-limits. The magnitude of the association is the substantive quantity: where cluster counts permit we report it with a cluster-robust interval, and where they do not we report the point estimate with a permutation test of the null, reading significance there rather than a trustworthy interval.


### Predictions and out-of-regime falsification {#sec-non_local-predictions}

We commit in advance to three predictions, following from the mechanism of @sec-non_local rather than fitted to the data, together with a validation check on the first. The first is a baseline the other two are read against, and the second and third are falsifications, one within the non-local sample and one against the local sample; both falsifications turn on the timing logic already given, that a freshly acquired credential postdates the 2016 occupation and so can reach forward to the destination but not back to the origin. The identifying content is in the second and third, not the first: the baseline association is confounded on its own, and it is the gap and the collapse that separate the mechanism from a standing specificity disposition, down to the single asymmetric-forward residual named in @sec-non_local as the one the design does not discharge.

First, on the non-local sample, the EOS of the acquired education is positively associated with the OES of the destination occupation. This is the baseline, not yet a finding: on its own it is consistent with the pipeline mechanism and equally with a standing disposition toward specificity that sorts a worker into specific credentials and specific occupations alike, and it does not discriminate the two. Its role is to establish the existence, level, and sign of the linkage that the gap then differences and the local-sample collapse falsifies against; a within-worker forward gap is uninterpretable without the baseline association it is a contrast within. What identifies the mechanism is the conjunction of this baseline with the two predictions below, not the baseline itself.

Second, on the non-local sample, the destination association exceeds the origin association: the EOS--OES gap is positive. This is the within-sample falsification. A gap of zero would say the acquired education is no more tied to where the worker went than to where they started, which is what a standing disposition toward specificity, linking both occupations alike, would produce; a positive gap is the signature of the credential doing the placing, and it is what marks the pattern as credential-driven rather than a trait of the worker.

Third, on the local sample restricted to workers who moved, the gap collapses. Absent a new credential the worker's education is stale, operative already in 2016 and unchanged through 2021, a standing attribute carried through the move rather than a force that produced it, so it links to origin and destination at similar strength and shows no forward advantage. This is the cross-sample falsification, and it uses the local movers. The contrast between the two samples is not between movement and stasis, since both are conditioned on movement, but between a credential acquired within the window and one predating it: the forward gap is present when the credential is new enough to have done the placing and absent when it is not. We commit to the collapse; a stale credential that showed the same forward reach as a fresh one would falsify the timing account on which the mechanism rests. The collapse also carries the weight of separating the mechanism from a forward-drifting disposition: such a drift is credential-independent and would raise the destination association above the origin one among these credential-less movers, so its absence here is what rules it out where the differencing of @sec-non_local could not. A gap that instead leaned backward, the stale education linking more strongly to origin than destination, would corroborate the account further, but we do not commit to it, and the committed prediction is the collapse of the forward advantage rather than its reversal.

A validation accompanies the demonstration. The association of the first prediction is between two marginals, the education's EOS and the occupation's OES, and a marginal association cannot by itself distinguish matching from coincidence: a high-EOS credential whose holder lands in an unrelated high-OES occupation, an optometry graduate who somehow becomes a dentist, contributes to the association exactly as a genuine pipeline does, though the credential does not feed that occupation at all. What separates the two is the lift of the worker's realized education--destination cell, $m(e, n_{2021})$, the one quantity that reads the pairing rather than either margin. We commit in advance to the association being concentrated in high-lift realized transitions: restricted to movers whose acquired credential distinctively feeds their destination, the association is present, and among movers whose realized cell is at or below baseline lift, a specific credential landing where it has no pipeline, it is not. It is the contrast between the two subsets that certifies matching, not the vanishing in the low-lift subset alone: the association's survival among the genuine pipelines rules out that its absence among the coincidental pairings is merely the lower power of a smaller or noisier subset. Were the association instead indifferent to realized lift, the two specificities would be co-occurring in workers without the channel that connects them, and the matching interpretation would fail. At access we also report the high-EOS, high-OES pairings by traffic with their lift, a legible catalogue in which the genuine pipelines appear at high lift and the coincidental pairings, if the interpretation is right, do not appear at all.

Two features of the samples bear on the cross-sample falsification, and both make it conservative. First, a subtle contamination runs the other way: a credential-gated move whose increment went unrecorded, the nurse-aide to licensed-practical-nurse case of @sec-cf, is routed local by the proxy classification though the mechanism would call it non-local. Such a leaked worker carries genuine forward reach into the local sample, which works against the predicted collapse rather than manufacturing it, since a leaked wormhole can only lift the local forward gap toward the non-local value and narrow the contrast. The lift-based reclassification of @sec-rp recovers these workers, and the proxy-versus-rescued contrast measures how much leakage there was; the collapse is predicted under both classifications and more sharply under the rescued one. Second, the credential-less mover sample is not thin but large: credential acquisition within a five-year window is rare, so non-acquirers vastly outnumber acquirers and the local-mover falsification runs on the populous side of that split. The falsification is therefore well powered to detect a credential-independent forward drift if one exists, and a collapse is strong confirmation rather than a failure-to-detect on a thin sample. What survives the falsification does so for a structural reason, not a power one: a forward-reaching disposition that is coextensive with credential acquisition, present only in the workers who retrain, is absent from the credential-less sample by construction, so no sample size can detect it there. The large sample decisively removes every credential-independent alternative and leaves exactly the credential-bundled residual, which is unfalsifiable by this route in principle rather than merely unmeasured, the asymmetric-forward disposition named in @sec-non_local as indistinguishable from the wormhole by construction. We report the local falsification at access with its realized size, but its force rests on the structural argument, not on clearing a power bar the large sample clears easily.

## Results \[PLACEHOLDER: awaiting data\]  {#sec-res}

### Local

\[Peak $\Delta KL$ and peak $\Delta CWTVD$ under each informative judge
($C_{\text{skill}}$, $C_{\text{hierarchy}}$) by cost metric, $\varepsilon$
and subsample; the two geometric panels are the bracket endpoints\]

\[Same, by 3 OES sub regimes, cost metric, $\varepsilon$ and subsample\]

\[Same, by 3 TEER sub regimes, cost metric, $\varepsilon$ and subsample\]

\[Age extension: same grid, by 3 age bands (15--25, 25--40, 40+) in place of OES, with the prediction that the winning metric shifts skill, hierarchy, binary across the bands; the ordering is the committed content, margins expected thin, and the 40+ binary result read against the sparse-cell power check of @sec-AE; the confusion between the age-band and OES sub-regime assignments reported alongside, as the OES--TEER $\kappa$ is, to quantify how far the two partitions cross-cut\]

\[Out-of-regime falsification: the non-local move rate reported first, as the descriptive first stage of @sec-local-falsification establishing whether the non-local matrix is movement-heavy; then the predicted metric's peak $\Delta CWTVD$ under the neutral blend in each of the Skill and Hierarchy sub-regimes, in regime and out, read below independence conditional on that first stage\]

### Non-local

\[two-panel traffic-weighted scatter: destination-OES against acquired-education EOS
beside origin-OES against EOS, non-local sample, points sized by pairing count\]

\[gap table: the EOS$\times$role interaction (destination minus origin association)
from the stacked within-worker specification, under both classifications; the
effective cluster count for education, occupation, and worker reported with the
table, and the inference selected by those counts, the analytic three-way
cluster-robust interval (clustered on education, occupation, and worker) where
all three counts permit, and the permutation test of @sec-non_local-inference on
the EOS$\times$role interaction where any margin is fragile, reported as a
$p$-value beside the point estimate with its realized granularity\]

\[matching validation: the demonstration association re-estimated split by the lift
of each mover's realized acquired-education--destination cell, high-lift versus
at-or-below-baseline, with the prediction that it is carried by the high-lift
split; and a traffic-ranked catalogue of the high-EOS, high-OES pairings with
their lift\]

\[cross-sample falsification: the same gap estimated on the local sample restricted
to movers, under both classifications, with the prediction that it collapses;
realized subsample size reported\]

\[latent-activation robustness: reinstated count and the rescued non-local gap
estimate at the primary value ($\tau = 2$, lift-floor $= 100$), beside the uncorrected
gap; full $\tau \in \{2, 3, 4\}$ sweep at fixed lift-floor in @sec-AD\]

## Limitations {#sec-limitations}

### Local Mobility

Two sample restriction mechanisms introduce a potential selection
concern for local mobility. Non-linkage and non-observability of
occupation are both disproportionately concentrated among
low-attachment workers, individuals with casual, seasonal, or
intermittent employment whose identifying information is also less
stable across census cycles, so the two filters largely co-select
the same low-skill population rather than introducing independent
exclusions. This shared exclusion is not neutral in occupational
space: it concentrates in the low-OES, low-skill region and removes
disproportionately short-range transitions, the adjacent low-skill
moves that casual employment generates. Its first-order effect is on
the representativeness of population-level transition estimates
(external validity).

Whether it also threatens the internal validity of the horse race
turns on a narrower condition than the occupational concentration
itself. The horse race is a relative comparison, so removing a region
of occupational space does not bias the adjudication between metrics
unless, within that region, the excluded transitions are
differentially rationalizable by one metric than the retained ones.
Concentration alone is insufficient: if excluded moves are easier for
every metric to fit, because they are short and short moves are easy
for any cost structure, all three metrics' fits fall together and the
ranking is undisturbed. The bias requires the excluded moves to be
easier for one particular metric, and the natural candidate is the
skill metric, since a casual worker taking the nearest available job
produces exactly the short-in-skill-space transition the skill metric
rewards. The live threat is therefore specific: excluding
low-attachment workers may deflate the skill metric's measured fit
relative to hierarchy and binary, biasing the race against skill.

This threat cannot be tested directly, because the excluded
transitions are by construction not in the sample: an unlinked or
occupation-missing worker has no origin--destination pair to score
against any cost matrix. Two indirect handles bound it instead, each
testing sensitivity along an observable attachment gradient and
extrapolating to the unobserved. First, labour-market attachment is
itself partially observable on the in-sample workers through the
census work-activity classification, which orders individuals by weeks
worked and full- versus part-time intensity over the reference year.
We re-estimate the local race on the low-attachment and high-attachment
strata and report whether the metric ranking, and the skill metric's
fit in particular, moves between them; a skill tilt among observed
casual workers is the best available evidence for the sign of the bias
from excluding the unobserved ones. Three caveats bound this handle.
The work-activity reference year is the calendar year before each
census, so the 2021 measure falls in 2020 and is contaminated by
pandemic labour disruption, for which reason we prefer the 2016 (2015
reference) measure where both are available. Work activity is a
long-form (25\% sample) variable, which is no additional restriction
here, since the linked frame is already the intersection of the two
long-form samples. The stratum that bears on the threat is unlikely to
be the scarce one: low-attachment workers are a large share of the
labour force, and the sample is conditioned on occupational
transitions, which low-attachment workers make more often than
stably-employed ones, so mobility conditioning tends to over-represent
rather than deplete them, and the low-OES region they occupy is where
transitions concentrate rather than thin out. We therefore expect the
low-attachment stratum to be adequately populated for the stratified
race, but do not rely on it: the realized cell counts in that stratum
are not known until access, and where they turn out too sparse to
detect a skill tilt of the magnitude at issue, the stratified race is
reported as underpowered rather than as evidence of no tilt, per the
detection-power logic of @sec-AE. Absence of a detectable tilt in a
stratum too small to detect one is not evidence the tilt is absent.
Second, where the linkage is probabilistic and carries a match score,
we sweep the linkage threshold and re-run the race across progressively
looser link sets, which admit disproportionately the marginally-linked,
identifier-unstable workers; a ranking stable across the sweep is
evidence the contest is insensitive to link quality, the
identifier-stability correlate of attachment.

Neither handle is a direct test, and their force rests on a
monotonicity assumption we state rather than verify: that the
unlinkable and occupation-missing workers lie further along the same
attachment gradient as the marginally-linked and observed-casual
populations, resembling more of the same rather than a qualitatively
different group. This is an identifying assumption of the same kind as
the asymmetric-forward disposition named for the non-local gap, one the
design bounds but does not discharge, and we report the two
sensitivities as evidence for it rather than as proof that the
exclusion is innocuous.

A notable omitted factor for local mobility  is \emph{path-specific} wage differentials. The additively separable component of wage gains, recoverable from origin and destination wage levels alone, is absorbed by the 
margins insofar as high-wage destinations attract larger inflows and 
low-wage origins larger outflows. The residual dyadic component, 
reflecting transition-specific wage premia and discounts beyond what 
origin and destination wage levels predict, is an omitted factor. 
Short-distance transitions are within a similar occupational 
neighbourhood, reducing employer uncertainty about candidate fit and 
productivity; long-distance transitions cross larger skill or 
hierarchical gaps, and employers may discount wage offers to 
compensate for that uncertainty, or workers may accept lower wages 
to gain entry to a distant occupation.

Thus, path-specific wage differentials are better understood as an 
alternative mechanism, rather than a confounder. If employers use the same metric that governs mobility to assess candidate fit uncertainty, discounting wages for transitions that are long under the winning metric, that discount reinforces worker selection toward shorter transitions under 
the same metric. Both channels predict identical patterns of 
transitions and are equally consistent with the winning metric being 
the correct description of the mobility geometry. The horse race 
identifies which distance geometry best rationalizes observed 
transitions in a reduced-form sense, but cannot separately attribute 
that rationalization to worker or employer preferences regarding 
distance.

### Non-Local Mobility {#sec-non_local-limits}

The non-local design faces limitations of four kinds, taken here in turn: a scope restriction on which moves the estimand covers, two measurement threats that bias the association downward, an inferential limitation from the small number of clusters the standard errors rest on, and a limit on what the gap identifies, which is a directional pattern rather than causation.

The non-local regime studies non-negative operative increments, as defined in @sec-rp: transitions where the operative credential rises above the worker's prior ceiling or respecifies at the same level. Substitutive moves, where an over-credentialed worker retrains below their ceiling and repositions on a sub-maximal credential, lie outside the estimand and are invisible to a highest-attainment record; the findings therefore speak to additive and lateral specific-human-capital activation, not to occupational mobility of every kind. This is a scope of the question rather than a gap in the answer, and it is consistent with the paper's premise that mobility is not one thing: the substitutive channel is a distinct phenomenon with its own economics, deserving its own treatment rather than absorption into this one.

The non-local partition is restricted to individuals who acquire a new highest educational attainment between 2016 and 2021 and report an occupation in both periods. A 2016 occupation is observed if and only if the individual was employed between May 1st and 7th, the 2016 census reference week. For the 2015/2016 school year, roughly half of 18-29 year old students reported concurrent work, with an (unweighted) ten percentage point decrease for university students, consistent with more intensive programs suppressing concurrent employment.^[source: Statistics Canada table 37-10-0104-01.] The Census reference week provides a partial offset: falling after the typical post-secondary academic year ends, it captures individuals who have recently finished exams and may have already begun summer employment. But this offset is weaker for intensive professional programs that are either year-round or have students for whom summer employment is atypical. The resulting truncation is not random: missingness is concentrated among the most occupationally specific educations, the high-EOS credentials that anchor the strongest pipelines, where the association the model predicts is sharpest. A worker missing a 2016 occupation is dropped, so these are disproportionately the specific-credential acquirers whose destination the mechanism most cleanly predicts, and their absence weakens the measured EOS--OES association rather than strengthening it, so the estimate is a lower bound. Notably, this missingness pattern is not purely a limitation: the disproportionate absence of 2016 occupations among the most specific credentials is itself consistent with the wormhole mechanism, since program intensity and specificity are precisely what suppresses concurrent employment. In the online companion we plot each education's occupational specificity against the share of its workers missing a 2016 occupation, quantifying the magnitude of this truncation.

```{r}
# #| label: tbl-working-while-studying
# #| tbl-cap: "Share of students employed during program, by age and institution type"
# 
#  tbbl <- cansim::get_cansim("3710010401")|>
#    janitor::clean_names()|>
#    mutate(age = str_extract(age, "\\d+"))|>
#    filter(ref_date=="2015/2016",
#           type_of_institution_attended %in% c("College","University"),
#           age %in% 18:30)|>
#    select(age, type_of_institution_attended, value)|>
#    pivot_wider(id_cols = age, names_from = type_of_institution_attended, values_from = value)
# 
#  tbbl_mean <- tbbl|>
#        summarise(age = "Unweighted Mean", across(c(College, University), mean))
# 
#  bind_rows(tbbl, tbbl_mean)|>
#    kbl(
#      digits = 1,
#      col.names = c("Age", "College (%)", "University (%)"),
#      booktabs = TRUE,
#      linesep = ""
#    ) |>
#    kable_styling(latex_options = c("HOLD_position")) |>
#   footnote(
#    general = "Statistics Canada Table 37-10-0104-01, 2015/2016.",
#    general_title = "Source:",
#    threeparttable = TRUE,
#    escape = FALSE
#  )|>
#    row_spec(nrow(tbbl)+1, bold = TRUE)  # bold mean row
```

A second measurement issue runs through the education record rather than the occupation record. Because education is highest attainment crossed with field, a worker who holds the credential that gated their occupation but has since earned a higher, unrelated credential is recorded under the later one. The canonical case is a generalist graduate degree stacked on a specialist bachelor's: an engineer who completes an MBA appears with a business education, not an engineering one; a nurse who completes a health-administration master's appears with a management education, not a nursing one. In each the gating credential is masked. This misattributes some specific-credential acquirers to generic educations, lowering their recorded EOS while leaving their destination OES intact, which weakens the measured association. Masking is not uniform across fields: it is concentrated in short-duration, stackable credentials, since a second master's is far cheaper to accumulate than a second of most credentials, and these are disproportionately the generic professional fields. The credentials most likely to do the masking are therefore the low-EOS ones, so masking tends to hide a specific gating credential behind a generic later one. The bias is nonetheless bounded: masking can only detach a specific credential from the specific destination it produced, never manufacture a pairing that did not occur, so the EOS--OES association and the gap are lower bounds on what an unmasked record would show.

The inference, as against the point estimate, rests on a small number of clusters. As set out in @sec-non_local-inference, the estimand lives on the intersection of the two long-form samples, a large sample of workers but one whose standard errors are governed by the number of distinct educations and occupations rather than the number of workers, and on the fragile of those two margins the effective cluster count can fall into few-clusters territory. This is the binding inferential limitation of the non-local design: where that count is small the analytic cluster-robust variance is unreliable on the fragile margin, and no single-margin resampling repairs it without dropping the clustering on the others, so primary inference falls back from the analytic three-way variance to the permutation test of @sec-non_local-inference, which is valid regardless of cluster counts but delivers a $p$-value on the null rather than an interval on the magnitude. A small realized count therefore costs the interval, not the point estimate, and weakens the strength of any claim the gap can support rather than being absorbable by sample size. The magnitude the gap identifies is unaffected; only the precision with which it is pinned down is at issue.

The gap identifies the credential's forward reach by differencing out any specificity the worker carried into the transition, and its robustness rests on a timing argument rather than on an untestable exclusion: a time-invariant disposition toward specificity loads on origin and destination alike and cancels, while the acquired credential cannot load on an origin that predates it. A disposition that drifts forward over a career is not cancelled by the differencing, since it shares the credential's direction, but it is separated by the local-mover falsification, where a credential-independent drift would surface and the mechanism predicts none. The one confounder the gap does not remove is a trait that links the acquired credential specifically to the destination and not the origin, but such a trait is the mechanism the model describes rather than an alternative to it, so the gap cannot separate the mechanism from a relabelling of itself. What the design does not deliver is causation. The gap establishes that the acquired credential's specificity attaches to the destination and not the origin, a directional pattern in realized transitions; whether the credential opened a path the worker then took, or the worker sought the destination and acquired the credential to enter it, the pattern is the same, and we do not claim to distinguish them.

## Conclusion

- What we found \[PLACEHOLDER: awaiting data\]

The central policy takeaway from this analysis is that labour market adjustment mechanisms are not one-size-fits-all: the appropriate intervention depends on which mobility regime contains the occupation experiencing a shortage. The three mobility sub-regimes suggest three distinct policy levers.

In the Skill sub-regime, reskilling and short-cycle training are the natural adjustment mechanisms. Workers in adjacent occupations already possess much of the relevant task content, and targeted reskilling can accelerate the reallocation of labour toward occupations in shortage without requiring formal retraining. The policy implication is that investment in modular, portable training that builds on existing skill profiles will generate the largest mobility response in this segment.

In the Hierarchy sub-regime, the relevant bottleneck is not skill transferability but the hierarchy's silo and platform structure. An occupational shortage within one silo cannot be addressed by upskilling workers from a different silo, even if the occupations share a near-identical skill profile. Instead, shortages are addressed within the silo, through lateral movement across a platform and, where the step up requires no credential, movement to an adjacent one. Where the upward step is credential-gated, it is not a within-silo mobility problem at all but a program-capacity one, and belongs to the lever of the Binary sub-regime below: the gating credential is acquired through a specific program whose entrants are drawn from across the labour market, so relief comes from expanding that program rather than from moving incumbents up the silo. Bridging programs, mentorship structures, and targeted recruitment across platforms of the same silo are the appropriate instruments for the uncredentialed steps.

In the Binary sub-regime, either a worker has the required education, or they do not. Neither reskilling nor within-silo movement can substitute for education. Adjustment in this sub-regime depends entirely on expanding formal training pipelines: increasing capacity in the relevant educational programs, reducing barriers to program entry, and shortening time-to-completion where possible. These are also the occupations that immigration policy should target, both in terms of admission and rapid credential recognition. 

If the hypothesized sub-regime structure is confirmed, the geometry of occupational mobility determines which policy lever to pull, and the wrong lever is not just ineffective but potentially counterproductive. 

```{=latex}
\appendix
```

## O\*NET to NOC 2021 Skill Mapping {#sec-AA}

OaSIS provides occupation-level skill profiles for Canadian
classifications and would be an attractive data source for constructing
$C_{\text{skill}}$. However, OaSIS does not expose how those profiles
are constructed: the mapping between O\*NET and NOC 2021 occupations is not observable, the weights linking underlying occupations are not published, and there is no information on how multiple O\*NET occupations are combined to form an occupation profile. This appendix describes an alternative mapping constructed explicitly from published crosswalks, allowing full transparency of inputs and weights, diagnostic evaluation of mapping quality, and reproducibility of all outputs.

The mapping chains four classifications: O\*NET 2019 $\rightarrow$ SOC
2018 $\rightarrow$ NOC 2016 $\rightarrow$ NOC 2021. Each stage introduces potential one-to-many relationships. A naive approach would assign equal weight to all admissible paths, but this leads to over-weighting of occupations that appear frequently in crosswalk chains and inclusion of unrelated occupations. Instead, path weights are constructed to reflect the degree of splitting at each stage: if a source occupation splits into $n_s$ destination occupations at stage $s$, each path through that stage receives weight $w_s = 1/n_s$, and the overall path weight is $w_{\text{path}} = w_1 \times w_2 \times w_3$. Path weights are then aggregated by origin O\*NET occupation and destination NOC 2021 code and normalized to sum to one within each NOC to yield base weights. The consequences of naive versus path weighting can be severe: for NOC 72600 (Air pilots, flight engineers and flying instructors), naive weights assign five-eighths of the skill profile to agricultural workers, while path weights assign 78\% to pilots.

A power transformation and renormalization is then applied to further
reduce the influence of mapping artifacts: $\tilde{w}_k \propto w_k^2$.
This concentrates weight on the dominant O\*NET contributors and
diminishes the influence of peripheral occupations that enter through
crosswalk chains with many splits. For NOC 72600, downweighting
increases the pilot share of the skill profile to 96\%. Each O\*NET
occupation is then represented as a point in a reduced skill space using the first 12 principal components of the O\*NET skill content variables, and for each NOC 2021 code the weighted centroid of its contributing O\*NET occupations is computed using the downweighted mapping, yielding a 12-dimensional skill profile for each NOC occupation.

Two diagnostics assess the quality of each occupation's mapping in skill space. Herfindahl concentration measures how concentrated the mapping weights are across contributing O\*NET occupations: $H = \sum_k \tilde{w}_k^2$. For example, a Herfindahl of $\frac{2}{3}$ indicates that the occupation skill profile is effectively dominated by a single O\*NET occupation with a weight of $\approx .8$.  In contrast, when $H=\frac{1}{3}$ the largest contributor's share is between $\frac13$ and $\sqrt{\frac13}=.58$. Scaled Similarity measures the internal coherence of the contributing occupations in skill space: the weighted average distance of each contributing O\*NET occupation from the occupation centroid is computed, then normalized to a $[0.05, 1]$ interval by dividing by the maximum value across all occupations, multiplying by 0.95, and subtracting from 1, so that $SS = 1 - 0.95 \times \bar{d} / \max_j \bar{d}_j$. The lower bound of 0.05 is chosen to match the approximate lower bound of the Herfindahl index, making the two measures' floors commensurable. The mapping quality composite is the product $H \times SS$, which is high only when the mapping is both concentrated and internally coherent. The three interpretation cases are: high $H$ with any $SS$ (no problem, almost all weight on a single O\*NET occupation); low $H$ with high $SS$ (no problem, diffuse but coherent); and low $H$ with low $SS$ (problematic, a diffuse average of disparate occupations whose skill profile is unreliable). NOC 41200 (University professors and lecturers) is a canonical example of the third case.

The six occupations ranked below the verified benchmark 72600 (see @sec-data-public) are excluded from the local and non-local samples. The full mapping pipeline is implemented in [@onet_to_noc] and is constructed entirely from O\*NET data and published crosswalks.

Because NOC is in places a finer classification than O\*NET, some distinct NOC unit groups inherit the same O\*NET profile and so sit at zero skill distance from one another: see @tbl-zero-skill-groups for a summary of the groups.  Grouping occupations by shared profile yields `r zeros_components$no` such groups, spanning `r zeros_tabyl$total[1]` occupations and accounting for the `r zeros_components$num_zero_off_diagonal` zero off-diagonal cells noted in @sec-cost-metrics:

```{r}
#| label: tbl-zero-skill-groups
#| tbl-cap: "Zero skill distance groups."

zeros_tabyl|>
  select(group_size, n, percent)|>
  gt() |>
  fmt_percent(columns = c(percent), decimals = 1) |>
  cols_label(
    group_size = "Group Size",
    n = "Number",
    percent = "Percent"
  ) |>
  tab_options(table.font.size = "90%", data_row.padding = px(4)) 
```

The group of twelve is the processing and manufacturing supervisors, which NOC distinguishes by industry (supervisors of food and beverage processing, of plastic and rubber products, of motor vehicle assembling, and so on) but which O\*NET represents as a single first-line production supervisor occupation, onto which all twelve map with a Herfindahl concentration of one. Their shared profile is therefore not a symptom of noisy mapping but the opposite: the twelve are among the most cleanly mapped occupations in the matrix, and their zero mutual distances reflect O\*NET's occupational resolution rather than any deficiency in the crosswalk. Eleven of the twelve share the same broad category, where the hierarchical metric also places them close (distance one or two), so the two geometries agree that these occupations are near. The twelfth, 72022 (supervisors, printing and related occupations), sits in a different broad category, so the hierarchical metric assigns it the cross-category sentinel of nine to each of the other eleven while the skill metric assigns zero. This is the widest disagreement the two geometries admit, on a set of occupations the horse race assigns to the Hierarchy sub-regime; whether the skill reading or the hierarchical reading better rationalizes the observed transitions among these supervisors is precisely what the local model adjudicates.

## Lift construction {#sec-AB}

Lift is the cell-level term inside the education-side KL divergence $\text{KL}(P(\text{NOC}\mid e)\,\|\,P(\text{NOC}))$: where that divergence summarises how distinctive an education's whole occupation distribution is, lift reads a single education--occupation cell, how distinctively this education feeds this occupation. It is high where an education channels into an occupation far more than baseline, near one where it is no more likely than average to lead there, and below one where it leads there less than baseline.

Lift $m(e,n) = P(\text{NOC}=n\mid \text{education}=e)/P(\text{NOC}=n)$ is estimated from the census education$\times$NOC table (98-10-0404-01, described in @sec-data-public), and two features of that table make the raw ratio unstable in thin cells: the conditional $P(n\mid e)$ is a sample proportion, noisy when few workers hold education $e$, and the published counts are randomly rounded to base five, so a small cell's value is dominated by rounding. We address the two separately.

For the sampling noise, we smooth the conditional with a Dirichlet prior centred on the marginal occupation distribution:
$$\hat P(n\mid e) = \frac{c_{en} + \alpha\, P(n)}{T_e + \alpha},$$
where $c_{en}$ is the count of education-$e$ workers in occupation $n$, $T_e = \sum_n c_{en}$, and $\alpha$ is the prior strength in pseudo-workers. The prior is the marginal $P(n)$, so an education with no information is pulled toward "no pipeline" ($m \to 1$), and the pull is proportional to ignorance: a thin education ($T_e$ small relative to $\alpha$) is shrunk hard, a well-populated one ($T_e \gg \alpha$) is left essentially untouched. Because the prior is the same $P(n)$ that appears in the denominator of lift, the shrinkage target is exactly the ratio's neutral point, and a fully shrunk cell has lift one by construction. We shrink only the numerator; the denominator $P(n)$, estimated from the full table, is left raw, so that genuine pipelines into rare occupations, where a small $P(n)$ produces a large lift, are preserved rather than flattened. We set $\alpha = 15$, chosen so that the ranking of the strongest pipelines is stable: raising $\alpha$ far higher, to 1000, leaves the top of the ranking unchanged, since those pipelines are driven by the raw denominator rather than by numerator noise, while $\alpha = 15$ leaves well-populated cells such as optometry (education total near seven thousand) shrunk by under a percent.

For the rounding, we require a minimum realised cell count. A cell with $c_{en} = 15$ has a relative standard error near $1/\sqrt{15} \approx 26\%$ from sampling alone, above Statistics Canada's own "use with caution" threshold, and base-five rounding adds error of the same order; below roughly this count the estimated lift reflects noise and rounding rather than a real pipeline. Inspection of the public table shows that the correspondence a high lift is meant to capture is itself unreliable well above that point: among high-lift cells, those with counts in the low tens are frequently incoherent pairings that clear the ratio by small-sample chance, while by a count of a few hundred the high-lift pairs are almost uniformly genuine pipelines. We therefore set the lift-floor at $c_{en} \ge 100$, where the sampling relative standard error is near ten percent, comfortably below Statistics Canada's caution threshold, and the high-lift correspondence is coherent. Cells below the lift-floor are treated as unmeasured rather than as low-lift, since the ratio there is not a reliable estimate of anything, and workers whose education--destination cell falls below it carry no lift value. This is a reliability standard on the construction of the quantity, fixed in advance on the public table, blind to any interior transition. The unfiltered ratio, which is a mismeasurement rather than a weaker measurement, is not an alternative we entertain.

The lift-floor is fixed at $100$ rather than swept, and two facts on the public table establish that $100$ is neither too low nor worth raising. It is not too low: the sampling relative standard error at the lift-floor is near ten percent and the Dirichlet prior displaces the estimate by a comparable amount, both small against the spread of log lift, which runs across roughly thirteen units from the anti-associated tail to the strongest pipelines, so a cell at the lift-floor is neither noise-dominated nor prior-dominated relative to the variation the measure must resolve. It is not worth raising: moving the lift-floor from $100$ to $200$ removes about two percent of the activated sample, a slice already shown reliable by the same standard, so a higher lift-floor would discard usable observations without adding reliability the estimates lack. Because $100$ is defensible on reliability alone and the sample it admits is already reliable, the lift-floor is held fixed and robustness is assessed on the activation threshold $\tau$ (@sec-AD), the parameter of the rule rather than of the measurement.

Lift is read in natural logarithm.^[We use \emph{lift}, the term for this quantity in association-rule learning and market-basket analysis. The same ratio of a conditional probability to its marginal appears across fields under other names: a revealed-comparative-advantage index, a relative risk, and, in logarithm, the pointwise mutual information of education and occupation. Each expresses one idea, how much more (or less) than independence a given education feeds a given occupation.] The log is the natural scale for a multiplicative quantity whose neutral point is one: equal ratios become equal distances, the long right tail of rare-destination pipelines is compressed onto a comparable scale with ordinary moves, and its neutral point at $\log m = 0$ marks a pipeline no stronger than baseline, which is the point the activation rule of @sec-AD reads against. The Dirichlet prior guarantees the smoothed lift is strictly positive, so the log is always defined.

## CWTVD resolving power  {#sec-AC}

@sec-local adjudicates the horse race on the likelihood criterion and the
geometry-aware criterion, the latter scored under each informative judge,
and a win requires leading on all of them. The role of $\Delta CWTVD$ as a
co-equal geometry-aware criterion rests on a property established here by
simulation: among the two correlated informative geometries, $\Delta CWTVD$
resolves which governs more sharply than $\Delta KL$ does. This appendix
substantiates that property and states why it can be established only by
simulation.
 
The [online companion](https://rpmartin.github.io/not-one-thing-companion#sec-higher-resolution) plots improvement relative to
independence against $\varepsilon$ for each candidate metric, across three
mixed data-generating processes (columns) and the three scoring rows:
$\Delta CWTVD$ scored under $C_{\mathrm{hierarchy}}$, $\Delta CWTVD$ scored
under $C_{\mathrm{skill}}$, and $\Delta KL$. Read down a column, the two
$\Delta CWTVD$ rows agree on which geometry peaks highest in every mixture:
skill in the skill-heavy $75/25$ column, hierarchy in the hierarchy-heavy
$25/75$ column, whichever the true mixture favours at $50/50$. The winner
is the same under either scoring endpoint in each of these simulated
mixtures; we read this endpoint agreement as observed in the simulation,
not as a guarantee that the peak winner is constant across the interior
blends, which the fixed-$\varepsilon$ linearity would give only before the
peak over $\varepsilon$ is taken. This agreement is the
robustness the two-endpoint criterion is meant to certify, not a
degeneracy: the endpoints coincide on the winner while remaining far enough
apart in construction, $\mathrm{corr}(C_{\mathrm{skill}},
C_{\mathrm{hierarchy}}) = 0.3$, that coincidence is informative.

The contrast with $\Delta KL$ is the resolving-power property. Under either
$\Delta CWTVD$ scoring the gap between the skill and hierarchy peaks tracks
the imbalance of the true mixture, widening as the mixture departs from
balance. $\Delta KL$ holds the two curves close to coincident across all
three mixtures, resolving the imbalanced cases only weakly. A second
difference appears at the benchmark: under every mixture the $\Delta KL$
curve for $C_{\mathrm{binary}}$ retains a positive tail well past its peak,
the same geometry-blind credit for the aggregate stay/move split that
drives the out-of-regime KL behaviour in @sec-local-falsification; both
$\Delta CWTVD$ rows strip that credit out and the $C_{\mathrm{binary}}$
tail collapses to the independence line. Both features point the same way:
$\Delta CWTVD$ separates the two correlated geometries more sharply and
denies the benchmark credit it has not earned geometrically.

The simulations also exhibit the wrong-bet penalty that the out-of-regime
falsification relies on. In the $25/75$ column scored under
$C_{\mathrm{hierarchy}}$, skill is the minority geometry graded by its
rival's judge, and its peak falls below the $C_{\mathrm{binary}}$
benchmark; scored under $C_{\mathrm{skill}}$ it recovers above binary. A
metric making a geometric bet is penalised by distance for making the wrong
one, where binary, making no off-diagonal bet, cannot be. This is the
in-sample analogue of the below-independence crossing predicted in
@sec-local-falsification, with binary in place of independence as the
reference, and it corroborates the mechanism under known
data-generating processes rather than testing it.
 

This resolving-power property is established under known data-generating
processes rather than read off the estimation sample, for the structural
reason given in @sec-local: the property is a statement about the mapping
from a known true mixture to the measured gap between criteria, and a
single observed labour market delivers one gap drawn from one unknown
mixture, with no reference mixture against which its magnitude could be
judged. The Monte Carlo is therefore the complete justification the claim
admits, not preliminary evidence awaiting empirical corroboration. What it justifies is the mapping from mixture to gap, not the behaviour of the peak at finite $N$; that the $\mathrm{argmax}$ does not open a gap by sampling alone where the truth separates nothing is a separate question, settled at the realized count by the calibration of @sec-AE. The
figures here substantiate the property; @sec-local states what follows for
trusting $\Delta CWTVD$ when it parts from $\Delta KL$ on the Canadian
data.
 
## Threshold sensitivity for latent-activation reinstatement {#sec-AD}

The local/non-local partition routes on the recorded education change, a proxy for whether a large human-capital increment became operative within the window (@sec-rp). Latent activation is one of the proxy's two known failures, and the only one that commits to a threshold: reclassification on the reverse error turns on $\log m = 0$, a property of the measure, so it has no parameter whose sensitivity needs reporting. This appendix therefore treats latent activation alone, identifying those workers from what is observed and reinstating them into the non-local sample as a robustness check on the EOS--OES estimates. It is a correction to measurement error in the partition variable, not a headline result, and the reinstated estimate is reported only beside the uncorrected one.

The reason the exercise is worth running, rather than signing the selection by
argument, is that the sign is genuinely ambiguous a priori. The non-local sample
loses whichever latent activators failed to produce a recorded change, and which
workers those are depends on how the holding-out ends. A graduate who takes the
first offer activates promptly, their move coincides with the recorded
credential, and they stay non-local. A graduate who holds out activates late,
often only after the credential has aged into a prior cycle as recorded
attainment, so their move shows no change and drops to local. If the wait ends
in the specific match it was for, the droppers landed in the high-OES occupation
their high-EOS credential points to, so they are the very cases the gap predicts,
and their loss attenuates the association. If the wait instead ends in
capitulation, the worker settling for a generic occupation that does not draw on
the credential they acquired, the droppers are the non-matching cases, and their
loss selects the sample toward matches, inflating it. Both outcomes begin in the same
behaviour and the record distinguishes neither, since in both the credential is
old and the move is unaccompanied by a recorded change. No argument from the record resolves
this, but the reinstatement does, because the dropped workers, their acquired
educations, and their destinations are observed.

A worker with unchanged education $e$ is reinstated when the credential is
dormant at the origin and a genuine pipeline at the destination:
$$\log m(e, n_{2016}) < 0 \quad\text{and}\quad \log m(e, n_{2021}) \ge \tau ,$$
with the destination cell required to clear the lift-floor of @sec-AB.
The first condition fixes the origin as genuinely non-matching, below-baseline
lift, and is a gate held at baseline. The second declares the destination a
pipeline of strength at least $\tau$ on the log-lift scale. These two conditions
are the whole of the rule. The lift-floor is not a further condition a
worker must meet to count as activated; it is a precondition for the lift value
to be trustworthy at all, and it belongs to the lift construction of @sec-AB
rather than to the rule. The rule therefore has a single parameter, $\tau$, and
that is the axis the sweep below varies, with the lift-floor held fixed at its
reliability-justified value.

One direction of possible circularity is worth naming. The dormant-origin gate
conditions only on the 2016 occupation, which plays no part in the destination
OES the gap is built on, so it is clean. The pipeline condition uses the 2021
occupation, which is destination-side and so shares information with the
destination OES the gap reads. Membership is the contrast of the two, so workers
reaching the same 2021 occupation are still separated by their 2016 state; but
because the destination side is not wholly clean, the correction's credibility
rests on the sensitivity of the reinstated estimate to $\tau$ and to the lift-floor,
which is what the sweep reports.

We report the reinstated count alongside the rescued non-local gap estimate, beside
the uncorrected gap, across the threshold $\tau \in \{2, 3, 4\}$ on log lift, with
$\tau = 2$ the primary value. The lift-floor is held fixed at $100$ rather
than varied: @sec-AB shows that at this count neither sampling noise nor the
Dirichlet prior displaces the estimate materially against the spread of log lift,
so cells at the lift-floor are already reliable, and raising it would discard a
reliable slice of the sample without buying reliability the estimates lack. The
activation threshold, by contrast, is the parameter of the rule itself, and it
is the axis the robustness sweep varies. The sweep is reported in full; we do not
select a value after seeing the estimates. The threshold is fixed on the
definition of activation, a genuine dormant-to-pipeline switch, and is not tuned
toward moving the gap; moving it is neither the aim nor a success condition. The
maintained expectation is the opposite: latent activation is a leak we expect to
be small, so a small reinstated count together with a gap stable against the
uncorrected estimate is the affirmative outcome, evidence that the record-change
partition is not materially contaminated by this failure. A gap stable across
the sweep, and across tightening the dormant-origin gate below baseline, indicates
the correction does not turn on the destination association or on the threshold; a gap
that moves only at permissive $\tau$ locates the sensitivity; a large reinstated
count that moved the gap substantially would instead be evidence the partition
had been materially wrong, the one outcome that would count against the
uncorrected estimate. Because the count is reported rather than engineered, a small count is
read as latent activation being empirically rare, itself consistent with the
mechanism, since a worker already holding their highest credential is unlikely
to leave it dormant and then activate it within a five-year window.

Where the reinstated count is large enough to be informative, the reinstated
workers also resolve the sign left ambiguous above: whether latent activators are
genuine wormholes a selective search stranded in the local sample, or off-pipeline
moves a quality gradient leaves behind, is read directly from whether their
reinstatement sharpens or dilutes the non-local gap. Where the count is small,
that sign is left unresolved, but immaterially so, since a leak too small to move
the estimate does not need its direction fixed.

## Sparse-cell calibration and detection power {#sec-AE}

Both cells this appendix guards predict the same outcome, that distance ceases to order destinations and binary wins, but they reach it by independent routes and occupy different places in the argument, and it is worth stating the relation before the checks, since the two can look like one thing measured twice. The high-OES Binary subset reaches the hard-to-predict pole through the occupation: where an occupation's human capital is occupation-specific, the few who leave scatter without geometric order. The 40-and-over band reaches the same pole through the worker: late in a career the reservation wage has risen so far that the surviving moves carry no systematic geometry. Same predicted signature, but OES is a property of the occupation and age a property of the worker, so the two partitions cross-cut rather than coincide; a high-OES occupation holds workers of every age, and a 40-and-over worker sits in every OES regime. Running both is therefore corroboration and not repetition, the same independence that makes the OES-versus-TEER check informative one level down: two orderings built on different logics that predict the same failure of geometry confirm it once through the occupation and once through the career. The independence is only as complete as the partitions diverge, so at access we report the confusion between the two assignments, as we do the OES--TEER $\kappa$, and read the corroboration as independent to the extent they cross-cut rather than overlap. The shared outcome is also why the two cells invite opposite checks. Binary winning in the OES subset is a headline result the grid is built to deliver, threatened by being manufactured; binary winning in the age band is a deliberately degraded corroboration, threatened instead by being vacuous, a default left when the graded rivals were never detectable. A result you bank must survive a false positive; a corroboration you hope for must survive being uninformative. That is the structure the rest of the appendix makes precise.

Two sub-regime readings depend on whether a horse race run on few observations can be trusted: the high-OES Binary subset of @sec-subregime-predictions, where thin off-diagonal mass raises the worry that a binary win is manufactured rather than real, and the 40-and-over age band of the same section, where a low move rate raises the mirror worry that a graded winner present in the data would go undetected and leave binary winning by default. Both are settled by simulation at the affected subset's realized count, and both procedures are fixed here in advance; neither can be executed before access, since each requires margins and effect sizes observed only in the linked file.

The two readings guard against opposite errors and so require opposite generating truths. The first is a false-positive question: when no geometry governs, how often does noise alone fake a decisive gap? We generate under a truth that sits exactly between the two informative geometries, the balanced cost $C = \tfrac{1}{2}(C_{\text{skill}}/7.145) + \tfrac{1}{2}(C_{\text{hierarchy}}/3)$, so that neither graded metric should win by construction. We solve the transport plan under this cost against the subset's realized margins at the reference $\varepsilon$, draw one thousand synthetic flows from it as a single multinomial over all cells at the subset's total count $N$, and on each synthetic flow run the horse race exactly as on real data, scoring each metric against that sample's own independence baseline and taking each metric's peak over the $\varepsilon$ sweep. We record the fraction of samples in which the peak gap between the two graded metrics exceeds the ten percent threshold. If fewer than five percent do, the threshold is safe against noise at this count and the subset's verdict is read normally; if more do, the subset is noise-limited and its verdict is not read. This is the reading committed for the high-OES Binary subset: it establishes that a decisive gap there is unlikely to be an artifact of sparse cells. A failure here does not condemn the populous Skill and Hierarchy subsets, which carry more off-diagonal informative mass and so face an easier separation of the two graded geometries; the check speaks to the count and margins it is run at and to no others. But it withdraws the reassurance the populous subsets otherwise inherit by example, that the sparsest regime in the design is already clean, and with it the licence to rest their verdicts on the transition count alone. Should the Binary calibration fail, any subset whose result we read decisively is calibrated in turn at its own count and margins before it is read, since the argmax noise the failure reveals is a function of the off-diagonal informative mass a subset carries and not of its total $N$.

The second is a power question, and it is what the 40-and-over band requires: if a graded metric genuinely governed at a realistic strength, would a race on this few moves recover it? Here the generating truth is not the balanced mixture but each graded geometry in turn. We generate under $C_{\text{skill}}$ alone at the effect size skill exhibits in the band where it governs, thin to the old band's realized move count, run the identical race, and record how often skill is recovered as the winner; we then repeat under $C_{\text{hierarchy}}$ alone at the strength hierarchy shows where it governs. The band is adequately powered only if both graded signals would have been recovered at its move count. If both would have and the data nonetheless favour binary, the binary result is substantive: the rivals had their chance and did not take it. If either graded signal would have been missed, a binary win cannot exclude a hidden signal of that kind, and we report the band as underpowered against that rival rather than as evidence that binary governs. The effect sizes anchoring these runs are observed only at access, in the bands where each graded metric wins, which is why the power check is a committed procedure rather than a result stated here.

The two halves share all machinery and differ only in the generating truth and the quantity recorded: the balanced mixture with a false-positive rate for calibration, each pure graded geometry with a recovery rate for power. A subset is read only if it passes the check its inference depends on, calibration for the high-OES Binary subset and power for the oldest age band.
 
## References
