---
title: "Occupational Mobility Is Not One Thing: Evidence from the Canadian Labour Market" 
author: "Richard Martin"
format:
  pdf:
    documentclass: article
    number-sections: true
    geometry: margin=1in
bibliography: references.bib
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \DeclareMathOperator*{\argmin}{argmin}
editor: 
  markdown: 
    wrap: 72
---

```{r, include=FALSE}
library(tidyverse)
library(here)
library(knitr)
library(kableExtra)
library(patchwork)
library(conflicted)
cdfs <- read_rds("out/dist_cdfs.rds")
calibration <- read_rds(here("out", "calibration.rds"))
```

## Pre-registration Notice

This paper is posted as a pre-registration prior to data access. Anything that looks like a result should be interpreted as a prior. This pre-registration of design is intended to defend against potential claims of post-hoc rationalization. The final paper will be posted upon completion of analysis.

## Abstract

Workers predominantly move between proximate occupations, but proximity means different things in different parts of the labour market. Using entropic regularized optimal transport to conduct a horse race among three distance metrics — skill, hierarchy, and binary — we show that no single geometry governs occupational mobility. Instead, mobility is organized into three sub-regimes defined by education specificity: skill distance governs low-specificity occupations where skills are broadly transferable, hierarchical distance governs mid-specificity occupations where credential ladders structure mobility, and binary switching costs govern high-specificity occupations where human capital is so specific it is virtually useless elsewhere. Each regime implies a distinct policy lever for addressing labour shortages.

## Introduction

Occupational mobility is not one thing. Workers tend to move between proximate occupations, but proximity is not self-evident: occupations differ along multiple dimensions simultaneously — the skills they require, their position in a credential hierarchy, and the degree to which they selectively admit workers with particular credentials. Each implies a different geometry of the labour market, and different predictions about who moves where. We test which geometry governs mobility — and find that the answer depends on where in the labour market you look.

We separate local mobility — transitions that occur without new educational attainment — from non-local mobility — transitions that occur when a new highest educational attainment is achieved. For local mobility, distance is the explanatory variable: we ask which of three candidate metrics (skill, hierarchy, and binary) best rationalizes observed occupational transitions. For non-local mobility, the role of distance is reversed: rather than asking which metric explains where workers go, we ask what determines how far they travel, treating transition distance as the outcome and measures of education and occupation specificity as its determinants.

We anticipate that no single metric will dominate globally. A labour market in which 
workers in highly credentialed professional occupations, managerial roles, and general 
labour all follow the same distance geometry would be surprising. To that end, we 
construct a measure of education specificity --- using KL divergence to capture how 
different an occupation's education distribution is from that of the labour market as 
a whole --- and split 504\footnote{The NOC 2021 contains 516 unit groups; 12 are excluded due to 
data issues and are discussed in Section~\ref{sec-data}.} occupations defined by Canada's National Occupational Classification (NOC) into three sub-regimes. The Horizontal sub-regime (low 
specificity) is hypothesized to be governed by skill distance: skills are generic and 
mobility is not constrained by hierarchical structure. The Vertical sub-regime (middle 
specificity) is hypothesized to be governed by hierarchical distance: skills are 
specific enough that credential ladder structure binds, with mobility flowing within 
labour market silos defined by the NOC broad group. The Minimal sub-regime (high 
specificity) is hypothesized to be governed by binary switching costs: human capital 
is so occupation-specific that the continuous distance structure of the other metrics 
is counterproductive, and mobility is better described as either staying or leaving.

The policy relevance is direct. Labour shortages in different segments of the market require different adjustment mechanisms, and the right mechanism depends on which distance metric governs mobility in that sub-regime. Shortages in Horizontal occupations may be addressed through reskilling that builds on transferable skills. Shortages in Vertical occupations are best addressed by increasing supply at the bottom rung of the relevant ladder, allowing upward movement through the credential hierarchy to relieve pressure at higher levels. Shortages in Minimal occupations, where the credential gate is narrow and mobility is structurally low, require supply-side interventions at the point of credential formation. Getting the geometry right is therefore not merely a methodological contribution — it determines which policy lever to pull.

For local mobility, we use entropic regularized optimal transport (EROT) to conduct the horse race. Gravity models are ill-suited to this purpose: origin and destination fixed effects absorb misspecification of the cost matrix so effectively that little structure is left to discriminate. EROT resolves this by holding the marginal structure fixed at observed values, separating the question of how much mobility occurs — determined by the margins — from the question of where mobility flows — shaped by the cost matrix. 

We evaluate fit using two complementary measures: relative KL divergence, which captures whether the predicted transition matrix assigns appropriate probability mass to observed transitions; and relative Cost-Weighted Total Variation Distance (CWTVD), which complements this by weighting misallocated mass by a neutral cost matrix so that errors involving geometrically distant pairs are penalized more heavily than errors involving nearby ones. A model that scores well on both has correctly allocated mass and allocated it to the right part of occupational space.

[FINDINGS PLACEHOLDER]

## Conceptual Framework

The central structural choice in this paper is the partition of occupational transitions into two qualitatively distinct regimes. A local transition occurs when a worker changes occupation without acquiring new educational attainment — the worker's highest credential in 2021 is the same as in 2016. A non-local transition occurs when a worker's highest educational attainment increases between the two census years. This partition reflects a fundamental difference in the mechanism governing mobility: for local transitions, human capital is fixed and distance governs destination choice; for non-local transitions, credential acquisition shifts the worker's position in occupational space, rendering distance secondary.

Two symmetric specificity measures play central roles in the analysis, one defined over occupations and one over credentials. Both are constructed as KL divergences of a conditional distribution from the relevant marginal, residualized on log cell size to remove a mechanical floor at low counts, and derived from the 2021 Census CIP/NOC table — a one-quarter sample of the Canadian population, of which at most one-sixteenth overlaps with the linked microdata used for estimation.

Occupation's education specificity is the KL divergence of $P(\text{education}|\text{NOC})$ from $P(\text{education})$, where education is defined as the cross product of CIP field of study and highest attainment level. A higher value indicates that the occupation draws from a more distinctive subset of the credential space than a randomly drawn worker would. Because the empirical education distribution can only place mass on a limited number of credentials at low occupation size, raw KL divergence has a mechanical floor; we therefore residualize $\log(\text{KL})$ on $\log(\text{occupation size})$ and define specificity as the residual. In the local partition this measure enters as a classifier defining the three sub-regimes; in the non-local partition it enters without modification as a continuous regressor for credential-driven distance traveled.

Education's occupation specificity is the KL divergence of $P(\text{NOC}|\text{education})$ from $P(\text{NOC})$, with the same residualization on log cell size applied. A higher value indicates that holders of a given credential are more concentrated in a distinctive subset of occupational space than a randomly drawn worker would be. This measure enters only the non-local partition, as a continuous regressor capturing the degree to which a credential points holders toward particular destination occupations.

Occupations are assigned to sub-regimes based on education specificity tertile, subject to two TEER override rules. The Horizontal sub-regime includes occupations with bottom tertile education specificity, expanded to include all TEER 0 (senior management) and TEER 5 (labour) occupations regardless of specificity tertile, reflecting that these occupations are structurally unconstrained by credential hierarchies. The Vertical sub-regime includes occupations with middle tertile education specificity, expanded to include TEER 3 and 4 occupations from the top tertile, reflecting that mid-TEER occupations retain upward mobility within their silo. The Minimal sub-regime includes the remaining top tertile occupations — those with both high education specificity and sufficiently high TEER that the credential gate is binding.

The local/non-local partition has a natural falsification structure. If the partition is meaningful, distance metrics should improve on the independence baseline within the local regime but not when applied out of regime to non-local transitions — where credential acquisition, not distance, is doing the explanatory work. Conversely, the specificity interaction should predict distance within the non-local regime but fail to do so out of regime, where human capital is fixed and distance governs. Each model is tested both where it should work and where it should not, and the partition's validity rests on finding the predicted double dissociation.

Occupational licensing represents an explicit form of credential selectivity (Kleiner and Krueger 2013), but selectivity also operates implicitly through the education mix that an occupation draws from in equilibrium. Occupation's education specificity captures this implicit structural feature directly from the data, without relying on curated lists of regulated occupations that are incomplete, jurisdiction-specific, and require manual updating.

The natural alternative to EROT for evaluating occupational mobility costs is the gravity model [@anderson_vanwincoop_2003; @silva2006log], which estimates origin and destination fixed effects jointly with a cost parameter. This joint estimation is precisely the problem in our context: fixed effects absorb variation in mobility rates that may reflect true cost structure, allowing a misspecified cost matrix to achieve good fit by compensating through the fixed effects. Using a gravity model would be akin to amputating a leg of each horse prior to the race. EROT resolves this by holding the marginal structure fixed at observed values, separating the question of how much mobility occurs from the question of where mobility flows. 

EROT shares the interior structure of a multinomial logit over destination occupations: 
conditional on origin, the probability of transitioning to destination $j$ is proportional 
to $e^{-C_{ij}/\varepsilon}$, as in a logit model where utility is defined as negative 
distance plus idiosyncratic taste shocks scaled by $\varepsilon$. The analogy is 
instructive but inexact: EROT enforces hard constraints on both margins, so the scaling 
vectors $u_i$ and $v_j$ are equilibrium objects rather than free parameters; and because 
the margins cannot adjust to absorb cost misspecification, a poorly chosen cost matrix 
has nowhere to hide. Reporting results across a grid of $\varepsilon$ values characterizes 
how sensitive the horse race outcome is to the assumed degree of idiosyncratic 
heterogeneity in mobility decisions. EROT is described in detail in 
Section \ref{sec-local}.

## Cost Metrics

The skill distance matrix $C_{\text{skill}}$ is constructed from 161 O\*NET
measures spanning skills, abilities, knowledge, and work activities,
mapped to NOC 2021 occupations via a purpose-built crosswalk. Each NOC
is represented as a weighted average of its constituent O\*NET
occupations, with weights determined by the crosswalk mapping.

O\*NET provides two scales for each measure: Importance (how important
the attribute is to the occupation) and Level (the degree to which the
attribute is required). Both scales are standardized to a common 0--10
range and combined via geometric mean, yielding a single score that
reflects both the relevance and the intensity of each attribute. The
geometric mean is preferred over the arithmetic mean as it penalizes
cases where one scale is near zero — an attribute that is important but
required at negligible level, or required at high level but deemed
unimportant, receives a lower composite score than one scoring
moderately on both dimensions. The two scales are highly correlated in
practice, so this choice is unlikely to be consequential for the
resulting distance matrix, but we note it for transparency.

Raw O\*NET measures are standardized prior to dimensionality reduction.
Principal components analysis is applied to the 161-dimensional skill
profiles, and the number of components retained is determined
empirically: we identify the minimum k such that the Spearman rank
correlation between pairwise distances in the k-dimensional PCA space
and pairwise distances in the full 161-dimensional space exceeds 0.99.
This criterion — rank preservation of pairwise distances rather than
variance explained — is chosen because the object of interest is the
distance matrix itself, not the underlying skill dimensions. Twelve
principal components satisfy this criterion.

The skill distance between any two occupations is then the Euclidean
distance between their coordinates in the retained 12-dimensional PCA
space. The resulting skill distance matrix is visualized as an
interactive heatmap in the online companion, allowing inspection of
pairwise distances across the full set of NOC 2021 occupations.

A note on the crosswalk is warranted. An alternative approach would be
to use OaSIS, the official O\*NET to NOC mapping. We construct our own
crosswalk for a specific reason: falsifiability. The horse race requires
that measurement error in the skill distance matrix be attributable to a
known source — namely, imprecision in the O\*NET to NOC mapping. OaSIS
obscures the structure of that mapping, making it impossible to
characterize where and how measurement error enters $C_{\text{skill}}$.
Our crosswalk makes this structure transparent and therefore makes the
identification strategy falsifiable. The crosswalk is publicly available
at [ONET NOC 2021 crosswalk](https://github.com/bcgov/onet-noc2021-crosswalk).

The hierarchical distance matrix $C_{\text{hier}}$ is derived directly
from the NOC 2021 taxonomy, treating the occupational classification
structure as a tree metric. Distance between any two occupations is
defined as follows: zero if all five digits match (same unit group); one
if the first four digits match (same minor group, different unit group);
two if the first three digits match (same sub-major group, different
minor group); three plus the absolute difference in TEER level if only
the first digit matches (same broad occupational category, different
sub-major group); and nine for any cross-category transition. The
absolute TEER difference ranges from zero to five, so within-category
distances range from three to eight, preserving a strict separation from
the cross-category sentinel value of nine.

Intuitively, the hierarchical view of the labour market is that it is comprised of ten largely separate silos, each containing up to six platforms corresponding to TEER levels. Each platform has internal geography --- occupations within the same silo and platform are at distances of one, two, or three depending on how closely related they are within the NOC hierarchy. Moving between platforms always costs more than any within-platform move: inter-platform distances start at four and increase with the number of platforms traversed, up to eight. Moving between silos always costs more than any within-silo move: cross-silo transitions are assigned a sentinel value of nine, the minimum integer exceeding the maximum within-silo distance of eight. The hierarchical distance matrix is visualized as an interactive heatmap 
in the online appendix, allowing inspection of pairwise distances across 
all NOC 2021 occupations.

The binary switching cost matrix $C_{\text{binary}}$ assigns a cost of
zero to remaining in the same occupation and one to any transition,
presuming that any movement is extremely costly relative to staying. It
is a limiting case, even more extreme than the hierarchical distance ---
topologically, the labour market is comprised of 504 largely separate 
silos, with no internal structure of any kind. Any movement is equally 
costly regardless of destination.

$C_{\text{binary}}$ serves two roles in the analysis. First, it is the
theoretical straw man in the horse race: if occupational mobility costs
are purely a function of whether a switch occurs rather than how far,
then $C_{\text{binary}}$ should rationalize observed mobility patterns as
well as or better than $C_{\text{skill}}$ or $C_{\text{hier}}$. Second,
and more subtly, $C_{\text{binary}}$ is best suited to settings where
mobility is limited — where most workers remain in their origin
occupation. In such settings, the information contained in a richer cost
structure can cause model fit to worsen. Finding that $C_{\text{binary}}$
outperforms richer cost matrices in a particular sub-regime would
therefore be informative in its own right, suggesting that mobility in
that segment is too limited for cost structure to be identified from the
data.

## Normalization

The three cost matrices are defined on different scales by construction:
skill distances are continuous and spread over a wide range, hierarchical
distances are discrete with a sentinel value of nine reserved for
cross-category transitions, and binary distances take only the values
zero and one. Direct comparison in the horse race requires that the
matrices be anchored to a common scale; without normalization, EROT would
mechanically favour the matrix with the lowest typical pairwise
distances, regardless of descriptive accuracy.

We normalize skill and hierarchical distances to a common scale using a
quantile-matching approach, with binary serving as the reference. The
normalization proceeds as follows. Cross-category sentinel values of nine
are excluded from the hierarchical distribution, as they represent a
qualitatively distinct mobility regime rather than a point on the
continuous within-category distance scale. The conditional median of the
remaining hierarchical distances is four. We then locate four in the
unconditional hierarchical distribution — including the sentinel values —
which falls at the eighth percentile, reflecting the dominance of
cross-category pairs in the full pairwise distribution. Finally, we
locate the eighth percentile of the skill distance distribution, which is
8.59. Hierarchical distances are divided by four and skill distances by
8.59, anchoring both matrices at the same quantile of their respective
unconditional distributions.

```{r}
#| warning: false
#| echo: false
#| label: fig-cdfs
#| fig-cap: "Raw vs. Normalized CDFs"
cdfs
```

@fig-cdfs shows the empirical CDFs of all three matrices before
and after normalization. The raw distributions differ substantially in
scale and shape; after normalization, skill and hierarchical CDFs overlap
closely over the range of distances relevant to observed mobility,
diverging only in the right tail. The sensitivity of horse race results
to the normalization anchor is examined in @sec-AC

## Data {#sec-data}

The analysis uses the Census 2016--2021 longitudinal file, constructed by linking 
long-form respondents across the two census cycles via the Derived Record Depository
(DRD), Statistics Canada's national longitudinal database  of personal identifiers. 
Linkage is probabilistic; details of the matching keys and overall linkage rate will 
be verified upon data access and reported in the final paper.

The incomplete linkage introduces a potential selection concern distinct
from the observability selection discussed below. If unlinked individuals
are disproportionately concentrated in a part of occupational space where
one cost matrix has a strong prior advantage or disadvantage relative to
the others, the horse race outcome would reflect sample composition
rather than true metric performance. Unlinked individuals are likely to
be disproportionately recent immigrants, young workers, and high-movers
— individuals whose identifying information changed between census
cycles, reducing match probability. There is no obvious mechanism by
which these characteristics would differentially concentrate unlinked
individuals in occupational space in a way that systematically favours
or penalizes any particular cost matrix. Linkage-based selection is
therefore unlikely to threaten the internal validity of the horse race,
though this assumption cannot be directly verified without access to the
unlinked population.

The estimation sample is partitioned into subsets that vary in two
dimensions: whether the local/non-local regime assignment is correct,
and whether occupation measurement is clean. This design generates a
natural ordering of expected model fit that serves as an internal
validation of the partition's theoretical content.

The local clean sample is the primary identification sample for the
horse race, and the falsification sample for the specificity regression.
It comprises individuals with an observed occupation in both 2016 and
2021, no increase in highest educational attainment between the two
census years, no self-employment in 2016, no TEER--attainment mismatch
in 2016, and an O\*NET-to-NOC mapping quality score above the bottom
decile. This is the sample where the horse race should work best and
where the specificity regression should find nothing.

The non-local clean sample comprises individuals who acquire a new
highest educational attainment between 2016 and 2021 and report an
observed occupation in both periods, subject to the same measurement
restrictions: no self-employment in 2016, no TEER--attainment mismatch,
and an O\*NET-to-NOC mapping quality score above the bottom decile. It
is the primary estimation sample for the specificity regression, and the
falsification sample for the horse race. This is the sample where the
specificity regression should work best and where the horse race should
find nothing.

The cross-regime falsification thus runs symmetrically in both
directions. The non-local clean sample — where credential acquisition
governs destination choice — serves as a falsification for the horse
race: EROT should find no improvement over the independence baseline
here, since distance structure is irrelevant once human capital has
shifted. The local clean sample — where human capital is fixed — serves
as the mirror falsification for the specificity regression: credential
specificity should have no predictive power over distance traveled when
no new credential has been acquired. Each model is tested both where it
should work and where it should not, and the partition's validity rests
on finding the predicted double dissociation.

The mismeasurement diagnostic samples introduce measurement problems one
at a time, holding regime assignment correct. The occupational
mismeasurement sample retains individuals whose 2016 occupation is
likely mismeasured, identified by two criteria: self-employment in 2016,
where occupation codes are less reliably assigned; and TEER--attainment
mismatch, where the reported occupation is inconsistent with highest
educational attainment. Attainment mismatch could easily occur for either recent 
attainments or for those awaiting credential recognition.\footnote{For details regarding TEER--attainment mismatch criteria, 
see Appendix B.} Both criteria introduce noise into the origin 
occupation, and their effects on model fit are expected to be symmetric across 
cost matrices.

The mapping quality sample retains individuals whose 2016 occupation
falls in the bottom decile of the O\*NET-to-NOC mapping quality
composite; poor mapping introduces noise specifically into
$C_{\text{skill}}$ while leaving $C_{\text{hier}}$ unaffected, providing
a difference-in-differences diagnostic on metric-specific fit
degradation. Each of these samples is expected to show worse fit than
the corresponding clean sample, with the mapping quality sample showing
selective degradation of $C_{\text{skill}}$ relative to $C_{\text{hier}}$.

The mismeasurement residual samples comprise all observations not
assigned to any of the named samples above, split by regime. The local
mismeasured residual contains individuals in the local regime with
multiple simultaneous measurement problems — self-employment,
TEER--attainment mismatch, and/or poor O\*NET-to-NOC mapping in
combination. The non-local mismeasured residual contains individuals in
the non-local regime with multiple simultaneous measurement problems.
Within each regime the source of fit degradation is measurement noise
rather than regime misassignment, preserving the interpretability of the
regime axis while acknowledging that the specific contribution of each
noise source cannot be separately identified. Expected fit is worse than
the corresponding single-problem diagnostic samples but, crucially,
better than the out-of-regime clean samples — where the only problem is
regime misassignment. If that ordering holds, it constitutes direct
evidence that regime misassignment is more damaging to model fit than
the accumulation of measurement problems, strengthening the theoretical
case for the local/non-local partition.

The local partition is restricted to individuals with an observed
occupation in both 2016 and 2021, as both endpoints are required to
compute a transition distance. Non-observability has multiple sources —
including labour market entry and exit due to age — but these are not
expected to differentially affect the horse race. The relevant selection
concern is weak labour force attachment, which can produce a missing
occupation in either period and is positively related to skill level
[@statcan_lfs], implying that the excluded population is
disproportionately drawn from low-skill occupations characterized by
casual, seasonal, or intermittent employment.

This selection has differential implications for the three cost matrices
under evaluation. Skill distance is hypothesized to best rationalize
mobility in occupations where the education specificity of the origin
occupation is low — a category that encompasses many low-skill
occupations as well as management occupations. Low-skill workers in
these occupations are disproportionately excluded from the local
partition due to weaker labour force attachment. Sample selection bias
on the basis of occupation observability therefore disproportionately
disadvantages skill distance in the horse race: many of the observations
where this metric would be expected to excel are missing from the
estimation sample.

The non-local partition is restricted to individuals who acquire a new
credential between 2016 and 2021 and report an occupation in both
periods. The selection concern here is distinct from the local
partition: the excluded population is not disproportionately low-skill,
but rather disproportionately those whose human capital increased the
most. The most demanding credential programs — medicine, law, dentistry
— require sufficiently intensive study that concurrent employment is
largely precluded, meaning students in these programs are less likely to
report a 2016 occupation and thus excluded from the sample.

This selection mechanism has a specific implication for the non-local
partition's central empirical question: the relationship between CIP
program specificity and credential-driven distance traveled. The most
demanding programs tend to also be the most specific, and the human
capital increment they confer is large — generating the longest
credential-driven occupational jumps. Excluding these observations
therefore truncates the upper tail of the distance distribution
precisely where CIP specificity is highest. Estimates of the
relationship between program specificity and distance traveled should
accordingly be interpreted as lower bounds on the true relationship in
the full population of credential acquisitions.

## Local Model {#sec-local}

The entropy-regularized optimal transport problem minimizes the
Lagrangian, where the first term is the transport cost, the second is
the entropic regularization penalizing deterministic transport plans,
and the third and fourth enforce the marginal constraints $\sum_j
P_{ij} = a_i$ and $\sum_i P_{ij} = b_j$ via dual variables $f_i$ and
$g_j$.

$$\mathcal{L} = \sum_{ij} P_{ij} C_{ij} + \varepsilon \sum_{ij} P_{ij}(\log P_{ij} - 1) + \sum_i f_i(a_i - \sum_j P_{ij}) + \sum_j g_j(b_j - \sum_i P_{ij})$$

where $a_i$ and $b_j$ are the exact occupation stocks in 2016 and 2021
respectively. The framework is implicitly first-order Markov: the
margins absorb any factor that operates additively at the level of the
origin or destination occupation, but not factors that depend on the
specific origin-destination pair. The cost matrix is itself dyadic ---
it encodes the hypothesis that the specific pairing matters --- and the
horse race is asking which specification of dyadic structure best
captures the non-Markovian residual after the margins have absorbed the
first-order effects.

The first-order conditions yield:

$$\log P_{ij} = \frac{f_i}{\varepsilon} + \frac{g_j}{\varepsilon} - \frac{C_{ij}}{\varepsilon}$$

Exponentiating, the optimal transport plan takes the form $P^\star_{ij}
= u_i \, e^{-C_{ij}/\varepsilon} \, v_j$, where $u_i =
e^{f_i/\varepsilon}$ and $v_j = e^{g_j/\varepsilon}$ are scaling
vectors. Strict convexity of the regularized objective guarantees a
unique solution. The solution is therefore a row- and column-rescaling
of the Gibbs kernel $K_{ij} = e^{-C_{ij}/\varepsilon}$, and the
Sinkhorn algorithm recovers it by alternating row and column
normalization until both marginal constraints are satisfied to within a
specified tolerance. All computations use a log-domain Sinkhorn solver
for numerical stability, implemented in the \texttt{safesink} R package
[@safesink].

The regularization parameter $\varepsilon$ has a natural interpretation
in the random utility framework. The optimal transport plan $P^\star_{ij}
= u_i \, e^{-C_{ij}/\varepsilon} \, v_j$ is the static equilibrium of a
discrete choice model in which workers draw idiosyncratic utility shocks
scaled by $\varepsilon$ over destination occupations. Low values of
$\varepsilon$ concentrate the transport plan on cost-minimizing
assignments, approaching a deterministic reallocation in the limit; high
values diffuse mobility across destinations, approaching the independence
baseline $P_{ij} = a_i b_j$ as $\varepsilon \to \infty$. The RUM
analogy governs the interior structure of the transport plan --- how
probability mass is allocated across origin-destination pairs --- but
does not extend to the margins. In a standard logit model the choice
probabilities are outcomes of utility maximization with no constraint on
their aggregates; in EROT the origin and destination occupation
distributions are hard constraints that the transport plan must exactly
reproduce. This distinction is methodologically consequential: because
the margins cannot adjust to absorb cost misspecification, a poorly
chosen cost matrix cannot hide behind fixed effects or scale
recalibration --- its misfit appears directly in the objective.

A critical identification constraint follows directly from the form of
the optimal transport plan: only the ratio $C_{ij}/\varepsilon$ enters
$P^\star$. The scale of the cost matrix and the value of $\varepsilon$
are therefore not separately identified. This means that results
reported at a single value of $\varepsilon$ are uninterpretable without
an implicit normalization assumption --- and that assumption can be
gamed, since any cost matrix can be made to fit well by tuning
$\varepsilon$ to its scale. A horse race conducted at a single
temperature is not a horse race; it is a calibration exercise.

Evaluating across a grid of $\varepsilon$ values sidesteps this by looking for dominance that is robust across scales — a metric that wins at multiple temperatures cannot be attributed to a fortuitous scale alignment. We evaluate across a log-spaced grid of
$\varepsilon$ values ranging from 0.5 to 2, spanning from relatively
concentrated to relatively diffuse transport plans while remaining in
the region where the fit metrics retain discriminating power. The
criterion for a robust winner requires some care. At low values of
$\varepsilon$, cost-sensitive matrices concentrate probability mass too
aggressively relative to observed mobility, incurring an overconfidence
penalty that mechanically favours the flat binary matrix regardless of
which metric truly best rationalizes mobility. This low-temperature
binary lead is therefore uninformative about genuine descriptive
accuracy and should not be interpreted as evidence that binary is the
superior metric. The operative criterion is dominance over the full
epsilon grid after discounting this artefact: a cost matrix is declared
the winner if it dominates all competitors across the sweep, or
dominates all non-binary competitors across the sweep while losing to
binary only at low temperatures where the overconfidence penalty is
operative. Conversely, when binary genuinely best rationalizes mobility
--- as in sub-regimes where occupational transitions are not structured
by continuous distance --- it will dominate not just at low temperatures
but across the entire grid and by a margin that dwarfs the overconfidence
effect. The epsilon path thus serves a dual purpose: identifying the
winning metric and distinguishing genuine binary dominance from penalty
artefact.

The horse race is evaluated using two complementary fit metrics, both
normalized relative to the independence baseline $P^{\text{ind}}_{ij} =
a_i b_j$, so that 0 indicates no improvement over independence, 1
indicates perfect fit, and negative values indicate a transport plan that
fits observed mobility worse than independence.

The first metric is relative KL divergence improvement:

$$\Delta KL = \frac{KL(P^{\text{obs}} \parallel P^{\text{ind}}) - KL(P^{\text{obs}} \parallel \hat{P})}{KL(P^{\text{obs}} \parallel P^{\text{ind}})}$$

where $KL(P \parallel Q) = \sum_{ij} P_{ij} \log(P_{ij} / Q_{ij})$.
This measures the share of the divergence between observed transitions
and independence that is explained by the fitted transport plan. KL
divergence is sensitive to probability mass wherever it is misallocated,
but is geometry-blind: misallocating mass to a nearby occupation is
penalized identically to misallocating it to a distant one.

The second metric is relative Cost-Weighted Total Variation Distance
(CWTVD) improvement [@villani2009optimal; @rubner2000earth]:

$$\Delta CWTVD = \frac{CWTVD(P^{\text{obs}}, P^{\text{ind}}) - CWTVD(P^{\text{obs}}, \hat{P})}{CWTVD(P^{\text{obs}}, P^{\text{ind}})}$$

where $CWTVD(P^{\text{obs}}, \hat{P}) = \sum_{ij}
\frac{|P^{\text{obs}}_{ij} - \hat{P}_{ij}|}{2} \cdot C_{ij}$. CWTVD
extends classical total variation distance by weighting absolute
residuals elementwise by a cost matrix, penalizing probability mass
misallocated to geometrically distant origin-destination pairs more
heavily than mass misallocated to nearby ones. Unlike KL divergence,
CWTVD respects the metric structure of the outcome space; unlike the
Wasserstein distance, it does not solve an additional optimal transport
problem over the residual mass, making it computationally trivial given
an existing transport plan.

The scoring cost matrix $C$ used in CWTVD is the arithmetic mean of the
three candidate normalized cost matrices: $C_{\text{skill}}$,
$C_{\text{hier}}$, and $C_{\text{binary}}$. Using a neutral common metric
ensures that no candidate model is advantaged by the choice of scoring
matrix. The two metrics are complementary by design: KL divergence
captures overall probability mass misallocation regardless of geometry,
while CWTVD captures whether misallocation is concentrated in nearby or
distant pairs. A transport plan that scores well on both metrics fits
observed mobility both accurately and geometrically.

## Non-Local Model

The non-local model estimates the (conditional on positive) relationship
between credential and occupational specificity and distance traveled
following a credential acquisition. The estimation sample is restricted
to those with a positive distance between their 2016 and 2021
occupations.  The outcome is $\log(d_{ij})$, the log distance between the 2016 and
2021 occupations, computed separately using skill distance and
hierarchical distance. Running the regression on both metrics provides a
check on robustness to the choice of distance measure; the wormhole
mechanism (credential-driven jumps that allow workers to bypass the normal distance constraints of occupational space) should operate independently of which metric is used to
measure the jump.

The estimating equation is:

$$\log(d_{ij}) = \alpha + \beta \, \text{EOS}_c \times \text{OES}_j + \delta_i + \varepsilon_{ij}$$

where $\text{EOS}_c$ is education's occupation specificity for
credential $c$ acquired between 2016 and 2021 --- the KL divergence of
the conditional occupation distribution $P(\text{NOC} | \text{education})$
from the marginal $P(\text{NOC})$, residualized on log credential cell
size; $\text{OES}_j$ is occupation's education specificity for the 2021
destination occupation $j$ --- the KL divergence of the conditional
education distribution $P(\text{education} | \text{NOC})$ from the
marginal $P(\text{education})$, residualized on log occupation size; and
$\delta_i$ are origin occupation fixed effects.

The interaction $\text{EOS}_c \times \text{OES}_j$ enters without main
effects. This is not an oversight but a theoretical commitment: the
wormhole mechanism requires both an entrance and an exit simultaneously.
A specific credential that points holders toward a particular destination
($\text{EOS}_c$ high) enables a long jump only if the destination
selectively admits workers with that credential ($\text{OES}_j$ high);
and a selective destination ($\text{OES}_j$ high) generates a long
credential-driven jump only if the credential points holders there
($\text{EOS}_c$ high). Neither condition alone is theoretically predicted
to increase distance traveled, and including main effects would impose an
additive structure inconsistent with the wormhole mechanism. The product
$\text{EOS}_c \times \text{OES}_j$ is the natural sufficient statistic
for the joint condition.

The two specificity measures are conceptually distinct and can diverge in
economically meaningful ways. The cleaner case is a credential with high
$\text{EOS}_c$ and low $\text{OES}_j$: a nursing degree points almost
all holders toward nursing occupations, but nursing draws from multiple
credential pathways --- registered nurses, licensed practical nurses,
nurse practitioners --- so the destination occupation is not highly
selective with respect to any single credential. The credential has a
clear destination, but the destination has many entrances. The reverse
case --- low $\text{EOS}_c$, high $\text{OES}_j$ --- is less intuitively
obvious but equally important. A general engineering degree prepares
graduates for a wide range of occupations, so $\text{EOS}_c$ is low. Yet
if the destination is a highly specialized occupation such as petroleum
or nuclear engineering, that occupation draws almost exclusively from
particular technical disciplines, so $\text{OES}_j$ is high. In this
case the worker reaches a selective destination not because their
credential pointed them there, but through some other mechanism --- prior
experience, specialization within a general program, or networks. The
wormhole was not the route, and the product $\text{EOS}_c \times
\text{OES}_j$ correctly predicts no long credential-driven jump. The
interaction-only specification is therefore not merely a response to
collinearity concerns but a direct operationalization of the theoretical
mechanism.

Origin occupation fixed effects $\delta_i$ are included to close a
backdoor path from prior education choices to the specificity measures.
The 2016 occupation is downstream of the worker's educational trajectory
prior to the census period, and that trajectory influences both which
credential is subsequently acquired --- and hence $\text{EOS}_c$ --- and
the distance traveled. Conditioning on origin occupation absorbs this
unobserved selection, allowing the specificity measures to identify the
wormhole mechanism rather than proxying for occupational trajectory.

The non-local model is evaluated primarily by the sign, magnitude, and
significance of $\beta$, the coefficient on the interaction of
education's occupation specificity and occupation's education
specificity. The main hypotheses are: $\beta > 0$ in the non-local clean
sample, where the wormhole mechanism is operative; and $\beta \approx 0$
in the local clean falsification sample, where human capital is fixed
and credential specificity has no purchase on destination choice. The
$R^2$ of the regression serves as a summary fit measure and is tracked
across the full sample partition --- non-local clean, mismeasurement
diagnostics, mismeasured residual, and local clean falsification --- to
document the degradation in fit as samples depart from the conditions
under which the model is hypothesized to work. Results are reported for
both skill and hierarchical distance outcomes.

## Results  \[PLACEHOLDER --- awaiting data\]  {#sec-res}

\[KL and CWTVD improvement figures by cost metric, temperature and
subsample\]

\[KL and CWTVD improvement by sub regime, cost metric, temperature and
subsample\]

\[p-values and R squared from regressing log(distance>0) on specificity
interaction, and origin fixed effects.\]

## Limitations

For local mobility, a notable omitted factor is *path-specific* wage
differentials. The additively separable component of wage gains ---
recoverable from origin and destination wage levels alone --- is absorbed
by the margins insofar as high-wage destinations attract larger inflows,
and low-wage origins larger outflows. The residual dyadic component,
reflecting transition-specific wage premia and discounts beyond what
origin and destination wage levels predict, is an omitted factor.
Short-distance transitions are within a similar occupational
neighbourhood, reducing employer uncertainty about candidate fit and
productivity; long-distance transitions cross larger skill or
hierarchical gaps, and employers may discount wage offers to compensate
for that uncertainty, or workers may accept lower wages to gain entry to
a distant occupation.

The threat to the horse race would require that these residual dyadic
wage effects are differentially correlated with the competing cost
matrices --- that they track one metric more closely than the others.
There is no obvious mechanism by which employer uncertainty would operate
through skill distance systematically differently than through
hierarchical distance; the dyadic wage residual story is essentially
metric-agnostic.

Path-specific wage differentials therefore attenuate the observed
distance signal but are unlikely to distort the horse race result, which
depends only on relative fit across metrics rather than absolute fit
levels. Regardless, incorporating path-specific wage differentials into
the cost matrix or the estimation is not feasible with the Canadian
census data: occupation is recorded at the census date, but wages are
derived from prior-year tax records, and for occupational switchers the
observed wage reflects the origin occupation rather than the destination.

For non-local mobility, the specificity regression faces an analogous
but distinct identification challenge. The regression estimates the
reduced-form relationship between credential and occupational specificity
--- and their interaction --- and distance traveled following a
credential acquisition. Both the enablement and motivation channels
predict a positive relationship between specificity and distance: a
specific credential opens doors to particular distant occupations that a
general credential does not (enablement), and specific credentials tend
to command wage premia at those destinations, making long-distance moves
financially attractive (motivation). Because both channels operate
through the same credentials and occupations, the interaction term
conflates the two. Separating the enablement channel from the motivation
channel would require either an instrument that shifts wage premia
without shifting human capital reallocation, or a structural model of
credential choice and occupational mobility. Neither is feasible with
the available data. The regression should therefore be interpreted as
identifying the reduced-form predictive relationship between specificity
and distance traveled, without attribution to a specific mechanism.

The positive distance restriction removes many apprenticeship completers, since NOC makes no distinction between apprentices and journeypeople of the same occupation — those already working in the target trade in 2016 show zero measured distance upon completion and are excluded by construction. Workers who complete an entire apprenticeship program within the five-year window from a different 2016 occupation remain in the sample, so apprenticeship programs are underrepresented rather than absent. The direction of bias from this underrepresentation is ambiguous and operates through several channels.

Red seal trades credentials are likely high on both $\text{EOS}_c$ and
$\text{OES}_j$: completers are concentrated in the target trade
occupation (pushing $\text{EOS}_c$ up), and red seal trades occupations
draw almost exclusively from their own credential (pushing $\text{OES}_j$
up). This suggests that the excluded zeros are high on the interaction
term, and their exclusion is conservative with respect to the main
hypothesis --- including them would strengthen rather than weaken the
predicted positive relationship between the interaction and distance
traveled, for those who do show a positive distance.

However, a countervailing consideration tempers this optimism. The
technical similarity between trades occupations in O\*NET skill space
means that even apprenticeship completers who do show a positive distance
--- those who entered from a different occupation --- may show small
measured distances. Plumbers and carpenters, for instance, share
substantial task content in O\*NET profiles despite being formally
distinct occupations with separate licensing regimes and credential
requirements. This raises a broader question about whether O\*NET skill
profiles adequately capture the labour market barriers between trades
that are technically similar but institutionally distinct. To the extent
that skill distance understates the true occupational distance between
trades, the interaction coefficient for apprenticeship completers in the
sample is attenuated regardless of how high their credential specificity
is. The net direction of bias from apprenticeship underrepresentation is
therefore unclear, and resolving it would require either a longer panel
that captures more complete-program transitions or administrative data on
apprenticeship registration and completion linked to the census records.



## Conclusion

- What we found \[PLACEHOLDER --- awaiting data\]

The central policy takeaway from this analysis is that labour market
adjustment mechanisms are not one-size-fits-all: the appropriate
intervention depends on which mobility regime governs the occupations
experiencing shortage. The three-regime structure identified in the
horse race maps directly onto three distinct policy levers.

In low-gating occupations, where skill transferability governs mobility,
reskilling and short-cycle training are the natural adjustment
mechanisms. Workers in adjacent occupations already possess much of the
relevant task content, and targeted upskilling can accelerate the
reallocation of labour toward occupations in shortage without requiring
full credential acquisition. The policy implication is that investment
in modular, portable training that builds on existing skill profiles
will generate the largest mobility response in this segment.

In medium-gating occupations, where hierarchical distance governs
mobility, the relevant bottleneck is not skill transferability but
career ladder structure. Workers must move through intermediate rungs to
reach higher-TEER occupations, and the policy lever is directing workers
toward the first rung of the relevant ladder rather than attempting to
skip steps. Bridging programs, mentorship structures, and targeted
recruitment into entry-level positions in shortage occupations are the
appropriate instruments here. Attempting to accelerate adjustment
through reskilling alone will be ineffective if the hierarchical
structure of the occupation requires progressive credential
accumulation.

In high-gating occupations, where the binary constraint governs mobility
--- either a worker holds the required credential or they do not ---
neither reskilling nor career ladder navigation can substitute for
formal credential acquisition. Adjustment in this segment depends
entirely on expanding formal training pipelines: increasing capacity in
the relevant credential programs, reducing barriers to program entry,
and shortening time-to-credential where possible. The policy implication
is sobering: adjustment in high-gating occupations is structurally slow,
and labour shortages in this segment are unlikely to be resolved through
labour market flexibility alone.

## Appendix A: O\*NET to NOC 2021 Skill Mapping

OaSIS provides occupation-level skill profiles for Canadian
classifications and would be an attractive data source for constructing
$C_{\text{skill}}$. However, OaSIS does not expose how those profiles
are constructed: the mapping between O\*NET and NOC occupations is not
observable, the weights linking underlying occupations are not published,
and there is no information on how multiple O\*NET occupations are
combined to form a NOC profile. This appendix describes an alternative
mapping constructed explicitly from published crosswalks, allowing full
transparency of inputs and weights, diagnostic evaluation of mapping
quality, and reproducibility of all outputs.

The mapping chains four classifications: O\*NET 2019 $\rightarrow$ SOC
2018 $\rightarrow$ NOC 2016 $\rightarrow$ NOC 2021. Each stage introduces
potential one-to-many relationships. A naive approach would assign equal
weight to all admissible paths, but this leads to over-weighting of
occupations that appear frequently in crosswalk chains and inclusion of
unrelated occupations. Instead, path weights are constructed to reflect
the degree of splitting at each stage: if a source occupation splits into
$n_s$ destination occupations at stage $s$, each path through that stage
receives weight $w_s = 1/n_s$, and the overall path weight is
$w_{\text{path}} = w_1 \times w_2 \times w_3$. Path weights are then
aggregated by origin O\*NET occupation and destination NOC 2021 code and
normalized to sum to one within each NOC to yield base weights. The
consequences of naive versus path weighting can be severe: for NOC 72600
(Air pilots, flight engineers and flying instructors), naive weights
assign five-eighths of the skill profile to agricultural workers, while
path weights assign 78\% to pilots.

A power transformation and renormalization is then applied to further
reduce the influence of mapping artifacts: $\tilde{w}_k \propto w_k^2$.
This concentrates weight on the dominant O\*NET contributors and
diminishes the influence of peripheral occupations that enter through
crosswalk chains with many splits. For NOC 72600, downweighting
increases the pilot share of the skill profile to 96\%. Each O\*NET
occupation is then represented as a point in a reduced skill space using
the first 10 principal components of the O\*NET skill content variables,
and for each NOC 2021 code the weighted centroid of its contributing
O\*NET occupations is computed using the downweighted mapping, yielding a
10-dimensional skill profile for each NOC occupation.

Two diagnostics assess the quality of each NOC's mapping in skill space.
Herfindahl concentration measures how concentrated the mapping weights
are across contributing O\*NET occupations: $H = \sum_k \tilde{w}_k^2$.
For example, a Herfindahl in excess of $\frac{2}{3}$ indicates that the
NOC skill profile is effectively dominated by a single O\*NET occupation
with a minimum weight of 0.79; low values indicate a diffuse mapping.
Scaled Similarity measures the internal coherence of the contributing
occupations in skill space: the weighted average distance of each
contributing O\*NET occupation from the NOC centroid is computed, then
normalized to a $[0.05, 1]$ interval by dividing by the maximum value
across all NOCs, multiplying by 0.95, and subtracting from 1, so that
$SS = 1 - 0.95 \times \bar{d} / \max_j \bar{d}_j$. The lower bound of
0.05 is chosen to match the approximate lower bound of the Herfindahl
index, making the two measures' floors commensurable. The mapping quality
composite is the product $H \times SS$, which is high only when the
mapping is both concentrated and internally coherent. The three
interpretation cases are: high $H$ with any $SS$ (no problem --- almost
all weight on a single O\*NET occupation); low $H$ with high $SS$ (no
problem --- diffuse but coherent); and low $H$ with low $SS$
(problematic --- a diffuse average of disparate occupations whose skill
profile is unreliable). NOC 41200 (University professors and lecturers)
is a canonical example of the third case.

NOC 2021 occupations in the bottom decile of $H \times SS$ are excluded
from the local clean and non-local clean samples and isolated in the
mapping quality diagnostic subsample. Because the exclusion criterion is
based on mapping quality rather than occupation characteristics, it
introduces noise into $C_{\text{skill}}$ selectively while leaving
$C_{\text{hier}}$ and $C_{\text{binary}}$ unaffected --- a property
exploited in the mapping quality falsification design described in
Section @sec-data. The full mapping pipeline is implemented in
the \texttt{onet-noc2021-crosswalk} repository
(\url{https://github.com/bcgov/onet-noc2021-crosswalk}) and is
constructed entirely from O\*NET data and published crosswalks.

## Appendix B: 2016 TEER mismatch

Admisible TEERs are the Target TEER $\pm1$ plus TEER 0 (management), where the
assumption is that Management roles are independent of Educational attainment. 

\begin{table}[H]
\centering
\begin{tabular}{lcr}
\toprule
Highest Attainment & Target TEER & Admissible TEERs \\
\midrule
No certificate, diploma or degree & 5 & 4, 5, 0 \\
Secondary (high) school diploma & 4 & 3, 4, 5, 0 \\
Apprenticeship or trades certificate or diploma & 3 & 2, 3, 4, 0 \\
College / CEGEP / other non-university & 2 & 1, 2, 3, 0 \\
University certificate or diploma below bachelor & 2 & 1, 2, 3, 0 \\
Bachelor's degree & 1 & 1, 2, 0 \\
Above bachelor (Master's, PhD, MD, etc.) & 1 & 1, 2, 0 \\
\bottomrule
\end{tabular}
\caption{Admissible TEERs by highest attainment level}
\label{tab:teer_mismatch}
\end{table}

## Appendix C: Normalization Robustness  {#sec-AC}

Recall that in the main body of the paper, we anchored all distance metrics by 
the conditional median hierarchical distance.  In this appendix we explore whether 
the results are robust to using instead either the 25th or 75th conditional quantiles. 
Conditional quantiles are computed over within-category hierarchical distances,
excluding the cross-category sentinel value of nine. The unconditional percentile
locates the same value in the full pairwise distribution including sentinel values. 
Skill anchor is the corresponding quantile of the skill distance distribution.

```{r}
#| echo: false
calibration |>
  mutate(
    q_cond = scales::percent(q_cond),
    p_uncond = scales::percent(p_uncond, accuracy = 0.1),
    h_anchor = as.integer(h_anchor),
    s_anchor = round(s_anchor, 2)
  ) |>
  rename(
    `Conditional Quantile` = q_cond,
    `Hierarchical Anchor` = h_anchor,
    `Unconditional Percentile` = p_uncond,
    `Skill Anchor` = s_anchor
  ) |>
  kable(
    format = "latex",
    caption = "Normalization anchors under alternative calibration quantiles",
    align = c("c", "c", "c", "c"),
    booktabs = TRUE
  ) |>
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE
  )
```


## References
