---
title: "Occupational Mobility Is Not One Thing: Evidence from the Canadian Labour Market" 
author: "Richard Martin"
date: today
format:
  pdf:
    documentclass: article
    number-sections: true
    geometry: margin=1in
bibliography: references.bib
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \DeclareMathOperator*{\argmin}{argmin}
editor: 
  markdown: 
    wrap: 72
abstract: |
  Workers predominantly move between proximate occupations, but proximity means different things across the labour market. We use entropic regularized optimal   transport to conduct a horse race among three distance metrics — skill, hierarchy,   and binary — and show that no single geometry governs occupational mobility. Instead, the appropriate metric depends on each occupation's education specificity: skill where skills are highly transferable, hierarchy where silos (occupational broad   groups) and platforms (education and experience) structure mobility, and binary where   human capital is so occupation-specific that all alternative occupations appear equally distant. Each regime implies a distinct policy lever for addressing labour shortages.
---



```{r, include=FALSE}
library(tidyverse)
library(here)
library(knitr)
library(kableExtra)
library(patchwork)
library(conflicted)
cdfs <- read_rds(here("out","dist_cdfs.rds"))
calibration <- read_rds(here("out", "calibration.rds"))
```

## Pre-registration Notice {.unnumbered}

This paper is posted as a pre-registration of study design, prior to Statistics Canada Virtual Data Lab access.  Anything that looks like a result should be interpreted as a prior. Pre-registration is intended to defend against claims of post-hoc rationalization.

## Introduction

Occupational mobility is not one thing. Workers tend to move between proximate occupations, but proximity is not self-evident: occupations differ along multiple dimensions simultaneously --- canonically the tasks undertaken [@gathmann2010general], but also by required skills, hierarchical position, and selective education requirements. Each implies a different geometry of the labour market, and different predictions about who moves where. We test which geometry governs mobility --- and find that the answer depends on where in the labour market you look.

We separate local mobility (transitions that occur without new 
highest educational attainment) from non-local mobility (transitions that occur when a new highest educational attainment is achieved). For local mobility, distance is the explanatory variable — we ask which of three candidate metrics best rationalizes observed occupational transitions: skill (Euclidean in PCA-reduced O\*NET space), hierarchical (tree metric on the NOC taxonomy), and binary (discrete metric). For non-local mobility, the role of distance is reversed: rather than asking which metric explains where workers go, we ask what determines how far they travel, treating transition distance as the outcome and measures of education and occupation specificity as its determinants.

Regarding local mobility, we anticipate that no single metric will dominate globally. A labour market in which doctors, managers, and labourers all follow the same distance geometry would be surprising. We therefore partition the transition data into three sub-regimes based on the *origin* occupation's education specificity, each mapping to a hypothesized winning metric --- skill distance where human capital is general, hierarchical distance where silo structure binds, and binary switching costs where human capital is so occupation-specific that all alternative destinations appear equally distant. The Conceptual Framework develops this partition in detail.

The policy relevance is direct. Labour shortages in different segments of the market require different adjustment mechanisms, and the right mechanism depends on which distance metric governs mobility in that sub-regime. Shortages where human capital is general may be addressed through reskilling, building on transferable skills. Shortages where silo structure binds are best addressed by increasing supply at the bottom platform of the relevant silo, facilitating upward movement to address shortages at higher levels. Shortages where human capital is occupation-specific cannot be addressed through either mechanism: the relevant human capital is acquired only through specific education programs, so interventions must target access to and capacity within those programs directly. Getting the geometry right is therefore not merely a methodological contribution --- it determines which policy lever to pull.

For local mobility, we use entropic regularized optimal transport (EROT) [@peyre2019computational; @cuturi2013sinkhorn] to conduct the horse race, evaluating fit using relative KL divergence and relative Cost-Weighted Total Variation Distance (CWTVD). For non-local mobility, we estimate the reduced-form relationship between transition distance and two specificity measures --- the education's occupation specificity (EOS) of the new highest attainment, and the *destination* occupation's education specificity (OES). These measures proxy for the intensity of education-driven jumps: a new highest attainment with a high EOS opens doors to particular distant occupations, and a destination occupation with a high OES draws selectively from particular educations.

The two models provide complementary cross-regime falsification evidence. The specificity regression provides a clean test: measures of specificity should predict distance within the non-local regime but fail to do so in the local regime, where human capital is relatively stable. The horse race falsification is necessarily joint (see @sec-data): the falsification sample contains wrong-regime observations alongside measurement noise, so a finding that fit degrades outside the local clean sample is consistent with regime misassignment but does not rule out mismeasurement as the dominant source.

[FINDINGS PLACEHOLDER]

## Conceptual Framework

The central structural choice in this paper is the partition of 
occupational transitions into two qualitatively distinct regimes. A 
local transition occurs when a worker changes occupation without 
acquiring new educational attainment --- the worker's highest 
educational attainment in 2021 is the same as in 2016. A non-local 
transition occurs when a worker's highest educational attainment 
increases between the two census years. This partition reflects a 
fundamental difference in the mechanism governing mobility. For local 
transitions, the 2016 occupation is an informative starting point: 
human capital is relatively stable over the inter-census period, and 
distance from that starting point governs destination choice. For 
non-local transitions, the 2016 occupation is largely uninformative: 
a new highest educational attainment shifts the worker's position in 
occupational space in a way that is not captured by the 2016 
occupation, acting as a wormhole --- enabling discontinuous jumps that 
would otherwise be difficult to reconcile with their 2016 occupation.

Two symmetric specificity measures play central roles in the analysis, 
one defined over occupation (NOC) and one over *education*: the cross product 
of field of study and highest level of attainment. These 
measures are derived from Statistics Canada Table 98-10-0404-01 --- a 
one-quarter sample of the Canadian population, of which at most 
one-quarter overlaps with the linked microdata used for estimation.

To calculate an education's occupation specificity (EOS), we begin by 
calculating the KL divergence of $P(\text{NOC}|\text{education})$ from 
$P(\text{NOC})$. A higher value indicates that an education is 
distinctive in its occupation proportions. Because the empirical 
distribution can only place mass on a limited number of occupations for 
niche educations, KL divergence has a mechanical floor; we therefore 
residualize $\log(\text{KL})$ on $\log(T)$ where $T$ is the total 
number of workers with that education. The residual from this 
regression is defined to be EOS. This measure enters only the 
non-local partition as a continuous regressor proxying for wormhole 
entrance intensity --- a high EOS value increases the probability of 
a jump.

A symmetric procedure is followed to calculate an occupation's 
education specificity (OES). We begin by calculating the KL divergence 
of $P(\text{education}|\text{NOC})$ from $P(\text{education})$. A 
higher value indicates that an occupation is distinctive in its 
education proportions. Because the empirical distribution can only 
place mass on a limited number of educations for niche occupations, KL 
divergence has a mechanical floor; we therefore residualize 
$\log(\text{KL})$ on $\log(T)$ where $T$ is the total number of 
workers with that occupation. The residual from this regression is 
defined to be OES. In the local partition, OES of the \emph{origin} 
occupation is used as a classifier defining the three sub-regimes; in 
the non-local partition, OES of the *destination* occupation 
serves as a continuous regressor proxying for wormhole exit intensity 
--- conditional on a high EOS, a high OES increases the probability of 
a long jump.

OES captures where an occupation sits between the poles of general and 
specific human capital [@becker1964human]. Occupations drawing 
broadly from the education distribution require general human capital 
in Becker's sense --- skills whose returns are portable across 
employers and occupations --- while occupations drawing narrowly from 
specific credential pipelines require occupation-specific human 
capital, valuable within the occupation but not transferable beyond it. 
Occupational licensing [@kleiner2013] represents an explicit 
form of OES; licensing lists capture credential selectivity directly, 
but rely on curated registries that are often incomplete, 
jurisdiction-specific, and require manual updating. OES measures the 
same underlying phenomenon from the equilibrium education mix observed 
in the data, without requiring a curated list.

Canada's National Occupational Classification (NOC) organizes 
occupations into a five-digit hierarchy. The first digit identifies one of ten broad occupational categories (e.g., Natural and Applied 
Sciences, Health, Trades). The remaining digits subdivide each category into progressively finer groupings down to the unit group level. 
Crosscutting this hierarchy is the Training, Education, Experience and 
Responsibilities (TEER) system, which assigns each occupation a level 
from 0 to 5 reflecting the degree of training and experience required: 
TEER 0 denotes senior management, TEER 1 requires university education, and TEER 5 requires short-term work demonstration only. The broad 
category defines the silo boundary in the analysis; the TEER levels 
define the platforms within each silo.

The three sub-regimes are named after their hypothesized winning metric 
rather than after the OES dimension that defines them --- a convention 
adopted for clarity, not a presumption of results. The tertile 
boundaries are not theoretically derived; if the three-regime structure 
is genuine, the horse race results should be robust to moderate 
misclassification at the boundaries. The Skill sub-regime (low OES) is 
hypothesized to be governed by skill distance: skills are generic and 
mobility is not constrained by hierarchical structure. The Hierarchy 
sub-regime (middle OES) is hypothesized to be governed by hierarchical 
distance: skills are specific enough that silo structure binds, with 
mobility predominantly occurring within silos. The Binary sub-regime 
(high OES) is hypothesized to be governed by binary switching costs: 
human capital is so occupation-specific that the richer distance 
structure of the other metrics is counterproductive, and mobility is 
better described as either staying or, rarely, leaving at equal cost 
regardless of destination.

The Skill sub-regime includes occupations with bottom tertile education specificity, expanded to include all TEER 0 (senior management) and TEER 5 (labour) occupations regardless of specificity tertile. This override reflects two distinct mechanisms. TEER 0 occupations exhibit portable managerial human capital that is not confined to a single occupational silo — a school principal or library manager can move laterally across organizational contexts in a way that a physician or dentist cannot. TEER 5 occupations are frequently in different hierarchy silos, but occupy a flat valley bottom in skill space — movement along the valley floor is easier than hierarchical distance would suggest.

The Hierarchy sub-regime includes the remaining middle 
tertile occupations, where education specificity is sufficient to 
generate silo structure but not so high as to foreclose within-silo 
movement across platforms. The Binary sub-regime is comprised of the 
remaining occupations that have top-tertile OES.  Workers in these occupations are 
more likely to be in an absorbing state: their human capital is too 
specific to support further local mobility.
Thus a high OES predicts both a long initial jump — observable in the non-local sample when a new credential is acquired — and subsequent immobility — observable in the local sample when no new credential is acquired. At the worker level only one of these is observed in a five-year window, but the two predictions are tested in separate samples.

The local/non-local partition has a natural falsification structure. Each model is tested both where it should work and where it should not, and the regime-specific predictions provide stronger evidence for the causal interpretation of the underlying mechanisms than either falsification alone would. A single confounder would have to operate in opposite directions across regimes to account for both patterns; multiple confounders would each have to align fortuitously with the predicted cross-regime structure. Neither is impossible, but both are less parsimonious than the theoretical mechanism itself. The design does not eliminate the possibility of confounding, but it raises the bar considerably: the model's validity rests not on the absence of confounders but on the implausibility of a story that generates the predicted pattern of opposing failures across regimes.

## Cost Metrics

The skill distance matrix $C_{\text{skill}}$ is constructed from 161 
O\*NET measures spanning skills, abilities, knowledge, and work 
activities, mapped to NOC 2021 occupations via a purpose-built 
crosswalk @onet_to_noc. An alternative approach would be to 
use OaSIS occupation-level skill profiles, but we construct our own 
crosswalk for a specific reason: falsifiability. The horse race 
requires that measurement error in the skill distance matrix be 
attributable to a known source --- namely, imprecision in the O\*NET 
to NOC mapping. OaSIS obscures the structure of that mapping, making 
it impossible to characterize where and how measurement error enters 
$C_{\text{skill}}$; our crosswalk makes this structure transparent 
and therefore makes the identification strategy falsifiable. Each NOC 
is represented as a weighted average of its constituent O\*NET 
occupations, with weights determined by the crosswalk mapping.  For more details on the mapping see @sec-AA.

O\*NET provides two scales for each measure: Importance (how important the attribute is to the occupation) and Level (the degree to which the attribute is required). Both scales are standardized to a common 0--10 range and combined via geometric mean, yielding a single score that reflects both the relevance and the intensity of each attribute. The geometric mean is preferred over the arithmetic mean as it penalizes cases where one scale is near zero — an attribute that is important but required at negligible level, or required at high level but deemed unimportant, receives a lower composite score than one scoring moderately on both dimensions. The two scales are highly correlated in practice, so this choice is unlikely to be consequential for the resulting distance matrix, but we note it for transparency.

In high-dimensional spaces, Euclidean distances become increasingly uniform across pairs as dimensions accumulate, reducing their discriminating power. Principal components analysis is applied to the (standardized) 161-dimensional skill profiles, and the number of components retained is determined by a search over $k$: starting from $k=1$ we increment $k$ until the Spearman rank correlation between pairwise distances in the k-dimensional PCA space and pairwise distances in the full 161-dimensional space exceeds 0.99. This criterion --- rank preservation of pairwise distances rather than variance explained --- is chosen because the object of interest is the distance matrix itself, not the underlying skill dimensions. Twelve principal components satisfy this criterion. The skill distance between any two occupations is then the Euclidean distance between their coordinates in the retained 12-dimensional PCA space.  An interactive [skill distance heatmap](https://rpmartin.github.io/not-one-thing-companion/#sec-skill-heatmap) of the resulting matrix is available in the online companion.

The binary switching cost matrix $C_{\text{binary}}$ assigns a cost of zero to remaining in the same occupation and one to any transition. Structurally, the labour market is comprised of as many distinct silos as there are occupations, each with no internal structure of any kind. $C_{\text{binary}}$ serves two roles in the analysis. First, it is the theoretical straw man in the global horse race: if occupational mobility costs are purely a function of whether a switch occurs rather than how far, then $C_{\text{binary}}$ should rationalize observed mobility patterns better than $C_{\text{skill}}$ or $C_{\text{hierarchy}}$. Second, and more subtly,  $C_{\text{binary}}$ is best suited to settings where mobility is limited --- where most workers stay rather than move. In such settings, the information contained in a richer cost structure can cause model fit to worsen. This second role motivates the Binary sub-regime hypothesis: in occupations where human capital is so specific that all exits appear equally distant, binary is not merely the straw man but the correct description of the mobility geometry.

The hierarchical distance matrix $C_{\text{hierarchy}}$ is derived directly from the NOC 2021 taxonomy, treating the occupational classification structure as a tree metric. Distance between any two occupations is defined as follows: zero if all five digits match (same unit group); one if the first four digits match (same minor group, different unit group); two if the first three digits match (same sub-major group, different minor group); three plus the absolute difference in TEER level if only the first digit matches (same broad occupational category, different sub-major group); and nine for any cross-category transition. The absolute TEER difference ranges from zero to five, so within-category distances range from three to eight, preserving a strict separation from the cross-category sentinel value of nine.  An interactive [hierarchical distance heatmap](https://rpmartin.github.io/not-one-thing-companion/#sec-hier-heatmap) of the resulting matrix is available in the online companion.

NOC broad groups function as silos — bounded mobility clusters with limited entry points and internal progression along skill levels [@doeringer1971internal] — and TEER levels function as platforms within each silo, capturing the vertical skill stratification that governs upward movement. By construction, any between-platform move (minimum distance 4) is further than any within-platform move (maximum distance 3), and any cross-silo move (distance 9) is further than any within-silo move (maximum distance 8). The NOC broad group is a coarser unit than the occupational clusters described by @althauser1989, but the appropriate level of aggregation for the horse race: a finer partition would increase the preponderance of cross-category sentinel values in the hierarchical distance distribution, collapsing the metric's geometry toward binary and undermining its role as a distinct candidate in the horse race.

## Cost Metric Normalization

The three cost matrices are defined on different scales by construction: skill distances are continuous and spread over a wide range, hierarchical distances are discrete with a sentinel value of nine reserved for cross-category transitions, and binary distances take only the values zero and one. Direct comparison in the horse race requires that the matrices be anchored to a common scale; without normalization, the regularization parameter implies very different levels of concentration for each matrix, making the horse race results uninterpretable.

We normalize skill and hierarchical distances to a common scale by anchoring both at a substantively meaningful unit of hierarchical distance. The anchor is hierarchical distance 3 — the maximum distance without changing platforms. Distance 3 falls at the fourth percentile of the hierarchical distribution, reflecting the preponderance of cross-silo pairs. The corresponding fourth percentile of the skill distance distribution is 7.145. Dividing hierarchical distances by 3 and skill distances by 7.145 anchors both matrices at the same point in their respective distributions. @fig-cdfs shows the empirical CDFs of all three matrices before and after normalization. The raw distributions differ substantially in scale and shape; after normalization, skill and hierarchical CDFs overlap closely over the range of distances relevant to observed mobility, diverging only in the right tail. The sensitivity of horse race results to different anchors is examined in @sec-AC.


```{r}
#| warning: false
#| echo: false
#| label: fig-cdfs
#| fig-cap: "Cost Matrix Normalization: Raw vs. Normalized CDFs"
cdfs
```

## Mobility Data {#sec-data}

The analysis uses the Census 2016--2021 longitudinal file, constructed 
by linking long-form respondents across the two census cycles via the 
Derived Record Depository (DRD), Statistics Canada's national 
longitudinal database of personal identifiers. Linkage is 
probabilistic; details of the matching keys and overall linkage rate 
will be verified upon data access and reported in the final paper.

Canada's NOC 2021 hierarchy nominally contains 516 unit groups, but 
the five Senior Manager NOCs (00011--00015) are aggregated into NOC 
00018 in the CIP-NOC table used to derive EOS and OES, leaving 512 
occupations. Eight unit groups have no admissible O\*NET mapping and 
are excluded from all estimation samples, leaving 504 occupations in 
the analysis.

### Local Mobility Samples

For local mobility, the estimation sample is partitioned into two 
subsets: the local clean sample and its complement. The local clean sample is 
comprised of individuals with an observed occupation in both 2016 and 2021, no 
increase in highest educational attainment between the two census years, no self-employment in either year, no TEER--attainment mismatch in 2016, and an O\*NET-to-NOC mapping quality score above the bottom decile. This is the sample where the horse race should work best and where the specificity regression 
should find nothing.

The ideal falsification sample would be the non-local clean sample  --- 
identical selection criteria, with the sole difference being
that a new highest attainment occurred. Unfortunately, individuals whose
highest educational attainment increases between census years are rarely 
observed in TEER 0, 1, 2, or 3 occupations in 2016. Support heterogeneity confounds cross-sample fit comparisons: with different margins the fit metrics are measuring fundamentally different objects. Normalizing by the independence baseline does not restore comparability since the baseline itself is support-dependent --- $KL(P^{\text{obs}} \| P^{\text{ind}})$ varies with support composition, so $\Delta \text{KL}$ and $\Delta \text{CWTVD}$ cannot be meaningfully compared across samples with different supports.

Given that the non-local clean sample alone does not provide a comparable support, we pool the non-local clean sample with mismeasured observations and observations with poor O\*NET-to-NOC mapping quality to ensure that both samples have similar supports. Occupational mismeasurement is identified by two criteria. First, self-employment in either year, on the grounds that self-employment is associated with role multiplicity --- owners simultaneously performing managerial, technical, and administrative functions --- making the reported NOC unreliable for measuring occupational distance. Second, 
TEER--attainment mismatch, where the reported occupation is not 
representative of the individual's highest educational attainment. 
For the local observations this filter is applied to the 2016 
occupation --- the highest attainment may not be reflected there 
because either the attainment was very recent, or perhaps was a 
foreign credential not yet recognized.\footnote{For details 
regarding TEER--attainment mismatch criteria, see Appendix B.} 
The mapping quality criterion retains individuals whose 2016 
occupation falls in the bottom decile of the O\*NET-to-NOC mapping 
quality composite; poor mapping introduces noise specifically into 
$C_{\text{skill}}$ while leaving $C_{\text{hierarchy}}$ unaffected.

The complement of the local clean sample thus serves as a joint falsification containing three sources of expected fit degradation --- wrong regime, measurement noise, and metric-specific mapping noise --- and tests whether fit degrades, but cannot attribute degradation to any specific source. Mismeasurement is in this sense a feature rather than a bug: the mismeasured local observations cover the full labour market, ensuring broad origin support comparable to the local clean sample and making the fit metrics directly comparable.

### Non-Local Mobility Sample

The non-local clean sample comprises individuals who acquire a new 
highest educational attainment between 2016 and 2021 and report an 
observed occupation in both periods, subject to the same measurement 
restrictions as the local clean sample: no self-employment in either 
year, no TEER--attainment mismatch\footnote{For the 
non-local sample the TEER--attainment mismatch filter is applied to 
the 2021 occupation, where a credential acquired late in the 
inter-census window may not yet be reflected in the reported 
occupation.}, and an O\*NET-to-NOC mapping quality score above the bottom decile.

For the specificity regression, a four-sample diagnostic structure is
used. The non-local clean sample is the primary identification
sample, where the wormhole mechanism is operative and measurement is
clean. We then introduce measurement problems one at a time, holding
regime assignment correct. First, poor O\*NET-to-NOC mapping quality,
which should degrade $\beta$ coefficients and within-$R^2$ when
distance is measured in skill space while leaving hierarchical results
virtually unchanged --- a metric-specific prediction that serves as
a falsification of the mapping quality composite itself. Second,
occupation mismeasurement (self-employment or TEER--attainment
mismatch in 2021), which introduces noise into the outcome variable
regardless of which distance metric is used, producing symmetric
degradation across both. Finally, the local clean sample serves as
the falsification sample, where human capital is stable and all
$\beta$s are predicted to be approximately zero. $\beta$ coefficients
and within-$R^2$ are expected to degrade progressively from the first
sample to the third, then collapse toward zero in the falsification
sample. Since $\beta$ coefficients are not support-sensitive in the 
same way as $\Delta KL$ and $\Delta CWTVD$, support heterogeneity 
does not undermine these cross-sample comparisons; within-$R^2$ is 
reported alongside support size for each sample to facilitate 
interpretation.


## Local Model {#sec-local}

Mobility costs are latent: workers' switching costs are not observable 
in principle, so any distance matrix is necessarily a specification of 
an unobservable quantity. The three candidate matrices --- skill, 
hierarchical, and binary --- represent distinct hypotheses about the 
structure of those latent costs. The horse race asks which specification best rationalizes observed transitions, not which one correctly measures an observable.

The entropy-regularized optimal transport problem minimizes the following Lagrangian, where the first term is the transport cost, the second is the entropic regularization penalizing deterministic transport plans, and the third and fourth enforce the marginal constraints $\sum_j P_{ij} = a_i$ and $\sum_i P_{ij} = b_j$ via dual variables $f_i$ and
$g_j$.

$$\mathcal{L} = \sum_{ij} P_{ij} C_{ij} + \varepsilon \sum_{ij} P_{ij}(\log P_{ij} - 1) + \sum_i f_i(a_i - \sum_j P_{ij}) + \sum_j g_j(b_j - \sum_i P_{ij})$$

where $a_i$ and $b_j$ are the exact occupation stocks in 2016 and 2021
respectively. 

The first-order conditions yield:
\begin{equation}
\log P_{ij} = \frac{f_i}{\varepsilon} + \frac{g_j}{\varepsilon} - 
\frac{C_{ij}}{\varepsilon}
\end{equation}
Exponentiating, the optimal transport plan takes the form 
$P^\star_{ij} = u_i \, e^{-C_{ij}/\varepsilon} \, v_j$, where 
$u_i = e^{f_i/\varepsilon}$ and $v_j = e^{g_j/\varepsilon}$ are 
scaling vectors. Strict convexity of the regularized objective 
guarantees a unique solution. The solution is therefore a row- and 
column-rescaling of the exponential kernel $K_{ij} = e^{-C_{ij}/\varepsilon}$, 
where the Sinkhorn algorithm recovers it by alternating row and column 
normalization until both marginal constraints are satisfied to within a specified tolerance. All computations use a log-domain Sinkhorn solver 
for numerical stability, implemented in the \texttt{safesink} R package [@safesink].

In log space, the optimal transport plan decomposes into additive 
origin effects ($\log u_i$), additive destination effects ($\log v_j$), and a dyadic cost term ($-C_{ij}/\varepsilon$). The scaling vectors 
absorb any factor that operates at the level of the origin or 
destination occupation alone; the cost matrix captures the residual 
structure that depends on the specific origin-destination pairing. The 
horse race asks which specification of that dyadic structure best 
rationalizes observed transitions. The scaling vectors $u_i$ and $v_j$
have a natural interpretation as market-clearing shadow prices:
 $\log(v_j)-\log(u_j)$ measures net demand pressure at occupation $j$, capturing the degree to which the destination constraint binds more tightly than the origin constraint after controlling for cost structure. The online companion plots this statistic under skill vs. hierarchical distance, with points coloured by sub-regime; occupations far from the diagonal are those where the choice of geometry is consequential for policy.
 
The natural alternative for evaluating occupational mobility costs is 
the gravity model [@anderson_vanwincoop_2003; @silva2006log], which 
estimates origin and destination fixed effects jointly with the cost 
parameter(s). This joint estimation is precisely the problem in our 
context: fixed effects absorb variation in mobility rates that may 
reflect true cost structure, allowing misspecified cost matrices to 
achieve good fit by compensating through the fixed effects. Using a gravity model would be akin to randomly dosing the horses with various performance enhancing drugs — all competitors run faster, but the winner may be the fastest horse or simply the one whose treatment happened to help most, and the two cannot be disentangled.  EROT avoids this issue by holding the marginal structure fixed at observed values, separating the question of how much mobility occurs --- determined by the margins --- from the question of where mobility flows --- shaped by the cost matrix.

The regularization parameter $\varepsilon$ has a natural interpretation
in the random utility framework. The optimal transport plan $P^\star_{ij} = u_i \, e^{-C_{ij}/\varepsilon} \, v_j$ is the static equilibrium of a discrete choice model in which workers draw idiosyncratic utility shocks scaled by $\varepsilon$ over destination occupations. Low values of $\varepsilon$ concentrate the transport plan on cost-minimizing assignments, approaching a deterministic reallocation in the limit; high values diffuse mobility across destinations, approaching the independence baseline $P_{ij} = a_i b_j$ as $\varepsilon \to \infty$. The RUM analogy governs the interior structure of the transport plan --- how probability mass is allocated across origin-destination pairs --- but does not extend to the margins. In a standard logit model the choice probabilities are outcomes of utility maximization with no constraint on their aggregates; in EROT the origin and destination occupation distributions are hard constraints that the transport plan must exactly reproduce. This distinction is methodologically consequential: because the margins cannot adjust to absorb cost misspecification, a poorly chosen cost matrix cannot hide behind fixed effects or scale recalibration --- its misfit appears directly in the objective.  

A critical identification constraint follows directly from the form of
the optimal transport plan: only the ratio $C_{ij}/\varepsilon$ enters
$P^\star$. The scale of the cost matrix and the value of $\varepsilon$
are therefore not separately identified. This means that results
reported at a single value of $\varepsilon$ are uninterpretable without
an implicit normalization assumption --- and that assumption can be
gamed, since any cost matrix can be made to fit well by tuning
$\varepsilon$ to its scale. A horse race conducted at a single value is not a horse race; it is a calibration exercise.

Evaluating across a grid of $\varepsilon$ values sidesteps this by 
looking for dominance that is robust across scales --- a metric that 
wins at multiple values cannot be attributed to a fortuitous scale 
alignment. We evaluate across a log-spaced grid of $\varepsilon$ 
values ranging from 0.5 to 2, spanning from relatively concentrated 
to relatively diffuse transport plans while remaining in the region 
where the fit metrics retain discriminating power. The criterion for 
a robust winner requires some care. At low values of $\varepsilon$, 
cost-sensitive matrices concentrate probability mass too aggressively 
relative to observed mobility, incurring an overconfidence penalty 
that mechanically favours the flat binary matrix regardless of which 
metric truly best rationalizes mobility. This low-$\varepsilon$ binary 
lead is therefore uninformative about genuine descriptive accuracy and 
should not be interpreted as evidence that binary is the superior 
metric. The operative criterion is as follows: a non-binary metric is 
declared the winner if it dominates the other non-binary competitor across the sweep, even if it loses to binary at low values of $\varepsilon$ where the overconfidence penalty is operative. Binary requires a higher bar: genuine binary dominance requires persistence across the entire grid, not just at low $\varepsilon$. The $\varepsilon$ grid thus serves a dual purpose:  identifying the winning metric and distinguishing genuine binary dominance from the penalty artifact. The [online companion](https://rpmartin.github.io/not-one-thing-companion/#sec-simulation) demonstrates the low $\varepsilon$ artifact using simulation data (where the DGP is known).

The horse race is evaluated using two complementary fit metrics, both
normalized relative to the independence baseline $P^{\text{ind}}_{ij} =
a_i b_j$, so that 0 indicates no improvement over independence, 1
indicates perfect fit, and negative values indicate a transport plan that fits observed mobility worse than independence.

The first metric is relative KL divergence improvement:

$$\Delta KL = \frac{KL(P^{\text{obs}} \parallel P^{\text{ind}}) - KL(P^{\text{obs}} \parallel \hat{P})}{KL(P^{\text{obs}} \parallel P^{\text{ind}})}$$

where $KL(P \parallel Q) = \sum_{ij} P_{ij} \log(P_{ij} / Q_{ij})$.
This measures the share of the divergence between observed transitions
and independence that is explained by the fitted transport plan. KL
divergence is sensitive to probability mass wherever it is misallocated, but is geometry-blind: misallocating mass to a nearby occupation is penalized identically to misallocating it to a distant one.

The second metric is relative Cost-Weighted Total Variation Distance (CWTVD) improvement:

$$\Delta CWTVD = \frac{CWTVD(P^{\text{obs}}, P^{\text{ind}}) - CWTVD(P^{\text{obs}}, \hat{P})}{CWTVD(P^{\text{obs}}, P^{\text{ind}})}$$

where $CWTVD(P^{\text{obs}}, \hat{P}) = \sum_{ij}
\frac{|P^{\text{obs}}_{ij} - \hat{P}_{ij}|}{2} \cdot C_{ij}$. 

CWTVD is a cost-weighted extension of total variation distance: the 
classical TVD between discrete distributions, 
$\frac{1}{2}\sum_{ij}|P_{ij} - Q_{ij}|$, is extended by weighting 
each absolute residual by the corresponding entry of a cost matrix, 
penalizing probability mass misallocated to geometrically distant 
origin-destination pairs more heavily than mass misallocated to nearby 
ones. Unlike KL divergence, CWTVD respects the metric structure of the 
outcome space; unlike the Earth Mover's Distance [@rubner2000earth], it does not solve an additional optimal transport problem over the residual mass, making it computationally trivial given an existing transport plan.

The scoring cost matrix $C$ used in CWTVD is the arithmetic mean of the
three candidate normalized cost matrices: $C_{\text{skill}}$,
$C_{\text{hierarchy}}$, and $C_{\text{binary}}$. Using a neutral common metric ensures that no candidate model is advantaged by the choice of scoring matrix. The two metrics are complementary by design: KL divergence captures overall probability mass misallocation regardless of geometry, while CWTVD captures whether misallocation is concentrated in nearby or distant pairs. A transport plan that scores well on both metrics fits observed mobility both accurately and geometrically.

## Non-Local Model {#sec-non_local}

The non-local model estimates the conditional-on-positive relationship between $\log(d_{ij})$ --- the log distance between 2016 and 2021 occupations --- and the specificity measures EOS and OES. The outcome is computed separately using skill and hierarchical distance; the wormhole mechanism should operate independently of which metric measures the jump, providing a robustness check on the distance measure choice.

The theoretical mechanism is asymmetric. A high value of EOS is the primary condition for a wormhole jump — the entrance is logically prior to the exit. Conditional on a sufficiently high EOS, OES determines the distance of the jump. This suggests a joint specification that includes both the EOS main effect and an EOS × OES interaction. Whether the joint model can separately identify entrance and exit intensity depends on the worker-level correlation between EOS and OES: when the two measures are highly correlated, the interaction term EOS × OES approximates a quadratic in a single specificity index, and the coefficients on EOS and EOS × OES can no longer be cleanly interpreted as entrance and exit intensity separately. We therefore estimate three specifications — two univariate models and the joint model — and report the Spearman correlation between EOS and OES as a diagnostic informing how much structural weight to place on the joint model's coefficients. The three estimating equations are:

$$
\log(d_{ij}) = \alpha + \beta_1 \, \text{EOS} + \delta_i + \varepsilon_{ij}
$$ {#eq-eos}

$$
\log(d_{ij}) = \alpha + \beta_2 \, \text{OES} + \delta_i + \varepsilon_{ij}
$$ {#eq-oes}

$$
\log(d_{ij}) = \alpha + \beta_3 \, \text{EOS} + \beta_4 \, \text{EOS} \times \text{OES} + \delta_i + \varepsilon_{ij}
$$ {#eq-joint}


Prior to estimation, EOS and OES are standardized to mean zero and unit variance using the full linked sample. This ensures that coefficients are on a common scale both within and across the four subsamples, making the parameter estimates interpretable as a comparison of effect sizes, and the degradation pattern across samples interpretable as a comparison of fit rather than a reflection of differences in variable scale.

The primary hypothesis is that all $\beta$s are greater than zero. A secondary hypothesis posits that $\beta_1 > \beta_2$: the logical priority of wormhole entrance over exit predicts that education specificity has a stronger reduced-form association with distance traveled than occupation specificity. The Spearman correlation between $\text{EOS}$ and $\text{OES}$ is itself informative about the structure of education-occupation matching in the Canadian labour market, and is reported as a descriptive diagnostic alongside the regression results.

The model is evaluated across four samples as described in 
Section \ref{sec-data}, with results reported for both skill and 
hierarchical distance outcomes. The within-$R^2$ --- the share of 
variance in $\log(d_{ij})$ explained by the specificity measures 
after partialling out origin occupation fixed effects --- is not 
expected to be large in absolute terms: transition distance is 
inherently noisy and EOS and OES are aggregate measures that capture 
only the systematic component of specificity-driven mobility. The 
relevant comparison is the ordering of within-$R^2$ across samples, 
not its absolute level.

The positive distance restriction removes many apprenticeship completers, since NOC makes no distinction between apprentices and journeypeople of the same occupation --- those already working in the target trade in 2016 show zero measured distance upon completion and are excluded by construction. Rather than threatening identification, this restriction improves measurement validity: the excluded zeros reflect NOC coarseness rather than true immobility, and their inclusion would introduce noise into the outcome variable. Workers who complete an entire apprenticeship program within the five-year window from a different 2016 occupation remain in the sample, and for these workers the wormhole mechanism is cleanly observable. The limitation is one of external validity rather than internal validity: estimates do not generalise to within-trade apprenticeship completers, whose human capital increment is real but invisible to NOC.


## Results  \[PLACEHOLDER --- awaiting data\]  {#sec-res}

### Local

\[KL and CWTVD improvement figures by cost metric, $\varepsilon$ and
subsample\]

\[KL and CWTVD improvement by 3 sub regimes, cost metric, $\varepsilon$ and
subsample\]

### Non-local

\[tie-fighter plot for the $\beta$\]

\[ table for the $R^2$\]

## Limitations

### Local Mobility

Two sample restriction mechanisms introduce a potential selection 
concern for local mobility. Non-linkage and non-observability of 
occupation are both disproportionately concentrated among 
low-attachment workers --- individuals with casual, seasonal, or 
intermittent employment whose identifying information is also less 
stable across census cycles --- so the two filters largely co-select 
the same low-skill population rather than introducing independent 
exclusions. This shared exclusion affects the representativeness of 
population-level transition estimates (external validity) but does not 
threaten the internal validity of the horse race. For the horse race
results to be biased, the excluded workers' transitions would need to be 
systematically more or less rationalizable by a particular cost 
matrix than those of their observed counterparts in the same region 
of occupational space. No such mechanism is apparent: exclusion is 
driven by attachment instability and identifier churn, neither of 
which has a clear relationship to how well transitions conform to 
any particular cost structure conditional on occupation. The residual 
unlinked population among otherwise-observable workers --- recent 
immigrants, young workers, and high-movers --- similarly lacks any 
obvious occupational-space concentration that would differentially 
favour or penalize a particular metric.

A notable omitted factor for local mobility  is \emph{path-specific} wage differentials. The additively separable component of wage gains --- recoverable from origin and destination wage levels alone --- is absorbed by the 
margins insofar as high-wage destinations attract larger inflows and 
low-wage origins larger outflows. The residual dyadic component, 
reflecting transition-specific wage premia and discounts beyond what 
origin and destination wage levels predict, is an omitted factor. 
Short-distance transitions are within a similar occupational 
neighbourhood, reducing employer uncertainty about candidate fit and 
productivity; long-distance transitions cross larger skill or 
hierarchical gaps, and employers may discount wage offers to 
compensate for that uncertainty, or workers may accept lower wages 
to gain entry to a distant occupation.

Thus, path-specific wage differentials are better understood as an 
alternative mechanism, rather than a confounder. If employers use the same metric that governs mobility to assess candidate fit uncertainty --- discounting wages for transitions that are long under the winning metric --- that 
discount reinforces worker selection toward shorter transitions under 
the same metric. Both channels predict identical patterns of 
transitions and are equally consistent with the winning metric being 
the correct description of the mobility geometry. The horse race 
identifies which distance geometry best rationalizes observed 
transitions in a reduced-form sense, but cannot separately attribute 
that rationalization to worker or employer preferences regarding 
distance.

### Non-Local Mobility

The non-local partition is restricted to individuals who acquire a 
new highest educational attainment between 2016 and 2021 and report 
an occupation in both periods. The excluded population is not 
disproportionately low-skill, but rather disproportionately those 
whose human capital increased the most. The most demanding programs 
--- medicine, law, dentistry --- require sufficiently intensive study 
that concurrent employment is largely precluded, meaning students in 
these programs are less likely to report a 2016 occupation and thus 
are excluded from the sample. The most demanding programs tend also 
to be the most specific, and the human capital increment they confer 
is large --- generating the longest new-attainment-driven occupational 
jumps. Excluding these observations truncates the upper tail of the 
distance distribution precisely where education specificity is 
highest. Estimates of the relationship between education specificity 
and distance traveled should accordingly be interpreted as lower 
bounds on the true relationship in the full population of new 
educational attainment. Once the linked data is available, a 
diagnostic plot of EOS against the proportion of workers missing a 
2016 occupation --- computed by education type --- will quantify the 
magnitude of this truncation.

The specificity regression faces an analogous identification 
challenge to the horse race. EOS and OES are observed and enter the 
regression, but cannot distinguish between two mechanisms that 
predict identical patterns. A specific education opens doors to 
particular distant occupations that a general education does not --- 
the human capital match is narrow, so only a subset of distant 
destinations become reachable following attainment (enablement). At 
the same time, workers with specific educations may face stronger 
incentives to seek out distant occupations where those skills are 
scarce and rewarded, making long-distance moves financially attractive 
relative to staying within a crowded local neighbourhood (motivation). 
Because both channels operate through the same EOS and OES measures, 
the reduced-form estimates conflate the two and cannot be separately 
attributed to either. The regression should therefore be interpreted 
as identifying the reduced-form predictive relationship between 
specificity and distance traveled, without attribution to a specific 
mechanism --- though the estimate remains informative: if either 
channel is operative, the regression will detect it.




## Conclusion

- What we found \[PLACEHOLDER --- awaiting data\]

The central policy takeaway from this analysis is that labour market adjustment mechanisms are not one-size-fits-all: the appropriate intervention depends on which mobility regime contains the occupation experiencing a shortage. The three mobility sub-regimes suggest three distinct policy levers.

In the Skill sub-regime, reskilling and short-cycle training are the natural adjustment mechanisms. Workers in adjacent occupations already possess much of the relevant task content, and targeted reskilling can accelerate the reallocation of labour toward occupations in shortage without requiring formal retraining. The policy implication is that investment in modular, portable training that builds on existing skill profiles will generate the largest mobility response in this segment.

In the Hierarchy sub-regime, the relevant bottleneck is not skill transferability but the hierarchy's silo and platform structure. An occupational shortage within one silo cannot be addressed by upskilling workers from a different silo, even if the occupations share a near-identical skill profile. Instead, shortages must be addressed within-silo, through either lateral movement on a platform or vertical movement between platforms. The best way to address a shortage at a higher platform is by increasing the supply of workers to lower platforms, freeing incumbents to pursue upward movement through the silo — the occupational analogue of the internal labour market adjustment mechanism described by @doeringer1971internal. Bridging programs, mentorship structures, and targeted recruitment into within-silo entry-level positions are the appropriate instruments.

In the Binary sub-regime, either a worker has the required education, or they do not. Neither reskilling nor within-silo movement can substitute for education. Adjustment in this sub-regime depends entirely on expanding formal training pipelines: increasing capacity in the relevant educational programs, reducing barriers to program entry, and shortening time-to-completion where possible. These are also the occupations that immigration policy should target, both in terms of admission and rapid credential recognition. 

If the hypothesized sub-regime structure is confirmed, the geometry of occupational mobility determines which policy lever to pull --- and the wrong lever is not just ineffective but potentially counterproductive. 

This analysis answers the question: NOC $\rightarrow$ NOC. Who goes there?

## Appendix A: O\*NET to NOC 2021 Skill Mapping {#sec-AA}

OaSIS provides occupation-level skill profiles for Canadian
classifications and would be an attractive data source for constructing
$C_{\text{skill}}$. However, OaSIS does not expose how those profiles
are constructed: the mapping between O\*NET and NOC occupations is not
observable, the weights linking underlying occupations are not published, and there is no information on how multiple O\*NET occupations are combined to form a NOC profile. This appendix describes an alternative mapping constructed explicitly from published crosswalks, allowing full transparency of inputs and weights, diagnostic evaluation of mapping quality, and reproducibility of all outputs.

The mapping chains four classifications: O\*NET 2019 $\rightarrow$ SOC
2018 $\rightarrow$ NOC 2016 $\rightarrow$ NOC 2021. Each stage introduces potential one-to-many relationships. A naive approach would assign equal weight to all admissible paths, but this leads to over-weighting of occupations that appear frequently in crosswalk chains and inclusion of unrelated occupations. Instead, path weights are constructed to reflect the degree of splitting at each stage: if a source occupation splits into $n_s$ destination occupations at stage $s$, each path through that stage receives weight $w_s = 1/n_s$, and the overall path weight is $w_{\text{path}} = w_1 \times w_2 \times w_3$. Path weights are then aggregated by origin O\*NET occupation and destination NOC 2021 code and normalized to sum to one within each NOC to yield base weights. The consequences of naive versus path weighting can be severe: for NOC 72600 (Air pilots, flight engineers and flying instructors), naive weights assign five-eighths of the skill profile to agricultural workers, while path weights assign 78\% to pilots.

A power transformation and renormalization is then applied to further
reduce the influence of mapping artifacts: $\tilde{w}_k \propto w_k^2$.
This concentrates weight on the dominant O\*NET contributors and
diminishes the influence of peripheral occupations that enter through
crosswalk chains with many splits. For NOC 72600, downweighting
increases the pilot share of the skill profile to 96\%. Each O\*NET
occupation is then represented as a point in a reduced skill space using the first 10 principal components of the O\*NET skill content variables, and for each NOC 2021 code the weighted centroid of its contributing O\*NET occupations is computed using the downweighted mapping, yielding a 10-dimensional skill profile for each NOC occupation.

Two diagnostics assess the quality of each NOC's mapping in skill space. Herfindahl concentration measures how concentrated the mapping weights are across contributing O\*NET occupations: $H = \sum_k \tilde{w}_k^2$. For example, a Herfindahl of $\frac{2}{3}$ indicates that the NOC skill profile is effectively dominated by a single O\*NET occupation with a weight of $\approx .8$.  In contrast, when $H=\frac{1}{3}$ the largest contributor's share is between $\frac13$ and $\sqrt{\frac13}=.58$. Scaled Similarity measures the internal coherence of the contributing occupations in skill space: the weighted average distance of each contributing O\*NET occupation from the NOC centroid is computed, then normalized to a $[0.05, 1]$ interval by dividing by the maximum value across all NOCs, multiplying by 0.95, and subtracting from 1, so that $SS = 1 - 0.95 \times \bar{d} / \max_j \bar{d}_j$. The lower bound of 0.05 is chosen to match the approximate lower bound of the Herfindahl index, making the two measures' floors commensurable. The mapping quality composite is the product $H \times SS$, which is high only when the mapping is both concentrated and internally coherent. The three interpretation cases are: high $H$ with any $SS$ (no problem --- almost all weight on a single O\*NET occupation); low $H$ with high $SS$ (no problem --- diffuse but coherent); and low $H$ with low $SS$ (problematic --- a diffuse average of disparate occupations whose skill profile is unreliable). NOC 41200 (University professors and lecturers) is a canonical example of the third case.

Occupations in the bottom decile of $H \times SS$ are excluded
from the local clean and non-local clean samples. Because the exclusion criterion is based on mapping quality rather than occupation characteristics, it
introduces noise into $C_{\text{skill}}$ selectively while leaving
$C_{\text{hierarchy}}$ and $C_{\text{binary}}$ unaffected --- a property exploited in the mapping quality falsification design described in @sec-data. The full mapping pipeline is implemented in
[@onet_to_noc] and is constructed entirely from O\*NET data and published crosswalks.

## Appendix B: 2016 TEER mismatch {#sec-AB}

Admissible TEERs are the Target TEER $\pm1$ plus TEER 0 (management), where the assumption is that management roles are independent of educational attainment. 

\begin{table}[H]
\centering
\begin{tabular}{lcr}
\toprule
Highest Attainment & Target TEER & Admissible TEERs \\
\midrule
No certificate, diploma or degree & 5 & 4, 5, 0 \\
Secondary (high) school diploma & 4 & 3, 4, 5, 0 \\
Apprenticeship or trades certificate or diploma & 3 & 2, 3, 4, 0 \\
College / CEGEP / other non-university & 2 & 1, 2, 3, 0 \\
University certificate or diploma below bachelor & 2 & 1, 2, 3, 0 \\
Bachelor's degree & 1 & 1, 2, 0 \\
Above bachelor (Master's, PhD, MD, etc.) & 1 & 1, 2, 0 \\
\bottomrule
\end{tabular}
\caption{Admissible TEERs by highest attainment level}
\label{tab:teer_mismatch}
\end{table}

## Appendix C: Normalization Robustness  {#sec-AC}

In the main body of the paper, we anchor all distance metrics at hierarchical distance 3 — the maximum within-silo horizontal distance, representing movement within a broad occupational group without crossing TEER levels. In this appendix we examine robustness to anchoring at hierarchical distance 2 (same sub-major group, different minor group) and hierarchical distance 4 (same broad group, one TEER step), representing one step down into shorter horizontal movement and one step up into vertical movement respectively. The unconditional percentile locates each anchor in the full pairwise distribution. 

```{r}
#| echo: false
calibration |>
  mutate(
    p_uncond = scales::percent(p_uncond, accuracy = 0.1),
    h_anchor = as.integer(h_anchor),
    s_anchor = round(s_anchor, 2)
  ) |>
  rename(
    `Hierarchical Anchor` = h_anchor,
    `Unconditional Percentile` = p_uncond,
    `Skill Anchor` = s_anchor
  ) |>
  kable(
    format = "latex",
    caption = "Normalization anchors under alternative calibration quantiles",
    align = c("c", "c", "c"),
    booktabs = TRUE
  ) |>
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE
  )
```

## Appendix D: Revised Sample Design {#sec-appendix-d}

An earlier version of this pre-registration proposed a four-sample 
diagnostic structure for the horse race evaluation. The intended logic was that progressive degradation in fit metrics across the measurement quality dimension would validate the local/non-local partition --- if regime misassignment is more damaging to model fit than mismeasurement, that ordering constitutes direct evidence for the theoretical content of the partition.

The four samples were:

- **Local clean**: local regime, no measurement problems
- **Mapping quality**: local regime, poor O\*NET-to-NOC mapping (differentially hurts skill)
- **Occupational mismeasurement**: local regime, self-employment or TEER--attainment mismatch (hurts all metrics)
- **Non-local clean**: non-local regime, no measurement problems

This design was revised because $\Delta KL$ and $\Delta CWTVD$ are 
only comparable across samples with similar supports --- the set of 
occupations with positive counts in both origin and destination 
margins. The one-at-a-time diagnostic samples have heterogeneous 
supports: the non-local clean sample has severely restricted origin 
support since individuals acquiring new credentials are rarely 
observed in TEER 0, 1, 2, or 3 occupations in 2016, while the 
mismeasured local samples have broad supports covering the full 
labour market. With different sets of occupations in the margins, 
the fit metrics are measuring fundamentally different objects even 
when the cost matrix is identical. The diagnostic granularity of the 
four-sample structure was therefore illusory --- differences in fit 
metrics across samples cannot be attributed to specific measurement 
problems when support heterogeneity is a simultaneous confound.

The revised two-sample design --- local clean versus its complement 
--- addresses this problem. The complement pools all non-local 
observations with all mismeasured local observations; the mismeasured 
local observations cover the full labour market and restore broad 
origin support comparable to the local clean sample. The cost is that 
the complement serves as a joint falsification rather than a 
diagnostic one: it contains three sources of expected fit degradation 
--- wrong regime, measurement noise, and metric-specific mapping 
noise --- and the design tests whether the horse race is 
regime-specific without separately identifying which source dominates. 
This is a weaker claim than the original four-sample design promised, 
but it is a claim the design can actually deliver.

Note that the four-sample structure is retained for the specificity 
regression, where $\beta$ coefficients are not support-sensitive in 
the same way as $\Delta KL$ and $\Delta CWTVD$. The support 
comparability problem is specific to the fit metrics used in the 
horse race and does not affect the regression-based evaluation of 
the non-local model.

## References
