---
title: "Occupational Mobility Is Not One Thing: Evidence from the Canadian Labour Market" 
author: "Richard Martin"
date: today
format:
  pdf:
    documentclass: article
    number-sections: true
    geometry: margin=1in
bibliography: references.bib
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{threeparttable}
  - \DeclareMathOperator*{\argmin}{argmin}
editor: 
  markdown: 
    wrap: 72
execute:
  echo: false
  message: false
  warning: false
abstract: |
  Workers predominantly move between proximate occupations, but proximity means different things across the labour market. We use entropic regularized optimal   transport to conduct a horse race among three distance metrics — skill, hierarchy, and binary — and show that no single geometry governs occupational mobility. Instead, the appropriate metric depends on each occupation's education specificity: skill where skills are highly transferable, hierarchy where silos (occupational broad groups) and platforms (education and experience) structure mobility, and binary where human capital is so occupation-specific that neither skill nor hierarchy predict destinations. Each regime implies a distinct policy lever for addressing labour shortages.
---



```{r, include=FALSE}
library(tidyverse)
library(here)
library(knitr)
library(kableExtra)
library(patchwork)
library(gt)
library(conflicted)
conflicts_prefer(dplyr::filter)
cdfs <- read_rds(here("out","dist_cdfs.rds"))
calibration <- read_rds(here("out", "calibration.rds"))
noc_edu_spec  <- read_rds(here("out", "noc_edu_spec.rds"))
```

## Pre-registration Notice {.unnumbered}

This paper is posted as a pre-registration of study design, prior to Statistics Canada Virtual Data Lab access.  Anything that looks like a result should be interpreted as a prior. Pre-registration is intended to defend against claims of post-hoc rationalization.

## Introduction

Occupational mobility is not one thing. Workers tend to move between proximate occupations, but proximity is not self-evident: occupations differ along multiple dimensions simultaneously — canonically the tasks undertaken [@gathmann2010general], but also by required skills, hierarchical position, and selective education requirements. Each implies a different geometry of the labour market, and different predictions about who moves where. We test which geometry governs mobility — and find that the answer depends on where in the labour market you look.

We separate local mobility, the shorter transitions that occur without a substantial gain in human capital, from non-local mobility, the potentially much longer transitions that occur when a worker acquires substantially more human capital through education. A worker gains substantially either by reaching a higher level of attainment or by retraining into a distant field at the same level; a further credential in a closely related field at the same level adds little and we group it with local mobility. The two regimes are not parallel phenomena. Education shapes where a worker enters the occupational distribution, and where they enter determines which distance geometry governs subsequent movement. Non-local mobility is therefore the mechanism behind local mobility's heterogeneity: the specificity of the human capital acquired during education determines the initial jump across the labour market, and the jump's destination determines which metric governs transitions thereafter.

For local mobility, distance is the explanatory variable — we ask which of three candidate metrics best rationalizes observed occupational transitions: skill (Euclidean in PCA-reduced O*NET space), hierarchical (tree metric on occupation taxonomy), and binary (degenerate  metric). For non-local mobility, the role of distance is reversed: rather than asking which metric explains where workers go, we ask what determines how far they travel, treating distance as the outcome and measures of education and occupation specificity as its determinants.

Canada's National Occupational Classification (NOC) organizes occupations into a five-digit hierarchy. The first digit identifies one of ten broad occupational categories (e.g., Natural and Applied Sciences, Health, Trades). The remaining digits subdivide each category into progressively finer groupings down to the unit group level. Crosscutting this hierarchy is the Training, Education, Experience and Responsibilities (TEER) system, which assigns each occupation a level from 0 to 5 reflecting the degree of training and experience required: TEER 0 denotes senior management, TEER 1 requires university education, and TEER 5 requires short-term work demonstration only. For hierarchical distance, occupation broad categories delineate separate labour market silos, and TEER levels define platforms within each silo.

Regarding local mobility, we anticipate that no single metric will dominate globally. A labour market in which doctors, managers, and labourers all follow the same distance geometry would be surprising. We therefore partition the transition data into three sub-regimes based on the origin occupation's education specificity, each mapping to a hypothesised winning metric: skill distance where human capital is general, hierarchical distance where silo structure binds, and degenerate binary where human capital is so occupation-specific that neither skill nor hierarchical distance orders the destinations.

The policy relevance is direct. Labour shortages in different segments of the market require different adjustment mechanisms, and the right mechanism depends on which distance metric governs mobility in that sub-regime. Shortages where human capital is general may be addressed through reskilling, building on transferable skills. Shortages where silo structure binds are best addressed by increasing supply at the bottom platform of the relevant silo, facilitating upward movement to address shortages at higher levels. Shortages where human capital is occupation-specific cannot be addressed through either mechanism: the relevant human capital is acquired only through specific education programs, so interventions must target access to and capacity within those programs directly. Getting the geometry right is therefore not merely a methodological contribution — it determines which policy lever to pull.

For local mobility we use entropic regularized optimal transport (EROT) [@cuturi2013sinkhorn; @peyre2019computational] to conduct horse races among the candidate distance metrics, evaluating how well each rationalizes the full occupation-to-occupation transition matrix. The difficulty is that no single metric is self-evidently correct: skill, hierarchical, and binary distances each have theoretical support, and the natural regression-based test answers a different question from the one we want to settle. A gravity specification regressing log transitions on a candidate distance with origin and destination fixed effects, estimated by PPML to accommodate the count structure and zero flows, recovers the partial association between distance and flow once the margins are absorbed. That association characterizes how the metric covaries with transitions; it does not deliver the complete transition matrix the metric implies, which is what makes the geometries comparable to one another and to what is observed. EROT recovers that matrix. Given the occupation margins, each candidate metric yields a single predicted transition matrix, the entropic-regularized plan minimizing transport cost under that metric's cost structure, which we score against the observed matrix. Because every predicted plan satisfies the same margins by construction, the comparison across metrics isolates the within-margin allocation the geometry alone governs. Fixing the margins also makes the comparison a fair one, which we develop in @sec-local. 

For non-local mobility, we estimate, at the individual level, the reduced-form relationship between transition distance and two specificity measures. These measures capture the entrance and exit of wormholes through occupation space — the discontinuous, credential-driven jumps that constitute non-local movement. Education occupation specificity (EOS) measures how distinctive an education's associated occupation distribution is relative to the occupation distribution overall, proxying for a wormhole entrance. Occupation education specificity (OES) measures how distinctive an occupation's associated education distribution is relative to the education distribution overall, proxying for a wormhole exit. Each is a size-residualized divergence whose construction we give in @sec-cf; here it is enough that a high value flags a distinctive conditional distribution and rankings below are on that residualized measure. High values of both EOS and OES are predicted to produce long-distance jumps across the labour market; crucially, neither alone is sufficient — it is the conjunction of a specific education and a specific destination that opens a wormhole. For example, the occupation with the highest OES is 31111: Optometrists and the education with the highest EOS is 51.17 Optometry (OD) — a matched pair, the credential and the occupation it almost exclusively feeds. An optometry graduate whose 2021 occupation is not 31111 would be expected to travel a far shorter distance, the entrance opening onto no corresponding exit. At the other end of the spectrum, 24.01 Liberal arts and sciences, general studies and humanities has the lowest EOS and 65102: Store shelf stockers, clerks and order fillers has the lowest OES; this combination is expected to yield the shortest distance travelled.


[FINDINGS PLACEHOLDER]

## Conceptual Framework {#sec-cf}

The central structural choice in this paper is the partition of occupational transitions into two qualitatively distinct regimes, by the size of the human-capital increment a worker acquires between census years. We define *education* as the cross product of highest attainment and field of study, and we read a change in education for its increment rather than for its mere occurrence. A worker acquires a large increment by reaching a higher level of attainment, or by changing field at the same level across a large distance in the field-of-study classification, which is genuine retraining into an area where the original credential does not transfer. A worker whose education record changes only across a small distance at the same level gains a small increment in human capital. We therefore place large-increment changes in the non-local regime, and small or zero increment changes in the local regime. 

The increment, not the record change, is what the partition turns on, and this matters because a same-level field change is observationally ambiguous. The census records highest attainment and one field of study, not the number of credentials a worker holds, so a worker reporting one field in 2016 and a related field in 2021 may have earned a second degree or may have reported a single degree two ways. We do not need to resolve which. A second credential in a closely related field at the same level is a negligible human-capital increment whether or not it was genuinely earned, so both readings imply a worker who has not been repositioned in occupation space, and both belong with local mobility. Routing on increment size rather than on the bare fact of a record change makes the partition robust to this ambiguity.

The partition reflects a fundamental difference in the mechanism governing mobility. For local transitions the 2016 occupation is an informative starting point, since human capital is relatively stable over the inter-census period and distance from that starting point governs destination choice. For non-local transitions the 2016 occupation is largely uninformative: a large human-capital increment shifts the worker's position in occupational space in a way the 2016 occupation does not capture, acting as a wormhole that enables discontinuous jumps which would otherwise be difficult to reconcile with where the worker started.

Two symmetric specificity measures play central roles in the analysis,
one defined over occupation and one over education. These measures are
derived from Statistics Canada Table 98-10-0404-01, which is based on
the long-form Census --- a random one-quarter sample of Canadian
households. The linked 2016--2021 file used for estimation is
constructed by linking long-form respondents across the two census
cycles, and since each cycle independently samples one quarter of
households, the linked file covers approximately one sixteenth of all
Canadian households. The overlap between the two sources is therefore
at most one quarter --- the share of the long-form 2021 respondents
who also appeared in the 2016 long-form. This limited overlap means
the specificity measures are derived largely from independent data
relative to the estimation sample, reducing the risk that EOS and OES
mechanically predict transition distances through shared sampling
rather than through the credential-selectivity mechanism. To calculate
an education's occupation specificity (EOS), we begin by calculating
the KL divergence of $P(\text{NOC}|\text{education})$ from
$P(\text{NOC})$. A higher value indicates that an education is
distinctive in its occupation proportions. Because the empirical
distribution can only place mass on a limited number of occupations for
niche educations, KL divergence has a mechanical floor; we therefore
residualize $\log(\text{KL})$ on $\log(T)$ where $T$ is the total
number of workers with that education. The residual from this
regression is EOS. Because it is a regression residual, EOS is signed
and centred near zero: a high value is a divergence above what the
education's size predicts, not a high raw divergence, and the
highest-EOS and lowest-EOS rankings used above are rankings on this
residual. This measure enters only the non-local partition as
a regressor proxying for wormhole entrance intensity. A symmetric
procedure is followed to calculate an occupation's education
specificity (OES): the KL divergence of $P(\text{education}|\text{NOC})$
from $P(\text{education})$ is residualized via regression $\log(\text{KL})
\sim \log(T)$. OES serves dual purposes: in the local partition, OES
of the *origin* occupation is used as a classifier defining three
sub-regimes; in the non-local partition, OES of the *destination*
occupation serves as a regressor proxying for wormhole exit intensity.

OES proxies for where an occupation sits between the poles of general and 
specific human capital [@becker1964human]. Occupations drawing broadly from 
the education distribution require general human capital in Becker's sense, 
skills whose returns are portable across employers and occupations, while 
occupations drawing narrowly from specific credential pipelines require 
occupation-specific human capital, valuable within the occupation but not 
transferable beyond it. Occupational licensing [@kleiner2013] is one 
institutional source of high specificity, and licensing lists offer a direct proxy for credential selectivity, but they rely on curated registries that 
are often incomplete, jurisdiction-specific, and require manual updating. OES proxies for the same underlying selectivity from the equilibrium education mix observed in the data, without requiring a curated list.

The three sub-regimes are named for their hypothesised winning metric
rather than for the dimension that defines them --- a labelling
convention adopted for readability, not a presumption of results. The
primary partition is by OES rank. The group sizes are imported from the
TEER groupings described below --- 96 Binary, 86 Skill, and 322
Hierarchy --- fixing the boundaries before any transition data is
examined. This borrowing is purely a sizing convention: TEER supplies
the three counts, not which occupations populate them. Assignment
within those counts is determined entirely by OES rank, with the 96
occupations with the highest OES forming the Binary sub-regime, the 86
with the lowest forming the Skill sub-regime, and the remaining 322
forming the Hierarchy sub-regime. Importing the sizes from TEER rather
than cutting the OES distribution at natural breakpoints is a
pre-commitment device: it prevents the boundaries from being chosen,
consciously or otherwise, to fit the transition data. @fig-regimes
confirms that OES varies continuously across the occupational
distribution with no visible clustering, so no natural breakpoints
exist to exploit in any case.

The partition follows directly from the human capital specificity
argument. Occupations with high OES draw narrowly from specific
credential pipelines and map to the Binary sub-regime, where the prediction is that no occupational geometry orders destinations. Occupations with low OES draw broadly from the education distribution; their human capital is
general in Becker's sense, and skill similarity governs mobility. The
majority sit between these poles, where the silo and platform structure
of the NOC hierarchy governs transitions. The resulting split ---
roughly 19\% Binary, 64\% Hierarchy, and 17\% Skill --- reflects a
prior that hierarchical distance governs the bulk of the Canadian
labour market, with skill and binary metrics governing the tails.

The TEER partition that supplies the group sizes has an independent theoretical motivation paralleling the OES logic. TEER enters this paper in three distinct roles: the level semantics introduced in the introduction (0 senior management through 5 short demonstration), the specificity ordering used here to motivate and size the partition, and the tree-metric distance use formalized in @sec-cost-metrics. The partition rests on the second of these. TEER 1 occupations, such as physicians, dentists, and lawyers, carry the most occupation-specific human capital and map to the Binary sub-regime, which makes no prediction about where a move leads. At the other end of the spectrum, we hypothesize that TEER 0 and 5 occupations carry so little occupation-specific human capital that they can freely traverse the broad occupational groups, and they map to the Skill sub-regime. TEER 2, 3, and 4 fall in between, with human capital specific enough to impede cross-silo transitions yet not so specific as to produce occupational lock-in, and map to the Hierarchy sub-regime. The two orderings rest on different logics, yet both motivate the same three-way conceptual split, so the TEER partition serves as a robustness check on the primary OES-based partition. The partitions agree on 61% of occupations; Cohen's $\kappa = 0.27$ reads lower because the construction forces equal group sizes, inflating the chance-agreement baseline against which $\kappa$ is measured. The check is whether the horse-race results hold under the TEER assignment, not whether the two assignments coincide.

We treat the OES partition as the primary specification and the TEER partition as a confirmatory check. TEER is a coarse, discrete proxy for specificity, built for an unrelated administrative purpose; OES is a continuous measure derived directly from the equilibrium education mix observed for each occupation. The two partitions agree where TEER's administrative categories happen to track specificity, and diverge where they do not. @tbl-partition-crosstab quantifies that divergence: row-level concordance is weakest for TEER-Skill (35 of 86 occupations, 41%), intermediate for TEER-Binary (49 of 96, 51%), and strongest for TEER-Hierarchy (225 of 322, 70%). We read this ordering as evidence that TEER 0 and TEER 5 are the noisiest administrative buckets relative to the underlying specificity construct: a non-trivial share of nominally general-human-capital occupations turn out, once measured directly via OES, to draw from a more selective credential pipeline than their TEER level implies. The case of 41210, College and other vocational instructors — TEER 1, yet low OES — illustrates the same problem from the other direction.

This yields a falsifiable secondary prediction. If TEER is a noisier proxy for the same underlying construct that OES measures directly, the OES-based horse race should produce a better fit for the predicted-winning metric, since misclassified occupations pull the TEER-based average toward origins where the predicted geometry does not govern. We predict this gap to be ordered Skill > Binary > Hierarchy, tracking the concordance ordering above — largest where TEER misclassification is most severe, smallest where the two partitions already agree closely. A failure of this ordering would undercut the claim that TEER and OES are measuring the same underlying construct at different resolutions rather than two distinct classifications.


```{r}
#| label: tbl-partition-crosstab
#| tbl-cap: "Sub-regime confusion matrix."

library(kableExtra)
library(dplyr)

lvls <- c("Skill", "Hierarchy", "Binary")   # low -> high education specificity

m <- noc_edu_spec |>
  mutate(across(c(`Based on TEER`,
                  `Based on Occupation Education Specificity`),
                \(x) factor(x, levels = lvls))) |>
  with(table(`Based on TEER`, `Based on Occupation Education Specificity`)) |>
  as.data.frame.matrix() |>
  tibble::rownames_to_column("TEER")

shade <- "#EAF0F4"

kbl(m, booktabs = TRUE, align = "lccc", row.names = FALSE,
    col.names = c("TEER", "Skill", "Hierarchy", "Binary"), linesep = "") |>
  add_header_above(c(" " = 1, "Occupation education specificity" = 3),
                   bold = TRUE) |>
  kable_styling(latex_options = "hold_position",
                full_width = FALSE, position = "center") |>
  column_spec(2, bold       = c(TRUE,    FALSE,   FALSE)) |>
  column_spec(3, bold       = c(FALSE,   TRUE,    FALSE)) |>
  column_spec(4, bold       = c(FALSE,   FALSE,   TRUE)) |>
  footnote(
    general = paste(
      "Counts are occupations (n = 504).",
      "Both partitions share the same group sizes by construction",
      "(86 Skill, 322 Hierarchy, 96 Binary);",
      "bold indicates concordance",
      "(309/504 = .61, Cohen's kappa = 0.27).",
      "The opposite-pole corners hold a single occupation between them:",
      "41210, College and other vocational instructors."
    ),
    general_title = "", footnote_as_chunk = TRUE,
    escape = FALSE, threeparttable = TRUE
  )
```


```{r}
#| warning: false
#| echo: false
#| label: fig-regimes
#| fig-cap: "Mobility Sub-Regimes"

noc_edu_spec |>
  pivot_longer(cols = contains("Based on"), names_to = "criteria", values_to = "Sub-Regime") |>
  ggplot(aes(T, specificity, text = noc, fill = `Sub-Regime`)) +
  scale_fill_viridis_d() +
  geom_point(shape = 21, size = 1.5, stroke = .5) +
  scale_x_log10(labels = scales::comma) +
  facet_wrap(~criteria) +
  theme(text = element_text(size = 9)) +
  labs(x = "Occupation Size",
       y = "Occupation's Education Specificity")
```

Occupational mobility tends to decline with age, but age is a proxy for
the underlying mechanism rather than the mechanism itself.
@mincer1981labor show empirically that age effects on mobility largely
disappear once tenure is controlled, and attribute the residual to
specific human capital accumulation rather than aging per se: as workers
commit to an occupational path, the wedge between their current wage and
outside options widens, making mobility increasingly costly. Tenure is
their operational measure of that wedge, but the wedge is unobservable
in principle --- what they estimate is the hazard of separation as a
function of job duration, not specific capital directly. OES approaches
the same underlying quantity from the supply side: rather than inferring
specific capital from the duration of job spells, it reads off the
credential selectivity of the occupation from the equilibrium education
mix. A newly trained and a 20-year veteran physician occupy the same
high-OES occupation, and should be similarly immobile; a newly appointed
and a long-tenured senior manager occupy the same low-OES occupation and
should remain similarly mobile. Tenure varies within each pair but the
mobility prediction does not --- because what constrains mobility is the
credential selectivity of the occupation, not the worker's time in it.
This is a distinction tenure cannot identify as a stable property of the
occupation, but OES can.

The local/non-local partition has a natural falsification structure.
Each model is tested both where it should work and where it should not,
and the regime-specific predictions provide stronger evidence for the
causal interpretation of the underlying mechanisms than either
falsification alone would. A single confounder would have to operate
in opposite directions across regimes to account for both patterns;
multiple confounders would each have to align fortuitously with the
predicted cross-regime structure. Neither is impossible, but both are
less parsimonious than the theoretical mechanism itself. The design
does not eliminate the possibility of confounding, but it raises the
bar considerably: the model's validity rests not on the absence of
confounders but on the implausibility of a story that generates the
predicted pattern of opposing failures across regimes.

## Cost Metrics {#sec-cost-metrics}

The skill distance matrix $C_{\text{skill}}$ is constructed from 161 
O\*NET measures spanning skills, abilities, knowledge, and work 
activities, mapped to NOC 2021 occupations via a purpose-built 
crosswalk @onet_to_noc. An alternative approach would be to 
use OaSIS occupation-level skill profiles, but we construct our own 
crosswalk for a specific reason: falsifiability. The horse race 
requires that measurement error in the skill distance matrix be 
attributable to a known source --- namely, imprecision in the O\*NET 
to NOC 2021 mapping. OaSIS obscures the structure of that mapping, making 
it impossible to characterize where and how measurement error enters 
$C_{\text{skill}}$; our crosswalk makes this structure transparent 
and therefore makes the identification strategy falsifiable. Each NOC 
is represented as a weighted average of its constituent O\*NET 
occupations, with weights determined by the crosswalk mapping.  For more details on the mapping see Appendix A.

O\*NET provides two scales for each measure: Importance (how important the attribute is to the occupation) and Level (the degree to which the attribute is required). Both scales are standardized to a common 0--10 range and combined via geometric mean, yielding a single score that reflects both the relevance and the intensity of each attribute. The geometric mean is preferred over the arithmetic mean as it penalizes cases where one scale is near zero — an attribute that is important but required at negligible level, or required at high level but deemed unimportant, receives a lower composite score than one scoring moderately on both dimensions. The two scales are highly correlated ($\rho=.96$), so this choice is unlikely to be consequential for the resulting distance matrix but we note it for transparency.

The 161 O\*NET skill items are correlated, so raw Euclidean distance in the full item space implicitly upweights dimensions measured by many items over those measured by few. We apply PCA to the standardised profiles to remove this redundancy, then retain the number of components needed to preserve pairwise distance rank ordering: starting from $k = 1$, we increment until the Spearman rank correlation between pairwise distances in the $k$-dimensional PCA space and in the full 161-dimensional space exceeds 0.99. This criterion — rank preservation rather than variance explained — is chosen because the object of interest is the distance matrix itself, not the underlying skill dimensions. Twelve components satisfy it. The skill distance between any two occupations is then the Euclidean distance between their coordinates in this 12-dimensional space. An interactive heatmap of the resulting distance matrix is available in the [online companion.](https://rpmartin.github.io/not-one-thing-companion/#sec-skill-heatmap)

The binary switching cost matrix $C_{\text{binary}}$ assigns a cost of zero to 
remaining in the same occupation and one to any transition. Structurally, the 
labour market is comprised of as many distinct silos as there are occupations, 
each with no internal structure of any kind. $C_{\text{binary}}$ serves two 
roles in the analysis. First, it is the theoretical straw man in the global 
horse race: if occupational mobility costs are purely a function of whether a 
switch occurs rather than how far, then $C_{\text{binary}}$ should rationalize 
observed mobility patterns better than $C_{\text{skill}}$ or 
$C_{\text{hierarchy}}$. Second, $C_{\text{binary}}$ is best suited to settings 
where occupational distance does not order destinations, so a transition, when 
it occurs, carries the same cost regardless of where it lands. Such flatness 
plausibly suppresses mobility as well, though the horse race turns on 
destination structure rather than the move rate, which all three metrics 
accommodate. Where destinations are undifferentiated in this way, the 
structured metrics' additional geometry does not improve on $C_{\text{binary}}$. 
This second role motivates the Binary sub-regime hypothesis: where human 
capital is specific enough that no occupational geometry orders exits, binary 
is not a degenerate benchmark but a candidate description of the mobility 
geometry, tested against the structured metrics on equal footing. Having no
off-diagonal structure to allocate, binary has a single lever, the aggregate
stay/move split, so it behaves as a one-parameter stay/move model whose
fitted concentration encodes the mobility rate rather than a featureless
benchmark. @sec-local uses this characterization when comparing where the
three metrics peak.

The hierarchical distance matrix $C_{\text{hierarchy}}$ is derived directly from the NOC 2021 taxonomy, treating the occupational classification structure as a tree metric. Distance between any two occupations is defined as follows: zero if all five digits match (same unit group); one if the first four digits match (same minor group, different unit group); two if the first three digits match (same sub-major group, different minor group); three plus the absolute difference in TEER level if only the first digit matches (same broad occupational category, different sub-major group); and nine for any cross-category transition. The absolute TEER difference ranges from zero to five, so within-category distances range from three to eight, preserving a strict separation from the cross-category sentinel value of nine.  An interactive hierarchical distance heatmap of the resulting matrix is available in the [online companion.](https://rpmartin.github.io/not-one-thing-companion/#sec-hier-heatmap)

Occupation broad groups function as silos — bounded mobility clusters with limited entry points and internal progression along skill levels [@doeringer1971internal] — and TEER levels function as platforms within each silo, capturing the vertical skill stratification that governs upward movement. By construction, any between-platform move (minimum distance 4) is further than any within-platform move (maximum distance 3), and any cross-silo move (distance 9) is further than any within-silo move (maximum distance 8). The occupation broad group is a coarser unit than the occupational clusters described by @althauser1989, but the appropriate level of aggregation for the horse race: a finer partition would increase the preponderance of cross-category sentinel values in the hierarchical distance distribution, collapsing the metric's geometry toward binary and undermining its role as a distinct candidate in the horse race.

## Cost Metric Normalization

The three cost matrices are defined on different scales by construction: skill 
distances are continuous and spread over a wide range, hierarchical distances 
are discrete with a sentinel value of nine reserved for cross-category 
transitions, and binary distances take only the values zero and one. Because 
the cost matrix scale and $\varepsilon$ are not separately identified, the 
choice of anchor shifts where a metric's peak falls on the $\varepsilon$ axis 
but not the peak fit it attains; the winning metric is therefore invariant to 
normalization. Normalization matters instead for the legibility of the sweep 
plots: on a common $\varepsilon$ grid, matrices on different scales imply 
different effective $\varepsilon$, so the curves are not visually comparable 
unless first placed on a common scale. For the same reason we anchor at a 
substantively meaningful unit, so that position on the $\varepsilon$ axis 
remains interpretable.

We normalize skill and hierarchical distances to a common scale by anchoring 
both at a meaningful unit of hierarchical distance: distance 3, the maximum 
distance without changing platforms. Distance 3 falls at the fourth percentile 
of the hierarchical distribution, reflecting the preponderance of cross-silo 
pairs. The corresponding fourth percentile of the skill distance distribution 
is 7.145. Dividing hierarchical distances by 3 and skill distances by 7.145 
anchors both matrices at the same point in their respective distributions. 
Binary distances are not subject to this anchoring; their two-point support is 
shown for reference. @fig-cdfs gives the empirical CDFs before and after 
normalization. Normalization pins the skill and hierarchical distributions to 
the same point at the anchor, but their shapes remain distinct: hierarchical 
distance is discrete, with the small minority of within-platform pairs spread 
below the anchor and the bulk of pairs concentrated at the cross-category 
sentinel, while skill distance is continuous and dispersed across a wider 
range. This residual difference in shape is the separability the horse race 
exploits; normalization equalizes scale, not geometry.


```{r}
#| warning: false
#| echo: false
#| label: fig-cdfs
#| fig-cap: "Empirical CDFs of the three cost matrices before and after normalization; normalization places skill and hierarchical on a common scale for the shared $\\varepsilon$ sweeps without equalizing their shape."
cdfs
```

## Mobility Data {#sec-data}

The analysis uses the Census 2016--2021 longitudinal file, constructed 
by linking long-form respondents across the two census cycles via the 
Derived Record Depository (DRD), Statistics Canada's national 
longitudinal database of personal identifiers. Linkage is 
probabilistic; details of the matching keys and overall linkage rate 
will be verified upon data access.

[Linkage Details Here]

In the following analysis we make use of a \emph{measurement complete} 
subset of the linked data where we observe occupation, field of study, 
highest educational attainment, class of worker and age in both 2016 and 2021.

[Missingness Details Here]

Canada's NOC 2021 hierarchy nominally contains 516 unit groups, but 
the five Senior Manager NOCs (00011--00015) are aggregated into NOC 
00018 in the CIP-NOC table used to derive EOS and OES, leaving 512 
occupations. Eight unit groups have no admissible O*NET mapping and 
are excluded from all estimation samples, leaving 504 occupations in 
the analysis.

These 504 occupations are the universe over which the sub-regime
partition is defined: the OES ranking that fixes the three sub-regimes
(96 Binary, 86 Skill, 322 Hierarchy) is computed over all 504, and
@tbl-partition-crosstab and @fig-regimes are likewise on the full 504.
The mapping-quality exclusion enters one step later, at estimation
rather than partition. The bottom decile of the $H \times SS$ mapping
quality composite (Appendix A) flags 50 poorly mapped occupations, and
the local clean and non-local clean samples are built from transitions
among the remaining 454. A transition is dropped whenever either
endpoint is one of the 50, so the clean estimation samples run on the
454 cleanly mapped occupations while the partition they inherit was
cut on the full 504. % RICH VERIFY: 504-then-cut, 50 poorly mapped, 454 clean — confirm headcounts against pipeline.

There are two core subsamples used in both the local and non-local mobility analyses. Their role reverses depending on whether they serve as the primary identification sample or the falsification sample. The two samples are separated by the size of the human-capital increment a worker acquires between 2016 and 2021, as developed in @sec-cf. We operationalise the increment from the education record, the cross product of highest attainment and field of study, using the field-of-study classification's two-digit series as the unit of field distance.\[To be determined whether CIP in the linked file is harmonized or versioned.\] A worker acquires a large increment by reaching a higher level of attainment, or by changing field at the same level across two-digit series, which is retraining into an area where the original credential does not transfer. A worker whose education changes only within a two-digit series at the same level acquires a small increment and is grouped with the local sample.

Grouping the within-series same-level changes with local mobility is robust to a measurement ambiguity in the education record, for the reason given in @sec-cf: a second credential in a closely related field at the same level is a negligible increment whether or not it was genuinely earned, so the routing does not depend on resolving which occurred. Here we add only what the data section is positioned to say, about the composition of the routed cell. That composition tilts toward report instability in any case, since a redundant credential in a similar field is an unattractive investment: a worker seeking to build on an existing credential would typically pursue a higher level rather than a similar credential at the same level.

The \emph{local clean sample} comprises individuals whose education increment is small by this criterion, with no self-employment, no TEER--attainment mismatch, and an O*NET-to-NOC mapping quality score above the bottom decile. This is the sample where distance-based models should best rationalise mobility: human capital is stable over the inter-census period, the 2016 occupation is an informative anchor, and education acquired before 2016 is already reflected in where the worker starts. It is old news, not a useful predictor of subsequent movement.

The \emph{non-local clean sample} comprises individuals whose education increment is large by this criterion, subject to the same measurement restrictions as the local clean sample. This is the sample where the specificity regression should have the best fit: a large increment may open a wormhole to a distant occupation, the 2016 occupation is no longer an informative anchor, and education acquired within the inter-census window is fresh news, the active ingredient in long-distance jumps. In contrast, the local model will struggle to rationalize these transitions on the basis of distance.

In addition, for both models we investigate the effect of measurement error in occupation, when the data is otherwise appropriate for the model being fit. We assume that occupation measurement error is likely when either individuals are self-employed, or exhibit a TEER--attainment mismatch, as described in Appendix B. Self-employed workers frequently perform multiple roles simultaneously, making a single occupation an unreliable summary of their occupational position. The TEER--attainment correspondence is not incidental: occupation unit groups are defined in part by their typical education requirements,
so a reported occupation that is inconsistent with reported attainment is more likely to reflect a temporal lag — a credential acquired too recently to be reflected in the current occupation, or a foreign credential not yet formally recognized — than a stable and genuine labour market outcome. In either case the recorded occupation is an unreliable measure of the worker's actual occupational position at the time of the census. This occupation mismeasurement introduces noise into the measurement of distance regardless of metric, producing symmetric fit degradation.

For non-local mobility we can perform one further diagnostic: the effect of poor skill mapping from O\*NET to NOC, when the data is otherwise appropriate for the model being fit. Poor skill mapping is identified by the bottom decile of the O\*NET-to-NOC mapping quality composite. Poor mapping introduces noise specifically into $C_{\text{skill}}$ while leaving 
$C_{\text{hierarchy}}$ unaffected --- a metric-specific prediction 
that serves as a falsification of the mapping quality composite 
itself. This diagnostic is not available for local mobility. As noted
above, the local clean subset is built from transitions among the 454
cleanly mapped occupations, whereas the mapping quality subset is built
from transitions involving at least one of the 50 poorly mapped
occupations, so the two subsets have disjoint cells by construction.
Since $\Delta KL$ and $\Delta CWTVD$ require a common set of possible origin-destination pairs to be comparable, no valid comparison is possible between the local clean and mapping quality subsets.

## Local Model {#sec-local}

Mobility costs are latent: workers' switching costs are not observable 
in principle, so any distance matrix is necessarily a specification of 
an unobservable quantity. The three candidate matrices --- skill, 
hierarchical, and binary --- represent distinct hypotheses about the 
structure of those latent costs.

The entropy-regularized optimal transport problem minimizes the following Lagrangian, where the first term is the transport cost, the second is the entropic regularization penalizing deterministic transport plans, and the third and fourth enforce the marginal constraints $\sum_j P_{ij} = a_i$ and $\sum_i P_{ij} = b_j$ via dual variables $f_i$ and
$g_j$.

$$\mathcal{L} = \sum_{ij} P_{ij} C_{ij} + \varepsilon \sum_{ij} P_{ij}(\log P_{ij} - 1) + \sum_i f_i(a_i - \sum_j P_{ij}) + \sum_j g_j(b_j - \sum_i P_{ij})$$

where $a_i$ and $b_j$ are the empirical marginal distributions in 2016 and 2021 respectively. 

The first-order conditions yield:
\begin{equation}
\log P_{ij} = \frac{f_i}{\varepsilon} + \frac{g_j}{\varepsilon} - 
\frac{C_{ij}}{\varepsilon}
\end{equation}
Exponentiating, the optimal transport plan takes the form 
$P^\star_{ij} = u_i \, e^{-C_{ij}/\varepsilon} \, v_j$, where 
$u_i = e^{f_i/\varepsilon}$ and $v_j = e^{g_j/\varepsilon}$ are 
scaling vectors. Strict convexity of the regularized objective 
guarantees a unique solution. The solution is therefore a row- and 
column-rescaling of the exponential kernel $K_{ij} = e^{-C_{ij}/\varepsilon}$, where the Sinkhorn algorithm recovers it by alternating row and column 
normalization until both marginal constraints are satisfied to within a specified tolerance. All computations use a log-domain Sinkhorn solver 
for numerical stability, implemented in the \texttt{safesink} R package [@safesink].

In log space, the optimal transport plan decomposes into additive origin effects
($\log u_i$), additive destination effects ($\log v_j$), and a dyadic cost term
($-C_{ij}/\varepsilon$). The scaling vectors correspond to Lagrange multipliers on
the marginal constraints and adjust to ensure the implied joint distribution matches
the observed origin and destination occupation stocks, absorbing any components of
the data that operate at the level of origins or destinations alone given the
specified cost structure. The cost matrix captures the residual structure that
depends on origin--destination pairing. The horse race asks which specification of
that dyadic structure best rationalizes observed transitions.

The gravity model [@anderson_vanwincoop_2003; @silva2006log] estimates origin and destination fixed effects jointly with the cost parameters, and that joint estimation is precisely the problem in our setting. The fixed effects absorb variation in mobility rates that may reflect true cost structure, which lets a misspecified cost matrix achieve good fit by compensating through the fixed effects. Using a gravity model would be akin to giving the slower horses performance-enhancing drugs in inverse proportion to their natural speed: the race is closer than it should be, and the outcome can be reversed entirely. EROT fixes the margins at observed values and yields a fair race instead. Each cost matrix must earn its fit in the residual dyadic structure, with no compensation for its own misspecification, which separates the question of where mobility flows, shaped by the cost matrix and the focus of this paper, from the question of why workers move. Wages, amenities, and other occupation-level incentives operate through the margins and are deliberately outside the scope of the identification strategy.




The regularization parameter $\varepsilon$ has a natural interpretation
in the random utility framework. The optimal transport plan $P^\star_{ij} = u_i \, e^{-C_{ij}/\varepsilon} \, v_j$ is the static equilibrium of a discrete choice model in which workers draw idiosyncratic utility shocks scaled by $\varepsilon$ over destination occupations. Low values of $\varepsilon$ concentrate the transport plan on cost-minimizing assignments, approaching a deterministic reallocation in the limit; high values diffuse mobility across destinations, approaching the independence baseline $P_{ij} = a_i b_j$ as $\varepsilon \to \infty$. The RUM analogy governs the interior structure of the transport plan --- how probability mass is allocated across origin-destination pairs --- but does not extend to the margins. In a standard logit model the choice probabilities are outcomes of utility maximization with no constraint on their aggregates; in EROT the origin and destination occupation distributions are hard constraints that the transport plan must exactly reproduce. This distinction is methodologically consequential: because the margins cannot adjust to absorb cost misspecification, a poorly chosen cost matrix cannot hide behind fixed effects or scale recalibration --- its misfit appears directly in the objective. 

A critical identification constraint follows directly from the form of 
the optimal transport plan: only the ratio $C_{ij}/\varepsilon$ enters 
$P^\star$. The scale of the cost matrix and the value of $\varepsilon$ 
are therefore not separately identified, and results reported at a 
single value of $\varepsilon$ are uninterpretable without an implicit 
normalization assumption. We evaluate across a wide log-spaced grid of 
$\varepsilon$ values from 0.1 to 1000, deliberately broader than the range 
expected to be informative, to ensure the informative region is fully 
contained within the evaluated range and its boundaries are visually apparent.

The horse race is evaluated using two complementary fit metrics, both
normalized relative to the independence baseline $P^{\text{ind}}_{ij} =
a_i b_j$, so that 0 indicates no improvement over independence, 1
indicates perfect fit, and negative values indicate a transport plan that fits observed mobility worse than independence. 

The first metric is relative KL divergence improvement:

$$\Delta KL = \frac{KL(P^{\text{obs}} \parallel P^{\text{ind}}) - KL(P^{\text{obs}} \parallel \hat{P})}{KL(P^{\text{obs}} \parallel P^{\text{ind}})}$$

where $KL(P \parallel Q) = \sum_{ij} P_{ij} \log(P_{ij} / Q_{ij})$.
This measures the share of the divergence between observed transitions
and independence that is explained by the fitted transport plan. KL
divergence is sensitive to probability mass wherever it is misallocated, but is geometry-blind: misallocating mass to a nearby occupation is penalized identically to misallocating it to a distant one.

The second metric is relative Cost-Weighted Total Variation Distance (CWTVD) improvement:

$$\Delta CWTVD = \frac{CWTVD(P^{\text{obs}}, P^{\text{ind}}) - CWTVD(P^{\text{obs}}, \hat{P})}{CWTVD(P^{\text{obs}}, P^{\text{ind}})}$$

where $CWTVD(P^{\text{obs}}, \hat{P}) = \sum_{ij}
\frac{|P^{\text{obs}}_{ij} - \hat{P}_{ij}|}{2} \cdot C_{ij}$. 

CWTVD is a cost-weighted extension of total variation distance: the 
classical TVD between discrete distributions, 
$\frac{1}{2}\sum_{ij}|P_{ij} - Q_{ij}|$, is extended by weighting 
each absolute residual by the corresponding entry of a cost matrix, 
penalizing probability mass misallocated to geometrically distant 
origin-destination pairs more heavily than mass misallocated to nearby 
ones. The scoring cost matrix $C$ used in CWTVD is the arithmetic mean of the three candidate normalized cost matrices: $C_{\text{skill}}$, $C_{\text{hierarchy}}$, and $C_{\text{binary}}$. The ideal scoring matrix would be the latent cost matrix itself, but were that available there would be no horse race to run. The arithmetic mean is the natural neutral alternative: no candidate geometry is privileged, since each cost matrix enters the blend with equal weight. The two geometrically informative matrices are genuinely separable --- the cell-level correlation between $C_{\text{skill}}$ and $C_{\text{hierarchy}}$ is 0.3 --- so the blend is not simply collapsing two near-identical matrices into one.

We adjudicate the horse race on two criteria, and a metric wins only by
leading on both. The first is peak relative KL improvement. For each
candidate cost matrix we maximise $\Delta KL$ over the $\varepsilon$
grid, and the reason for using it is that the EROT plan
$P^\star_{ij} = u_i\, e^{-C_{ij}/\varepsilon}\, v_j$ is the equilibrium of
a multinomial-logit discrete-choice model. Comparing candidates by their
maximised likelihood is then identical to comparing their peak relative
KL improvement, so peak $\Delta KL$ is maximum-likelihood model
selection rather than a fit heuristic. The second criterion is peak
relative CWTVD improvement over the same grid. KL is blind to the
geometry of the outcome space, scoring a unit of misallocated mass the
same whether it lands on a near pair or a far one. CWTVD weights each
misallocation by distance, and across mixed data-generating processes it
separates the two correlated informative geometries,
$\mathrm{corr}(C_{\mathrm{skill}}, C_{\mathrm{hierarchy}}) = 0.3$, more
sharply than KL does, with the gap between them tracking how far the true
mixture sits from balance. The [online companion](https://rpmartin.github.io/not-one-thing-companion) documents this directly.
 
The two justifications are independent. KL carries the likelihood
interpretation but cannot see geometry; CWTVD sees geometry but carries no
likelihood interpretation. Neither contains the other, so we hold them on
equal footing rather than naming one primary, and a win must satisfy both.
 
The case for CWTVD differs from the case for KL in what the estimation data
can and cannot show. The
resolving-power property is established under known data-generating
processes, as one establishes consistency or test power for an estimator
rather than reading it off the sample the estimator is later applied to.
It can be established only that way. The property is a claim about the
mapping from a known true mixture to the gap CWTVD reports between metrics,
and checking that mapping requires holding the truth fixed and reading off
the measurement, which presupposes knowing the truth. The estimation
sample is a single transition matrix drawn from one unknown mixture, so it
yields one gap and no reference mixture against which that gap's size could
be judged. No quantity in this sample, or in any single observed labour
market, can confirm or disconfirm the property. The simulation is not
preliminary evidence awaiting empirical confirmation but the only support
the claim admits. The simulation validates the instrument and the
estimation sample takes the measurement, and we make no prediction that
the Canadian data will confirm CWTVD's advantage. The data answers a
different question, which geometry governs Canadian mobility, and the
resolving-power property is what licenses trusting CWTVD's answer to that
question when it parts from KL's.
 
We compare peaks rather than improvement at any single $\varepsilon$
because of the same non-identification that motivated the grid above:
only the ratio $C_{ij}/\varepsilon$ enters $P^\star$, so a metric's
improvement at a fixed $\varepsilon$ has no interpretation without an
implicit normalization, and the peak is the scale-free summary that
removes the ambiguity. Each candidate traces an
interior maximum over the grid: at low $\varepsilon$ the plan
over-concentrates on the diagonal relative to the observed move rate, and
as $\varepsilon \to \infty$ it diffuses to the independence baseline
$P_{ij} = a_i b_j$, where improvement returns to zero. The grid from $0.1$
to $1000$ is wide enough that all three candidates peak in its interior on
both criteria, with the uninformative tails visible at each end. Binary
peaks in the interior too. As the one-parameter stay/move model described
in @sec-cost-metrics, its only lever is the aggregate stay/move split, so
its peak sits at the $\varepsilon$ where the diffusion of mass off the
diagonal matches the empirical move rate, and it peaks earlier than skill
and hierarchy, which keep improving as falling $\varepsilon$ sharpens an
off-diagonal allocation that pays off.
 
The winner of the race is the cost matrix whose peak exceeds the
runner-up's by at least 10\% on both criteria at once. Requiring both is
more selective than crowning the leader on either alone. A metric wins only
if the likelihood endorses its fit and the geometry-aware criterion
certifies it geometrically sound, so a plan that puts mass in roughly the
right cells but at the wrong distances, scoring high on $\Delta KL$ and low
on $\Delta CWTVD$, does not win, nor does its mirror image. Three outcomes
follow, and we report whichever obtains:
 
\begin{enumerate}
  \item \textbf{Decisive win.} One metric leads the runner-up by 10\%
  on both criteria. The race resolves in its favour.
  \item \textbf{Split.} Different metrics lead on the two criteria, or one
  metric leads on both but clears 10\% on only one. The race is
  unresolved. We report the split by name, stating which geometry the
  likelihood favours and which the geometry-aware criterion favours,
  because that disagreement is itself a description of how observed
  mobility departs from each candidate.
  \item \textbf{No signal.} No metric leads by 10\% on either criterion.
  No candidate geometry rationalizes the transitions distinctly better
  than the others.
\end{enumerate}
 
The 10\% threshold is an effect-size floor, set in advance. We treat two
peak improvements differing by less than ten percentage points as
substantively equivalent regardless of whether the difference is
statistically distinguishable, on the grounds that a smaller difference
does not warrant naming one geometry the governor of mobility in that
regime. At the sample sizes the linked census file provides, even the
smallest sub-regime holds thousands of transitions, so sampling noise in
the peak gap is well below this floor and the binding constraint is the
substantive one rather than the statistical one. The one regime where this
ordering could fail is the high-OES Binary sub-regime, whose low mobility
leaves its off-diagonal cells sparse. We therefore confirm at access, by
simulating a balanced skill/hierarchy mixture at that regime's transition
count, that a balanced mixture rarely produces a peak gap above 10\%; if it
does, the Binary regime's race is noise-limited and we interpret it as
such. This is the only place the floor's statistical adequacy is in
question, and the only simulation the threshold requires.
 
A split is an informative no-decision rather than a coin to be flipped. A
disagreement between a likelihood-grounded and a geometry-aware criterion
localizes how a metric misfits, and breaking it in favour of one criterion
by fiat would discard exactly the signal the two criteria were chosen to
expose.

### Out-of-Regime Falsification {#sec-local-falsification}


The out-of-regime criterion is stated against independence, on the
geometry-aware measure. In regime, the predicted winning metric,
$C_{\mathrm{skill}}$ in the Skill sub-regime and $C_{\mathrm{hierarchy}}$
in the Hierarchy sub-regime, peaks above independence on both $\Delta KL$
and $\Delta CWTVD$, since its geometry rationalizes the transitions and
improvement over independence is what the horse race measures. Out of
regime, where a new highest attainment resets the worker's position and
renders the 2016 occupation uninformative, that metric is predicted to fall
below independence on $\Delta CWTVD$.
 
The out-of-regime criterion rests on $\Delta CWTVD$ alone, and the reason
is that $\Delta KL$ cannot register the failure being tested. The
falsification asks whether the predicted metric has made the wrong
geometric bet, placing mass at the wrong distances. $\Delta CWTVD$ weights
each misallocation by distance and so penalises exactly this, falling below
independence when a short-move metric meets a movement-heavy matrix.
$\Delta KL$ is blind to geometry and scores any plan that captures the
diagonal stay/move structure, which every metric shares. A wrong-geometry
metric therefore collects partial KL credit for the diagonal even while its
off-diagonal bet is wrong, and its peak $\Delta KL$ can sit slightly above
independence rather than below it. This is visible in the simulations: in
the online companion, under a binary data-generating process the skill and
hierarchy metrics peak below independence on $\Delta CWTVD$ but rise to a
small positive peak on $\Delta KL$ at moderate $\varepsilon$. Requiring the
geometry-blind measure to confirm a geometric failure would ask the wrong
instrument to register it. We therefore state the out-of-regime criterion
on $\Delta CWTVD$, the measure built to see what the falsification is about.
 
Setting the line at independence rather than at binary discharges the
conditionality that a binary benchmark would leave open. A wrongly
predicting metric always sits below binary out of regime, but whether its
$\Delta CWTVD$ also falls below independence depends on the non-local
transition matrix being movement-heavy: only when most credential-changers
move, and move far, does a short-move metric place enough mass on vacated
near pairs to fit worse than independence. We predict the non-local matrix
is movement-heavy, because the wormhole mechanism is precisely the claim
that a new credential drives long discontinuous jumps. The
below-independence criterion therefore tests two things at once. Its
crossing requires both that the predicted metric makes the wrong geometric
bet out of regime and that the non-local matrix is movement-heavy as the
mechanism predicts, so a metric whose $\Delta CWTVD$ falls below zero
jointly falsifies the local model out of regime and confirms the long-jump
prediction. A metric that stayed above independence out of regime would
either be governing where it should not or would reveal a non-local matrix
lighter than the mechanism requires, and either reading counts against the
design.
 
Independence is a fixed point, so the criterion is unambiguous. The
predicted metric's $\Delta CWTVD$ peak is either below zero or it is not,
with no margin to calibrate and no grey zone to adjudicate. This is the
respect in which the falsification differs from the race. The race resolves
a close contest between two correlated geometries by asking whether one
peak exceeds the other by 10\%, a margin set because the contestants are
near. The falsification asks whether a single metric crosses a fixed line,
which it either does or does not.
 
Peaking above independence in regime is a weaker bar than winning the
race, and the gap between them is real. A metric clears the in-regime side
of the falsification by improving on independence at all; it wins the race
only by leading every rival by 10\% on both measures. Both skill and
hierarchy peak well above independence in either of their sub-regimes,
since they are correlated and each partially rationalizes the other's
transitions, yet only one wins. Improving on independence is necessary for
a geometry to be governing at all; leading by 10\% on both measures is what
crowns it.
 
The mechanism that drives the predicted metric's $\Delta CWTVD$ below
independence is the same one that makes binary a useful point of
comparison. All three cost matrices assign zero self-distance,
$C_{ii} = 0$, so all three concentrate mass on the diagonal and capture the
dominant stay/move feature of the transition matrix. That diagonal
contribution is common to every plan, and what distinguishes the metrics is
the off-diagonal allocation. Binary makes no off-diagonal bet, since every
switch is equidistant under it, so it cannot be wrong about where movers
go. Skill and hierarchy do make that bet, and out of regime it is the wrong
one, because they place mass on geometrically proximate pairs that wormhole
movers avoid while the data concentrates on far pairs. A metric predicting
short moves into a movement-heavy matrix withholds mass from exactly where
the data sits, and $\Delta CWTVD$ penalises that misallocation by distance,
driving it below independence. The signature of this failure on
$\Delta CWTVD$ is a curve that rises monotonically toward the independence
baseline from below without crossing it, improving as $\varepsilon$ grows
only because diffusion toward independence is the one thing that helps a
metric whose geometry points the wrong way. The online companion shows this
under a binary data-generating process, where skill and hierarchy stay
below zero on $\Delta CWTVD$ across the grid while their $\Delta KL$, for
the diagonal-credit reason given above, rises instead to a small positive
peak at moderate $\varepsilon$ --- which is why the out-of-regime criterion
rests on $\Delta CWTVD$ rather than on both measures.
 
The sub-regime models are falsified the same way, with one exception. In
regime each sub-regime is expected to produce a metric that peaks above
independence on both measures, and out of regime that metric's $\Delta
CWTVD$ peak is predicted to fall below independence. We apply this to the
Skill and Hierarchy sub-regimes only. The Binary sub-regime is exempt from
the distance-based out-of-regime test for a reason internal to its own
hypothesis, and the exemption does not depend on what its out-of-regime
cell turns out to contain. Its in-regime claim is that no distance geometry
governs, so it never generates a prediction that distance ceases to govern
out of regime; there is no distance bet to be caught making wrongly, with
or without movers in the cell. Separately, the lock-in mechanism that
defines the sub-regime predicts that cell is near-empty, because workers
already in a high-OES occupation rarely acquire a new highest attainment,
since occupation-specific human capital is reached through a single
credentialing pipeline and confers no incentive to re-credential. A
non-empty cell therefore falsifies the lock-in premise without thereby
reinstating a distance test the hypothesis never licensed: the
distance-based leg stays inapplicable, and the non-emptiness is read as
evidence against lock-in in its own right. % RICH DECISION POINT: conservative branch — non-empty cell falsifies lock-in but does NOT reopen the out-of-regime distance test. Alternative: let a non-empty cell reinstate the Skill/Hierarchy-style below-independence test on those movers. Swap if you want the riskier pre-registration.
The out-of-regime content of the Binary hypothesis otherwise lies on the
entry side and is carried by the destination-OES coefficient in the
non-local specificity regression rather than by a horse race.

## Non-Local Model {#sec-non_local}

The non-local model estimates the conditional-on-positive relationship between how far a worker travels through occupation space and the specificity measures EOS and OES. We measure that distance two ways, with different estimators and complementary roles. The primary outcome is $\log(d_{ij})$, the log skill distance between 2016 and 2021 occupations, estimated by OLS. The second is a binary indicator for whether the move crosses a occupation broad group — a silo — estimated by a linear probability model (LPM). This is the operational form of the hierarchical metric for non-local jumps, where the cross-silo margin is expected to carry essentially all of the hierarchical signal among credential-changers and the within-silo gradient is too sparse to estimate as a graded distance. The two outcomes are not interchangeable robustness checks but draw on disjoint inputs: the skill outcome depends on the O\*NET-to-NOC crosswalk, whereas the cross-silo outcome depends only on the NOC taxonomy and is immune to crosswalk error. Concordant specificity effects across the two therefore rule out that the wormhole result is an artifact of the skill mapping — the input to which this paper is most exposed.

The theoretical mechanism is symmetric: a wormhole jump requires both an entrance and an exit. A specific education (high EOS) channels workers toward a narrow set of occupations that would otherwise be unreachable; a specific destination occupation (high OES) draws selectively from particular credential pipelines. Neither alone is sufficient for a discontinuous jump — a specific education without a specific destination lands the worker somewhere generic (e.g. Optometry graduate *not* working as Optometrist), and a specific destination without a specific education implies arrival via general credentials (e.g. Ferdinand Waldo Demara). The long-distance jumps that characterise wormhole travel occur only when both ends are specific simultaneously. A bilinear $\text{EOS}\times\text{OES}$ term fails to encode this: with mean-centred measures the product is equally large and positive when both lie well below average, spuriously predicting long jumps for the least-specific movers. We therefore use a rectified interaction, $\max(\text{EOS},0)\cdot\max(\text{OES},0)$, which is non-zero only when both specificities are strictly positive, so the multiplicative term fires in the high–high region alone; this rectified interaction is the theoretically primary coefficient. We estimate three specifications — two univariate models and a joint model with both main effects and the rectified interaction — and report the Spearman correlation between EOS and OES as a diagnostic.

Whether the joint model can separately identify entrance and exit intensity depends on the worker-level correlation between EOS and OES, which governs how densely the high–high region is populated — the only region from which the rectified interaction is identified. When the two measures are highly correlated, that region collapses toward the diagonal and $\max(\text{EOS},0)\cdot\max(\text{OES},0)$ approximates a rectified quadratic in a single specificity index, so the main effects and the interaction can no longer be cleanly decomposed; the same correlation that populates the region enough to identify the interaction also blurs the separation of the two main effects within it. The univariate models remain interpretable regardless of the correlation — each estimates the reduced-form association between a single specificity measure and distance traveled. A low correlation supports a structural reading of the joint model's coefficients; a high correlation limits interpretation to the univariate models, with the joint model reported for completeness. The correlation is itself informative about the structure of education-occupation matching in the Canadian labour market, independent of its implications for identification.
Let $y_{ij}$ denote the outcome — $\log(d_{ij})$ in skill space, estimated by OLS, or the cross-silo indicator, estimated by LPM — and let $\bar{d}_j$ be the destination's mean inbound distance under the independence baseline (its average distance from a randomly drawn origin; for the cross-silo outcome, the baseline probability that a random origin lies outside $j$'s broad group). The remoteness control is defined in the same space as the outcome it accompanies, so it swaps with the outcome: skill-space mean inbound distance for the OLS specification, baseline cross-silo probability for the LPM. This matters for the immunity claim above. The cross-silo specification depends on the NOC taxonomy through both its outcome and its control and so touches the O\*NET crosswalk at no point, whereas the skill specification is crosswalk-dependent on both sides.

\begin{equation}
y_{ij} = \alpha + \beta_1\text{EOS} + \gamma\bar{d}j + \delta_i + \varepsilon{ij}
\label{eq-eos}
\end{equation}
\begin{equation}
y_{ij} = \alpha + \beta_2\text{OES} + \gamma\bar{d}j + \delta_i + \varepsilon{ij}
\label{eq-oes}
\end{equation}
\begin{equation}
y_{ij} = \alpha + \beta_3\text{EOS} + \beta_4\text{OES} + \beta_5\left[\max(\text{EOS},0)\cdot\max(\text{OES},0)\right] + \gamma\bar{d}j + \delta_i + \varepsilon{ij}
\label{eq-joint}
\end{equation}

The specifications include origin occupation fixed effects $\delta_i$ and a destination remoteness control $\bar{d}_j$, but not destination fixed effects. Destination fixed effects are infeasible: OES is a property of the destination occupation and is collinear with a full set of destination dummies, which would absorb the very variation OES is meant to capture. The remoteness control instead nets out a destination's intrinsic "farness" — an occupation lying far from all others in skill space, or in a small broad group that most origins fall outside, registers long jumps mechanically, with no credential mechanism involved. Conditioning on $\bar{d}_j$ ensures the OES coefficient reflects credential-gated channeling rather than baseline remoteness. This is the conservative reading: if remoteness is itself part of how specificity operates, the control absorbs some of the mechanism. Because OES and $\bar{d}_j$ are both destination-level, they draw on the same between-destination variation, so a strong association between specificity and remoteness will widen the OES standard error. Origin remoteness needs no separate term — $\delta_i$ absorbs it with every other origin-level determinant of distance.

Prior to estimation, EOS and OES are standardized to mean zero and unit variance using the full linked sample. This places coefficients on a common scale within and across the four subsamples, making the estimates interpretable as a comparison of effect sizes and the degradation pattern across samples interpretable as a comparison of fit rather than of variable scale. Because the measures are standardized on the full sample, the kink in the rectified interaction sits at the common zero threshold in every subsample; it is fixed rather than estimated, which preserves standard inference. Standard errors will be clustered on education and destination occupation.

All specificity coefficients are expected to be non-negative, with the rectified interaction $\beta_5$ the primary prediction. Under strict complementarity the main effects $\beta_3$ and $\beta_4$ are expected to be weak relative to $\beta_5$: a specific education whose realised destination is generic, or a general education reaching a specific destination, should travel little farther than a wholly unspecific move, so large main effects would instead indicate that specificity lengthens jumps additively rather than only in combination. The remoteness coefficient $\gamma$ is expected positive.

Age is excluded from the specification. The well-documented decline in occupational mobility with age [@topel1992job] is, for jump distance, a life-cycle composition effect that operates through the 2016 origin occupation: workers who acquire a new attainment later in life tend to start from a higher skill and hierarchy position than those completing the same credential early, so they sit closer to the credential's destination and travel a shorter measured distance. That channel is a property of where a worker starts, not of the wormhole mechanism, and the origin fixed effects $\delta_i$ absorb it along with every other origin-level determinant of distance. A separate age term is therefore neither needed nor clean: conditional on origin it carries no signal the mechanism predicts, and because age plausibly correlates with the specificity measures through selection into particular credential types, including it risks partialling out legitimate specificity variation. We make no prediction about an age coefficient.

The model is evaluated across four samples as described in @sec-data, with results reported for both the skill and cross-silo outcomes. The within-$R^2$ --- the share of variance in the outcome explained by the specificity measures after partialling out origin fixed effects and destination remoteness --- is not expected to be large in absolute terms: transition distance is inherently noisy and EOS and OES are aggregate measures that capture only the systematic component of specificity-driven mobility. The relevant comparison is the ordering of within-$R^2$ across samples, not its absolute level.

The positive distance restriction removes many apprenticeship completers, since NOC makes no distinction between apprentices and journeypeople of the same occupation --- those already working in the target trade in 2016 show zero measured distance upon completion and are excluded by construction. Rather than threatening identification, this restriction improves measurement validity: the excluded zeros reflect occupation coarseness rather than true immobility, and their inclusion would introduce noise into the outcome variable. Workers who complete an entire apprenticeship program within the five-year window from a different 2016 occupation remain in the sample, and for these workers the wormhole mechanism is cleanly observable. The limitation is one of external validity rather than internal validity: estimates do not generalise to within-trade apprenticeship completers, whose human capital increment is real but invisible to NOC.


### Out-of-Regime Falsification

The non-local model is falsified in the local clean sample, where the highest
education attainment is unchanged. The criterion is the cross-sample degradation of the specificity coefficients and the within-$R^2$, not their vanishing in absolute terms. Because EOS and OES are standardized on the full linked sample, the $\beta$ estimates and the within-$R^2$ are directly comparable across subsamples, and the prediction is that both collapse from the non-local to the local sample. We do not require the local coefficients to be exactly zero: a specific education can compress local move distances weakly through its main effect even absent a new credential, so the claim is degradation toward zero relative to the non-local estimates, consistent with the ordering-not-level standard applied to the within-$R^2$ throughout.

The collapse has two sources, and both are consistent with the wormhole
mechanism. Without a new highest attainment the wormhole channel is shut: the
worker's education is already reflected in their 2016 occupation, the starting
point of any local move, so it carries no additional predictive content for that
move, whereas the non-local education is acquired within the window and does its
repositioning where it can be observed. Second, high-OES destinations are
credential-gated by construction and are therefore largely unreachable without a
new attainment, so OES variation is truncated in the local sample and a
near-zero coefficient on destination OES there reflects range restriction as
much as a switched-off mechanism. Neither reading threatens the falsification. A
spurious EOS/OES--distance relationship---one driven by a confound rather than
by credential-contingent enablement---would appear in both samples, since the
confound would not know which workers had changed education; its appearance in
neither the local coefficients nor anywhere a confound could hide is what marks
the relationship as credential-contingent rather than mechanical.

Two features of the local sample bear on its standing as a falsification sample.
It is large---most workers do not acquire a new highest attainment over the
inter-census period---so the null it delivers is precise rather than
underpowered, a well-powered absence of relationship rather than silence. It is
also subject to a modest contamination, because the partition variable is only a
proxy for the mechanism: what matters is whether the 2016 occupation is
informative, and "no inter-census change in attainment" stands in for that.
The proxy errs for workers who earned their highest attainment shortly before the
2016 reference week---formally local, but with a transitional rather than
credential-matched 2016 occupation, and so behaving like non-local movers whose
jump straddles the census boundary. The leakage is small: the great majority of
the local sample earned their attainment well before 2016, and two filters
already remove the clearest cases---the TEER--attainment consistency filter
screens recent graduates in credential-mismatched jobs, and the
measurement-complete requirement excludes the high-EOS graduates of intensive
programs, who disproportionately lack a 2016 occupation at all. What leakage
remains biases toward the null: leaked movers import wormhole behaviour into the
local sample, degrading the horse-race fit and injecting specificity into the
regression, both of which work against the predicted dissociation rather than
producing it. This is the final reason the criterion is stated as degradation
and not as an absolute zero, which would be hostage to precisely this
contamination.

## Results \[PLACEHOLDER --- awaiting data\]  {#sec-res}

### Local

\[KL and CWTVD improvement figures by cost metric, $\varepsilon$ and
subsample\]

\[KL and CWTVD improvement by 3 OES sub regimes, cost metric, $\varepsilon$ and
subsample\]

\[KL and CWTVD improvement by 3 TEER sub regimes, cost metric, $\varepsilon$ and
subsample\]

### Non-local

\[tie-fighter plot for the $\beta$\]

\[ table for the $R^2$\]

## Limitations

### Local Mobility

Two sample restriction mechanisms introduce a potential selection 
concern for local mobility. Non-linkage and non-observability of 
occupation are both disproportionately concentrated among 
low-attachment workers --- individuals with casual, seasonal, or 
intermittent employment whose identifying information is also less 
stable across census cycles --- so the two filters largely co-select 
the same low-skill population rather than introducing independent 
exclusions. This shared exclusion affects the representativeness of 
population-level transition estimates (external validity) but does not 
threaten the internal validity of the horse race. For the horse race
results to be biased, the excluded workers' transitions would need to be 
systematically more or less rationalizable by a particular cost 
matrix than those of their observed counterparts in the same region 
of occupational space. No such mechanism is apparent: exclusion is 
driven by attachment instability and identifier churn, neither of 
which has a clear relationship to how well transitions conform to 
any particular cost structure conditional on occupation. The residual 
unlinked population among otherwise-observable workers --- recent 
immigrants, young workers, and high-movers --- similarly lacks any 
obvious occupational-space concentration that would differentially 
favour or penalize a particular metric.

A notable omitted factor for local mobility  is \emph{path-specific} wage differentials. The additively separable component of wage gains --- recoverable from origin and destination wage levels alone --- is absorbed by the 
margins insofar as high-wage destinations attract larger inflows and 
low-wage origins larger outflows. The residual dyadic component, 
reflecting transition-specific wage premia and discounts beyond what 
origin and destination wage levels predict, is an omitted factor. 
Short-distance transitions are within a similar occupational 
neighbourhood, reducing employer uncertainty about candidate fit and 
productivity; long-distance transitions cross larger skill or 
hierarchical gaps, and employers may discount wage offers to 
compensate for that uncertainty, or workers may accept lower wages 
to gain entry to a distant occupation.

Thus, path-specific wage differentials are better understood as an 
alternative mechanism, rather than a confounder. If employers use the same metric that governs mobility to assess candidate fit uncertainty --- discounting wages for transitions that are long under the winning metric --- that discount reinforces worker selection toward shorter transitions under 
the same metric. Both channels predict identical patterns of 
transitions and are equally consistent with the winning metric being 
the correct description of the mobility geometry. The horse race 
identifies which distance geometry best rationalizes observed 
transitions in a reduced-form sense, but cannot separately attribute 
that rationalization to worker or employer preferences regarding 
distance.

### Non-Local Mobility

The non-local partition is restricted to individuals who acquire a new highest educational attainment between 2016 and 2021 and report an occupation in both periods. A 2016 occupation is observed if and only if the individual was employed between May 1st and 7th --- the 2016 census reference week. For the 2015/2016 school year, roughly half of 18-29 year old students reported concurrent work, with an (unweighted) ten percentage point decrease for university students, consistent with more intensive programs suppressing concurrent employment.^[source: Statistics Canada table 37-10-0104-01.] The Census reference week provides a partial offset --- falling after the typical post-secondary academic year ends, it captures individuals who have recently finished exams and may have already begun summer employment --- but this offset is weaker for intensive professional programs that are either year-round or have students for whom summer employment is atypical. The resulting truncation is not random: missingness is concentrated at the top of the EOS distribution, precisely where the wormhole mechanism is strongest and jumps are longest. Thus estimates are attenuated toward zero and should be interpreted as lower bounds on the true relationship, given that truncation is concentrated precisely where the signal is strongest. Notably, this missingness pattern is not purely a limitation: disproportionate absence of 2016 occupations among high-EOS educations is itself consistent with the wormhole mechanism, since program intensity and specificity are precisely what suppresses concurrent employment. In the online companion we plot EOS against the share of workers missing a 2016 occupation by education type, quantifying the magnitude of this truncation.

```{r}
# #| label: tbl-working-while-studying
# #| tbl-cap: "Share of students employed during program, by age and institution type"
# 
#  tbbl <- cansim::get_cansim("3710010401")|>
#    janitor::clean_names()|>
#    mutate(age = str_extract(age, "\\d+"))|>
#    filter(ref_date=="2015/2016",
#           type_of_institution_attended %in% c("College","University"),
#           age %in% 18:30)|>
#    select(age, type_of_institution_attended, value)|>
#    pivot_wider(id_cols = age, names_from = type_of_institution_attended, values_from = value)
# 
#  tbbl_mean <- tbbl|>
#        summarise(age = "Unweighted Mean", across(c(College, University), mean))
# 
#  bind_rows(tbbl, tbbl_mean)|>
#    kbl(
#      digits = 1,
#      col.names = c("Age", "College (%)", "University (%)"),
#      booktabs = TRUE,
#      linesep = ""
#    ) |>
#    kable_styling(latex_options = c("HOLD_position")) |>
#   footnote(
#    general = "Statistics Canada Table 37-10-0104-01, 2015/2016.",
#    general_title = "Source:",
#    threeparttable = TRUE,
#    escape = FALSE
#  )|>
#    row_spec(nrow(tbbl)+1, bold = TRUE)  # bold mean row
```


The specificity regression faces an analogous identification 
challenge to the horse race. EOS and OES are observed and enter the 
regression, but cannot distinguish between two mechanisms that 
predict identical patterns. A specific education opens doors to 
particular distant occupations that a general education does not --- 
the human capital match is narrow, so only a subset of distant 
destinations become reachable following attainment (enablement). At 
the same time, workers with specific educations may face stronger 
incentives to seek out distant occupations where those skills are 
scarce and rewarded, making long-distance moves financially attractive 
relative to staying within a crowded local neighbourhood (motivation). 
Because both channels operate through the same EOS and OES measures, 
the reduced-form estimates conflate the two and cannot be separately 
attributed to either. The regression should therefore be interpreted 
as identifying the reduced-form predictive relationship between 
specificity and distance traveled, without attribution to a specific 
mechanism --- though the estimate remains informative: if either 
channel is operative, the regression will detect it.

## Conclusion

- What we found \[PLACEHOLDER --- awaiting data\]

The central policy takeaway from this analysis is that labour market adjustment mechanisms are not one-size-fits-all: the appropriate intervention depends on which mobility regime contains the occupation experiencing a shortage. The three mobility sub-regimes suggest three distinct policy levers.

In the Skill sub-regime, reskilling and short-cycle training are the natural adjustment mechanisms. Workers in adjacent occupations already possess much of the relevant task content, and targeted reskilling can accelerate the reallocation of labour toward occupations in shortage without requiring formal retraining. The policy implication is that investment in modular, portable training that builds on existing skill profiles will generate the largest mobility response in this segment.

In the Hierarchy sub-regime, the relevant bottleneck is not skill transferability but the hierarchy's silo and platform structure. An occupational shortage within one silo cannot be addressed by upskilling workers from a different silo, even if the occupations share a near-identical skill profile. Instead, shortages must be addressed within-silo, through either lateral movement on a platform or vertical movement between platforms. The best way to address a shortage at a higher platform is by increasing the supply of workers to lower platforms, freeing incumbents to pursue upward movement through the silo — the occupational analogue of the internal labour market adjustment mechanism described by @doeringer1971internal. Bridging programs, mentorship structures, and targeted recruitment into within-silo entry-level positions are the appropriate instruments.

In the Binary sub-regime, either a worker has the required education, or they do not. Neither reskilling nor within-silo movement can substitute for education. Adjustment in this sub-regime depends entirely on expanding formal training pipelines: increasing capacity in the relevant educational programs, reducing barriers to program entry, and shortening time-to-completion where possible. These are also the occupations that immigration policy should target, both in terms of admission and rapid credential recognition. 

If the hypothesized sub-regime structure is confirmed, the geometry of occupational mobility determines which policy lever to pull --- and the wrong lever is not just ineffective but potentially counterproductive. 

```{=latex}
\appendix
```


## O\*NET to NOC 2021 Skill Mapping {#sec-AA}

OaSIS provides occupation-level skill profiles for Canadian
classifications and would be an attractive data source for constructing
$C_{\text{skill}}$. However, OaSIS does not expose how those profiles
are constructed: the mapping between O\*NET and NOC 2021 occupations is not observable, the weights linking underlying occupations are not published, and there is no information on how multiple O\*NET occupations are combined to form an occupation profile. This appendix describes an alternative mapping constructed explicitly from published crosswalks, allowing full transparency of inputs and weights, diagnostic evaluation of mapping quality, and reproducibility of all outputs.

The mapping chains four classifications: O\*NET 2019 $\rightarrow$ SOC
2018 $\rightarrow$ NOC 2016 $\rightarrow$ NOC 2021. Each stage introduces potential one-to-many relationships. A naive approach would assign equal weight to all admissible paths, but this leads to over-weighting of occupations that appear frequently in crosswalk chains and inclusion of unrelated occupations. Instead, path weights are constructed to reflect the degree of splitting at each stage: if a source occupation splits into $n_s$ destination occupations at stage $s$, each path through that stage receives weight $w_s = 1/n_s$, and the overall path weight is $w_{\text{path}} = w_1 \times w_2 \times w_3$. Path weights are then aggregated by origin O\*NET occupation and destination NOC 2021 code and normalized to sum to one within each NOC to yield base weights. The consequences of naive versus path weighting can be severe: for NOC 72600 (Air pilots, flight engineers and flying instructors), naive weights assign five-eighths of the skill profile to agricultural workers, while path weights assign 78\% to pilots.

A power transformation and renormalization is then applied to further
reduce the influence of mapping artifacts: $\tilde{w}_k \propto w_k^2$.
This concentrates weight on the dominant O\*NET contributors and
diminishes the influence of peripheral occupations that enter through
crosswalk chains with many splits. For NOC 72600, downweighting
increases the pilot share of the skill profile to 96\%. Each O\*NET
occupation is then represented as a point in a reduced skill space using the first 12 principal components of the O\*NET skill content variables, and for each NOC 2021 code the weighted centroid of its contributing O\*NET occupations is computed using the downweighted mapping, yielding a 12-dimensional skill profile for each NOC occupation.

Two diagnostics assess the quality of each occupation's mapping in skill space. Herfindahl concentration measures how concentrated the mapping weights are across contributing O\*NET occupations: $H = \sum_k \tilde{w}_k^2$. For example, a Herfindahl of $\frac{2}{3}$ indicates that the occupation skill profile is effectively dominated by a single O\*NET occupation with a weight of $\approx .8$.  In contrast, when $H=\frac{1}{3}$ the largest contributor's share is between $\frac13$ and $\sqrt{\frac13}=.58$. Scaled Similarity measures the internal coherence of the contributing occupations in skill space: the weighted average distance of each contributing O\*NET occupation from the occupation centroid is computed, then normalized to a $[0.05, 1]$ interval by dividing by the maximum value across all occupations, multiplying by 0.95, and subtracting from 1, so that $SS = 1 - 0.95 \times \bar{d} / \max_j \bar{d}_j$. The lower bound of 0.05 is chosen to match the approximate lower bound of the Herfindahl index, making the two measures' floors commensurable. The mapping quality composite is the product $H \times SS$, which is high only when the mapping is both concentrated and internally coherent. The three interpretation cases are: high $H$ with any $SS$ (no problem --- almost all weight on a single O\*NET occupation); low $H$ with high $SS$ (no problem --- diffuse but coherent); and low $H$ with low $SS$ (problematic --- a diffuse average of disparate occupations whose skill profile is unreliable). NOC 41200 (University professors and lecturers) is a canonical example of the third case.

Occupations in the bottom decile of $H \times SS$ are excluded
from the local clean and non-local clean samples. Because the exclusion criterion is based on mapping quality rather than occupation characteristics, it
introduces noise into $C_{\text{skill}}$ selectively while leaving
$C_{\text{hierarchy}}$ and $C_{\text{binary}}$ unaffected --- a property exploited in the mapping quality falsification design described in @sec-data. The full mapping pipeline is implemented in
[@onet_to_noc] and is constructed entirely from O\*NET data and published crosswalks.

## TEER attainment consistency filter {#sec-AB}

Admissible TEERs are the Target TEER $\pm1$ plus TEER 0 (management), where the assumption is that management roles are independent of educational attainment. This filter is applied to both endpoints (2016 and 2021) to ensure the distance between reported occupations is well-measured.   

\begin{table}[H]
\centering
\begin{tabular}{lcr}
\toprule
Highest Attainment & Target TEER & Admissible TEERs \\
\midrule
No certificate, diploma or degree & 5 & 4, 5, 0 \\
Secondary (high) school diploma & 4 & 3, 4, 5, 0 \\
Apprenticeship or trades certificate or diploma & 3 & 2, 3, 4, 0 \\
College / CEGEP / other non-university & 2 & 1, 2, 3, 0 \\
University certificate or diploma below bachelor & 2 & 1, 2, 3, 0 \\
Bachelor's degree & 1 & 1, 2, 0 \\
Above bachelor (Master's, PhD, MD, etc.) & 1 & 1, 2, 0 \\
\bottomrule
\end{tabular}
\caption{Admissible TEERs by highest attainment level}
\label{tab:teer_mismatch}
\end{table}


## CWTVD resolving power and scoring-matrix robustness  {#sec-AC}

@sec-local adjudicates the horse race on two criteria of
equal standing, $\Delta KL$ and $\Delta CWTVD$, and a win requires
leading on both. The role of $\Delta CWTVD$ as a co-equal criterion
rests on a property established here by simulation: among the two
correlated informative geometries, $\Delta CWTVD$ resolves which
governs more sharply than $\Delta KL$ does. This appendix substantiates
that property, states why it can be established only by simulation, and
then checks that the resulting sub-regime rankings do not depend on the
particular cost matrix used to score $\Delta CWTVD$.
 
The online companion plots improvement relative to independence
against $\varepsilon$ for each candidate metric, across three mixed
data-generating processes (columns) and both scoring criteria (rows).% RICH FIX REF: was \ref{fig:mixture-sweep} (companion figure, no in-doc target) — replace with the companion's actual figure/anchor name.
Read
down each column: at a $75/25$ skill/hierarchy mixture the peak gap between
the skill and hierarchy curves is wider under $\Delta CWTVD$ (top row)
than under $\Delta KL$ (bottom row); at $25/75$ the ordering reverses and
the gap again opens wider under $\Delta CWTVD$; at $50/50$ the gap is at
its narrowest under both criteria. The gap under $\Delta CWTVD$ thus
tracks the imbalance of the true mixture, widening as the mixture departs
from balance and shrinking to its minimum at balance. $\Delta KL$, by
contrast, holds the two curves close to coincident across all three
mixtures, resolving the imbalanced cases only weakly. A second difference
appears at the benchmark: under every mixture the $\Delta KL$ curve for
$C_{\mathrm{binary}}$ retains a positive tail well past its peak, the same
geometry-blind credit for the aggregate stay/move split that drives the
out-of-regime KL behaviour in @sec-local-falsification; $\Delta CWTVD$
strips that credit out and the $C_{\mathrm{binary}}$ tail collapses to
the independence line. Both features point the same way:
$\Delta CWTVD$ separates the two correlated geometries more sharply and
denies the benchmark credit it has not earned geometrically.
 

This resolving-power property is established under known data-generating
processes rather than read off the estimation sample, for the structural
reason given in @sec-local: the property is a statement about the mapping
from a known true mixture to the measured gap between criteria, and a
single observed labour market delivers one gap drawn from one unknown
mixture, with no reference mixture against which its magnitude could be
judged. The Monte Carlo is therefore the complete justification the claim
admits, not preliminary evidence awaiting empirical corroboration. The
figures here substantiate the property; @sec-local states what follows for
trusting $\Delta CWTVD$ when it parts from $\Delta KL$ on the Canadian
data.
 

The scoring matrix $C$ used in $\Delta CWTVD$ is the arithmetic mean of
the three candidate normalized cost matrices. Because $C_{\mathrm{binary}}$
is constant off-diagonal, the blend's cell-level weighting is structurally
determined by $C_{\mathrm{skill}}$ and $C_{\mathrm{hierarchy}}$ regardless
of aggregation method; the relevant check is therefore whether the
sub-regime rankings survive when each candidate matrix is used as the sole
judge. The companion plots $\Delta CWTVD$ against
$\varepsilon$ for all three metrics, by sub-regime (columns) and scoring
matrix (rows), on the in-regime subset; diagonal panels (a plan scored
under its own cost matrix) are omitted.% RICH FIX REF: was \ref{fig:cwtvd-judge} (companion figure, no in-doc target) — replace with the companion's actual figure/anchor name. The $C_{\mathrm{skill}}$ and
$C_{\mathrm{hierarchy}}$ rows are the substantive robustness checks: if the
predicted winner in each sub-regime leads under the blended judge, it
should also lead under the opposing geometrically informative judge. The
$C_{\mathrm{binary}}$ row isolates how much of $\Delta CWTVD$'s signal
reflects the distance gradient specifically, over and above the
off-diagonal structure any geometry-aware metric would register.



## References
