---
title: "Occupational Mobility Is Not One Thing: Evidence from the Canadian Labour Market" 
author: "Richard Martin"
date: today
format:
  pdf:
    documentclass: article
    number-sections: true
    geometry: margin=1in
bibliography: references.bib
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{threeparttable}
  - \DeclareMathOperator*{\argmin}{argmin}
editor: 
  markdown: 
    wrap: 72
execute:
  echo: false
  message: false
  warning: false
abstract: |
  Workers predominantly move between proximate occupations, but proximity means different things across the labour market. We use entropic regularized optimal   transport to conduct a horse race among three distance metrics (skill, hierarchy, and binary), and show that no single geometry governs occupational mobility. Instead, the appropriate metric depends on each occupation's education specificity: skill where human capital is general, hierarchy where silos (occupational broad groups) and platforms (education and experience) structure mobility, and binary where human capital is so occupation-specific that neither skill nor hierarchy predict destinations. Each regime implies a distinct policy lever for addressing labour shortages.
--- 



```{r, include=FALSE}
#packages
library(tidyverse)
library(here)
library(knitr)
library(kableExtra)
library(patchwork)
library(gt)
library(janitor)
library(irr)
library(kableExtra)
library(conflicted)
conflicts_prefer(dplyr::filter)
#functions
format_counts <- function(tbbl) {
  tbbl %>%
    ungroup() %>%
    mutate(part = paste(n, `TEER`)) %>%
    pull(part) %>%
    (\(x) {
      paste0(
        "(",
        paste(x[-length(x)], collapse = ", "),
        ", and ",
        x[length(x)],
        ")"
      )
    })()
}
#constants
lvls <- c("Skill", "Hierarchy", "Binary")   # low -> high education specificity
shade <- "#EAF0F4"
#data
cdfs <- read_rds(here("out","dist_cdfs.rds"))
calibration <- read_rds(here("out", "calibration.rds"))
noc_edu_spec  <- read_rds(here("out", "noc_edu_spec.rds"))
teer_counts <- tabyl(noc_edu_spec$`Based on TEER`)|>
  rename(TEER="noc_edu_spec$`Based on TEER`")
kappa <- round(kappa2(noc_edu_spec[,c("Based on TEER", 
                                "Based on Occupation Education Specificity")])$value,2)
confusion <- noc_edu_spec |>
  mutate(across(c(`Based on TEER`,
                  `Based on Occupation Education Specificity`),
                \(x) factor(x, levels = lvls))) |>
  with(table(`Based on TEER`, `Based on Occupation Education Specificity`)) 
prop_same <- sum(diag(confusion))/sum(confusion)
m <- confusion|>
  as.data.frame.matrix() |>
  tibble::rownames_to_column("TEER")
disagree <- noc_edu_spec|>filter(`Based on TEER`=="Binary", `Based on Occupation Education Specificity`=="Skill")|>pull(noc)
```

## Pre-registration Notice {.unnumbered}

This paper is posted as a pre-registration of study design, prior to Statistics Canada Virtual Data Lab access.  Anything that looks like a result should be interpreted as a prior. Pre-registration is intended to defend against claims of post-hoc rationalization.

## Introduction

Occupational mobility is not one thing. Workers tend to move between proximate occupations, but proximity is not self-evident: occupations differ along multiple dimensions simultaneously, canonically the tasks undertaken [@gathmann2010general], but also by required skills, hierarchical position, and selective education requirements. Each implies a different geometry of the labour market, and different predictions about who moves where. We test which geometry governs mobility, and find that the answer depends on where in the labour market you look.

We separate local mobility, the shorter transitions that occur without a substantial gain in human capital, from non-local mobility, the potentially much longer transitions that follow a large gain in human capital. The two are not parallel phenomena: non-local mobility is the mechanism behind local mobility's heterogeneity. The specificity of the human capital a worker brings to bear shapes the distance of their non-local jump, and where they land governs which metric best rationalizes their subsequent local transitions.

For local mobility, distance is the explanatory variable: we ask which of three candidate metrics best rationalizes observed occupational transitions: skill (Euclidean in PCA-reduced O*NET space), hierarchical (tree metric on occupation taxonomy), and binary (degenerate metric). For non-local mobility, the role of distance is reversed: rather than asking which metric explains where workers go, we ask what determines the distance of a non-local transition, treating distance as the outcome and the pipeline strength of the worker's education--destination pair as its determinant.

Canada's National Occupational Classification (NOC) organizes occupations into a five-digit hierarchy. The first digit identifies one of ten broad occupational categories (e.g., Natural and Applied Sciences, Health, Trades). Crosscutting this hierarchy, the second digit is the Training, Education, Experience and Responsibilities (TEER): a level from 0 to 5 reflecting vertical stratification. For example, TEER 0 indicates management, whereas TEER 1–5 reflect decreasing levels of required education (TEER 1 university education at the high end, and TEER 5 no formal education at the low end). The remaining three digits subdivide each category into progressively finer groupings down to the unit group level. For hierarchical distance, occupation broad categories delineate separate labour market silos, and TEER levels define platforms within each silo.

Regarding local mobility, we anticipate that no single metric will dominate globally. A labour market in which managers, electricians and doctors all follow the same distance geometry would be surprising. We therefore partition the transition data into three sub-regimes based on the origin occupation’s education specificity, corresponding to three hypothesised distance structures: skill distance where human capital is general, hierarchical distance where silo structure binds, and a degenerate binary regime where human capital is so occupation-specific that distance no longer meaningfully orders destinations.

The policy relevance is direct. Labour shortages in different segments of the market require different adjustment mechanisms, and the right mechanism depends on which distance metric governs. Shortages where human capital is general may be addressed through reskilling, augmenting existing transferable skills. Shortages where silo structure binds are addressed within the silo through lateral moves across a platform and experience-based moves to an adjacent one, wherever the step requires no capacity-constrained credential. Shortages where human capital is occupation-specific cannot be filled by moving workers from other occupations, and this includes credential-gated positions wherever they sit. The relevant human capital is acquired only through specific education programs, whose entrants are drawn from across the labour market rather than from within any one silo. Interventions must therefore target access to and capacity within those programs directly. Getting the geometry right is therefore not merely a methodological contribution: it determines which policy lever to pull.

For local mobility we use entropic regularized optimal transport (EROT) [@cuturi2013sinkhorn; @peyre2019computational] to conduct horse races among candidate distance metrics, evaluating how well each rationalizes the full occupation-to-occupation transition matrix. A gravity model [@anderson_vanwincoop_2003], estimated by PPML [@silva2006log], shares EROT's functional form but estimates its origin and destination effects having observed the full transition matrix. Because these fixed effects are fitted using the same interior transitions on which model performance is evaluated, they can absorb discrepancies arising from a misspecified geometry, compressing differences in fit across competing distance metrics.

In contrast, EROT is solved using only a candidate cost metric and the margins of the transition matrix, without access to the interior flows. Its scaling vectors are pinned to the margins rather than fit to the cells, so discrepancies due to a misspecified geometry must appear in the predicted interior. No metric is shown the transition matrix when determining these scaling factors: the entire matrix, diagonal and off-diagonal alike, is predicted from the cost structure and the margins. The diagonal carries no discriminating signal, since all metrics assign zero self-distance and therefore treat staying identically; metrics differ only in their allocation of off-diagonal mass, which is where the comparison is determined. Each geometry's score is then a verdict on the cost structure alone, based on the allocation it did not use in estimation, which is what makes the comparison clean; we develop this in @sec-local.

For non-local mobility, we estimate, at the individual level, the reduced-form relationship between transition distance and the pipeline strength of a worker's education--destination pair. A wormhole is a path a credential opens through occupation space: a specific education leads to a specific occupation that would otherwise be unreachable, and the long jump is the traversal of that path. We measure pipeline strength by lift^[We use \emph{lift}, the term for this quantity in association-rule learning and market-basket analysis. The same ratio of a conditional probability to its marginal appears across fields under other names: a revealed-comparative-advantage index, a relative risk, and, in logarithm, the pointwise mutual information of education and occupation. Each expresses one idea, how much more (or less) than independence a given education feeds a given occupation.], $m(e,n) = P(n\mid e)/P(n)$, how much more an education $e$ feeds occupation $n$ than $n$'s overall prevalence would predict. Lift is a property of the pair, so a worker travels far only by realizing a high-lift path, and travels a short distance in two cases: a generic education offers no high-lift path to take, and a specific credential whose holder lands off its distinctive path realizes low lift wherever else they go. Distance travelled between occupations is predicted to increase with realized lift.

For example, atmospheric sciences and meteorology (40.04) and meteorologists and climatologists (21103) score the highest of all combinations, with a lift of 11,700: the credential opens a path to that occupation and to little else. At the other extreme, the lowest-lift^[Lowest among reliably measured pairs. Lift is computed only for education--occupation cells with at least 30 workers, below which sampling and base-five rounding dominate the ratio, see Appendix B.] combination is registered nursing (51.38) and electricians (72200), with a lift of 0.01: a highly specific credential whose holder is recorded in an unrelated, credential-gated occupation, which predicts the shortest distance travelled from the 2016 occupation.

[FINDINGS PLACEHOLDER]

## Regime Partition {#sec-rp}

The central structural choice in this paper is the partition of occupational transitions into two regimes, by the size of the human-capital increment that becomes operative for a worker between census years. We define *education* as the cross product of highest attainment and field of study. What matters is not when a credential was earned but when its specific human capital becomes operative, reaching an occupation that uses it, so a credential earned earlier but put to use only within the window is a large increment on the same footing as one freshly earned. The increment is latent; its proxy is the change in the education record between cycles. A recorded rise in level, or a same-level field change across a two-digit CIP boundary, marks a large increment and routes non-local; a smaller or absent record change routes local. Near-field changes within a two-digit CIP are negligible increments and route local under either the genuine or the misreported reading, so the split is invariant to the marginal misreporting that dominates.

We study non-negative operative increments only, which puts two kinds of case outside our object of interest. A recorded *decrease* in attainment level, whatever lies behind it, revocation, lost licence, or inconsistent report, is a negative increment, excluded from every specification by a rule fixed in advance on the recorded level alone. The exclusion matters because a decrease tends to accompany an unpredictable career change, the kind of move that fits the Binary metric and misfits the structured ones, so retaining decreases would tend to inflate Binary's win for a reason unrelated to specificity. *Substitutive retraining*, an over-credentialed worker who repositions on a newly earned lower credential, is likewise not the additive mobility we study, and needs no separate handling: recorded correctly its highest attainment is unchanged, so it routes local and leaks toward the null like any other unrecorded activation; recorded as a decrease, through the confusion of highest attainment with the most recent or most salient credential, it is excluded by the rule above. Neither fate lets it into the non-local sample, so it never contaminates the specificity estimate; the local leak is handled on the same toward-null footing as every other unrecorded activation.

Within scope, the latent increment and its proxy can diverge in two directions, and lift allows us to identify some classification errors. Recall that lift $m(e,n) = P(n\mid e)/P(n)$ captures whether a worker's 2021 occupation $n$ is one their education $e$ distinctively feeds. A worker with unchanged education is reinstated to non-local when the credential is dormant at the origin and operative at the destination: $$m(e, n_{2016}) < 1 \quad\text{and}\quad m(e, n_{2021}) - m(e, n_{2016}) \ge \tau.$$ Conversely a worker with an education change is reinstated to local when the credential is dormant at the destination: $m(e, n_{2021})<1$. Both are the same disagreement between latent and proxy, and the robustness check resolves both the same way: reclassify on lift and re-estimate, reporting the result beside the uncorrected one.

Lift can not rescue the classification errors that leave no signature, but fortunately these bias toward the null in both directions. A worker with unchanged education stays local when the credential activated but carried the worker into an occupation it does not distinctively feed. This is a genuine non-local move which the local geometry cannot rationalize: the local horse-race fit degrades. The opposite case, a worker with an education change, stays non-local when the credential did not actually activate but nevertheless appears operative at the destination: $m(e,n_{2021})>1$. This is a genuine local move, contrary to the predicted lift--distance relationship, which both flattens the estimated relationship and lowers its explanatory power.

Each residual degrades the explanatory power of its own side: the uncorrected local sample understates how well the local geometry fits, and the uncorrected non-local sample understates how well lift explains distance. This is why the local criterion is stated as degradation, not as an absolute zero that such contamination could hold hostage. A result that survives both the reclassification and this residual bias is therefore not an artifact of the proxy; @sec-AD gives the detection rule and its threshold sensitivity.

The two regimes differ in the mechanism governing mobility, not merely in the size of the increment. For local transitions the 2016 occupation anchors destination choice; for non-local transitions a large human capital increment repositions the worker in a way the 2016 occupation does not capture, the wormhole of the introduction. @sec-data develops what this difference implies for each estimation sample.

The local/non-local partition has a natural falsification structure.
Each model is tested both where it should work and where it should not,
and the regime-specific predictions provide stronger evidence for the
causal interpretation of the underlying mechanisms than either
falsification alone would. A single confounder would have to operate
in opposite directions across regimes to account for both patterns;
multiple confounders would each have to align fortuitously with the
predicted cross-regime structure. Neither is impossible, but both are
less parsimonious than the theoretical mechanism itself. The design
does not eliminate the possibility of confounding, but it raises the
bar considerably: the model's validity rests not on the absence of
confounders but on the implausibility of a story that generates the
predicted pattern of opposing failures across regimes.




## Specificity and the Sub-regimes {#sec-cf}

Two specificity constructs play central roles in the analysis, both from the 2021 Census (Table 98-10-0404-01), a random one-quarter sample of Canadian households.^[The linked 2016--2021 file used for estimation is constructed by linking long-form respondents across the two census cycles, and since each cycle independently samples one quarter of households, the linked file covers approximately one sixteenth of all Canadian households. The overlap between the two sources is therefore at most one quarter, the share of the long-form 2021 respondents who also appeared in the 2016 long-form. This limited overlap means the specificity constructs are derived largely from independent data relative to the estimation sample, reducing the risk that they mechanically predict transition distances through shared sampling rather than through the credential-selectivity mechanism.] Occupation education specificity (OES) is a size-residualized KL divergence measuring how distinctive an occupation's education distribution is; it sorts the local sub-regimes on origin OES. Lift is the cell-level education--occupation quantity that enters the non-local regression. Both are read from the same education$\times$NOC table.

Lift is $m(e,n) = P(\text{NOC}=n \mid \text{education}=e)/P(\text{NOC}=n)$, the ratio of an education's conditional occupation probability to that occupation's unconditional share. It is the cell-level term inside the education-side KL divergence $\text{KL}(P(\text{NOC}\mid e)\,\|\,P(\text{NOC}))$: where that divergence summarises how distinctive an education's whole occupation distribution is, lift reads a single education--occupation cell, how distinctively this education feeds this occupation. Lift is high where an education channels into an occupation far more than baseline, near one where it is no more likely than average to lead there, and below one where it leads there less than baseline. Because lift is a ratio of probabilities estimated from finite, disclosure-rounded cells, it is noisy for thin education--occupation pairs; it is smoothed toward its neutral value by education size and computed only for cells reliably measured, entering the regression in natural logarithm. The construction is detailed in @sec-AB.

OES is the size-residualized KL divergence of $P(\text{education}|\text{NOC})$ from $P(\text{education})$, residualized on $\log(T)$ to remove the mechanical floor by which a niche occupation's finite education distribution inflates the raw divergence. Because it is a regression residual, OES is signed and centred near zero: a high value is a divergence above what the occupation's size predicts. It classifies the three local sub-regimes on *origin* OES, a proxy for how credential-specific the worker's 2016 occupation is.

OES proxies for where an occupation sits between the poles of general and 
specific human capital [@becker1964human]. Occupations drawing broadly from 
the education distribution require general human capital in Becker's sense, 
skills whose returns are portable across employers and occupations, while 
occupations drawing narrowly from specific credential pipelines require 
occupation-specific human capital, valuable within the occupation but not 
transferable beyond it. Occupational licensing [@kleiner2013] is one 
institutional source of high specificity, and licensing lists offer a direct proxy for credential selectivity, but they rely on curated registries that 
are often incomplete, jurisdiction-specific, and require manual updating. OES proxies for the same underlying selectivity from the equilibrium education mix observed in the data, without requiring a curated list.

The three sub-regimes are named for their hypothesised winning metric
rather than for the dimension that defines them, a labelling
convention adopted for readability, not a presumption of results. The
primary partition is by OES rank. The group sizes are imported from the
TEER groupings described below `r format_counts(teer_counts)`, fixing the 
boundaries before any transition data is examined. This borrowing is purely 
a sizing convention: TEER supplies
the three counts, not which occupations populate them. Assignment
within those counts is determined entirely by OES rank, with the 
`r teer_counts$n[teer_counts$TEER=="Binary"]` occupations with the
highest OES forming the Binary sub-regime, the `r teer_counts$n[teer_counts$TEER=="Skill"]`
with the lowest forming the Skill sub-regime, and the remaining `r teer_counts$n[teer_counts$TEER=="Hierarchy"]`
forming the Hierarchy sub-regime. Importing the sizes from TEER rather
than cutting the OES distribution at natural breakpoints is a
pre-commitment device: it prevents the boundaries from being chosen,
consciously or otherwise, to fit the transition data. @fig-regimes
confirms that OES varies continuously across the occupational
distribution with no visible clustering, so no natural breakpoints
exist to exploit in any case.

The partition follows directly from the human capital specificity
argument. Occupations with high OES draw narrowly from specific
credential pipelines and map to the Binary sub-regime, where the prediction is that no occupational geometry orders destinations. For example, absent a new highest attainment, the occupational movement an optometrist is hard to predict: most remain optometrists, and among the few who do move, neither skill nor hierarchical distance predicts where they go. In contrast, occupations with low OES draw broadly from the education distribution; their human capital is general in Becker's sense, and skill similarity governs mobility. For example, absent a new highest attainment, a 2016 manager's subsequent mobility is rationalized by skill distance: the skill profile of most managers (even across NOC silos) is quite similar, yielding many short-distance options.

The remaining occupations sit between these poles, where the silo and platform structure of the NOC hierarchy governs transitions. A nurse aide (NOC 33102) is an example: the closest destinations are the same-platform occupations within its minor group, such as dental assistant, medical laboratory assistant, and pharmacy assistant, all lateral moves among TEER 3 assisting occupations in health. Hierarchical distance also predicts, at higher cost, upward movement to a different platform within the health silo, such as licensed practical nurse (NOC 32101). Where the platform above is credential-gated, as licensed practical nurse is, the jump necessarily involves a specific educational increment, and that increment can go unrecorded: the new credential may be omitted from a self-reported education record, or it may be real but sub-maximal, correctly leaving the highest attainment field unchanged even though it enabled the jump. In either case a credential-gated move the human-capital story would call non-local appears local in the data, a source of regime contamination we return to in @sec-non_local_false. Where the platform above is instead reachable only through accumulated experience, no such increment exists and no contamination arises; the missing attainment change is correct.

The TEER partition that supplies the group sizes has an independent theoretical motivation paralleling the OES logic. TEER 1 occupations, such as physicians, dentists, and lawyers, carry the most occupation-specific human capital, which makes it difficult to predict alternative occupations.  At the other end of the spectrum, we hypothesize that TEER 0, 4, and 5 occupations carry so little occupation-specific human capital that movement can freely traverse the broad occupational groups. TEER 2 and 3 fall in between, with human capital specific enough to impede cross-silo transitions yet not so specific as to produce occupational lock-in. The two orderings rest on different logics, yet both motivate the same three-way conceptual split. The TEER partition serves as a robustness check on the primary OES-based partition. The partitions agree on `r scales::percent(prop_same)` of occupations, with Cohen's $\kappa$=`r kappa`. The agreement is modest, which is what the design expects and all it requires: TEER and OES are built on different logics, so the check is whether the horse-race results hold under the TEER assignment, not whether the two assignments coincide.

We treat the OES partition as the primary specification and the TEER partition as a confirmatory check. TEER is a coarse, discrete proxy for specificity, built for an unrelated administrative purpose; OES is a continuous measure derived directly from the equilibrium education mix observed for each occupation. The two partitions agree where TEER's administrative categories happen to track specificity, and diverge where they do not. @tbl-partition-crosstab quantifies that divergence.  In particular, there are `r length(disagree)` occupations that, based on being TEER 1 would be in the binary sub-regime, but based on OES they belong in the skill sub-regime: `r disagree|>paste(collapse = ", ")`. We expect these to see substantially more movement than, say, optometrists, with mobility governed by skill similarity rather than credential lock-in.


```{r}
#| label: tbl-partition-crosstab
#| tbl-cap: "Sub-regime confusion matrix."

kbl(m, booktabs = TRUE, align = "lccc", row.names = FALSE,
    col.names = c("TEER", "Skill", "Hierarchy", "Binary"), linesep = "") |>
  add_header_above(c(" " = 1, "Occupation education specificity" = 3),
                   bold = TRUE) |>
  kable_styling(latex_options = "hold_position",
                full_width = FALSE, position = "center") |>
  column_spec(2, bold       = c(TRUE,    FALSE,   FALSE)) |>
  column_spec(3, bold       = c(FALSE,   TRUE,    FALSE)) |>
  column_spec(4, bold       = c(FALSE,   FALSE,   TRUE)) |>
  footnote(
    general = paste(
      "Counts are occupations (n = 504).",
      "Both partitions share the same group sizes by construction"
    ),
    general_title = "", footnote_as_chunk = TRUE,
    escape = FALSE, threeparttable = TRUE
  )
```


```{r}
#| warning: false
#| echo: false
#| label: fig-regimes
#| fig-cap: "Mobility Sub-Regimes"

noc_edu_spec |>
  pivot_longer(cols = contains("Based on"), names_to = "criteria", values_to = "Sub-Regime") |>
  ggplot(aes(T, specificity, text = noc, fill = `Sub-Regime`)) +
  scale_fill_viridis_d() +
  geom_point(shape = 21, size = 1.5, stroke = .5) +
  scale_x_log10(labels = scales::comma) +
  facet_wrap(~criteria) +
  theme(text = element_text(size = 9)) +
  labs(x = "Occupation Size",
       y = "Occupation's Education Specificity")
```

Occupational mobility tends to decline with age, but age is a proxy for
the underlying mechanism rather than the mechanism itself.
@mincer1981labor show empirically that age effects on mobility largely
disappear once tenure is controlled, and attribute the residual to
specific human capital accumulation rather than aging per se: as workers
commit to an occupational path, the wedge between their current wage and
outside options widens, making mobility increasingly costly. Tenure is
their operational measure of that wedge, but the wedge is unobservable
in principle, what they estimate is the hazard of separation as a
function of job duration, not specific capital directly. OES approaches
the same underlying quantity from the supply side: rather than inferring
specific capital from the duration of job spells, it reads off the
credential selectivity of the occupation from the equilibrium education
mix. A newly trained and a 20-year veteran physician occupy the same
high-OES occupation, and should be similarly immobile; a newly appointed
and a long-tenured senior manager occupy the same low-OES occupation and
should remain similarly mobile. Tenure varies within each pair but the
mobility prediction does not, because what constrains mobility is the
credential selectivity of the occupation, not the worker's time in it.
This is a distinction tenure cannot identify as a stable property of the
occupation, but OES can.

## Cost Metrics {#sec-cost-metrics}

The skill distance matrix $C_{\text{skill}}$ is constructed from 161 
O\*NET measures spanning skills, abilities, knowledge, and work 
activities, mapped to NOC 2021 occupations via a purpose-built 
crosswalk @onet_to_noc. An alternative approach would be to 
use OaSIS occupation-level skill profiles, but we construct our own 
crosswalk for a specific reason: falsifiability. The horse race 
requires that measurement error in the skill distance matrix be 
attributable to a known source: imprecision in the O\*NET 
to NOC 2021 mapping. OaSIS obscures the structure of that mapping, making 
it impossible to characterize where and how measurement error enters 
$C_{\text{skill}}$; our crosswalk makes this structure transparent 
and therefore makes the identification strategy falsifiable. Each NOC 
is represented as a weighted average of its constituent O\*NET 
occupations, with weights determined by the crosswalk mapping.  For more details on the mapping see Appendix A.

O\*NET provides two scales for each measure: Importance (how important the attribute is to the occupation) and Level (the degree to which the attribute is required). Both scales are standardized to a common 0--10 range and combined via geometric mean, yielding a single score that reflects both the relevance and the intensity of each attribute. The geometric mean is preferred over the arithmetic mean as it penalizes cases where one scale is near zero, an attribute that is important but required at negligible level, or required at high level but deemed unimportant, receives a lower composite score than one scoring moderately on both dimensions. The two scales are highly correlated ($\rho=.96$), so this choice is unlikely to be consequential for the resulting distance matrix but we note it for transparency.

The 161 O\*NET skill items are correlated, so raw Euclidean distance in the full item space implicitly upweights dimensions measured by many items over those measured by few. We apply PCA to the standardised profiles to remove this redundancy, then retain the number of components needed to preserve pairwise distance rank ordering: starting from $k = 1$, we increment until the Spearman rank correlation between pairwise distances in the $k$-dimensional PCA space and in the full 161-dimensional space exceeds 0.99. This criterion, rank preservation rather than variance explained, is chosen because the object of interest is the distance matrix itself, not the underlying skill dimensions. Twelve components satisfy it. The skill distance between any two occupations is then the Euclidean distance between their coordinates in this 12-dimensional space. An interactive heatmap of the resulting distance matrix is available in the [online companion.](https://rpmartin.github.io/not-one-thing-companion/#sec-skill-heatmap)

A small number of off-diagonal cells of $C_{\text{skill}}$ are exactly zero, because a few NOC unit groups map to the same O\*NET occupation: NOC resolves distinctions that O\*NET does not, so occupations O\*NET treats as one are assigned identical skill profiles and zero mutual skill distance. This affects 280 of the roughly 253,000 ordered off-diagonal pairs (0.11\%), falling into 36 groups of mutually skill-identical occupations. The zeros are not measurement error: where O\*NET does not separate two occupations, no O\*NET-based skill measure can, and zero distance is the correct statement of that limit. We keep them, and characterize the groups in Appendix A.

The binary switching cost matrix $C_{\text{binary}}$ assigns a cost of zero to remaining in the same occupation and one to any transition. Under entropically regularized optimal transport, this implies that all off-diagonal transitions share the same switching cost, so conditional on leaving an occupation, relative transition probabilities are governed entirely by the destination scaling factors, which are pinned by the destination margins. In other words, the origin occupation provides no additional information for ordering alternative destinations once a move occurs.

$C_{\text{binary}}$ serves two roles in the analysis. First, it is the theoretical straw man in the global horse race: the hypothesis that mobility costs depend only on whether a switch occurs rather than on where it leads. If this hypothesis holds, $C_{\text{binary}}$ should rationalize observed mobility better than either $C_{\text{skill}}$ or $C_{\text{hierarchy}}$. Second, $C_{\text{binary}}$ is a plausible description of labour markets in which occupational distance carries little information about the destinations of movers. In such settings, conditioning on a move, destinations are allocated according to the destination scalings, which track overall prevalence rather than proximity to the origin occupation, so the additional geometry imposed by the structured metrics does not improve predictive performance.

The horse race therefore turns on destination structure rather than mobility rates, since all three metrics accommodate the observed propensity to remain in the same occupation. This second role motivates the Binary sub-regime hypothesis: where human capital is sufficiently specific that occupational distance does not systematically order exits, binary is not merely a degenerate benchmark but a substantive model of mobility. Because it imposes no origin-specific structure on off-diagonal transitions, its only free mechanism is the aggregate stay--move split, governed by the single parameter $\varepsilon$, which sets the overall degree of mobility. @sec-local uses this characterization when comparing where the three metrics peak.

The hierarchical distance matrix $C_{\text{hierarchy}}$ is derived directly from the NOC 2021 taxonomy, treating the occupational classification structure as a tree metric. Distance between any two occupations is defined as follows: zero if all five digits match (same unit group); one if the first four digits match (same minor group, different unit group); two if the first three digits match (same sub-major group, different minor group); three plus the absolute difference in TEER level if only the first digit matches (same broad occupational category, different sub-major group); and nine for any cross-category transition. The absolute TEER difference ranges from zero to five, so within-category distances range from three to eight, preserving a strict separation from the cross-category sentinel value of nine.  An interactive hierarchical distance heatmap of the resulting matrix is available in the [online companion.](https://rpmartin.github.io/not-one-thing-companion/#sec-hier-heatmap)

Occupation broad groups function as silos (bounded mobility clusters with limited entry points and internal progression along skill levels [@doeringer1971internal]), and TEER levels function as platforms within each silo, capturing the vertical skill stratification that governs upward movement. By construction, any between-platform move (minimum distance 4) is further than any within-platform move (maximum distance 3), and any cross-silo move (distance 9) is further than any within-silo move (maximum distance 8). The occupation broad group is a coarser unit than the occupational clusters described by @althauser1989, but the appropriate level of aggregation for the horse race: a finer partition would increase the preponderance of cross-category sentinel values in the hierarchical distance distribution, collapsing the metric's geometry toward binary and undermining its role as a distinct candidate in the horse race.

## Cost Metric Normalization

The three cost matrices are defined on different scales by construction: skill 
distances are continuous and spread over a wide range, hierarchical distances 
are discrete with a sentinel value of nine reserved for cross-category 
transitions, and binary distances take only the values zero and one. Because 
the cost matrix scale and $\varepsilon$ are not separately identified, the 
choice of anchor shifts where a metric's peak falls on the $\varepsilon$ axis 
but not the peak fit it attains; the winning metric is therefore invariant to 
normalization. Normalization matters instead for the legibility of the sweep 
plots: on a common $\varepsilon$ grid, matrices on different scales imply 
different effective $\varepsilon$, so the curves are not visually comparable 
unless first placed on a common scale. For the same reason we anchor at a 
substantively meaningful unit, so that position on the $\varepsilon$ axis 
remains interpretable.

We normalize skill and hierarchical distances to a common scale by anchoring 
both at a meaningful unit of hierarchical distance: distance 3, the maximum 
distance without changing platforms. Distance 3 falls at the fourth percentile 
of the hierarchical distribution, reflecting the preponderance of cross-silo 
pairs. The corresponding fourth percentile of the skill distance distribution 
is 7.145. Dividing hierarchical distances by 3 and skill distances by 7.145 
anchors both matrices at the same point in their respective distributions. 
Binary distances are not subject to this anchoring; their two-point support is 
shown for reference. @fig-cdfs gives the empirical CDFs before and after 
normalization. Normalization pins the skill and hierarchical distributions to 
the same point at the anchor, but their shapes remain distinct: hierarchical 
distance is discrete, with the small minority of within-platform pairs spread 
below the anchor and the bulk of pairs concentrated at the cross-category 
sentinel, while skill distance is continuous and dispersed across a wider 
range. This residual difference in shape is the separability the horse race 
exploits; normalization equalizes scale, not geometry.


```{r}
#| warning: false
#| echo: false
#| label: fig-cdfs
#| fig-cap: "Empirical CDFs of the three cost matrices before and after normalization; normalization places skill and hierarchical on a common scale for the shared $\\varepsilon$ sweeps without equalizing their shape."
cdfs
```

## Mobility Data {#sec-data}

The analysis uses the Census 2016--2021 longitudinal file, constructed 
by linking long-form respondents across the two census cycles via the 
Derived Record Depository (DRD), Statistics Canada's national 
longitudinal database of personal identifiers. Linkage is 
probabilistic; details of the matching keys and overall linkage rate 
will be verified upon data access.

[Linkage Details Here]

In the following analysis we make use of a \emph{measurement complete} 
subset of the linked data where we observe occupation, field of study, 
highest educational attainment, class of worker and age in both 2016 and 2021.

[Missingness Details Here]

Canada's NOC 2021 hierarchy nominally contains 516 unit groups, but 
the five Senior Manager NOCs (00011--00015) are aggregated into NOC 
00018 in the CIP-NOC table used to derive the specificity constructs, leaving 512 
occupations. Eight unit groups have no admissible O*NET mapping and 
are excluded from all estimation samples, leaving 504 occupations in 
the analysis.

These 504 occupations are the universe over which the sub-regime
partition is defined: the OES ranking that fixes the three sub-regimes
`r format_counts(teer_counts)` is computed over all 504, and
@tbl-partition-crosstab and @fig-regimes are likewise on the full 504.
The mapping-quality exclusion enters one step later, at estimation
rather than partition. <!-- REVIEW #1: state the base the decile is computed on. The 8 no-O*NET-mapping exclusions and the 50 bottom-decile-mapped occupations must be disjoint sets for 504 - 50 = 454 to hold as written. If the H×SS composite is scored after the 8 are already dropped, the decile is of ~504 (giving ~50), say so explicitly. If it is scored on 512, the arithmetic and the disjointness both need restating. A referee will check 516 - 4 - 8 = 504 and then ask where the 50 come from. -->
The bottom decile of the $H \times SS$ mapping
quality composite (Appendix A) flags 50 poorly mapped occupations, and
the local clean and non-local clean samples are built from transitions
among the remaining 454. A transition is dropped whenever either
endpoint is one of the 50, so the clean estimation samples run on the
454 cleanly mapped occupations while the partition they inherit was
cut on the full 504.

There are two core subsamples used in both the local and non-local mobility analyses. Their role reverses depending on whether they serve as the primary identification sample or the falsification sample. The two samples are separated by the size of the human-capital increment a worker acquires between 2016 and 2021, as developed in @sec-rp. We operationalise the increment from the education record, the cross product of highest attainment and field of study, using the field-of-study classification's two-digit series as the unit of field distance.\[To be determined whether CIP in the linked file is harmonized or versioned.\] A worker brings a large increment into play by reaching a higher level of attainment, or by changing field at the same level across two-digit series, which is retraining into an area where the original credential does not transfer. A change in the recorded education between cycles is read as specific human capital becoming operative within the window, whether freshly earned or earned earlier and only now put to use, since the record dates neither. A worker whose education changes only within a two-digit series at the same level brings a small increment into play and is grouped with those with no change: these same level, same 2 digit CIP new attainments make up X% of all new attainments.

The \emph{local clean sample} comprises individuals whose education increment is small by this criterion, with no self-employment and an O*NET-to-NOC mapping quality score above the bottom decile. This is the sample where distance-based models should best rationalise mobility: no specific human capital becomes newly operative over the inter-census period, so the 2016 occupation is an informative anchor. Any education already in place is reflected in where the worker starts and, so long as it is not newly brought to bear, does not drive subsequent movement.

The \emph{non-local clean sample} comprises individuals whose education increment is large by this criterion, subject to the same measurement restrictions as the local clean sample. This is the sample where the specificity regression should have the best fit: a large increment may open a wormhole to a distant occupation, the 2016 occupation is no longer an informative anchor, and specific human capital becoming operative within the inter-census window is the active ingredient in long-distance jumps. In contrast, the local model will struggle to rationalize these transitions on the basis of distance.

In addition, for both models we investigate the effect of measurement error in occupation, when the data is otherwise appropriate for the model being fit. We treat occupation measurement error as likely when individuals are self-employed: self-employed workers frequently perform multiple roles simultaneously, making a single occupation an unreliable summary of their occupational position. We do not treat an inconsistency between reported occupation and reported attainment as measurement error. Under the latent--operative distinction developed in @sec-rp, such an inconsistency is more naturally read as an activation lag, a credential held but not yet operative at the 2016 reference week, or newly operative and not yet reflected in the 2021 occupation, than as a misreported occupation. This is signal, not noise: the 2016 case is precisely the dormant credential the reinstatement of @sec-AD recovers, and the 2021 case is a genuine in-scope increment whose destination the reference week has not yet caught. Screening these cases as mismeasured would delete the very phenomenon the partition is built to study. We rely instead on the accuracy of the census occupation coding itself, Statistics Canada's flagship instrument, for which the assumption that a reported NOC is a genuine labour-market outcome rather than a coding error is well founded; residual miscoding is second-order relative to activation lag and enters distance measurement as symmetric attenuating noise.

For non-local mobility we can perform one further diagnostic: the effect of poor skill mapping from O\*NET to NOC, when the data is otherwise appropriate for the model being fit. Poor skill mapping is identified by the bottom decile of the O\*NET-to-NOC mapping quality composite. Poor mapping introduces noise specifically into $C_{\text{skill}}$ while leaving 
$C_{\text{hierarchy}}$ unaffected, a metric-specific prediction 
that serves as a falsification of the mapping quality composite 
itself. This diagnostic is not available for local mobility. As noted
above, the local clean subset is built from transitions among the 454
cleanly mapped occupations, whereas the mapping quality subset is built
from transitions involving at least one of the 50 poorly mapped
occupations, so the two subsets have disjoint cells by construction.
Since $\Delta KL$ and $\Delta CWTVD$ require a common set of possible origin-destination pairs to be comparable, no valid comparison is possible between the local clean and mapping quality subsets.

## Local Model {#sec-local}

Mobility costs are latent: workers' switching costs are not observable 
in principle, so any distance matrix is necessarily a specification of 
an unobservable quantity. The three candidate matrices, skill, 
hierarchical, and binary, represent distinct hypotheses about the 
structure of those latent costs.

The entropy-regularized optimal transport problem minimizes the following Lagrangian, where the first term is the transport cost, the second is the entropic regularization penalizing deterministic transport plans, and the third and fourth enforce the marginal constraints $\sum_j P_{ij} = a_i$ and $\sum_i P_{ij} = b_j$ via dual variables $f_i$ and
$g_j$.

$$\mathcal{L} = \sum_{ij} P_{ij} C_{ij} + \varepsilon \sum_{ij} P_{ij}(\log P_{ij} - 1) + \sum_i f_i(a_i - \sum_j P_{ij}) + \sum_j g_j(b_j - \sum_i P_{ij})$$

where $a_i$ and $b_j$ are the empirical marginal distributions in 2016 and 2021 respectively. 

The first-order conditions yield:
\begin{equation}
\log P_{ij} = \frac{f_i}{\varepsilon} + \frac{g_j}{\varepsilon} - 
\frac{C_{ij}}{\varepsilon}
\end{equation}
Exponentiating, the optimal transport plan takes the form 
$P^\star_{ij} = u_i \, e^{-C_{ij}/\varepsilon} \, v_j$, where 
$u_i = e^{f_i/\varepsilon}$ and $v_j = e^{g_j/\varepsilon}$ are 
scaling vectors. Strict convexity of the regularized objective 
guarantees a unique solution. The solution is therefore a row- and 
column-rescaling of the exponential kernel $K_{ij} = e^{-C_{ij}/\varepsilon}$, where the Sinkhorn algorithm recovers it by alternating row and column 
normalization until both marginal constraints are satisfied to within a specified tolerance. All computations use a log-domain Sinkhorn solver 
for numerical stability, implemented in the \texttt{safesink} R package [@safesink].

In log space, the optimal transport plan decomposes into additive origin effects
($\log u_i$), additive destination effects ($\log v_j$), and a dyadic cost term
($-C_{ij}/\varepsilon$). The scaling vectors correspond to Lagrange multipliers on
the marginal constraints and adjust to ensure the implied joint distribution matches
the observed origin and destination occupation stocks, absorbing any components of
the data that operate at the level of origins or destinations alone given the
specified cost structure. The cost matrix captures the residual structure that
depends on origin--destination pairing. The horse race asks which specification of
that dyadic structure best rationalizes observed transitions.

The gravity model [@anderson_vanwincoop_2003; @silva2006log] estimates origin and destination fixed effects jointly with the cost parameters, and that joint estimation is precisely the problem in our setting. Because the fixed effects are fit to the transitions, they carry the slack to absorb variation that a wrong cost matrix leaves misfit, so a misspecified geometry achieves good fit by compensating through them rather than by being right. Using a gravity model would be akin to giving the slower horses performance-enhancing drugs in inverse proportion to their natural speed: the race is closer than it should be, and the finishing order may be perturbed. EROT fixes the margins at observed values, and its scaling vectors are pinned to those margins rather than fit to the cells, so they cannot take up a wrong geometry's misfit. Each cost matrix must earn its fit in the residual dyadic structure, with no compensation for its own misspecification, which separates the question of where mobility flows, shaped by the cost matrix and the focus of this paper, from the question of why workers move. Wages, amenities, and other occupation-level incentives operate through the margins and are deliberately outside the scope of the identification strategy.

One might ask why we do not simply run PPML with the margins constrained to their observed values, since EROT and the gravity model share the functional form $u_i \, e^{-C_{ij}/\varepsilon} \, v_j$: the entropic plan is Anderson--van Wincoop gravity with $\beta = -1/\varepsilon$ and the scaling vectors in the role of the fixed effects. The constraint is not an extra step, because PPML with a full set of origin and destination fixed effects already reproduces the observed margins at its optimum; the first-order conditions for the fixed effects are the margin-matching equations. What the referee's constrained regression would still do, and EROT would not, is estimate a separate distance coefficient $\beta$ per cost matrix, jointly with the fixed effects, so that a misspecified geometry can shrink its own $\beta$ toward zero while the fixed effects absorb the misfit, leaving a coefficient that cannot distinguish a geometry that does not order destinations from one that does but was rescued at the margins. EROT removes exactly that freedom: it pins $\beta = -1/\varepsilon$ on a grid common to every metric, so no candidate estimates its own distance-sensitivity; it scores the constructed plan against the observed off-diagonal cells rather than maximising the plan's own likelihood, which is what lets the geometry-aware $\Delta CWTVD$ criterion enter at all, a criterion PPML's likelihood has no analogue for; and it imposes the margins unconditionally, for every $C$ and every $\varepsilon$, rather than achieving them at a jointly fitted optimum, so margin-matching is orthogonal to the geometry question rather than co-solved with it. Constrained PPML with those three defects repaired is not an alternative to the entropic transport formulation; it is that formulation.

The fairness of the race has a second aspect that distinguishes it from
an in-sample regression fit. A cost matrix is fixed a priori, not
estimated from the transitions, and given the margins and a value of
$\varepsilon$ the transport plan is then fully determined. The plan is a
prediction of the off-diagonal transition structure that never consults
that structure: nothing in its construction is chosen by looking at where
workers actually moved, so a metric that fits the observed transitions
well has not earned the fit by bending toward the data the way a regression
with enough parameters does. It committed to a geometry in advance and the
geometry happened to order destinations the way workers moved. Good fit is
therefore not cheap here in the way good in-sample fit is cheap in
regression, and the usual worry about overfitting the outcome does not
arise, because the outcome was not fit. The one degree of freedom is
$\varepsilon$, a single scale parameter swept over a grid and common to
every metric, so no candidate is granted more flexibility than another.
The diagonal is the one place all metrics coincide regardless of geometry:
every cost matrix assigns zero self-distance, $C_{ii}=0$, so every plan
concentrates mass on the diagonal and reproduces the aggregate stay/move
split. That shared diagonal credit is common ground rather than a
discriminating signal, and the race is decided on the off-diagonal
allocation, which is exactly the part no cost matrix was allowed to see.




The regularization parameter $\varepsilon$ has a natural interpretation
in the random utility framework. The optimal transport plan $P^\star_{ij} = u_i \, e^{-C_{ij}/\varepsilon} \, v_j$ is the static equilibrium of a discrete choice model in which workers draw idiosyncratic utility shocks scaled by $\varepsilon$ over destination occupations. Low values of $\varepsilon$ concentrate the transport plan on cost-minimizing assignments, approaching a deterministic reallocation in the limit; high values diffuse mobility across destinations, approaching the independence baseline $P_{ij} = a_i b_j$ as $\varepsilon \to \infty$. The RUM analogy governs the interior structure of the transport plan, how probability mass is allocated across origin-destination pairs, but does not extend to the margins. In a standard logit model the choice probabilities are outcomes of utility maximization with no constraint on their aggregates; in EROT the origin and destination occupation distributions are hard constraints that the transport plan must exactly reproduce. This distinction is methodologically consequential: because the margins cannot adjust to absorb cost misspecification, a poorly chosen cost matrix cannot hide behind fixed effects or scale recalibration, its misfit appears directly in the objective. 

A critical identification constraint follows directly from the form of 
the optimal transport plan: only the ratio $C_{ij}/\varepsilon$ enters 
$P^\star$. The scale of the cost matrix and the value of $\varepsilon$ 
are therefore not separately identified, and results reported at a 
single value of $\varepsilon$ are uninterpretable without an implicit 
normalization assumption. We evaluate across a wide log-spaced grid of 
$\varepsilon$ values from 0.1 to 1000, deliberately broader than the range 
expected to be informative, to ensure the informative region is fully 
contained within the evaluated range and its boundaries are visually apparent.

The horse race is evaluated using two complementary fit metrics, both
normalized relative to the independence baseline $P^{\text{ind}}_{ij} =
a_i b_j$, so that 0 indicates no improvement over independence, 1
indicates perfect fit, and negative values indicate a transport plan that fits observed mobility worse than independence. 

The first metric is relative KL divergence improvement:

$$\Delta KL = \frac{KL(P^{\text{obs}} \parallel P^{\text{ind}}) - KL(P^{\text{obs}} \parallel \hat{P})}{KL(P^{\text{obs}} \parallel P^{\text{ind}})}$$

where $KL(P \parallel Q) = \sum_{ij} P_{ij} \log(P_{ij} / Q_{ij})$.
This measures the share of the divergence between observed transitions
and independence that is explained by the fitted transport plan. KL
divergence is sensitive to probability mass wherever it is misallocated, but is geometry-blind: misallocating mass to a nearby occupation is penalized identically to misallocating it to a distant one.

The second metric is relative Cost-Weighted Total Variation Distance (CWTVD) improvement:

$$\Delta CWTVD = \frac{CWTVD(P^{\text{obs}}, P^{\text{ind}}) - CWTVD(P^{\text{obs}}, \hat{P})}{CWTVD(P^{\text{obs}}, P^{\text{ind}})}$$

where $CWTVD(P^{\text{obs}}, \hat{P}) = \sum_{ij}
\frac{|P^{\text{obs}}_{ij} - \hat{P}_{ij}|}{2} \cdot C_{ij}$. 

CWTVD is a cost-weighted extension of total variation distance: the 
classical TVD between discrete distributions, 
$\frac{1}{2}\sum_{ij}|P_{ij} - Q_{ij}|$, is extended by weighting 
each absolute residual by the corresponding entry of a cost matrix, 
penalizing probability mass misallocated to geometrically distant 
origin-destination pairs more heavily than mass misallocated to nearby 
ones. The cost matrix $C$ that weights the CWTVD residuals is not fixed to a single blend. For adjudication we score $\Delta CWTVD$ under each of the two informative cost matrices in turn, $C_{\text{skill}}$ and $C_{\text{hierarchy}}$, treating each as sole judge. These two are the endpoints of the family of blends the score could use: because $C_{\text{binary}}$ is flat off-diagonal and contributes no off-diagonal weighting, that family is the convex hull of $C_{\text{skill}}$ and $C_{\text{hierarchy}}$, and a metric that leads at both endpoints leads under every blend between them. Scoring at the two endpoints therefore certifies geometric soundness across the whole family, where a single blended judge would certify it only at one interior point. The two informative matrices are genuinely separable: the cell-level correlation between $C_{\text{skill}}$ and $C_{\text{hierarchy}}$ is about 0.3, so the endpoints are far enough apart to bracket a real interval rather than two near-identical judges.

We adjudicate the horse race on the likelihood criterion and the
geometry-aware criterion, the latter evaluated at both informative
endpoints, and a metric wins only by leading on all of them. The
likelihood criterion is peak relative KL improvement. For each
candidate cost matrix we maximise $\Delta KL$ over the $\varepsilon$
grid, and the reason for using it is that the EROT plan
$P^\star_{ij} = u_i\, e^{-C_{ij}/\varepsilon}\, v_j$ is the equilibrium of
a multinomial-logit discrete-choice model. Comparing candidates by their
maximised likelihood is then identical to comparing their peak relative
KL improvement, so peak $\Delta KL$ is maximum-likelihood model
selection rather than a fit heuristic. The geometry-aware criterion is
peak relative CWTVD improvement over the same grid, scored under each
informative judge, $C_{\text{skill}}$ and $C_{\text{hierarchy}}$, in
turn. KL is blind to the geometry of the outcome space, scoring a unit
of misallocated mass the same whether it lands on a near pair or a far
one. CWTVD weights each misallocation by distance, and across mixed
data-generating processes it separates the two correlated informative
geometries, $\mathrm{corr}(C_{\mathrm{skill}}, C_{\mathrm{hierarchy}}) =
0.3$, more sharply than KL does, with the gap between them tracking how
far the true mixture sits from balance. The [online companion](https://rpmartin.github.io/not-one-thing-companion) documents this directly.
 
The two justifications are independent. KL carries the likelihood
interpretation but cannot see geometry; CWTVD sees geometry but carries no
likelihood interpretation. Neither contains the other, so we hold them on
equal footing rather than naming one primary, and a win must satisfy the
likelihood criterion and the geometry-aware criterion at both endpoints.

The exclusion of recorded decreases (@sec-rp) enters the horse race with a
direction we state in advance. A decrease is an unpredictable career change,
which fits the Binary metric and misfits the two structured metrics, so
retaining decreases would raise Binary's peak and lower the others'. Excluding
them is therefore conservative in the Binary sub-regime, where it removes
observations that support the hypothesised winner, and flattering in the Skill
and Hierarchy sub-regimes, where it removes observations that misfit the
hypothesised winner. Because the exclusion helps us in two of the three
sub-regimes, we report the horse race both ways at the primary OES partition:
the winning metric and its peak improvement with decreases excluded, our
primary specification, and with decreases retained. The exclusion is on a
pre-outcome membership rule, a recorded level decrease, applied uniformly and
without reference to fit, so the comparison is a check rather than a choice; we
pre-commit to the prediction that the winner is unchanged in every sub-regime,
that Binary's peak rises modestly under retention, and that the structured
metrics' peaks fall modestly, each by the signed amount the career-change
reading implies. A winner that survived retention despite the misfit it adds in
the Skill and Hierarchy sub-regimes is not an artifact of the exclusion.
 
The case for CWTVD differs from the case for KL in what the estimation data
can and cannot show. The
resolving-power property is established under known data-generating
processes, as one establishes consistency or test power for an estimator
rather than reading it off the sample the estimator is later applied to.
It can be established only that way. The property is a claim about the
mapping from a known true mixture to the gap CWTVD reports between metrics,
and checking that mapping requires holding the truth fixed and reading off
the measurement, which presupposes knowing the truth. The estimation
sample is a single transition matrix drawn from one unknown mixture, so it
yields one gap and no reference mixture against which that gap's size could
be judged. No quantity in this sample, or in any single observed labour
market, can confirm or disconfirm the property. The simulation is not
preliminary evidence awaiting empirical confirmation but the only support
the claim admits. The simulation validates the instrument and the
estimation sample takes the measurement, and we make no prediction that
the Canadian data will confirm CWTVD's advantage. The data answers a
different question, which geometry governs Canadian mobility, and the
resolving-power property is what licenses trusting CWTVD's answer to that
question when it parts from KL's.
 
We compare peaks rather than improvement at any single $\varepsilon$
because of the same non-identification that motivated the grid above:
only the ratio $C_{ij}/\varepsilon$ enters $P^\star$, so a metric's
improvement at a fixed $\varepsilon$ has no interpretation without an
implicit normalization, and the peak is the scale-free summary that
removes the ambiguity. Each candidate traces an
interior maximum over the grid: at low $\varepsilon$ the plan
over-concentrates on the diagonal relative to the observed move rate, and
as $\varepsilon \to \infty$ it diffuses to the independence baseline
$P_{ij} = a_i b_j$, where improvement returns to zero. The grid from $0.1$
to $1000$ is wide enough that all three candidates peak in its interior on
each criterion, with the uninformative tails visible at each end. Binary
peaks in the interior too. As the one-parameter stay/move model described
in @sec-cost-metrics, its only lever is the aggregate stay/move split, so
its peak sits at the $\varepsilon$ where the diffusion of mass off the
diagonal matches the empirical move rate, and it peaks earlier than skill
and hierarchy, which keep improving as falling $\varepsilon$ sharpens an
off-diagonal allocation that pays off.
 
For KL the peak is more than a convenient scale-free summary. Since the
EROT plan is a multinomial-logit equilibrium and $\Delta KL$ is monotone in
the maximised log-likelihood, maximising $\Delta KL$ over the grid profiles
the likelihood over $\varepsilon$, the single unidentified scale parameter.
Peak $\Delta KL$ is therefore the profile-likelihood model-selection
criterion, and comparing peaks across metrics compares maximised
likelihoods. Peak $\Delta CWTVD$ carries no such derivation. CWTVD has no
likelihood for the peak to profile, so taking its maximum is the same
scale-free summary applied by analogy to KL, not an operation a
nuisance-parameter argument singles out: other scale-free summaries of the
$\Delta CWTVD$ curve would serve equally on that argument alone. What
licenses the peak for CWTVD is instead the resolving-power property
established in @sec-AC, that the peak $\Delta CWTVD$ gap between correlated
geometries tracks the true mixture more sharply than the KL gap does. The
justification for peaking CWTVD is thus empirical, resting on the
simulation that grounds the CWTVD case generally, where the justification
for peaking KL is analytic. We adopt the peak for both, holding them
co-equal, while noting the asymmetry in what each peak rests on.
 
The winner of the race is the cost matrix whose peak exceeds the nearest
rival's by at least 10\% on every criterion at once: peak $\Delta KL$, and
peak $\Delta CWTVD$ under each of the two informative judges. Requiring all
three is more selective than crowning the leader on any alone. A metric
wins only if the likelihood endorses its fit and the geometry-aware
criterion certifies it sound at both extremes of the blend family, so a
plan that puts mass in roughly the right cells but at the wrong distances,
scoring high on $\Delta KL$ and low on either geometric criterion, does not
win. Because the nearest rival can differ between the two judges, a winner
must lead whichever rival is closest under each. Three outcomes follow, and
we report whichever obtains:

\begin{enumerate}
  \item \textbf{Decisive win.} One metric leads the nearest rival by 10\%
  on all three criteria. The race resolves in its favour.
  \item \textbf{Endpoint split.} The two geometric criteria favour
  different metrics: one leads under $C_{\text{skill}}$-scoring, another
  under $C_{\text{hierarchy}}$-scoring. The winner is blend-dependent.
  Rather than a defect this is the design's most informative outcome, since
  a single blended judge would have crowned whichever metric led at the
  midpoint and concealed that the endpoints disagree. We report both, and
  the disagreement locates the mobility geometry of that regime as
  genuinely intermediate between the two rather than governed by either
  alone.
  \item \textbf{Unresolved.} No metric clears 10\% on all three criteria
  and the two geometric criteria do not cleanly favour different metrics.
  This covers both the case where the likelihood and geometry-aware
  criteria disagree about a single winner and the case where no candidate
  leads by 10\% on any criterion. We report which criteria favour which
  metric, since even an unresolved race describes how observed mobility
  departs from each candidate, but we name no governor.
\end{enumerate}
 
The 10\% threshold is an effect-size floor, set in advance and applied
uniformly to all three criteria. We treat two peak improvements differing
by less than ten percentage points as substantively equivalent regardless
of whether the difference is statistically distinguishable, on the grounds
that a smaller difference does not warrant naming one geometry the governor
of mobility in that regime. The floor is substantive, not statistical, so
it does not flex with the number of criteria: a gap below 10\% on a
geometric criterion is substantively unimpressive whether it stands alone
or beside the others, and correcting the bar for the count of criteria
would confuse an effect-size floor with a significance threshold. At the
sample sizes the linked census file provides, even the smallest sub-regime
holds thousands of transitions, so sampling noise in each peak gap is well
below this floor and the binding constraint is the substantive one. The one
regime where this could fail is the high-OES Binary sub-regime, whose low
mobility leaves its off-diagonal cells sparse. We therefore confirm at
access, by simulating a balanced skill/hierarchy mixture at that regime's
transition count, that a balanced mixture rarely produces a peak gap above
10\% on either geometric criterion; if it does, the Binary regime's race is
noise-limited and we interpret it as such.
 
An unresolved race is an informative no-decision rather than a coin to be
flipped. A disagreement between the likelihood-grounded and the
geometry-aware criteria, or between the two geometric endpoints, localizes
how a metric misfits, and breaking it by fiat would discard exactly the
signal the criteria were chosen to expose.

### Out-of-Regime Falsification {#sec-local-falsification}


The falsification scores $\Delta CWTVD$ under a single neutral cost matrix
rather than under the two informative judges used to adjudicate the race.
That matrix is $C$, the arithmetic mean of the three normalized cost
matrices $C_{\text{skill}}$, $C_{\text{hierarchy}}$, and
$C_{\text{binary}}$, in which no candidate geometry is privileged. The race
needs two judges because it resolves a close comparative contest between
correlated geometries, and bracketing the blend family is what certifies a
winner across it. The falsification needs the opposite: not a comparison
but a single fixed line that a predicted metric either crosses or does not.
A neutral blend supplies exactly one such line, and scoring the crossing
under two judges instead would reintroduce the grey zone the fixed line is
meant to remove. This is the one role for which the blend, retired from
adjudication, remains the right instrument.

The out-of-regime criterion is stated against independence, on this
neutral geometry-aware measure. In regime, the predicted winning metric,
$C_{\mathrm{skill}}$ in the Skill sub-regime and $C_{\mathrm{hierarchy}}$
in the Hierarchy sub-regime, peaks above independence on $\Delta KL$
and on $\Delta CWTVD$ under the neutral blend, since its geometry
rationalizes the transitions and improvement over independence is what the
horse race measures. Out of regime, where a new highest attainment resets
the worker's position and renders the 2016 occupation uninformative, that
metric is predicted to fall below independence on $\Delta CWTVD$.
 
The out-of-regime criterion rests on $\Delta CWTVD$ alone, and the reason
is that $\Delta KL$ cannot register the failure being tested. The
falsification asks whether the predicted metric has made the wrong
geometric bet, placing mass at the wrong distances. $\Delta CWTVD$ weights
each misallocation by distance and so penalises exactly this, falling below
independence when a short-move metric meets a movement-heavy matrix.
$\Delta KL$ is blind to geometry and scores any plan that captures the
diagonal stay/move structure, which every metric shares. A wrong-geometry
metric therefore collects partial KL credit for the diagonal even while its
off-diagonal bet is wrong, and its peak $\Delta KL$ can sit slightly above
independence rather than below it. This is visible in the simulations: in
the [online companion](https://rpmartin.github.io/not-one-thing-companion), under a binary data-generating process the skill and
hierarchy metrics peak below independence on $\Delta CWTVD$ but rise to a
small positive peak on $\Delta KL$ at moderate $\varepsilon$. Requiring the
geometry-blind measure to confirm a geometric failure would ask the wrong
instrument to register it. We therefore state the out-of-regime criterion
on $\Delta CWTVD$, the measure built to see what the falsification is about.
 
Setting the line at independence rather than at binary discharges the
conditionality that a binary benchmark would leave open. A wrongly
predicting metric always sits below binary out of regime, but whether its
$\Delta CWTVD$ also falls below independence depends on the non-local
transition matrix being movement-heavy: only when most credential-changers
move, and move far, does a short-move metric place enough mass on vacated
near pairs to fit worse than independence. We predict the non-local matrix
is movement-heavy, because the wormhole mechanism is precisely the claim
that a new credential drives long discontinuous jumps. The
below-independence criterion therefore tests two things at once. Its
crossing requires both that the predicted metric makes the wrong geometric
bet out of regime and that the non-local matrix is movement-heavy as the
mechanism predicts, so a metric whose $\Delta CWTVD$ falls below zero
jointly falsifies the local model out of regime and confirms the long-jump
prediction. A metric that stayed above independence out of regime would
either be governing where it should not or would reveal a non-local matrix
lighter than the mechanism requires, and either reading counts against the
design.
 
Independence is a fixed point, so the criterion is unambiguous. The
predicted metric's $\Delta CWTVD$ peak is either below zero or it is not,
with no margin to calibrate and no grey zone to adjudicate. This is the
respect in which the falsification differs from the race. The race resolves
a close contest between two correlated geometries by asking whether one
peak exceeds the other by 10\%, a margin set because the contestants are
near. The falsification asks whether a single metric crosses a fixed line,
which it either does or does not.
 
Peaking above independence in regime is a weaker bar than winning the
race, and the gap between them is real. A metric clears the in-regime side
of the falsification by improving on independence at all; it wins the race
only by leading every rival by 10\% on both measures. Both skill and
hierarchy peak well above independence in either of their sub-regimes,
since they are correlated and each partially rationalizes the other's
transitions, yet only one wins. Improving on independence is necessary for
a geometry to be governing at all; leading by 10\% on both measures is what
crowns it.
 
The mechanism that drives the predicted metric's $\Delta CWTVD$ below
independence is the same one that makes binary a useful point of
comparison. All three cost matrices assign zero self-distance,
$C_{ii} = 0$, so all three concentrate mass on the diagonal and capture the
dominant stay/move feature of the transition matrix. That diagonal
contribution is common to every plan, and what distinguishes the metrics is
the off-diagonal allocation. Binary makes no off-diagonal bet, since every
switch is equidistant under it, so it cannot be wrong about where movers
go. Skill and hierarchy do make that bet, and out of regime it is the wrong
one, because they place mass on geometrically proximate pairs that wormhole
movers avoid while the data concentrates on far pairs. A metric predicting
short moves into a movement-heavy matrix withholds mass from exactly where
the data sits, and $\Delta CWTVD$ penalises that misallocation by distance,
driving it below independence. The signature of this failure on
$\Delta CWTVD$ is a curve that rises monotonically toward the independence
baseline from below without crossing it, improving as $\varepsilon$ grows
only because diffusion toward independence is the one thing that helps a
metric whose geometry points the wrong way. The [online companion](https://rpmartin.github.io/not-one-thing-companion) shows this
under a binary data-generating process, where skill and hierarchy stay
below zero on $\Delta CWTVD$ across the grid while their $\Delta KL$, for
the diagonal-credit reason given above, rises instead to a small positive
peak at moderate $\varepsilon$, which is why the out-of-regime criterion
rests on $\Delta CWTVD$ rather than on both measures.
 
The sub-regime models are falsified the same way, with one exception. In
regime each sub-regime is expected to produce a metric that peaks above
independence on both measures, and out of regime that metric's $\Delta
CWTVD$ peak is predicted to fall below independence. We apply this to the
Skill and Hierarchy sub-regimes only. The Binary sub-regime is exempt from
the distance-based out-of-regime test for a reason internal to its own
hypothesis, and the exemption does not depend on what its out-of-regime
cell turns out to contain. Its in-regime claim is that no distance geometry
governs, so it never generates a prediction that distance ceases to govern
out of regime; there is no distance bet to be caught making wrongly, with
or without movers in the cell. Separately, the lock-in mechanism that
defines the sub-regime predicts that cell is near-empty, because workers
already in a high-OES occupation rarely acquire a new highest attainment,
since occupation-specific human capital is reached through a single
credentialing pipeline and confers no incentive to re-credential. A
non-empty cell therefore falsifies the lock-in premise without thereby
reinstating a distance test the hypothesis never licensed: the
distance-based leg stays inapplicable, and the non-emptiness is read as
evidence against lock-in in its own right. The out-of-regime content of the Binary hypothesis otherwise lies on the
entry side and is carried by the destination-OES coefficient in the
non-local specificity regression rather than by a horse race.

## Non-Local Model {#sec-non_local}

The non-local model estimates the conditional-on-positive relationship between how far a worker travels through occupation space and the pipeline strength of their transition, measured by lift. We measure that distance two ways, with different estimators and complementary roles. The primary outcome is $\log(d_{ij})$, the log skill distance between 2016 and 2021 occupations, estimated by OLS. The second is a binary indicator for whether the move crosses an occupation broad group, a silo, estimated by a linear probability model (LPM). The hierarchical metric is ordinal, not cardinal: its values, one through the cross-category sentinel of nine, are ranks with an assigned spacing rather than real distances, so they cannot enter an OLS outcome as a graded distance. Its distribution is also dominated by the cross-category sentinel, with within-silo values a thin minority, so the one contrast with mass on both sides is whether a move crosses a silo at all. The binary cross-silo indicator is therefore the operational form the metric supports for non-local jumps.

The cross-silo margin is the right hierarchical form because of where high-lift destinations sit. A high-lift move is by construction a move into a credential-fed occupation, and credential-fed occupations are concentrated in the skilled silos rather than in the generic entry silos such as sales and service, which contain few. The high-lift destinations within sales and service are a small and specific set of licensed or skilled-service occupations, funeral directors, chefs, insurance and real-estate agents, each fed by a dedicated credential, while the generic occupations that make up most of the silo are reached at close to baseline prevalence from every education. A high-lift credential earned by a worker in the generic part of such a silo therefore has almost no within-silo destination to channel into, so the move it opens is cross-silo by the structure of where credentials lead rather than by the length of the jump.

The mechanism is directional: a wormhole is a path a credential opens through occupation space. A specific education channels workers toward a narrow set of occupations that would otherwise be unreachable, and the long jump occurs when a worker travels along that channel into the occupation the credential distinctively feeds. This is a property of the education--destination pair, not of either alone, and one quantity measures it: the lift $m(e,n) = P(n\mid e)/P(n)$ of the destination occupation $n$ for the worker's education $e$, defined in @sec-cf as the cell-level term of the education-side specificity divergence. Read directly rather than aggregated, it makes the conjunction requirement automatic: a generic education offers no high-lift path to take, and a specific credential whose holder lands off its distinctive path realizes low lift wherever else they go, so only travel along a genuine pipeline registers as high. There is no need to build the conjunction from separate entrance and exit measures and no interaction term to specify; the pairing that a wormhole requires is exactly what lift measures.

This directness disposes of two cases that the separate-measures approach had to handle by construction. A worker who reaches a specific destination without the credential that feeds it arrives on general human capital, and lift reads their education--destination pair as low: the destination's specificity does not make it a wormhole for them, because their education does not lead there. And ability-driven search failure is handled the same way: two graduates with the same credential need not fare alike, one reaching the specific destination the credential opens and another, after a harder search, settling for a generic occupation nearby; the second has low destination lift and is correctly read as a short, non-pipeline move, entering the regression at its observed short distance rather than requiring a separate exclusion. The same low destination lift marks a relabelled or credential-unrelated move for reclassification to local under the wedge of @sec-rp, so the non-local regression and the wedge reclassification key on one quantity.

Let $y_{ij}$ denote the outcome ($\log(d_{ij})$ in skill space, estimated by OLS, or the cross-silo indicator, estimated by LPM), and let $\bar{d}_j$ be the destination's mean inbound distance under the independence baseline (its average distance from a randomly drawn origin; for the cross-silo outcome, the baseline probability that a random origin lies outside $j$'s broad group). The remoteness control is defined in the same space as the outcome it accompanies, so it swaps with the outcome: skill-space mean inbound distance for the OLS specification, baseline cross-silo probability for the LPM.

\begin{equation}
y_{ij} = \alpha + \beta\, \log(m(e_i, j)) + \gamma\bar{d}_j + \delta_i + \varepsilon_{ij}
\label{eq-lift}
\end{equation}

The specification includes origin occupation fixed effects $\delta_i$ and a destination remoteness control $\bar{d}_j$, but not destination fixed effects. The reason is structural, not one of feasibility. Lift is a property of the education--destination pair, so all workers who reach a given destination through the same education carry the identical lift; within a strong pipeline, where a destination is reached predominantly through one education, lift is very nearly constant, and the variation in distance among those workers comes from their differing origins, not from lift. Destination fixed effects would therefore identify the lift coefficient only from within-destination variation in lift, which is precisely what a pipeline lacks: the more exclusive the pipeline, the less within-destination lift variation it has, so destination fixed effects are blindest exactly where the mechanism is strongest. The identifying variation is between destinations, high-lift pipelines against low-lift ones, which is also what the wormhole claim is about. The remoteness control preserves that variation while netting out the one confound of concern, a destination lying far from all others in skill space, or in a small broad group most origins fall outside, which registers long jumps mechanically with no credential mechanism involved. Conditioning on $\bar{d}_j$ ensures the lift coefficient reflects credential-gated channeling rather than baseline remoteness; this is the conservative reading, since if remoteness is itself part of how specificity operates, the control absorbs some of the mechanism. Origin remoteness needs no separate term; $\delta_i$ absorbs it with every other origin-level determinant of distance.

Both outcome specifications share the lift coefficient $\beta$, expected to be positive: travel along a stronger education--occupation pipeline is longer, and more likely to cross a silo, since the pipeline is what carries a worker to a distant occupation the credential distinctively opens. The two coefficients read on a common per-1\%-of-lift basis, differing only in the units of their response. For the OLS skill-distance outcome, a 1\% increase in lift is associated with approximately a $\beta$ percent increase in distance, the log--log coefficient reading as an elasticity. For the cross-silo LPM, the same 1\% increase is associated with approximately a $\beta/100$ percentage-point increase in the probability of crossing a silo, the raw per-log-unit coefficient divided by 100 to convert it to a per-1\% effect. The remoteness coefficient $\gamma$ is expected to be positive. Standard errors will be clustered on education and destination occupation.

Age is excluded from the specification. The well-documented decline in occupational mobility with age [@topel1992job] is, for jump distance, a life-cycle composition effect that operates through the 2016 origin occupation: workers who acquire a new attainment later in life tend to start from a higher skill and hierarchy position than those completing the same credential early, so they sit closer to the credential's destination and travel a shorter measured distance. That channel is a property of where a worker starts, not of the wormhole mechanism, and the origin fixed effects $\delta_i$ absorb it along with every other origin-level determinant of distance. A separate age term is therefore neither needed nor clean: conditional on origin it carries no signal the mechanism predicts, and because age plausibly correlates with lift through selection into particular credential types, including it risks partialling out legitimate pipeline variation.

The model is evaluated across four samples as described in @sec-data, with results reported for both the skill and cross-silo outcomes. The within-$R^2$, the share of variance in the outcome explained by lift after partialling out origin fixed effects and destination remoteness, is not expected to be large in absolute terms: transition distance is inherently noisy and lift is an aggregate measure that captures only the systematic component of specificity-driven mobility. The relevant comparison is the ordering of within-$R^2$ across samples, not its absolute level.

The positive-distance restriction removes many apprenticeship completers, since NOC makes no distinction between apprentices and journeypeople of the same occupation: those already working in the target trade in 2016 show zero measured distance upon completion and are excluded by construction. Rather than threatening identification, this restriction improves measurement validity: the excluded zeros reflect occupation coarseness rather than true immobility, and their inclusion would introduce noise into the outcome variable. Workers who complete an entire apprenticeship program within the five-year window from a different 2016 occupation remain in the sample, and for these workers the wormhole mechanism is cleanly observable. The limitation is one of external validity rather than internal validity: estimates do not generalise to within-trade apprenticeship completers, whose human capital increment is real but invisible to NOC.

Beyond the apprenticeship case, the positive-distance restriction is also a selection on the outcome, but most of what it excludes is genuine immobility rather than a selection artifact. A true stayer contributes no distance to explain, so excluding stayers sets the scope of the conditional-on-positive estimand rather than selecting the surviving movers on anything, and the external-validity caveat above already covers it. The internal-validity concern is confined to the residual coarseness zeros, the real moves NOC records as zero distance, because OES predicts how often those occur as well as the distance conditional on moving. Insofar as the excluded zeros reflect coarseness rather than genuine immobility, the surviving positive subsample is selected on a margin the regressor also drives, with high-OES origins contributing their movers as a selected upper tail that the destination-side remoteness control does not absorb. The likely direction of that residual bias is attenuation, on the same logic as the local-sample leakage: the coarseness zeros are short measured moves, so resolving the coarseness would restore short high-OES moves that flatten the estimated specificity--distance gradient rather than steepen it. We therefore read the restricted estimate as attenuated, consistent with the lower-bound interpretation applied to the truncation in the limitations, and report it as such rather than as an unbiased effect.




### Out-of-Regime Falsification {#sec-non_local_false}

The non-local model is falsified in the local clean sample, where the highest
education attainment is unchanged. The criterion is the cross-sample degradation of the lift coefficient and the within-$R^2$, not their vanishing in absolute terms. Because lift enters on a common natural-log scale in every subsample, the $\beta$ estimate and the within-$R^2$ are directly comparable across subsamples, and the prediction is that both collapse from the non-local to the local sample. We do not require the local coefficient to be exactly zero: a high-lift education can compress local move distances weakly even absent a new credential, so the claim is degradation toward zero relative to the non-local estimate, consistent with the ordering-not-level standard applied to the within-$R^2$ throughout.

The collapse has two sources, and both are consistent with the wormhole
mechanism. Without a new highest attainment the wormhole channel is shut: the
worker's education is already reflected in their 2016 occupation, the starting
point of any local move, so it carries no additional predictive content for that
move, whereas the non-local education is acquired within the window and does its
repositioning where it can be observed. Second, high-lift transitions require a
worker to have moved into the occupation their credential distinctively feeds,
which without a new attainment is largely already reflected in the 2016
occupation, so high-lift realisations are truncated in the local sample and a
near-zero lift coefficient there reflects range restriction as much as a
switched-off mechanism. Neither reading threatens the falsification. A
spurious lift--distance relationship, one driven by a confound rather than
by credential-contingent enablement, would appear in both samples, since the
confound would not know which workers had changed education; its appearance in
neither the local coefficient nor anywhere a confound could hide is what marks
the relationship as credential-contingent rather than mechanical.

Two features of the local sample bear on its standing as a falsification sample.
It is large (most workers do not acquire a new highest attainment over the
inter-census period), so the null it delivers is precise rather than
underpowered, a well-powered absence of relationship rather than silence. It is
also subject to a modest contamination, because the partition variable is only a
proxy for the mechanism: what matters is whether the 2016 occupation is
informative, and "no substantial inter-census change in attainment" stands in for that.
Recorded highest attainment is in turn a lossy proxy for
the worker's actual position in human-capital space: any increment the record
fails to capture routes a genuinely repositioned worker into the local sample.
The failure takes several forms. A worker may earn a credential that does not
raise highest attainment, a sub-baccalaureate certificate, or a second field at
the same level, so that real training leaves the partition variable unmoved; a
nurse aide completing a practical-nursing certificate is one such case. Or the
timing may be wrong: a worker who earned their highest attainment shortly before
the 2016 reference week is formally local but holds a transitional rather than
credential-matched 2016 occupation, and so behaves like a non-local mover whose
jump straddles the census boundary. What unites these cases is not their cause
but their effect: the record misses a real increment, so the leakage cannot be
enumerated source by source, and the defense does not rest on cataloguing it.

The leakage is small, and the measurement-complete requirement removes one clear
class of it, excluding the graduates of intensive, occupationally specific programs, who
disproportionately lack a 2016 occupation at all. What matters more than any
filter is that whatever leakage remains biases toward the null, for a reason that
holds regardless of why the record failed: a leaked mover imports wormhole
behaviour into the local sample, degrading the horse-race fit and injecting
specificity into the regression, both of which work against the predicted
dissociation rather than producing it. This is the final reason the criterion is
stated as degradation and not as an absolute zero, which would be hostage to
precisely this contamination.

## Results \[PLACEHOLDER: awaiting data\]  {#sec-res}

### Local

\[Peak $\Delta KL$ and peak $\Delta CWTVD$ under each informative judge
($C_{\text{skill}}$, $C_{\text{hierarchy}}$) by cost metric, $\varepsilon$
and subsample; the two geometric panels are the bracket endpoints\]

\[Same, by 3 OES sub regimes, cost metric, $\varepsilon$ and subsample\]

\[Same, by 3 TEER sub regimes, cost metric, $\varepsilon$ and subsample\]

### Non-local

\[tie-fighter plot for the $\beta$\]

\[ table for the $R^2$\]

\[latent-activation robustness: non-local slope with reinstated latent
activators (@sec-AD) beside the uncorrected slope, across the lift-threshold
grid\]

## Limitations

### Local Mobility

Two sample restriction mechanisms introduce a potential selection 
concern for local mobility. Non-linkage and non-observability of 
occupation are both disproportionately concentrated among 
low-attachment workers, individuals with casual, seasonal, or 
intermittent employment whose identifying information is also less 
stable across census cycles, so the two filters largely co-select 
the same low-skill population rather than introducing independent 
exclusions. This shared exclusion affects the representativeness of 
population-level transition estimates (external validity) but does not 
threaten the internal validity of the horse race. For the horse race
results to be biased, the excluded workers' transitions would need to be 
systematically more or less rationalizable by a particular cost 
matrix than those of their observed counterparts in the same region 
of occupational space. No such mechanism is apparent: exclusion is 
driven by attachment instability and identifier churn, neither of 
which has a clear relationship to how well transitions conform to 
any particular cost structure conditional on occupation. The residual 
unlinked population among otherwise-observable workers, recent 
immigrants, young workers, and high-movers, similarly lacks any 
obvious occupational-space concentration that would differentially 
favour or penalize a particular metric.

A notable omitted factor for local mobility  is \emph{path-specific} wage differentials. The additively separable component of wage gains, recoverable from origin and destination wage levels alone, is absorbed by the 
margins insofar as high-wage destinations attract larger inflows and 
low-wage origins larger outflows. The residual dyadic component, 
reflecting transition-specific wage premia and discounts beyond what 
origin and destination wage levels predict, is an omitted factor. 
Short-distance transitions are within a similar occupational 
neighbourhood, reducing employer uncertainty about candidate fit and 
productivity; long-distance transitions cross larger skill or 
hierarchical gaps, and employers may discount wage offers to 
compensate for that uncertainty, or workers may accept lower wages 
to gain entry to a distant occupation.

Thus, path-specific wage differentials are better understood as an 
alternative mechanism, rather than a confounder. If employers use the same metric that governs mobility to assess candidate fit uncertainty, discounting wages for transitions that are long under the winning metric, that discount reinforces worker selection toward shorter transitions under 
the same metric. Both channels predict identical patterns of 
transitions and are equally consistent with the winning metric being 
the correct description of the mobility geometry. The horse race 
identifies which distance geometry best rationalizes observed 
transitions in a reduced-form sense, but cannot separately attribute 
that rationalization to worker or employer preferences regarding 
distance.

### Non-Local Mobility

The non-local regime studies non-negative operative increments, as defined in @sec-rp: transitions where the operative credential rises above the worker's prior ceiling or respecifies at the same level. Substitutive moves, where an over-credentialed worker retrains below their ceiling and repositions on a sub-maximal credential, lie outside the estimand and are invisible to a highest-attainment record; the findings therefore speak to additive and lateral specific-human-capital activation, not to occupational mobility of every kind. This is a scope of the question rather than a gap in the answer, and it is consistent with the paper's premise that mobility is not one thing: the substitutive channel is a distinct phenomenon with its own economics, deserving its own treatment rather than absorption into this one.

The non-local partition is restricted to individuals who acquire a new highest educational attainment between 2016 and 2021 and report an occupation in both periods. A 2016 occupation is observed if and only if the individual was employed between May 1st and 7th, the 2016 census reference week. For the 2015/2016 school year, roughly half of 18-29 year old students reported concurrent work, with an (unweighted) ten percentage point decrease for university students, consistent with more intensive programs suppressing concurrent employment.^[source: Statistics Canada table 37-10-0104-01.] The Census reference week provides a partial offset: falling after the typical post-secondary academic year ends, it captures individuals who have recently finished exams and may have already begun summer employment. But this offset is weaker for intensive professional programs that are either year-round or have students for whom summer employment is atypical. The resulting truncation is not random: missingness is concentrated among the most occupationally specific educations, precisely the credentials that anchor the highest-lift pipelines, where the wormhole mechanism is strongest and jumps are longest. Thus estimates are attenuated toward zero and should be interpreted as lower bounds on the true relationship, given that truncation is concentrated precisely where the signal is strongest. Notably, this missingness pattern is not purely a limitation: the disproportionate absence of 2016 occupations among the most specific credentials is itself consistent with the wormhole mechanism, since program intensity and specificity are precisely what suppresses concurrent employment. In the online companion we plot each education's occupational specificity against the share of its workers missing a 2016 occupation, quantifying the magnitude of this truncation.

```{r}
# #| label: tbl-working-while-studying
# #| tbl-cap: "Share of students employed during program, by age and institution type"
# 
#  tbbl <- cansim::get_cansim("3710010401")|>
#    janitor::clean_names()|>
#    mutate(age = str_extract(age, "\\d+"))|>
#    filter(ref_date=="2015/2016",
#           type_of_institution_attended %in% c("College","University"),
#           age %in% 18:30)|>
#    select(age, type_of_institution_attended, value)|>
#    pivot_wider(id_cols = age, names_from = type_of_institution_attended, values_from = value)
# 
#  tbbl_mean <- tbbl|>
#        summarise(age = "Unweighted Mean", across(c(College, University), mean))
# 
#  bind_rows(tbbl, tbbl_mean)|>
#    kbl(
#      digits = 1,
#      col.names = c("Age", "College (%)", "University (%)"),
#      booktabs = TRUE,
#      linesep = ""
#    ) |>
#    kable_styling(latex_options = c("HOLD_position")) |>
#   footnote(
#    general = "Statistics Canada Table 37-10-0104-01, 2015/2016.",
#    general_title = "Source:",
#    threeparttable = TRUE,
#    escape = FALSE
#  )|>
#    row_spec(nrow(tbbl)+1, bold = TRUE)  # bold mean row
```


The specificity regression faces an analogous identification 
challenge to the horse race. Lift is observed and enters the 
regression, but cannot distinguish between two mechanisms that 
predict identical patterns. A specific education opens doors to 
particular distant occupations that a general education does not: 
the human capital match is narrow, so only a subset of distant 
destinations become reachable following attainment (enablement). At 
the same time, workers with specific educations may face stronger 
incentives to seek out distant occupations where those skills are 
scarce and rewarded, making long-distance moves financially attractive 
relative to staying within a crowded local neighbourhood (motivation). 
Because both channels operate through the same pipelines that lift 
measures, the reduced-form estimate conflates the two and cannot be separately 
attributed to either. The regression should therefore be interpreted 
as identifying the reduced-form predictive relationship between 
pipeline strength and distance traveled, without attribution to a specific 
mechanism, though the estimate remains informative: if either 
channel is operative, the regression will detect it.

## Conclusion

- What we found \[PLACEHOLDER: awaiting data\]

The central policy takeaway from this analysis is that labour market adjustment mechanisms are not one-size-fits-all: the appropriate intervention depends on which mobility regime contains the occupation experiencing a shortage. The three mobility sub-regimes suggest three distinct policy levers.

In the Skill sub-regime, reskilling and short-cycle training are the natural adjustment mechanisms. Workers in adjacent occupations already possess much of the relevant task content, and targeted reskilling can accelerate the reallocation of labour toward occupations in shortage without requiring formal retraining. The policy implication is that investment in modular, portable training that builds on existing skill profiles will generate the largest mobility response in this segment.

In the Hierarchy sub-regime, the relevant bottleneck is not skill transferability but the hierarchy's silo and platform structure. An occupational shortage within one silo cannot be addressed by upskilling workers from a different silo, even if the occupations share a near-identical skill profile. Instead, shortages must be addressed within-silo, through either lateral movement on a platform or vertical movement between platforms. The best way to address a shortage at a higher platform is by increasing the supply of workers to lower platforms, freeing incumbents to pursue upward movement through the silo, the occupational analogue of the internal labour market adjustment mechanism described by @doeringer1971internal. Bridging programs, mentorship structures, and targeted recruitment into within-silo entry-level positions are the appropriate instruments.

In the Binary sub-regime, either a worker has the required education, or they do not. Neither reskilling nor within-silo movement can substitute for education. Adjustment in this sub-regime depends entirely on expanding formal training pipelines: increasing capacity in the relevant educational programs, reducing barriers to program entry, and shortening time-to-completion where possible. These are also the occupations that immigration policy should target, both in terms of admission and rapid credential recognition. 

If the hypothesized sub-regime structure is confirmed, the geometry of occupational mobility determines which policy lever to pull, and the wrong lever is not just ineffective but potentially counterproductive. 

```{=latex}
\appendix
```


## O\*NET to NOC 2021 Skill Mapping {#sec-AA}

OaSIS provides occupation-level skill profiles for Canadian
classifications and would be an attractive data source for constructing
$C_{\text{skill}}$. However, OaSIS does not expose how those profiles
are constructed: the mapping between O\*NET and NOC 2021 occupations is not observable, the weights linking underlying occupations are not published, and there is no information on how multiple O\*NET occupations are combined to form an occupation profile. This appendix describes an alternative mapping constructed explicitly from published crosswalks, allowing full transparency of inputs and weights, diagnostic evaluation of mapping quality, and reproducibility of all outputs.

The mapping chains four classifications: O\*NET 2019 $\rightarrow$ SOC
2018 $\rightarrow$ NOC 2016 $\rightarrow$ NOC 2021. Each stage introduces potential one-to-many relationships. A naive approach would assign equal weight to all admissible paths, but this leads to over-weighting of occupations that appear frequently in crosswalk chains and inclusion of unrelated occupations. Instead, path weights are constructed to reflect the degree of splitting at each stage: if a source occupation splits into $n_s$ destination occupations at stage $s$, each path through that stage receives weight $w_s = 1/n_s$, and the overall path weight is $w_{\text{path}} = w_1 \times w_2 \times w_3$. Path weights are then aggregated by origin O\*NET occupation and destination NOC 2021 code and normalized to sum to one within each NOC to yield base weights. The consequences of naive versus path weighting can be severe: for NOC 72600 (Air pilots, flight engineers and flying instructors), naive weights assign five-eighths of the skill profile to agricultural workers, while path weights assign 78\% to pilots.

A power transformation and renormalization is then applied to further
reduce the influence of mapping artifacts: $\tilde{w}_k \propto w_k^2$.
This concentrates weight on the dominant O\*NET contributors and
diminishes the influence of peripheral occupations that enter through
crosswalk chains with many splits. For NOC 72600, downweighting
increases the pilot share of the skill profile to 96\%. Each O\*NET
occupation is then represented as a point in a reduced skill space using the first 12 principal components of the O\*NET skill content variables, and for each NOC 2021 code the weighted centroid of its contributing O\*NET occupations is computed using the downweighted mapping, yielding a 12-dimensional skill profile for each NOC occupation.

Two diagnostics assess the quality of each occupation's mapping in skill space. Herfindahl concentration measures how concentrated the mapping weights are across contributing O\*NET occupations: $H = \sum_k \tilde{w}_k^2$. For example, a Herfindahl of $\frac{2}{3}$ indicates that the occupation skill profile is effectively dominated by a single O\*NET occupation with a weight of $\approx .8$.  In contrast, when $H=\frac{1}{3}$ the largest contributor's share is between $\frac13$ and $\sqrt{\frac13}=.58$. Scaled Similarity measures the internal coherence of the contributing occupations in skill space: the weighted average distance of each contributing O\*NET occupation from the occupation centroid is computed, then normalized to a $[0.05, 1]$ interval by dividing by the maximum value across all occupations, multiplying by 0.95, and subtracting from 1, so that $SS = 1 - 0.95 \times \bar{d} / \max_j \bar{d}_j$. The lower bound of 0.05 is chosen to match the approximate lower bound of the Herfindahl index, making the two measures' floors commensurable. The mapping quality composite is the product $H \times SS$, which is high only when the mapping is both concentrated and internally coherent. The three interpretation cases are: high $H$ with any $SS$ (no problem, almost all weight on a single O\*NET occupation); low $H$ with high $SS$ (no problem, diffuse but coherent); and low $H$ with low $SS$ (problematic, a diffuse average of disparate occupations whose skill profile is unreliable). NOC 41200 (University professors and lecturers) is a canonical example of the third case.

Occupations in the bottom decile of $H \times SS$ are excluded
from the local clean and non-local clean samples. Because the exclusion criterion is based on mapping quality rather than occupation characteristics, it
introduces noise into $C_{\text{skill}}$ selectively while leaving
$C_{\text{hierarchy}}$ and $C_{\text{binary}}$ unaffected, a property exploited in the mapping quality falsification design described in @sec-data. The full mapping pipeline is implemented in
[@onet_to_noc] and is constructed entirely from O\*NET data and published crosswalks.

Because NOC is in places a finer classification than O\*NET, some distinct NOC unit groups inherit the same O\*NET profile and so sit at zero skill distance from one another. Grouping occupations by shared profile yields 36 such groups, spanning 97 occupations and accounting for the 280 zero off-diagonal cells noted in @sec-cost-metrics: 27 pairs, three triples, three quadruples, two quintuples, and one group of twelve. The group of twelve is the processing and manufacturing supervisors, which NOC distinguishes by industry (supervisors of food and beverage processing, of plastic and rubber products, of motor vehicle assembling, and so on) but which O\*NET represents as a single first-line production supervisor occupation, onto which all twelve map with a Herfindahl concentration of one. Their shared profile is therefore not a symptom of noisy mapping but the opposite: the twelve are among the most cleanly mapped occupations in the matrix, and their zero mutual distances reflect O\*NET's occupational resolution rather than any deficiency in the crosswalk. Eleven of the twelve share the same broad category, where the hierarchical metric also places them close (distance one or two), so the two geometries agree that these occupations are near. The twelfth, 72022 (supervisors, printing and related occupations), sits in a different broad category, so the hierarchical metric assigns it the cross-category sentinel of nine to each of the other eleven while the skill metric assigns zero. This is the widest disagreement the two geometries admit, on a set of occupations the horse race assigns to the Hierarchy sub-regime; whether the skill reading or the hierarchical reading better rationalizes the observed transitions among these supervisors is precisely what the local model adjudicates. Of the 97 occupations in these groups, four fall in the bottom mapping-quality decile and are dropped from the estimation samples; these are diffusely mapped twins, occupations that share a mixture of O\*NET sources rather than a single one, and the mapping-quality filter removes them for that diffuseness independently of their twinning. The remainder, including all twelve supervisors, are retained. We retain the zeros throughout, since flooring them would assert a skill distinction the O\*NET data does not support; as a robustness check, the zeros can be floored at the minimum resolvable skill distance and the horse race re-run.

## Lift construction {#sec-AB}

Lift $m(e,n) = P(\text{NOC}=n\mid \text{education}=e)/P(\text{NOC}=n)$ is estimated from the education$\times$NOC table (98-10-0404-01), and two features of that table make the raw ratio unstable in thin cells: the conditional $P(n\mid e)$ is a sample proportion, noisy when few workers hold education $e$, and the published counts are randomly rounded to base five, so a small cell's value is dominated by rounding. We address the two separately.

For the sampling noise, we smooth the conditional with a Dirichlet prior centred on the marginal occupation distribution:
$$\hat P(n\mid e) = \frac{c_{en} + \alpha\, P(n)}{T_e + \alpha},$$
where $c_{en}$ is the count of education-$e$ workers in occupation $n$, $T_e = \sum_n c_{en}$, and $\alpha$ is the prior strength in pseudo-workers. The prior is the marginal $P(n)$, so an education with no information is pulled toward "no pipeline" ($m \to 1$), and the pull is proportional to ignorance: a thin education ($T_e$ small relative to $\alpha$) is shrunk hard, a well-populated one ($T_e \gg \alpha$) is left essentially untouched. Because the prior is the same $P(n)$ that appears in the denominator of lift, the shrinkage target is exactly the ratio's neutral point, and a fully shrunk cell has lift one by construction. We shrink only the numerator; the denominator $P(n)$, estimated from the full table, is left raw, so that genuine pipelines into rare occupations, where a small $P(n)$ produces a large lift, are preserved rather than flattened. We set $\alpha = 15$, chosen so that the ranking of the strongest pipelines is stable: raising $\alpha$ far higher, to 1000, leaves the top of the ranking unchanged, since those pipelines are driven by the raw denominator rather than by numerator noise, while $\alpha = 15$ leaves well-populated cells such as optometry (education total near seven thousand) shrunk by under a percent.

For the rounding, we require a minimum realised cell count. A cell with $c_{en} = 15$ has a relative standard error near $1/\sqrt{15} \approx 26\%$ from sampling alone, above Statistics Canada's own "use with caution" threshold, and base-five rounding adds error of the same order; below roughly this count the estimated lift reflects noise and rounding rather than a real pipeline. We therefore compute lift only for cells with $c_{en} \ge 30$, where the sampling relative standard error is near eighteen percent, at Statistics Canada's caution threshold, and rounding error is under ten percent of the count. Cells below the floor are treated as unmeasured rather than as low-lift, since the ratio there is not a reliable estimate of anything. This is a reliability standard on the construction of the regressor, not an analytical choice about the estimation sample, and it is fixed in advance on the public table, blind to any transition distance. As a check on the floor value, the non-local estimates are reported across floors of ten, twenty, thirty, and fifty; every such specification applies a reliability floor, and the unfiltered ratio, which is a mismeasurement rather than a weaker measurement, is not among them.

Lift enters the regression in natural logarithm. The log is the natural scale for a multiplicative quantity whose neutral point is one: equal ratios become equal distances, the long right tail of rare-destination pipelines is compressed onto a comparable scale with ordinary moves, its neutral point at $\log m = 0$ marks a pipeline no stronger than baseline, and paired with the log outcome the coefficient reads as an elasticity. Because the same natural-log scale is applied in every subsample, the coefficient and within-$R^2$ are directly comparable across them. The Dirichlet floor guarantees the smoothed lift is strictly positive, so the log is always defined.

## CWTVD resolving power  {#sec-AC}

@sec-local adjudicates the horse race on the likelihood criterion and the
geometry-aware criterion, the latter scored under each informative judge,
and a win requires leading on all of them. The role of $\Delta CWTVD$ as a
co-equal geometry-aware criterion rests on a property established here by
simulation: among the two correlated informative geometries, $\Delta CWTVD$
resolves which governs more sharply than $\Delta KL$ does. This appendix
substantiates that property and states why it can be established only by
simulation.
 
The [online companion](https://rpmartin.github.io/not-one-thing-companion#sec-higher-resolution) plots improvement relative to
independence against $\varepsilon$ for each candidate metric, across three
mixed data-generating processes (columns) and the three criteria (rows):
$\Delta CWTVD$ scored under $C_{\mathrm{hierarchy}}$, $\Delta CWTVD$ scored
under $C_{\mathrm{skill}}$, and $\Delta KL$. Read down a column, the two
$\Delta CWTVD$ rows agree on which geometry peaks highest in every mixture:
skill in the skill-heavy $75/25$ column, hierarchy in the hierarchy-heavy
$25/75$ column, whichever the true mixture favours at $50/50$. The winner
is the same under either scoring endpoint, so a metric that leads under one
informative judge leads under the other and, by the bracketing argument in
@sec-local, under every blend between them. This agreement is the
robustness the two-endpoint criterion is meant to certify, not a
degeneracy: the endpoints coincide on the winner while remaining far enough
apart in construction, $\mathrm{corr}(C_{\mathrm{skill}},
C_{\mathrm{hierarchy}}) = 0.3$, that coincidence is informative.

The contrast with $\Delta KL$ is the resolving-power property. Under either
$\Delta CWTVD$ scoring the gap between the skill and hierarchy peaks tracks
the imbalance of the true mixture, widening as the mixture departs from
balance. $\Delta KL$ holds the two curves close to coincident across all
three mixtures, resolving the imbalanced cases only weakly. A second
difference appears at the benchmark: under every mixture the $\Delta KL$
curve for $C_{\mathrm{binary}}$ retains a positive tail well past its peak,
the same geometry-blind credit for the aggregate stay/move split that
drives the out-of-regime KL behaviour in @sec-local-falsification; both
$\Delta CWTVD$ rows strip that credit out and the $C_{\mathrm{binary}}$
tail collapses to the independence line. Both features point the same way:
$\Delta CWTVD$ separates the two correlated geometries more sharply and
denies the benchmark credit it has not earned geometrically.

The simulations also exhibit the wrong-bet penalty that the out-of-regime
falsification relies on. In the $25/75$ column scored under
$C_{\mathrm{hierarchy}}$, skill is the minority geometry graded by its
rival's judge, and its peak falls below the $C_{\mathrm{binary}}$
benchmark; scored under $C_{\mathrm{skill}}$ it recovers above binary. A
metric making a geometric bet is penalised by distance for making the wrong
one, where binary, making no off-diagonal bet, cannot be. This is the
in-sample analogue of the below-independence crossing predicted in
@sec-local-falsification, with binary in place of independence as the
reference, and it corroborates the mechanism under known
data-generating processes rather than testing it.
 

This resolving-power property is established under known data-generating
processes rather than read off the estimation sample, for the structural
reason given in @sec-local: the property is a statement about the mapping
from a known true mixture to the measured gap between criteria, and a
single observed labour market delivers one gap drawn from one unknown
mixture, with no reference mixture against which its magnitude could be
judged. The Monte Carlo is therefore the complete justification the claim
admits, not preliminary evidence awaiting empirical corroboration. The
figures here substantiate the property; @sec-local states what follows for
trusting $\Delta CWTVD$ when it parts from $\Delta KL$ on the Canadian
data.
 
## Latent-activation reinstatement {#sec-AD}

The local/non-local partition routes on the recorded education change, a proxy
for whether a large human-capital increment became operative within the window
(@sec-rp). Latent activation is the proxy's known failure: a worker whose
standing highest attainment was already in place in 2016 may activate it within
the window with no recorded change, and so routes local though the mechanism is
non-local. This appendix identifies those workers from what is observed and
reinstates them into the non-local sample as a robustness check on the
specificity--distance estimates. It is a correction to measurement error in the
partition variable, not a headline result, and the reinstated estimate is
reported only beside the uncorrected one.

The reason the exercise is worth running, rather than signing the selection by
argument, is that the sign is genuinely ambiguous a priori. The non-local sample
loses whichever latent activators failed to produce a recorded change, and which
workers those are depends on what makes a credential go latent. Under a search
channel, latency tracks the search itself: a graduate who takes the first offer
activates promptly, their move coincides with the recorded credential, and they
stay non-local, while a graduate who holds out for a distant match activates
late, often only after the credential has aged into a prior cycle as recorded
attainment, so their move shows no change and drops to local. Here the droppers
are disproportionately the long jumps, and their loss selects the non-local
sample toward shorter moves, attenuating the specificity--distance slope. Under
a quality channel the sign reverses: if latency reflects worker quality, with
high-quality graduates activating both fast and far while low-quality graduates
search long and settle near, the droppers are the short movers, and their loss
selects the sample toward longer moves, inflating the slope. The quality channel
is not even internally settled, since quality may as easily buy the patience to
hold out for a distant match, returning it to the search channel's sign. No
argument from the record resolves this, but the reinstatement does, because the
dropped workers and their distances are observed.

The signature of activation is a within-worker change in whether the occupation
uses the standing credential, with the credential held fixed. We read
credential--occupation match by the same lift $m(e,n) = P(\text{NOC}=n \mid \text{education}=e)/P(\text{NOC}=n)$ that serves as the non-local regressor, so the wedge reclassification and the non-local model are built on one quantity. Match is $m$ high, non-match $m$ low, with $m = 1$ the baseline at which an education feeds an occupation no more than the occupation's overall prevalence would predict.

A worker with unchanged education $e$ is reinstated when the credential is
dormant at the origin and operative at the destination:
$$m(e, n_{2016}) < 1 \quad\text{and}\quad m(e, n_{2021}) - m(e, n_{2016}) \ge \tau .$$
The first condition fixes the origin as genuinely non-matching, below-baseline
lift, and is a floor held fixed. The second declares the activation switch as a
lift increase across the window of at least $\tau$, the single pre-committed
threshold. The floor conditions only on the 2016 occupation, which does not
enter the specificity equation; the increase condition uses the 2021
occupation, which is destination-side and so shares information with destination
OES, a regressor. Membership is the contrast of the two, so workers reaching the
same 2021 occupation are separated on their 2016 state, but the destination leg
is not wholly clean and the sensitivity to $\tau$ and to the floor is what
certifies the correction is not manufactured by it.

We report the reinstated count alongside the reinstated non-local slope, beside
the uncorrected slope, across a grid of $\tau$ with the pre-committed value
marked. The threshold is fixed on the definition of activation, a genuine
dormant-to-operative lift switch, and is not tuned toward moving the slope;
moving it is neither the aim nor a success condition. The maintained expectation
is the opposite: latent activation is a leak we expect to be small, so a small
reinstated count together with a slope stable against the uncorrected estimate
is the affirmative outcome, evidence that the record-change partition is not
materially contaminated by this failure. A slope stable across the grid, and
across tightening the 2016 floor below one, indicates the correction does not
turn on the destination leg or on the threshold; a slope that moves only at
permissive $\tau$ locates the sensitivity; a large reinstated count that moved
the slope substantially would instead be evidence the partition had been
materially wrong, the one outcome that would count against the uncorrected
estimate. Because the count is reported rather than engineered, a small count is
read as latent activation being empirically rare, itself consistent with the
mechanism, since a worker already holding their highest credential is unlikely
to leave it dormant and then activate it within a five-year window.

Where the reinstated count is large enough to be informative, the observed
distances of the reinstated workers also resolve the sign left ambiguous above:
whether latent activators are the long jumps a selective search strands in the
local sample, or the short jumps a quality gradient leaves behind, is read
directly from whether their reinstatement raises or lowers the non-local slope.
Where the count is small, that sign is left unresolved, but immaterially so,
since a leak too small to move the estimate does not need its direction fixed.
 
## References
