## Education Specificity



# Step 0 --- Load and clean the table

We keep only the meaningful dimensions and drop the attainment aggregate.

``` {r, message=FALSE}
cip_noc <- bcgovpond::read_view("9810040401.csv") |>
  select(highest = starts_with("Highest"),
         cip     = starts_with("Major"),
         noc     = starts_with("Occupation"),
         value   = VALUE) |>
  filter(highest != "Total - Highest certificate, diploma or degree")
```

------------------------------------------------------------------------

# Step 1 --- Overall occupational distribution

We compute the overall occupation distribution across all education groups in the table. This represents the expected occupational outcome for an individual drawn from the educated workforce without conditioning on field of study.


``` {r}
p0 <- cip_noc |>
  summarise(total = sum(value), .by = noc) |>
  mutate(p0 = total / sum(total)) |>
  select(noc, p0) |>
  tibble::deframe()
```

------------------------------------------------------------------------

# Step 2 --- Distribution within each education bundle

For each (CIP × attainment):

-   Compute total workers T
-   Compute occupation shares p
-   Compare to baseline p0

``` {r}
educ_specificity <- cip_noc |>
  filter(value > 0) |>
  add_count(highest, cip, wt = value, name = "T") |>
  mutate(
    p  = value / T,
    p0 = p0[noc],
    kl_part = p * log(p / p0)
  ) |>
  summarise(
    KL = sum(kl_part),
    T  = first(T),
    .by = c(highest, cip)
  )
```

Interpretation:

-   KL ≈ 0 → looks like average worker → general education
-   Large KL → maps to unusual occupations → specialized

------------------------------------------------------------------------

# Step 3 --- Remove size effect

Small groups appear artificially specialized because small samples look
extreme. We remove this statistical artifact by modeling expected
divergence as a function of size.

``` {r}
fit_kl <- lm(log(KL) ~ log(T), data = educ_specificity)

educ_specificity <- educ_specificity |>
  mutate(
    specificity = log(KL) - predict(fit_kl)
  )
```

Now the index measures:

> How distinct the occupation mix is *relative to what we expect given
> sample size*

------------------------------------------------------------------------

# Step 4 --- Diagnostic plot

A valid index should not mechanically depend on size.

``` {r}
ggplot(educ_specificity, aes(T, specificity)) +
  geom_point(alpha = .3) +
  scale_x_log10(labels=scales::comma)
```

After removing size effects, dispersion increases with group size because larger education groups reveal true differences in occupational mapping, while small groups contain little information beyond sampling noise.


------------------------------------------------------------------------

# Interpretation

Positive values:
:   Education channels workers into unusually distinct occupations
    (pipeline)

Near zero:
:   Typical mapping for an education of that size

Negative values:
:   Diffuse/general education

------------------------------------------------------------------------

# Key insight

Traditional concentration measures fail because occupational distributions are estimated from finite counts, and small probabilities are either poorly measured or suppressed. These measures therefore depend mechanically on sample size and the thickness of the unobserved tail.
KL divergence instead depends on where the probability mass is located rather than how much mass lies in tiny cells. Residualizing by group size removes the remaining small-sample effect.

This yields a size-robust measure of occupational specificity rather than a disclosure-specific correction.


----------------------------------------------------------------------

# Specificity data

```{r}
educ_specificity|>
  arrange(desc(specificity))|>
  my_dt()
```
