---
title: "Online Companion to *Occupational Mobility Is Not One Thing: Evidence from the Canadian Labour Market*"
author: "Richard Martin"
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    toc-title: "Contents"
    code-fold: TRUE
    smooth-scroll: true
    self-contained: true
    theme: flatly
    page-layout: full
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(janitor)
library(plotly)
library(gt)
library(DT)
library(conflicted)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
source(here("R", "other.R"))
```

```{r load-data, include=FALSE}
k_vs_spearman    <- read_rds(here("out", "k_vs_spearman.rds"))
hier             <- read_rds(here("out", "hier_dist_raw.rds"))
noc_edu_spec  <- read_rds(here("out", "noc_edu_spec.rds"))
calibration      <- read_rds(here("out", "calibration.rds"))
skills_mat       <- read_rds(here("out", "skill_dist_raw.rds"))
sim_score_plots <- read_rds(here("out","sim_score_plots.rds"))
compare_lift <- read_rds(here("out","compare_lift.rds"))

skills_hc      <- hclust(as.dist(skills_mat), method = "average")
skills_ord     <- skills_hc$order
skills_mat_ord <- skills_mat[skills_ord, skills_ord]
skills_xlab    <- as.character(colnames(skills_mat_ord))
skills_ylab    <- as.character(rownames(skills_mat_ord))

dest_plt <- noc_edu_spec |>
  pivot_longer(cols = contains("Based on"), names_to = "criteria", values_to = "Sub Regime" )|>
  ggplot(aes(T, specificity, text = noc, fill = `Sub Regime`)) +
  scale_fill_viridis_d() +
  geom_point(shape = 21, size = 3, stroke = .5) +
  scale_x_log10(labels = scales::comma) +
  facet_wrap(~criteria) 


dest_spear <- round(cor(noc_edu_spec$T, noc_edu_spec$specificity, method = "spearman"), 2)
dest_pear  <- round(cor(noc_edu_spec$T, noc_edu_spec$specificity), 2)

```

## Overview {#sec-overview}

This companion is designed to ensure transparency and reproducibility for the paper
*Occupational Mobility Is Not One Thing: Evidence from the Canadian Labour Market*.
It provides technical detail on distance metrics, normalization, and specificity measures.

**Contents**

- **Skill distance**: O\*NET-based occupational skill vectors, dimensionality reduction
  via PCA, and the resulting continuous distance matrix.
- **Hierarchical distance**: Taxonomy-based discrete distances derived from the NOC
  structure, resulting in labour market silos with internal TEER platforms.
- **Cost matrix normalization**: Percentile-based anchoring across distance metrics.
- **Occupation's education specificity**: A size-adjusted measure of concentration in
  educational inflows into occupations.
- **Education's occupation specificity**: A size-adjusted measure of concentration in
  occupational outcomes across fields of study and attainment levels.

---

## Skill Distance {#sec-skill}

We use 161 occupational characteristics (Skills, Abilities, Knowledge, Work Activities)
from O\*NET to measure the distance between occupations. Because occupational skill
vectors are high-dimensional and highly correlated, Euclidean distances computed in the
full feature space may suffer from distance concentration. We therefore project the data
onto the 12 leading principal components, which achieve a 99% rank correlation of
pairwise distances relative to the full 161-dimensional space. This reduces noise from
low-variance directions while preserving the dominant geometry of the skill space.

### PCA Rank Preservation {#sec-skill-pca}

The figure below shows the Spearman rank correlation between pairwise distances in the
$k$-dimensional PCA space and the full 161-dimensional space, as a function of $k$. The
dashed line marks the 0.99 threshold; 12 components suffice.

```{r skill-pca, warning=FALSE}
ggplot(k_vs_spearman, aes(x = k, y = spearman)) +
  geom_hline(yintercept = .99, lty = 2) +
  geom_line() +
  geom_point() +
  scale_x_continuous(limits = c(0, 50)) +
  labs(
    title = "First 12 PCs achieve ≥0.99 rank preservation of pairwise distances.",
    y = "Spearman correlation (distance ranks)",
    x = "Number of PCs"
  ) +
  theme(text = element_text(size = 10))
```

### Raw Skill Distance Heatmap {#sec-skill-heatmap}

Occupations are ordered by AGNES. Diagonal entries are self-distances (zero). Dark clusters along the diagonal indicate groups of
occupations with similar skill profiles. Use left mouse button drag to zoom. Raw distances are normalized prior to use: [Cost Matrix Normalization](#sec-norm) 

```{r skill-heatmap}
#| fig-width: 5.6
#| fig-height: 8
plotly::plot_ly(
  x = skills_xlab,
  y = skills_ylab,
  z = skills_mat_ord,
  type = "heatmap",
  hovertemplate = paste(
    "<b>Origin:</b> %{y}<br>",
    "<b>Destination:</b> %{x}<br>",
    "<b>Distance:</b> %{z:.3f}",
    "<extra></extra>"
  )
) |>
  plotly::layout(
    xaxis = list(type = "category", showticklabels = FALSE, ticks = ""),
    yaxis = list(type = "category", showticklabels = FALSE, ticks = "")
)
```

---

## Hierarchical Distance {#sec-hier}

Hierarchical distances encode institutional and career barriers implied by the NOC
taxonomy. Distance between any two occupations is defined as follows:

1. All five digits match: distance = 0
2. First four digits match: distance = 1
3. First three digits match: distance = 2
4. First digit matches: distance = 3 + |ΔTEER|
5. Otherwise: distance = 9

The Spearman correlation between skill and hierarchical distances is 
`r round(cor(as.vector(skills_mat[upper.tri(skills_mat)]), as.vector(hier[upper.tri(hier)]), method = "spearman"), 3)`,
indicating that institutional proximity and skill similarity are positively but only
modestly related. The two metrics encode overlapping yet distinct occupational geometries.

### Distance Counts {#sec-hier-counts}

```{r hier-counts}
hier_count_plot <- table(hier) |>
  enframe() |>
  ggplot(aes(name, value)) +
  geom_col(alpha = .5) +
  scale_y_continuous(trans = "log10", labels = scales::comma) +
  theme_minimal() +
  labs(x = "Normalized Distance", y = "Counts")

plotly::ggplotly(hier_count_plot)
```

### Raw Hierarchical Distance Heatmap {#sec-hier-heatmap}

Occupations are ordered by AGNES. The block-diagonal structure reflects the ten NOC broad occupational categories (silos); yellow off-diagonal
entries are cross-silo pairs assigned the sentinel distance of 9.  Raw distances are normalized prior to use: [Cost Matrix Normalization](#sec-norm). Use left mouse button drag to zoom.

```{r hier-heatmap}
#| fig-width: 5.6
#| fig-height: 8
hier_hc      <- hclust(as.dist(hier), method = "average")
hier_ord     <- hier_hc$order
hier_mat_ord <- hier[hier_ord, hier_ord]
hier_xlab    <- as.character(colnames(hier_mat_ord))
hier_ylab    <- as.character(rownames(hier_mat_ord))

plotly::plot_ly(
  x = hier_xlab,
  y = hier_ylab,
  z = hier_mat_ord,
  type = "heatmap",
  hovertemplate = paste(
    "<b>Origin:</b> %{y}<br>",
    "<b>Destination:</b> %{x}<br>",
    "<b>Distance:</b> %{z:.3f}",
    "<extra></extra>"
  )
) |>
  plotly::layout(
    xaxis = list(type = "category", showticklabels = FALSE, ticks = ""),
    yaxis = list(type = "category", showticklabels = FALSE, ticks = "")
  )
```

---

## Cost Matrix Normalization {#sec-norm}

The normalization anchor is hierarchical distance 3 — the maximum within-silo horizontal distance, representing movement within a broad occupational group without crossing TEER levels. This anchor is theoretically motivated: any hierarchical distance above 3 involves either a TEER step or a cross-category transition, meaning distance 3 is the natural boundary between horizontal and vertical mobility within the silo structure. Hierarchical distances are divided by 3; skill distances are divided by 7.145, the value at the same unconditional percentile of the skill distance distribution. 

```{r norm-table}
calibration |>
  gt() |>
  fmt_number(columns = c(h_anchor, p_uncond, s_anchor), decimals = 3) |>
  cols_label(
    h_anchor = "Hierarchical Anchor",
    p_uncond = "Unconditional Percentile",
    s_anchor = "Skill Anchor"
  ) |>
  tab_options(table.font.size = "90%", data_row.padding = px(4)) |>
  tab_header(title = md("**Calibration Anchors Across Distance Metrics**"))
```

The median non-maximal hierarchical transition lies at approximately the 4th percentile of the full hierarchical distribution; the skill anchor is defined at this same percentile
to ensure comparable substitution intensity.


## Occupation's Education Specificity {#sec-occ-spec}

### Construction

**Step 1 — Overall Educational Distribution.** We compute the overall educational
distribution across all NOCs. This represents the expected education mix for a worker
drawn from the educated workforce without conditioning on occupation.

**Step 2 — Distribution Within Each NOC.** For each NOC we compute total workers $T$,
occupation-specific education shares $p$, and compare to the baseline distribution $p_0$
using KL divergence.

**Step 3 — Remove Size Effect.** Finite samples mechanically inflate divergence for small
occupations. We remove this artifact by residualizing $\log(\text{KL})$ on
$\log(T)$:

```{r occ-spec-code, echo=TRUE, eval=FALSE}
fit_kl <- lm(log(KL) ~ log(T), data = noc_edu_spec)

noc_edu_spec <- noc_edu_spec |>
  mutate(
    specificity = log(KL) - predict(fit_kl),
    TEER = str_sub(noc, 2, 2)
  )
```

The resulting specificity index measures how distinct an occupation's education mix is
relative to what is statistically expected given its size. KL divergence captures where
probability mass is concentrated relative to the aggregate baseline, rather than how much
mass lies in very small cells. Residualizing removes predictable finite-sample divergence,
isolating economically meaningful concentration. The index is orthogonal to occupation
size by construction, ensuring that specificity reflects structural concentration in
educational pathways rather than equilibrium scale effects.

### Specificity vs. Occupation Size {#sec-occ-spec-plot}

High specificity indicates few educational pathways into the occupation. Residualized
specificity is not correlated with log occupation size
(Pearson = `r dest_pear`, Spearman = `r dest_spear`).

```{r occ-spec-plot}
p <- ggplotly(dest_plt)
p$x$data <- rev(p$x$data)
p
```

### Occupation Specificity Rankings {#sec-occ-spec-table}

```{r occ-spec-table}
noc_edu_spec |>
  select(noc, specificity, contains("Based On")) |>
  arrange(desc(specificity)) |>
  my_dt()
```

## The $\varepsilon$ sweep and the cherry-picking problem {#sec-simulation}

Suppose that rather than evaluating across a grid of $\varepsilon$
values, we evaluated the models only at $\varepsilon = 0.5$. If the
DGP is hierarchy then binary would be identified as the winning
metric — it dominates both skill and hierarchy in terms of relative KL
improvement and relative CWTVD improvement. However, this win is an
artefact, not a signal.

```{r, warning=FALSE}
#| fig-width: 12
#| fig-height: 8
sim_score_plots$sub_regime_measures
```

All three cost matrices assign zero cost to the diagonal, so all three
transport plans over-concentrate mass on the diagonal at low
$\varepsilon$, predicting less movement than actually occurs. Skill and
hierarchy compound this by additionally over-concentrating on
geometrically cheap off-diagonal pairs, leaving insufficient mass for
transitions their cost structure scores as expensive. Binary's
over-concentration is confined to the diagonal alone, so it craters
less severely — and its mechanical peak, where diagonal concentration
just improves on independence before over-concentration takes over,
occurs at a lower $\varepsilon$ than the true metric's peak. This
mechanical peak is not evidence that binary is competitive with the
true metric; it is a property of the mathematics, not the data.

The winner of the horse race is the cost matrix whose peak relative
improvement — maximised over the full $\varepsilon$ grid — is highest.
This is the optimal transport analog to maximum likelihood model
selection: the EROT plan $P^\star_{ij} = u_i e^{-C_{ij}/\varepsilon}
v_j$ is the equilibrium of a multinomial logit discrete choice model,
and comparing models by their maximised likelihood is equivalent to
comparing their peak relative KL improvement. Peak CWTVD improvement
is reported alongside KL as a complementary criterion: in a horse race
about the geometry of mobility, closeness matters, and CWTVD penalises
misallocation in proportion to how far the mass is from where it
should be, where KL does not.  A result is declared a tie when the peak
improvements of the leading metrics are indistinguishable — the 50/50
skill/hierarchy panel below illustrates what a genuine tie looks like.

## The higher resolution of CWTVD {#sec-higher-resolution}

In the plot below, the peak score gap between the two leading cost metrics under relative CWTVD improvement (top row) widens as the data-generating mixture departs from balance — largest at 75/25 and 25/75, closing to a tie at the even 50/50 mixture, where a tie is the correct verdict. Relative KL improvement (bottom row) holds the two close to coincident across all three mixtures, resolving the imbalance only weakly. Note also the binary curve (red): under CWTVD its improvement collapses to the independence line after its peak, while under KL it retains a positive tail — KL rewards binary for reproducing the aggregate stay/move split regardless of where movers go, a credit CWTVD withholds.


```{r, warning=FALSE}
#| fig-width: 12
#| fig-height: 8
sim_score_plots$global_measures
```


## Destination lift by broad group {#sec-lift-by-silo}

The non-local model treats a high-lift move as one into an occupation a
credential distinctively feeds, and argues that such destinations are
concentrated in the credential-dense skilled silos rather than in the
generic-entry silos. The following plot substantiates that concentration
for the two endpoint cases: natural and applied sciences (NOC broad group 2),
a credential-dense silo, and sales and service (NOC broad group 6), the
generic-entry silo used as the illustration in the main text.

```{r, warning=FALSE}
#| fig-width: 12
#| fig-height: 8
compare_lift
```

The two empirical CDFs of destination lift cross once, near log lift zero, and
separate in the upper tail. Above the crossing the sales-and-service curve lies
above the sciences curve throughout, meaning a larger share of its
destination cells fall below any given high-lift threshold: at log lift 2,
roughly 95 percent of sales-and-service cells sit below, against roughly 80
percent for sciences, so sciences carries several times the high-lift tail
mass. The sales-and-service curve reaches one by log lift three or four, while
the sciences curve approaches one only past five. In words, sales and service
has almost no high-lift destinations, whereas sciences retains a substantial
high-lift tail of genuine credential pipelines.

The separation is real but moderate: the two distributions share a mode near
log lift zero and the crossing sits low, so the claim the figure supports is
the narrower one the mechanism requires, that the generic-entry silo has a
materially thinner high-lift tail, not that the two silos are disjoint. The few
high-lift destinations sales and service does contain are the licensed and
skilled-service occupations named in the main text.

Two features of the construction bear on reading the figure. Lift here is the
raw ratio, not size-residualized, so part of the tail difference reflects that
sciences occupations are on average rarer than sales-and-service occupations,
which mechanically admits higher lift through the smaller denominator; the
figure is therefore descriptive of the destination-lift distribution rather
than a size-adjusted measure of credential specificity. And both curves are
computed on the education--occupation cells that clear the reliability floor of
30 (see paper), so the tail is measured pipelines, not thin-cell noise.




