---
title: "Online Companion to *Occupational Mobility Is Not One Thing: Evidence from the Canadian Labour Market*"
author: "Richard Martin"
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    toc-title: "Contents"
    code-fold: TRUE
    smooth-scroll: true
    self-contained: true
    theme: flatly
    page-layout: full
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(janitor)
library(plotly)
library(gt)
library(DT)
library(conflicted)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
source(here("R", "other.R"))
```

```{r load-data, include=FALSE}
k_vs_spearman    <- read_rds(here("out", "k_vs_spearman.rds"))
hier             <- read_rds(here("out", "hier_dist_raw.rds"))
educ_specificity <- read_rds(here("out", "edu_noc_spec.rds"))
noc_specificity  <- read_rds(here("out", "noc_edu_spec.rds"))
calibration      <- read_rds(here("out", "calibration.rds"))
skills_mat       <- read_rds(here("out", "skill_dist_raw.rds"))
sim_score_plots <- read_rds(here("out","sim_score_plots.rds"))

skills_hc      <- hclust(as.dist(skills_mat), method = "average")
skills_ord     <- skills_hc$order
skills_mat_ord <- skills_mat[skills_ord, skills_ord]
skills_xlab    <- as.character(colnames(skills_mat_ord))
skills_ylab    <- as.character(rownames(skills_mat_ord))

dest_plt <- noc_specificity |>
  ggplot(aes(T, specificity, text = noc, fill = sub_regime)) +
  scale_fill_viridis_d() +
  geom_point(shape = 21, size = 3, stroke = .5) +
  scale_x_log10(labels = scales::comma) +
  labs(fill = "Sub Regime")

dest_spear <- round(cor(noc_specificity$T, noc_specificity$specificity, method = "spearman"), 2)
dest_pear  <- round(cor(noc_specificity$T, noc_specificity$specificity), 2)

spec_plt <- ggplot(educ_specificity, aes(T,
                                         specificity,
                                         fill = attain_bin,
                                         text = paste0("CIP:", cip,
                                                       "\n Highest:", highest))) +
  scale_fill_viridis_d() +
  geom_jitter(shape = 21, size = 1.5, stroke = .1) +
  scale_x_log10(labels = scales::comma) +
  labs(title = "Visibility of groups can be toggled by clicking legend below",
       fill = NULL)

spec_spear <- round(cor(educ_specificity$T, educ_specificity$specificity, method = "spearman"), 2)
spec_pear  <- round(cor(educ_specificity$T, educ_specificity$specificity), 2)
```

## Overview {#sec-overview}

This companion is designed to ensure transparency and reproducibility for the paper
*Occupational Mobility Is Not One Thing: Evidence from the Canadian Labour Market*.
It provides technical detail on distance metrics, normalization, and specificity measures.

**Contents**

- **Skill distance**: O\*NET-based occupational skill vectors, dimensionality reduction
  via PCA, and the resulting continuous distance matrix.
- **Hierarchical distance**: Taxonomy-based discrete distances derived from the NOC
  structure, resulting in labour market silos with internal TEER platforms.
- **Cost matrix normalization**: Percentile-based anchoring across distance metrics.
- **Occupation's education specificity**: A size-adjusted measure of concentration in
  educational inflows into occupations.
- **Education's occupation specificity**: A size-adjusted measure of concentration in
  occupational outcomes across fields of study and attainment levels.

---

## Skill Distance {#sec-skill}

We use 161 occupational characteristics (Skills, Abilities, Knowledge, Work Activities)
from O\*NET to measure the distance between occupations. Because occupational skill
vectors are high-dimensional and highly correlated, Euclidean distances computed in the
full feature space may suffer from distance concentration. We therefore project the data
onto the 12 leading principal components, which achieve a 99% rank correlation of
pairwise distances relative to the full 161-dimensional space. This reduces noise from
low-variance directions while preserving the dominant geometry of the skill space.

### PCA Rank Preservation {#sec-skill-pca}

The figure below shows the Spearman rank correlation between pairwise distances in the
$k$-dimensional PCA space and the full 161-dimensional space, as a function of $k$. The
dashed line marks the 0.99 threshold; 12 components suffice.

```{r skill-pca, warning=FALSE}
ggplot(k_vs_spearman, aes(x = k, y = spearman)) +
  geom_hline(yintercept = .99, lty = 2) +
  geom_line() +
  geom_point() +
  scale_x_continuous(limits = c(0, 50)) +
  labs(
    title = "First 12 PCs achieve ≥0.99 rank preservation of pairwise distances.",
    y = "Spearman correlation (distance ranks)",
    x = "Number of PCs"
  ) +
  theme(text = element_text(size = 10))
```

### Raw Skill Distance Heatmap {#sec-skill-heatmap}

Occupations are ordered by AGNES. Diagonal entries are self-distances (zero). Dark clusters along the diagonal indicate groups of
occupations with similar skill profiles. Use left mouse button drag to zoom. Raw distances are normalized prior to use: [Cost Matrix Normalization](#sec-norm) 

```{r skill-heatmap}
#| fig-width: 5.6
#| fig-height: 8
plotly::plot_ly(
  x = skills_xlab,
  y = skills_ylab,
  z = skills_mat_ord,
  type = "heatmap",
  hovertemplate = paste(
    "<b>Origin:</b> %{y}<br>",
    "<b>Destination:</b> %{x}<br>",
    "<b>Distance:</b> %{z:.3f}",
    "<extra></extra>"
  )
) |>
  plotly::layout(
    xaxis = list(type = "category", showticklabels = FALSE, ticks = ""),
    yaxis = list(type = "category", showticklabels = FALSE, ticks = "")
)
```

---

## Hierarchical Distance {#sec-hier}

Hierarchical distances encode institutional and career barriers implied by the NOC
taxonomy. Distance between any two occupations is defined as follows:

1. All five digits match: distance = 0
2. First four digits match: distance = 1
3. First three digits match: distance = 2
4. First digit matches: distance = 3 + |ΔTEER|
5. Otherwise: distance = 9

The Spearman correlation between skill and hierarchical distances is 
`r round(cor(as.vector(skills_mat[upper.tri(skills_mat)]), as.vector(hier[upper.tri(hier)]), method = "spearman"), 3)`,
indicating that institutional proximity and skill similarity are positively but only
modestly related. The two metrics encode overlapping yet distinct occupational geometries.

### Distance Counts {#sec-hier-counts}

```{r hier-counts}
hier_count_plot <- table(hier) |>
  enframe() |>
  ggplot(aes(name, value)) +
  geom_col(alpha = .5) +
  scale_y_continuous(trans = "log10", labels = scales::comma) +
  theme_minimal() +
  labs(x = "Normalized Distance", y = "Counts")

plotly::ggplotly(hier_count_plot)
```

### Raw Hierarchical Distance Heatmap {#sec-hier-heatmap}

Occupations are ordered by AGNES. The block-diagonal structure reflects the ten NOC broad occupational categories (silos); yellow off-diagonal
entries are cross-silo pairs assigned the sentinel distance of 9.  Raw distances are normalized prior to use: [Cost Matrix Normalization](#sec-norm). Use left mouse button drag to zoom.

```{r hier-heatmap}
#| fig-width: 5.6
#| fig-height: 8
hier_hc      <- hclust(as.dist(hier), method = "average")
hier_ord     <- hier_hc$order
hier_mat_ord <- hier[hier_ord, hier_ord]
hier_xlab    <- as.character(colnames(hier_mat_ord))
hier_ylab    <- as.character(rownames(hier_mat_ord))

plotly::plot_ly(
  x = hier_xlab,
  y = hier_ylab,
  z = hier_mat_ord,
  type = "heatmap",
  hovertemplate = paste(
    "<b>Origin:</b> %{y}<br>",
    "<b>Destination:</b> %{x}<br>",
    "<b>Distance:</b> %{z:.3f}",
    "<extra></extra>"
  )
) |>
  plotly::layout(
    xaxis = list(type = "category", showticklabels = FALSE, ticks = ""),
    yaxis = list(type = "category", showticklabels = FALSE, ticks = "")
  )
```

---

## Cost Matrix Normalization {#sec-norm}

The normalization anchor is hierarchical distance 3 — the maximum within-silo horizontal distance, representing movement within a broad occupational group without crossing TEER levels. This anchor is theoretically motivated: any hierarchical distance above 3 involves either a TEER step or a cross-category transition, meaning distance 3 is the natural boundary between horizontal and vertical mobility within the silo structure. Hierarchical distances are divided by 3; skill distances are divided by 7.145, the value at the same unconditional percentile of the skill distance distribution. Robustness to anchoring at distances 2 and 4 — one step shorter and one step into vertical movement respectively — is examined in Appendix C.

```{r norm-table}
calibration |>
  gt() |>
  fmt_number(columns = c(h_anchor, p_uncond, s_anchor), decimals = 3) |>
  cols_label(
    h_anchor = "Hierarchical Anchor",
    p_uncond = "Unconditional Percentile",
    s_anchor = "Skill Anchor"
  ) |>
  tab_options(table.font.size = "90%", data_row.padding = px(4)) |>
  tab_header(title = md("**Calibration Anchors Across Distance Metrics**"))
```

The median non-maximal hierarchical transition lies at approximately the 4th percentile
of the full hierarchical distribution; the skill anchor is defined at this same percentile
to ensure comparable substitution intensity.


## Occupation's Education Specificity {#sec-occ-spec}

### Construction

**Step 1 — Overall Educational Distribution.** We compute the overall educational
distribution across all NOCs. This represents the expected education mix for a worker
drawn from the educated workforce without conditioning on occupation.

**Step 2 — Distribution Within Each NOC.** For each NOC we compute total workers $T$,
occupation-specific education shares $p$, and compare to the baseline distribution $p_0$
using KL divergence.

**Step 3 — Remove Size Effect.** Finite samples mechanically inflate divergence for small
occupations. We remove this artifact by residualizing $\log(\text{KL})$ on
$\log(T)$:

```{r occ-spec-code, echo=TRUE, eval=FALSE}
fit_kl <- lm(log(KL) ~ log(T), data = noc_specificity)

noc_specificity <- noc_specificity |>
  mutate(
    specificity = log(KL) - predict(fit_kl),
    TEER = str_sub(noc, 2, 2)
  )
```

The resulting specificity index measures how distinct an occupation's education mix is
relative to what is statistically expected given its size. KL divergence captures where
probability mass is concentrated relative to the aggregate baseline, rather than how much
mass lies in very small cells. Residualizing removes predictable finite-sample divergence,
isolating economically meaningful concentration. The index is orthogonal to occupation
size by construction, ensuring that specificity reflects structural concentration in
educational pathways rather than equilibrium scale effects.

### Specificity vs. Occupation Size {#sec-occ-spec-plot}

High specificity indicates few educational pathways into the occupation. Residualized
specificity is not correlated with log occupation size
(Pearson = `r dest_pear`, Spearman = `r dest_spear`).

```{r occ-spec-plot}
p <- ggplotly(dest_plt)
p$x$data <- rev(p$x$data)
p
```

### Occupation Specificity Rankings {#sec-occ-spec-table}

```{r occ-spec-table}
noc_specificity |>
  select(noc, specificity, sub_regime) |>
  arrange(desc(specificity)) |>
  my_dt()
```

---

## Education's Occupation Specificity {#sec-educ-spec}

### Construction

Education's occupation specificity is constructed symmetrically to occupation's education
specificity, conditioning on education rather than occupation.

**Step 1 — Overall Occupational Distribution.** We compute the aggregate occupational
distribution across all educated workers, serving as the baseline absent conditioning on
field of study.

**Step 2 — Distribution Within Each Education.** For each (CIP × attainment) cell we
compute total graduates $T$, occupation shares $p$, and compare to the aggregate
occupational distribution $p_0$ using KL divergence.

**Step 3 — Remove Size Effect.** As with occupations, small cohorts mechanically inflate
divergence. We residualize $\log(\text{KL})$ on $\log(T)$:

```{r educ-spec-code, echo=TRUE, eval=FALSE}
fit_kl <- lm(log(KL) ~ log(T), data = educ_specificity)

educ_specificity <- educ_specificity |>
  mutate(specificity = log(KL) - predict(fit_kl))
```

The resulting index measures how narrowly a field of study feeds into occupations,
independent of cohort size. Professional and doctoral fields exhibit consistently high
occupational concentration; below that threshold, specificity varies substantially within
attainment levels, confirming that educational level alone does not determine pipeline
concentration.

### Specificity vs. Cohort Size {#sec-educ-spec-plot}

High specificity indicates few occupational pathways from the education. Residualized
specificity is not correlated with log cohort size
(Pearson = `r spec_pear`, Spearman = `r spec_spear`).

```{r educ-spec-plot}
spec_plotly <- plotly::ggplotly(spec_plt)

for (i in seq_along(spec_plotly$x$data)) {
  spec_plotly$x$data[[i]]$name <- gsub(",1", "", spec_plotly$x$data[[i]]$name)
  spec_plotly$x$data[[i]]$name <- gsub("^\\(|\\)$", "", spec_plotly$x$data[[i]]$name)
}

spec_plotly |>
  plotly::layout(
    legend = list(orientation = "h", x = 0.5, xanchor = "center", y = -0.2)
  )
```

### Education Specificity Rankings {#sec-educ-spec-table}

```{r educ-spec-table}
educ_specificity |>
  select(highest, cip, specificity) |>
  arrange(desc(specificity)) |>
  my_dt()
```

## Cherry-picking simulation {#sec-simulation}

Suppose that rather than evaluating across a grid of $\varepsilon$ 
values, we evaluated the models only at $\varepsilon = 0.5$ and used 
only relative KL improvement to score them. In this case, when the 
true DGP is hierarchy, binary would be identified as the winning 
metric — but this win is merely an artifact of skill and hierarchy 
incurring an overconfidence penalty at low $\varepsilon$, where 
cost-sensitive matrices concentrate probability mass too aggressively 
relative to observed mobility. This artifact extends to CWTVD as well.

The operative winner-declaration rule does not require dominance 
across the full $\varepsilon$ grid for non-binary metrics. A 
non-binary metric is declared the winner if it dominates all 
non-binary competitors across the sweep, even if it loses to binary 
at low $\varepsilon$ where the overconfidence penalty is operative. 
Binary is the exception: a binary win requires dominance across the 
entire grid. Genuine binary dominance is distinguished from 
artefactual dominance by its persistence at mid-to-high $\varepsilon$ 
values where the overconfidence penalty dissipates.


```{r}
#| fig-width: 12
#| fig-height: 8
sim_score_plots$sub_regime_measures
```


